### 2016-08-11 doc/manuscript.Rmd
Added manuscript to doc

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/tagDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
    - /home/luke/Documents/bibliography/bib/bib-db/methods.bib
output: fost::default_manuscript
---

<!--

Outline:

- Primary analysis:
    - Heatmap
    - GEE
        - Include # of both uncorrected and corrected (FDR) significant associations
        - QIC modeling
        - DAG development
        - Look over the entire model (see which variables are significant, etc)
    - Interaction (manhattan plot)
    - Interaction plots (with total? + clinical?)
        - With Ethnicity, generate contrast plots of adjusted estimates
    - PLS with ISI and HOMA-IR
        - CV for determining number of comps
        - Loadings (of TAGFA)
        - Prediction (of predicted vs real)
        - Use extracted PLS components in GEE
        - Train and test (CV) to see how it predicts in a new dataset (ie. a future visit)
- Secondary analysis:
    - PLS-DA (dysgly)
        - Train and test (CV)
        - Loadings (of TAGFA) 
        - Groupings
        - Prediction (misclassification)
    - AROC
        - dysgly with clinical TAG vs the 'best' fatty acid to compare to

Possible interpretations:

- muscle IR -> higher TAG
- But longitudinal, since no interaction with time, so maybe the higher TAG is
because of higher IR (since this population is at risk of being IR).

-->

# Title page

Title:

Author: Luke W. Johnston (1), MSc; Stewart B. Harris (2), MD; Ravi Retnakaran
(3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,6), PhD

<!-- During time of study. -->
Affiliation:

(1) Department of Nutritional Sciences, University of Toronto, Toronto, Ontario, Canada.
(2) Centre for Studies in Family Medicine, University of Western Ontario, London, Ontario, Canada.
(3) Division of Endocrinology, University of Toronto, Toronto, ON, Canada.
(4) Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto, Ontario, Canada.
(5) Department of Physiology, University of Toronto, Toronto, Ontario, Canada.
(6) Dalla Lana School of Public Health, University of Toronto, Toronto, Ontario, Canada.

Corresponding author:

- Name: Anthony J. Hanley
- Current address:  
    Department of Nutritional Sciences  
    Faculty of Medicine  
    University of Toronto  
    FitzGerald Building, 150 College Street, Room 341  
    Toronto, ON, Canada, M5S 3E2
- Phone number: 416.978.3616
- Fax number:
- Email: anthony.hanley@utoronto.ca

Disclaimers:

Funding support:

# Abstract

# Background

Outline:

- TAG strong predictor of diabetes and CVD risk
- Not well studied how composition plays a role
- Limited longitudinal data

Our objective was to examine the longitudinal role of the specific composition
of the TAG serum fraction on OGTT-derived measures of insulin sensitivity and
beta-cell function.

# Subjects and Methods

```{r load_data, message=FALSE}
devtools::load_all('.')
set_options()
load_data()
ds <- project_data
```

```{r caption_setup}
## Include captions below using `captioner` package
fig <- captioner::captioner(prefix = 'Figure')
cite_f <- pryr::partial(fig, display = 'cite')
sfig <- captioner::captioner(prefix = 'Supplemental Figure')
cite_sf <- pryr::partial(sfig, display = 'cite')
tab <- captioner::captioner(prefix = 'Table')
cite_t <- pryr::partial(tab, display = 'cite')
stab <- captioner::captioner(prefix = 'Supplemental Table')
cite_st <- pryr::partial(stab, display = 'cite')
# usage: cite_st('basicChar')

fig_gee <- fig('gee', 'caption here')
```

```{r set_datasets}
pls_df_homa <- prep_pls_data(ds, 'linvHOMA')
pls_df_isi <- prep_pls_data(ds, 'lISI')
```

```{r analyze_gee, cache=TRUE}
gee_results <- analyze_gee(ds)
gee_results_int <- analyze_gee(ds, intvar = 'VN')
```

```{r analyze_pls}
pls_results <- list()
pls_results$homa$train <- analyze_pls(pls_df_homa$train, 'linvHOMA')
pls_results$homa$test <- analyze_pls(pls_df_homa$test, 'linvHOMA')
pls_results$isi$train <- analyze_pls(pls_df_isi$train, 'lISI')
pls_results$isi$test <- analyze_pls(pls_df_isi$test, 'lISI')
```

```{r analyze_cor}
cor_results <- analyze_corr(ds)
```

# Results

## GEE

Main effect

Interaction effect

## PLS

Accuracy of predictive ability of PLS components with real measure {{use test maybe?}}.
Correlation of pred vs real in test set.

## PLS-DA

Misclassification of dysgly {{using test?}}, misclass by group too ('yes' vs 'no')

## AROC

AROC of clinical TAG vs TAGFA palmitic acid etc.

# Discussion

Outline:

- General findings
- Limitations of study
- Previous lit and limits of those
- How this study fills those gaps and new things learned
- Potential clinical application

# Acknowledgements

## Funding/support

## Conflict of interest

# Tables and figures

```{r table_basic}
table_basic()
```

```{r fig_heatmap}
plot_heatmap(cor_results)
```

```{r fig_dist}

```

```{r table_dist}
table_distribution()
```

```{r table_corr}
#table_corr(cor_results)
```

```{r fig_gee}
plot_gee_main(gee_results)
```

```{r table_gee_main}
table_gee_main(gee_results, '')
```

```{r fig_gee_int}
seer::view_interaction(gee_results_int, 
                       groups = 'unit~Yterms',
                       ylab = 'Triacylglycerol fatty acids') +
        graph_theme()
```

```{r fig_interacts}
# plot_interactions(ds, fa, outcome)
# plot_interactions(ds, fa, outcome)
# plot_interactions(ds, fa, outcome)
```

```{r fig_pls_cvncomp}
# Or as a table?
plot_pls_cv_ncomp(pls_results$homa$train)
plot_pls_cv_ncomp(pls_results_homa)
```

```{r fig_pls_loadings}
# As a single figure
seer::view_pls_xloadings(pls_results$homa$train) +
    graph_theme()
seer::view_pls_xloadings(pls_results$homa$test) +
    graph_theme()
seer::view_pls_xloadings(pls_results$isi$train) +
    graph_theme()
seer::view_pls_xloadings(pls_results$isi$test) +
    graph_theme()
```

```{r fig_pls_prediction}
# maybe not a figure, maybe just a correlation?
predplot(pls_results$isi$train, ncomp = 1:2)
predplot(pls_results$homa$train, ncomp = 1:2)
plot_pls_predict(pls_results$isi$train)
plot_pls_predict(pls_results_homa)
plot_pls_predict(test_set, pls_predict_set)
```

```{r fig_pls_comp_gee}
plot_gee_main(gee_results_plscomps)
```

```{r table_gee_pls}
table_gee_main(gee_results_plscomps)
```

```{r fig_plsda_cvncomps}
# can't remember if this is possible...
plot_plsda_cv_ncomp(plsda_results)
```

```{r fig_plsda_loadings}
plot_plsda_loadings(plsda_results)
```

```{r fig_plsda_groupings}
plot_plsda_groupings(plsda_results)
```

```{r fig_aroc}
plot_aroc(aroc_results) #?
```

# Supplemental Methods

# References

### 2016-08-19 R/utils.R
Added util func to format the proper rounding

@@ -5,6 +5,10 @@ trim_ws <- function (x) {
     gsub("^\\s+|\\s+$", "", x)
 }
 
+format_rounding <- function(x, digits) {
+    trim_ws(format(round(., digits = digits), nsmall = digits))
+}
+
 graph_theme <- function(base.plot, ticks = TRUE, minor.grid.lines = FALSE, legend.pos = 'bottom') {
     graph.theme <-
         ggplot2::"%+replace%"(

### 2016-08-19 R/analyze.R
Moved GEE functions all into one function

@@ -1,87 +1,12 @@
 # Analyze -----------------------------------------------------------------
 
-#' Run GEE models
-#'
-#' @param data The project data
-#' @param y outcomes (IS, BCF)
-#' @param x predictors (TAGFA)
-#' @param covariates to adjust for
-#' @param intvar interaction variable
-#' @param rename_x Function to rename x variables
-#' @param rename_y Function to rename y variables
-#' @export
-analyze_gee <- function(data = project_data,
-                        y = outcomes,
-                        x = list(
-                            tg_conc = tg_conc,
-                            tg_pct = tg_pct,
-                            tg_total = tg_totals
-                        ),
-                        covariates = c('VN', 'BaseAge', 'Sex', 'Ethnicity', 'Waist'),
-                        intvar = NULL,
-                        rename_x = renaming_fats,
-                        rename_y = renaming_outcomes) {
-
-    int <- !is.null(intvar)
-    if (int) {
-        extract_term <- ':'
-    } else {
-        extract_term <- 'Xterm$'
-    }
-    data %>%
-        prep_gee_data() %>%
-        mason::design('gee') %>%
-        mason::add_settings(family = stats::gaussian(),
-                            corstr = 'ar1', cluster.id = 'SID') %>%
-        mason::add_variables('yvars', outcomes) %>%
-        mason::add_variables('xvars', x[['tg_pct']]) %>%
-        mason::add_variables('covariates', covariates) %>% {
-            if (int) {
-                mason::add_variables(., 'interaction', intvar)
-            } else {
-                .
-            }
-        } %>%
-        mason::construct() %>%
-        mason::add_variables('xvars', x[['tg_conc']]) %>%
-        mason::construct() %>%
-        mason::add_variables('xvars', x[['tg_total']]) %>%
-        mason::construct() %>%
-        mason::scrub() %>%
-        mason::polish_filter(extract_term, 'term') %>%
-        dplyr::mutate(unit = ifelse(grepl('pct', Xterms), 'mol%',
-                                    ifelse(grepl('^tg\\d', Xterms), 'nmol/mL',
-                                           'Totals'))) %>%
-        mason::polish_transform_estimates(function(x) (exp(x) - 1) * 100) %>%
-        mason::polish_renaming(rename_x, 'Xterms') %>%
[…]

### 2016-08-19 R/gee.R

@@ -1,0 +1,167 @@
+# Functions for the GEE analysis
+#
+# Grab or combine data ----------------------------------------------------
+
+#' Prepare the project data for analysis through GEE.
+#'
+#' @param data project data
+#' @export
+prep_gee_data <- function(data) {
+    no_fattyacids <- data %>%
+        dplyr::select(-dplyr::matches('pct_tg\\d+|^tg\\d+'),
+                      -TotalNE,-TotalTG, -BaseTAG, -lBaseTAG)
+
+    scaled_variables <- data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select(SID, TotalNE, TotalTG, BaseTAG, lBaseTAG,
+                      dplyr::matches('pct_tg\\d+|^tg\\d+')) %>%
+        dplyr::mutate_each(dplyr::funs(as.numeric(scale(.))), -SID)
+
+    dplyr::full_join(
+            no_fattyacids,
+            scaled_variables,
+            by = 'SID'
+        ) %>%
+        dplyr::group_by(VN) %>%
+        dplyr::mutate(
+            Waist = as.numeric(scale(Waist)),
+            ALT = as.numeric(scale(ALT)),
+            BaseAge = as.numeric(scale(BaseAge)),
+            MET = as.numeric(scale(MET))
+        ) %>%
+        dplyr::ungroup() %>%
+        dplyr::arrange(SID, VN)
+}
+
+# Analyze -----------------------------------------------------------------
+
+#' Run GEE models on the prepared project data.
+#'
+#' @param data The project data
+#' @param y outcomes (IS, BCF)
+#' @param x predictors (TAGFA)
+#' @param covariates to adjust for
+#' @param intvar interaction variable
+#' @param rename_x Function to rename x variables
+#' @param rename_y Function to rename y variables
+#' @export
+analyze_gee <- function(data = project_data,
+                        y = outcomes,
+                        x = list(
+                            tg_conc = tg_conc,
+                            tg_pct = tg_pct,
+                            tg_total = tg_totals
+                        ),
+                        covariates = c('VN', 'BaseAge', 'Sex', 'Ethnicity', 'Waist'),
+                        intvar = NULL,
+                        rename_x = renaming_fats,
+                        rename_y = renaming_outcomes) {
+
+    int <- !is.null(intvar)
+    if (int) {
+        extract_term <- ':'
+    } else {
[…]

### 2016-08-19 R/prep_data.R

@@ -3,38 +3,8 @@
 prep_pls_data <- function(data, y, x) {
     data %>%
         dplyr::filter(VN == 0) %>%
-        dplyr::select_(y, x) %>%
+        dplyr::select_(.dots = c(y, x)) %>%
         stats::na.omit()
-}
-
-prep_gee_data <- function(data) {
-    no_fattyacids <- data %>%
-        dplyr::select(-dplyr::matches('pct_tg\\d+|^tg\\d+'),
-                      -TotalNE,-TotalTG, -BaseTAG, -lBaseTAG)
-
-    scaled_variables <- data %>%
-        dplyr::filter(VN == 0) %>%
-        dplyr::select(SID, TotalNE, TotalTG, BaseTAG, lBaseTAG,
-                      dplyr::matches('pct_tg\\d+|^tg\\d+')) %>%
-        dplyr::mutate_each(dplyr::funs(as.numeric(scale(.))), -SID)
-
-    gee_ready_data <-
-        dplyr::full_join(
-            no_fattyacids,
-            scaled_variables,
-            by = 'SID'
-        ) %>%
-        dplyr::group_by(VN) %>%
-        dplyr::mutate(
-            Waist = as.numeric(scale(Waist)),
-            ALT = as.numeric(scale(ALT)),
-            BaseAge = as.numeric(scale(BaseAge)),
-            MET = as.numeric(scale(MET))
-        ) %>%
-        dplyr::ungroup() %>%
-        dplyr::arrange(SID, VN)
-
-    return(gee_ready_data)
 }
 
 prep_dysglycemia_data <- function(data) {

### 2016-08-19 R/gee.R
Updated gee vignette and checked that all functions work

@@ -122,29 +122,29 @@ analyze_gee <- function(data = project_data,
 #'
 #' @export
 table_gee_main <- function(results, caption = NULL, digits = 1) {
-    gee_table_prep <- results %>%
+    table_data <- results %>%
         dplyr::mutate_each(dplyr::funs(format_rounding(., digits = digits)),
                            estimate, conf.low, conf.high) %>%
         dplyr::mutate(
             p.binary = ifelse(p.value <= 0.05, '\\*', ''),
-            estimate.ci = paste0(estimate, ' (', conf.low, ', ', conf.high, ')', p.binary)
+            estimate.ci = paste0(estimate, ' (', conf.low, ', ', conf.high, ')',
+                                 p.binary)
         ) %>%
         dplyr::select(Yterms, Xterms, unit, estimate.ci) %>%
-        tidyr::spread(Yterms, estimate.ci)
+        tidyr::spread(Yterms, estimate.ci) %>%
+        dplyr::mutate(Xterms = as.character(Xterms))
 
-    gee_table <-
-        dplyr::bind_rows(
-            data.frame(Xterms = paste0('**', unique(gee_table_prep$unit)[3], '**')),
-            dplyr::filter(gee_table_prep, unit == 'Totals'),
-            data.frame(Xterms = paste0('**', unique(gee_table_prep$unit)[1], '**')),
-            dplyr::filter(gee_table_prep, unit == 'nmol/mL'),
-            data.frame(Xterms = paste0('**', unique(gee_table_prep$unit)[2], '**')),
-            dplyr::filter(gee_table_prep, unit == 'mol%')
+    dplyr::bind_rows(
+            tibble::data_frame(Xterms = paste0('**Totals**')),
+            dplyr::filter(table_data, unit == 'Totals'),
+            tibble::data_frame(Xterms = paste0('**nmol/mL**')),
+            dplyr::filter(table_data, unit == 'nmol/mL'),
+            tibble::data_frame(Xterms = paste0('**mol%**')),
+            dplyr::filter(table_data, unit == 'mol%')
         ) %>%
         dplyr::select(-unit) %>%
-        dplyr::rename('Fatty acid' = Xterms)
-
-    pander::pander(gee_table, missing = '', caption = caption)
+        dplyr::rename('Fatty acid' = Xterms) %>%
+        pander::pander(missing = '', caption = caption)
 }
 
 # Plotting ----------------------------------------------------------------

### 2016-08-19 R/utils.R

@@ -6,7 +6,7 @@ trim_ws <- function (x) {
 }
 
 format_rounding <- function(x, digits) {
-    trim_ws(format(round(., digits = digits), nsmall = digits))
+    trim_ws(format(round(x, digits = digits), nsmall = digits))
 }
 
 graph_theme <- function(base.plot, ticks = TRUE, minor.grid.lines = FALSE, legend.pos = 'bottom') {

### 2016-08-19 R/analyze.R
Moved pls functions into their own file

@@ -1,18 +1,5 @@
 # Analyze -----------------------------------------------------------------
 
-analyze_pls <- function(data = project_data,
-                        y, x = tg_pct, ncomp = 2,
-                        type = c('train', 'test', 'full')) {
-    data %>%
-        prep_pls_data(y = y, x = x) %>%
-        mason::design('pls') %>%
-        mason::add_settings(ncomp = ncomp, cv.data = TRUE, cv.seed = 12345) %>%
-        mason::add_variables('yvars', y) %>%
-        mason::add_variables('xvars', x) %>%
-        mason::construct(cv = type) %>%
-        mason::scrub()
-}
-
 analyze_corr <-
     function(data = project_data,
              x = c(outcomes, 'BMI', 'Waist', 'Age', 'lALT', 'lTAG', 'Chol', 'HDL', 'LDL'),
@@ -26,7 +13,7 @@ analyze_corr <-
         mason::add_variables('xvars', x) %>%
         mason::construct() %>%
         mason::scrub() %>%
-        mason::polish_renaming(renaming_fa, 'Vars2') %>%
+        mason::polish_renaming(renaming_fats, 'Vars2') %>%
         mason::polish_renaming(function(x)
             gsub('l(ALT|TAG|IGIIR|invHOMA|ISI|ISSI2)', '\\1', x) %>%
                 renaming_outcomes(), 'Vars1') %>%

### 2016-08-19 R/pls.R

@@ -1,0 +1,25 @@
+# Functions for the PLS analysis
+#
+# Grab or combine data ----------------------------------------------------
+
+prep_pls_data <- function(data, y, x) {
+    data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select_(.dots = c(y, x)) %>%
+        stats::na.omit()
+}
+
+# Analyze -----------------------------------------------------------------
+
+analyze_pls <- function(data = project_data,
+                        y, x = tg_pct, ncomp = 2,
+                        type = c('train', 'test', 'full')) {
+    data %>%
+        prep_pls_data(y = y, x = x) %>%
+        mason::design('pls') %>%
+        mason::add_settings(ncomp = ncomp, cv.data = TRUE, cv.seed = 12345) %>%
+        mason::add_variables('yvars', y) %>%
+        mason::add_variables('xvars', x) %>%
+        mason::construct(cv = type) %>%
+        mason::scrub()
+}

### 2016-08-19 R/prep_data.R

@@ -1,12 +1,5 @@
 # Grab or combine data ----------------------------------------------------
 
-prep_pls_data <- function(data, y, x) {
-    data %>%
-        dplyr::filter(VN == 0) %>%
-        dplyr::select_(.dots = c(y, x)) %>%
-        stats::na.omit()
-}
-
 prep_dysglycemia_data <- function(data) {
     dysgly.data <-
         dplyr::left_join(

### 2016-08-19 R/corr.R
Moved correlation functions to own file

@@ -1,5 +1,14 @@
+# Functions to run the correlation analysis
+#
 # Analyze -----------------------------------------------------------------
 
+#' Correlation analysis.
+#'
+#' @param data Project data
+#' @param x The covariates and outcomes
+#' @param y The fatty acids.
+#'
+#' @export
 analyze_corr <-
     function(data = project_data,
              x = c(outcomes, 'BMI', 'Waist', 'Age', 'lALT', 'lTAG', 'Chol', 'HDL', 'LDL'),
@@ -28,4 +37,22 @@ analyze_corr <-
         dplyr::mutate(Vars2 = factor(Vars2, unique(Vars2)),
                       Vars1 = factor(Vars1, unique(Vars1)),
                       Correlations = round(Correlations, 2))
+    }
+
+# Plotting ----------------------------------------------------------------
+
+#' Correlation heatmap plot.
+#'
+#' @param results Correlation results
+#'
+#' @export
+plot_heatmap <- function(results) {
+     results %>%
+        seer::view_heatmap(
+            y = 'Vars2',
+            x = 'Vars1',
+            ylab = 'Triacylglycerol fatty acids (nmol/mL)',
+            number.colours = 5,
+            values.size = 4) +
+        graph_theme(ticks = FALSE, legend.pos = 'right')
 }

### 2016-08-23 R/setup.R
Add settings and packages used

@@ -27,7 +27,7 @@ set_options <- function() {
     # For the document (knitr)
     knitr::opts_chunk$set(
         warning = FALSE, message = FALSE, collapse = TRUE,
-        fig.showtext = TRUE
+        fig.showtext = TRUE, comment = '#>'
     )
 }
 

### 2016-08-23 R/prep_data.R
Created function to clean QIC output

@@ -30,3 +30,16 @@ prep_dysglycemia_data <- function(data) {
 
     return(dysgly.data)
 }
+
+
+#' Clean data output from the QIC model selection functions.
+#'
+#' @param data QIC output data.
+#'
+#' @export
+clean_qic <- function(data) {
+    data %>%
+        tibble::rownames_to_column() %>%
+        dplyr::select(Model = rowname, QIC = IC, Delta = delta) %>%
+        dplyr::mutate(id = seq_len(nrow(.)))
+}

### 2016-08-23 R/tables.R
Created completed function for making QIC table (includes incomplete funs)

@@ -1,0 +1,60 @@
+# Tables ------------------------------------------------------------------
+
+table_basic <- function(data = project_data, caption = NULL) {
+    data %>%
+        dplyr::mutate(Ethnicity = ifelse(VN == 0, as.character(Ethnicity), NA),
+                      Ethnicity = as.factor(Ethnicity),
+                      Sex = ifelse(VN == 0, as.character(Sex), NA),
+                      Sex = as.factor(Sex)) %>%
+        carpenter::outline_table('f.VN') %>%
+        carpenter::add_rows(c('HOMA', 'ISI', 'IGIIR', 'ISSI2'), carpenter::stat_medianIQR, digits = 1) %>%
+        carpenter::add_rows(c('ALT', 'TAG', 'Chol', 'HDL', 'BaseTotalNE', 'MET',
+                              'Age', 'BMI', 'Waist'),
+                            carpenter::stat_meanSD, digits = 1) %>%
+        carpenter::add_rows(c('Ethnicity', 'Sex'), carpenter::stat_nPct, digits = 0) %>%
+        carpenter::renaming('rows', renaming_table_rows) %>%
+        carpenter::renaming("header", c('Measure', 'Baseline', '3-yr', '6-yr')) %>%
+        carpenter::build_table(caption = caption)
+}
+
+table_distribution <- function(data = project_data, caption = NULL) {
+    fa <- function(pattern, x = c(tg_conc, 'TotalTG'))
+        grep(pattern, x, value = TRUE)
+    tgfa <- c(fa('3$'), fa('6$'), fa('7$'), fa('9$'), fa('0$'), fa('TotalTG$'))
+
+    data %>%
+        dplyr::filter(VN == 0) %>%
+        tidyr::gather(FA, Value, dplyr::matches('tg\\d')) %>%
+        dplyr::mutate(unit = as.factor(ifelse(grepl('pct_', FA), 'mol%', 'nmol/mL')),
+                      FA = gsub('pct_', '', FA)) %>%
+        tidyr::spread(FA, Value) %>%
+        dplyr::mutate(TotalTG = ifelse(unit == 'mol%', NA, TotalTG),
+                      unit = factor(unit, levels = c('nmol/mL', 'mol%'), ordered = TRUE)) %>%
+        carpenter::outline_table('unit') %>%
+        carpenter::add_rows(tgfa, carpenter::stat_meanSD, digits = 1) %>%
[…]

### 2016-08-23 R/variables.R
Added final covariates to use.

@@ -2,3 +2,4 @@ tg_conc <- grep('^tg\\d+', vars, value = TRUE)
 tg_pct <- grep('^pct_tg\\d+', vars, value = TRUE)
 tg_totals <- c('TotalTG', 'TAG')
 outcomes <- c('linvHOMA', 'lISI', 'lIGIIR', 'lISSI2')
+covariates <- c('VN', 'Waist', 'BaseAge', 'Ethnicity', 'Sex', 'ALT', 'MET', 'TotalNE')

### 2016-08-23 R/gee.R
Minor fixes to GEE fun (were bugs)

@@ -52,7 +52,7 @@ analyze_gee <- function(data = project_data,
                             tg_pct = tg_pct,
                             tg_total = tg_totals
                         ),
-                        covariates = c('VN', 'BaseAge', 'Sex', 'Ethnicity', 'Waist'),
+                        covars = covariates,
                         intvar = NULL,
                         rename_x = renaming_fats,
                         rename_y = renaming_outcomes) {
@@ -68,9 +68,9 @@ analyze_gee <- function(data = project_data,
         mason::design('gee') %>%
         mason::add_settings(family = stats::gaussian(),
                             corstr = 'ar1', cluster.id = 'SID') %>%
-        mason::add_variables('yvars', outcomes) %>%
+        mason::add_variables('yvars', y) %>%
         mason::add_variables('xvars', x[['tg_pct']]) %>%
-        mason::add_variables('covariates', covariates) %>% {
+        mason::add_variables('covariates', covars) %>% {
             if (int) {
                 mason::add_variables(., 'interaction', intvar)
             } else {

### 2016-08-24 R/pls.R
Added more and completed the pls functions.

@@ -2,6 +2,13 @@
 #
 # Grab or combine data ----------------------------------------------------
 
+#' Prepare the project data for PLS analysis.
+#'
+#' @param data Project data.
+#' @param y Y or outcome variable.
+#' @param x X or exposure variables of interest.
+#'
+#' @export
 prep_pls_data <- function(data, y, x) {
     data %>%
         dplyr::filter(VN == 0) %>%
@@ -11,15 +18,50 @@ prep_pls_data <- function(data, y, x) {
 
 # Analyze -----------------------------------------------------------------
 
+#' Compute PLS analysis.
+#'
+#' @param data Project data.
+#' @param y Outcome variable.
+#' @param x Exposure variables.
+#' @param ncomp Number of components.
+#'
+#' @export
 analyze_pls <- function(data = project_data,
-                        y, x = tg_pct, ncomp = 2,
-                        type = c('train', 'test', 'full')) {
+                        y, x = tg_pct, ncomp = NULL) {
     data %>%
         prep_pls_data(y = y, x = x) %>%
         mason::design('pls') %>%
-        mason::add_settings(ncomp = ncomp, cv.data = TRUE, cv.seed = 12345) %>%
+        mason::add_settings(ncomp = ncomp, validation = 'CV', cv.data = TRUE, cv.seed = 5436) %>%
         mason::add_variables('yvars', y) %>%
         mason::add_variables('xvars', x) %>%
-        mason::construct(cv = type) %>%
+        mason::construct() %>%
         mason::scrub()
+}
+
+# Plots -------------------------------------------------------------------
+
+#' Plot the X loadings of the PLS results.
+#'
+#' @param data PLS output results.
+#'
+#' @export
+plot_pls <- function(data) {
+    seer::view_pls_xloadings(data, renaming.x = renaming_fats) +
+        graph_theme(minor.grid.lines = FALSE)
+}
+
+# Calculations ------------------------------------------------------------
+
+#' Calculate the correlation between the predicted and measured outcome.
+#'
+#' @param model The PLS output results.
+#' @param test The test data to predict from.
+#' @param ncomps The component to predict on.
+#'
+#' @export
+calc_pred_corr <- function(model, test, ncomps = 1) {
+    predicted <- predict(model, ncomp = ncomps, newdata = test)
+    measured <- as.matrix(model.response(model.frame(formula(model),
+                                                     data = test)))
+    cor(predicted, measured)
 }

### 2016-08-24 R/plot.R
Added plot for TAG distribution

@@ -1,0 +1,30 @@
+# Plotting ----------------------------------------------------------------
+
+#' Plots the distribution of the TAGFA composition.
+#'
+#' @param data Project data.
+#'
+#' @export
+plot_tagfa <- function(data = project_data) {
+    data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select(SID, dplyr::matches('^tg\\d+')) %>%
+        tidyr::gather(Measure, Value,-SID) %>%
+        stats::na.omit() %>%
+        dplyr::mutate(Measure = renaming_fats(Measure)) %>%
+        dplyr::mutate(
+            order1 = substr(Measure, nchar(Measure), nchar(Measure)),
+            order1 = ifelse(order1 == 0, 10, order1),
+            order1 = ifelse(order1 == 'l', 20, order1),
+            order1 = as.integer(order1)
+        ) %>%
+        dplyr::arrange(dplyr::desc(order1)) %>%
+        dplyr::select(-order1) %>%
+        dplyr::mutate(Measure = Measure %>%
+                   factor(., levels = unique(.))) %>%
+        seer::view_boxplots(
+            'Measure', 'Value',
+            xlab = 'Concentration (nmol/mL)',
+            ylab = 'Triacylglyceride fatty acids') +
+        graph_theme(ticks = FALSE)
+}

### 2016-08-25 R/pls.R
Updated pls correlation fun to include p-value (incase)

@@ -24,14 +24,15 @@ prep_pls_data <- function(data, y, x) {
 #' @param y Outcome variable.
 #' @param x Exposure variables.
 #' @param ncomp Number of components.
+#' @param cv Whether to use CV.
 #'
 #' @export
 analyze_pls <- function(data = project_data,
-                        y, x = tg_pct, ncomp = NULL) {
+                        y, x = tg_pct, ncomp = NULL, cv = TRUE) {
     data %>%
         prep_pls_data(y = y, x = x) %>%
         mason::design('pls') %>%
-        mason::add_settings(ncomp = ncomp, validation = 'CV', cv.data = TRUE, cv.seed = 5436) %>%
+        mason::add_settings(ncomp = ncomp, validation = 'CV', cv.data = cv, cv.seed = 5436) %>%
         mason::add_variables('yvars', y) %>%
         mason::add_variables('xvars', x) %>%
         mason::construct() %>%
@@ -60,8 +61,9 @@ plot_pls <- function(data) {
 #'
 #' @export
 calc_pred_corr <- function(model, test, ncomps = 1) {
-    predicted <- predict(model, ncomp = ncomps, newdata = test)
-    measured <- as.matrix(model.response(model.frame(formula(model),
-                                                     data = test)))
-    cor(predicted, measured)
+    predicted <- stats::predict(model, ncomp = ncomps, newdata = test)
+    measured <- as.matrix(stats::model.response(
+        stats::model.frame(formula(model), data = test))
+        )
+    broom::tidy(stats::cor.test(predicted, measured))[c(1, 3)]
 }

### 2016-08-25 R/tables.R
Added roxygen docs to table.R

@@ -1,5 +1,11 @@
 # Tables ------------------------------------------------------------------
 
+#' Create a table of the basic characteristics of PROMISE.
+#'
+#' @param data Project data.
+#' @param caption Table caption.
+#'
+#' @export
 table_basic <- function(data = project_data, caption = NULL) {
     data %>%
         dplyr::mutate(Ethnicity = ifelse(VN == 0, as.character(Ethnicity), NA),
@@ -17,7 +23,13 @@ table_basic <- function(data = project_data, caption = NULL) {
         carpenter::build_table(caption = caption)
 }
 
-table_distribution <- function(data = project_data, caption = NULL) {
+#' Create a table of the values for the conc and mol% of TAGFA.
+#'
+#' @param data Project data.
+#' @param caption Table caption.
+#'
+#' @export
+table_tagfa <- function(data = project_data, caption = NULL) {
     fa <- function(pattern, x = c(tg_conc, 'TotalTG'))
         grep(pattern, x, value = TRUE)
     tgfa <- c(fa('3$'), fa('6$'), fa('7$'), fa('9$'), fa('0$'), fa('TotalTG$'))

### 2016-09-01 R/fetch_data.R
Created binary var for Ethnicity

@@ -85,7 +85,12 @@ fetch_data <- function() {
                     ),
                     c('Other', 'European', 'Other', 'Latino/a',
                       'Other', 'South Asian')
-                )
+                ),
+            BiEthnicity = plyr::mapvalues(
+                Ethnicity,
+                c('Other', 'European', 'Latino/a', 'South Asian'),
+                c('Non-European', 'European', 'Non-European', 'Non-European')
+            )
         ) %>%
         dplyr::arrange(SID, VN) %>%
         dplyr::filter(!is.na(TotalTG))

### 2016-09-01 doc/manuscript.Rmd
Added an onload function to load up settings/data
Mid-state of manuscript writing (at least to have a baseline in history)

@@
 
+Note: keep the story simple and clean, without getting too dense
+
 Outline:
@@
         - Look over the entire model (see which variables are significant, etc)
-    - Interaction (manhattan plot)
-    - Interaction plots (with total? + clinical?)
-        - With Ethnicity, generate contrast plots of adjusted estimates
     - PLS with ISI and HOMA-IR
-        - CV for determining number of comps
-        - Loadings (of TAGFA)
-        - Prediction (of predicted vs real)
-        - Use extracted PLS components in GEE
-        - Train and test (CV) to see how it predicts in a new dataset (ie. a future visit)
+        - CV for determining number of comps (best is 2)
+        - Loadings (of TAGFA) (tho all are the same, use HOMA/ISI. BCF have
+        lower corr between measured and predicted) (Use COrr of both components
+        (1 and 2))
+        - Prediction (of predicted vs real) (in the pls.rmd file, not in
+        manuscript (or maybe in ESM?))
+        - Use extracted PLS components in GEE (?)
+        - Train and test (CV) to see how it predicts in a new dataset (ie. a
+        future visit) (done. In pls.rmd file. As ESM?)
 - Secondary analysis:
@@
 
+NEFA may be a mediating factor in association, but that adjustment will not
+attenuate this assoc since TAGFA composition is also dictated by other sources
+for DNL (e.g. recent meal, DNL from glucose stores, etc.).
+
 # Subjects and Methods
@@
 devtools::load_all('.')
-set_options()
-load_data()
-ds <- project_data
+ds0yr <- project_data %>% 
+    dplyr::filter(VN == 0)
 ```
@@
 fig_gee <- fig('gee', 'caption here')
-```
-
-```{r set_datasets}
-pls_df_homa <- prep_pls_data(ds, 'linvHOMA')
-pls_df_isi <- prep_pls_data(ds, 'lISI')
+fig_consort <- fig('consort', 'caption here')
 ```
@@
 ```{r analyze_gee, cache=TRUE}
-gee_results <- analyze_gee(ds)
-gee_results_int <- analyze_gee(ds, intvar = 'VN')
+gee_df <- analyze_gee()
 ```
 
-```{r analyze_pls}
-pls_results <- list()
-pls_results$homa$train <- analyze_pls(pls_df_homa$train, 'linvHOMA')
-pls_results$homa$test <- analyze_pls(pls_df_homa$test, 'linvHOMA')
-pls_results$isi$train <- analyze_pls(pls_df_isi$train, 'lISI')
-pls_results$isi$test <- analyze_pls(pls_df_isi$test, 'lISI')
+```{r analyze_others}
+# CV results are shown in pls.Rmd vignette
+pls_homa <- analyze_pls(y = 'linvHOMA', cv = FALSE)
+pls_isi <- analyze_pls(y = 'lISI', cv = FALSE)
+
+# Correlation between predicted and actual
+pls_homa_t <- analyze_pls(y = 'linvHOMA')
+pls_isi_t <- analyze_pls(y = 'lISI')
+
+cor_df <- analyze_corr()
 ```
 
-```{r analyze_cor}
-cor_results <- analyze_corr(ds)
-```
+Participants from London and Toronto, Canada, were recruited into the 
+Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
+for recruitment into PROMISE required having one or more risk factors for type 2
+diabetes mellitus, including obesity, hypertension, family history of diabetes, 
+and/or a history of gestational diabetes or birth of a macrosomic infant. 
+Participants aged 30 years and older (n=736) attended the baseline visit
+between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
+examination visits completed to date (2004-2006, 2007-2009, and 2010-2013). 
+Participants are contacted annually by telephone. The
+current study sample used data on participants who did not have diabetes at baseline,
+who returned for one or more of the follow-up examinations, and who had samples
+available for fatty acid measurements (n=477). A diagram of the sample size at
+each visit is shown in `r cite_sf('consort')`. At each examination,
+participants undergo metabolic characterization, anthropometric measurements,
+and questionnaires on lifestyle and sociodemographics. Research ethics approval
+was obtained from Mount Sinai Hospital and the University of Western Ontario,
+and all participants provided written informed consent.
+Data collection methods were standardized across the 2 centres and research
+nurses were centrally trained.
+
+## Blood measure assessments
+
+At each examination, an 8-12 hour fasting blood sample was drawn from each
+participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
+minute and 2 hour blood draw. All blood samples were processed and frozen at
+-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
+procedures. Cholesterol, HDL, and clinically-measured triacylglycerides (TAG)
+were measured using Roche Modular's enzymatic colorimetric tests (Mississauga,
+ON).  Both insulin and glucose were derived from the OGTT at fasting, 30 minute,
+and 2 hour time points. Specific insulin was measured using the Elecsys 1010
+(Roche Diagnostics, Basel, Switzerland) immunoassay analyzer and
+electrochemiluminescence immunoassay. This assay shows 0.05% cross-reactivity to
+intact human pro-insulin and the Des 31,32 circulating split form (Linco Res.
+Inc), and has a coefficient of variation (CV) of 9.3%. Glucose was determined
+using an enzymatic hexokinase (Roche Modular, Roche Diagnostics) with a
+detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The inter-assay %CV is <1.1%
+and intra-assay %CV is < 1.9%.  All assays were performed at the Banting and
+Best Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose
+(IFG), impaired glucose tolerance (IGT), and diabetes were categorized using the
+2006 WHO criteria [@WHO2006].
+
+TAGFA composition was quantified using stored fasting serum samples from the 
+baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
+exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be 
+stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
+of heptadecanoic acid (17:0) was added to the serum as an internal standard
+prior to extracting total lipids according to the method of Folch [@Folch1957a].
+Each serum lipid fraction---non-esterified fatty acids (NEFA), cholesteryl
+ester, phospholipid, and TAGFA---was isolated using thin layer chromatography;
+each fraction was visualized under UV light after lightly spraying with
+8-anilino-1-naphthalene sulfonic acid (0.1% wt/vol) and then converted to fatty
+acid methyl esters with 14% boron trifluoride in methanol at 100°C for 1 h.
+Fatty acid methyl esters were separated and quantified using a Varian-430 gas
+chromatograph (Varian, Lake Forest, CA, USA) equipped with a Varian Factor Four
+capillary column and a flame ionization detector, which were injected in
+splitless mode. Fatty acid concentrations (nmol/ml) were calculated by
+proportional comparison of gas chromatography peak areas to that of the internal
+standards [@Nishi2014a]. There were 22 fatty acids measured in the TAGFA
+fraction. Given their diverse biology, as well as the complexity of the
+analyses, findings for other lipid fractions in this cohort are reported
+separately (see ref @Johnston2016a and for the analysis of the phospholipid and
+cholesteryl ester fractions and {{next citation}} for the analysis of the NEFA
+fraction).
+
+## Outcome variables
+
+Insulin sensitivity and beta-cell function indices were computed using the OGTT 
+glucose and insulin data. Insulin sensitivity was assessed using 1/HOMA-IR (1
+divided by HOMA-IR) [@Matthews1985] and the Insulin Sensitivity Index (ISI)
+[@Matsuda1999]. HOMA-IR largely reflects hepatic insulin resistance, while ISI
+reflects whole-body insulin sensitivity [@AbdulGhani2007].  Beta-cell function
+was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR)
+and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
+is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
+the disposition index (but using OGTT values). Each index has been validated
+against gold standard measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].
+
+## Anthropometrics and sociodemographics
+
+Height, weight, and waist circumference (WC) were measured at all clinic 
+examinations using standard procedures.  WC was measured at the natural waist, 
+defined as the narrowest part of the torso between the umbilicus and the xiphoid
+process. BMI was calculated by dividing weight (kg) by height (m) squared.
+Sociodemographic information, including age, sex, and ethnicity, were determined
+using questionnaires administered at each examination. In the lifestyle
+questionnaire, physical activity was determined using a version of the
+Modifiable Activity Questionnaire (MAQ) [@Kriska1990].  The MAQ collects 
+information on leisure and occupational activity, including intensity, 
+frequency, and duration, over the past year.  Each reported activity from the 
+MAQ was weighted by its metabolic intensity allowing for the estimation of 
+MET-hours per week.
+
+## Statistical analysis
+
+The primary outcome variables for this analysis were 
+1/HOMA-IR, ISI, IGI/IR, and ISSI-2, which were log-transformed for
+the statistical modeling. The primary predictor variables for this analysis were
+22 individual TAGFA as mole percent (mol%) of the total fraction and as
+concentration (nmol/mL). Pearson correlation coefficients were computed to
+assess the relationships of individual TAGFA with other continuous variables.
+
+Of note is clinically measured TAG was only quantified on a portion of the
+6-year visit (n = ~120 {{ find true number }}).
+
+For the primary analysis, generalized estimating equation (GEE) models 
+[@Zeger1986a] were used to determine the longitudinal associations between the 
+outcome variables and the predictor variables. Given the longitudinal design 
+an auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE models, 
+though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
+is well suited to longitudinal cohort studies given it's capacity to handle
+missed visits. The predictor variables and continuous covariates were scaled
+(mean centered and standardized). The TAGFA variables were classified as
+*time-independent* (held constant) as they were measured only at the baseline
+visit, while the outcome variables and covariates were set as *time-dependent*.
+No imputation was conducted on missing values.
+
+Covariate selection was based on previous literature, directed acyclic graph 
+[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
+(QIC) {{citation?}}. `r cite_st('qic')` shows the covariates compared using QIC. While the
+final GEE model selected as best fitting differed between insulin sensitivity
+and beta-cell function measures, the differences in QIC values were less than
+10 between the models, suggesting similar fit. As such, we selected the model
+that had the fewest covariates and that had similar fit between the outcome
+measures ({{##}}). The final GEE model was adjusted for {{covariates}}. After scaling,
+log-transforming, and exponentiating, the GEE estimates are
+interpreted as an expected percent difference in the outcome variable for every 
+standard deviation (SD) increase in the predictor variable given the covariates 
+are held constant. Lastly for the GEE models, we tested for an interaction with
+sex, ethnicity, or time by the predictor term for each outcome variable.
+
+While GEE accounts for the longitudinal design of the data, this approach is 
+limited in that it cannot analyze the inherent multivariate nature of the 
+composition of the TAGFA fraction. To confirm the GEE results in a multivariate 
+environment (all TAGFA), partial least squares regression (PLS) {{citation}} was used to
+identify the patterns of TAGFA composition against insulin sensitivity and
+beta-cell function. For more detailed explanation of GEE and PLS, please see the
+{{supplemental}} Methods.
+
+All analyses were performed using R `r getRversion()` [@Rbase], along with the R
+packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a] and pls
+`r packageVersion('pls')` for PLS {{cite?}}. The R code and extra analyses for
+this manuscript is available at {{code doi}}. Results were considered
+statistically significance at p<0.05, after adjusting for multiple testing using
+the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a]. STROBE was used
+as a guideline for reporting [@Vandenbroucke2007b].
+
+The number of components to extract from PLS was determined by 10-fold CV.
+
+Dataset was split into a training set and a testing set to determine the
+predictability of the given PLS model. Final results presented use the full dataset.a
 
@@
 
-## GEE
+```{r inline}
+dys <- project_data %>% 
+    prep_dys_data() %>% 
+    calc_convert_dys()
+
+tagfa <- calc_tagfa_percent()
+outcome_chg <- calc_outcome_changes()
+calc_discr_npct(ds0yr$Sex, 'Female')
+calc_discr_npct(ds0yr$Ethnicity, 'European')
+estci <- calc_gee_estci(gee_df)
+fa_cor <- calc_cor()
+cor_pls_homa <- calc_pred_corr(pls_homa_t, pls_homa_t$test_data)
+cor_pls_isi <- calc_pred_corr(pls_isi_t, pls_isi_t$test_data)
+```
+
+## Basic characteristics of the PROMISE cohort
+
+Basic characteristics/demographics
+
+The outcome variables significantly decreased over the 6 years by a median 
+`r outcome_chg$chg` (p`r outcome_chg$p`; n=`r outcome_chg$n`).
+
+`r fig_tagfa` shows the distribution of each TAGFA (see `r stab_tagfa` for mean
+and SD {{ confirm }}). Three TAGFA contributed 
+`r format_rounding(sum(tagfa$pct))`% to the total TAG concentration:
+`r tagfa$c[1]`; `r tagfa$c[2]`; and, `r tagfa$c[3]`.
+
+`r fig_corr` shows a heatmap of the correlation of individual TAGFA with the
+outcome variables and several basic characteristics.
+As expected, nearly all TAGFA had very strong positive correlations (r=
+`r fa_cor$rng$TAG`) with clinically-measured TAG. There was also moderate
+negative correlations with HDL (r=`r fa_cor$rng$HDL`). For the outcome variables,
+the correlations for the insulin sensitivity measures were generally higher 
+(1/HOMA-IR: r=`r fa_cor$rng['1/HOMA-IR']`, ISI: r=`r fa_cor$rng$ISI`) than 
+for the beta-cell function measures (all r<0.30).
+
+## Generalized estimating equation models
 
@@
 
-Interaction effect
+There were {{ ## }} significant associations in the unadjusted GEE model shown
+in {{ `r sfig_gee` }}. Only a few associations were attenuated after full model
+adjustment primarily with the beta-cell function outcomes, as shown in 
+`r fig_gee`. As a concentration, nearly all TAGFA had a strong negative
+association on 1/HOMA-IR and ISI, and a few had strong negative associations
+with IGI/IR and ISSI-2. As a mol%, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7)
+had negative associations with 1/HOMA-IR and ISI (between {{ ## }}% lower
+insulin sensitivity for every SD increase in the TAGFA), while several more TAGFA had
+positive associations with 1/HOMA-IR and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9,
+18:2n-6, 20:2n-6, 20:4n-6, and 22:5n-3) predicting a {{ ## }}% higher insulin
+sensitivity for every SD increase in the TAGFA. 
+One TAGFA, 20:2n-6, had a very strong positive association with the
+insulin sensitivity measures, with a {{ ## }}% higher insulin sensitivity for
+every SD increase in 20:2n-6.
 
-## PLS
+Both clinically-measured TAG and total TAGFA concentration had very strong
+negative associations with all outcome variables.
+
+There were no significant interactions by time, sex, or ethnicity for any of the
+TAGFA (data not shown).
+
+## Clustering of TAGFA by metabolic measures
 
@@
 
-## PLS-DA
+The predictive ability of the PLS analysis using the TAGFA composition
 
-Misclassification of dysgly {{using test?}}, misclass by group too ('yes' vs 'no')
+`r fig_pls`
 
-## AROC
+The correlation between the predicted values of the outcome variables from the
+PLS analysis compared to the actual value was `r cor_pls_homa$r` 
+(p`r cor_pls_homa$p`) for 1/HOMA-IR and `r cor_pls_isi$r` (p`r cor_pls_isi$p`) 
+for ISI, suggested TAGFA composition has moderate predictive ability for 
+insulin sensitivity measures.
 
-AROC of clinical TAG vs TAGFA palmitic acid etc.
+`r calc_pred_corr()`
 
@@
 - Potential clinical application
+
+[@Rhee2011a] fasting TAG status better as discriminating between DM case vs
+control than 2hr-OGTT TAG
+
+[@Kotronen2009a] (this wasn't studied, but discussed) higher hepatic DNL may
+increase amount of 'shorter' chain FA (16:0, 14:0, etc.) into VLDL and hence
+into circulation. So higher of these FA may indicate higher DNL -> problems with
+liver function -> maybe hepatic IR -> cyclical?
+
+Discuss my PL and CE paper as well. Maybe also the NEFA one if it is published
+by then...
+
+The four FA that cluster together on the PLS also appear to be the same FA that
+make up most of DNL, especially in context of higher carb/refined carb in diet
+[@Wilke2009a].
 
@@
 
-```{r fig_heatmap}
-plot_heatmap(cor_results)
-```
-
 ```{r fig_dist}
-
+plot_tagfa()
 ```
@@
 ```{r table_dist}
-table_distribution()
+table_tagfa()
+```
+
+```{r fig_heatmap}
+plot_heatmap(cor_df)
 ```
@@
 ```{r fig_gee}
-plot_gee_main(gee_results)
+plot_gee_main(gee_df)
 ```
@@
 ```{r table_gee_main}
-table_gee_main(gee_results, '')
+table_gee_main(gee_df)
 ```
 
-```{r fig_gee_int}
-seer::view_interaction(gee_results_int, 
-                       groups = 'unit~Yterms',
-                       ylab = 'Triacylglycerol fatty acids') +
-        graph_theme()
-```
-
-```{r fig_interacts}
-# plot_interactions(ds, fa, outcome)
-# plot_interactions(ds, fa, outcome)
-# plot_interactions(ds, fa, outcome)
-```
-
-```{r fig_pls_cvncomp}
-# Or as a table?
-plot_pls_cv_ncomp(pls_results$homa$train)
-plot_pls_cv_ncomp(pls_results_homa)
-```
-
-```{r fig_pls_loadings}
-# As a single figure
-seer::view_pls_xloadings(pls_results$homa$train) +
-    graph_theme()
-seer::view_pls_xloadings(pls_results$homa$test) +
-    graph_theme()
-seer::view_pls_xloadings(pls_results$isi$train) +
-    graph_theme()
-seer::view_pls_xloadings(pls_results$isi$test) +
-    graph_theme()
-```
-
-```{r fig_pls_prediction}
-# maybe not a figure, maybe just a correlation?
-predplot(pls_results$isi$train, ncomp = 1:2)
-predplot(pls_results$homa$train, ncomp = 1:2)
-plot_pls_predict(pls_results$isi$train)
-plot_pls_predict(pls_results_homa)
-plot_pls_predict(test_set, pls_predict_set)
-```
-
-```{r fig_pls_comp_gee}
-plot_gee_main(gee_results_plscomps)
-```
-
-```{r table_gee_pls}
-table_gee_main(gee_results_plscomps)
-```
-
-```{r fig_plsda_cvncomps}
-# can't remember if this is possible...
-plot_plsda_cv_ncomp(plsda_results)
-```
-
-```{r fig_plsda_loadings}
-plot_plsda_loadings(plsda_results)
-```
-
-```{r fig_plsda_groupings}
-plot_plsda_groupings(plsda_results)
-```
-
-```{r fig_aroc}
-plot_aroc(aroc_results) #?
+```{r fig_pls}
+gridExtra::grid.arrange(
+    plot_pls(pls_homa) + ggplot2::ggtitle('A: 1/HOMA-IR'), 
+    plot_pls(pls_isi) + ggplot2::ggtitle('B: ISI'), 
+    ncol = 2)
 ```

### 2016-09-01 R/calculate.R
Calc fun for correlation data

@@ -130,3 +130,28 @@ calc_gee_estci <- function(results) {
         dplyr::arrange(Yterms, dplyr::desc(estimate)) %>%
         dplyr::select(Yterms, Xterms, e)
 }
+
+#' Extracts and computes various information from the correlation results.
+#'
+#' @param data Correlation results data.
+#'
+#' @export
+calc_cor <- function(data = cor_df) {
+    data %>%
+        dplyr::group_by(Vars1) %>%
+        {
+            high_cor <- dplyr::filter(., Correlations < -0.3 | Correlations > 0.3)
+            low_cor <- dplyr::filter(., Correlations > -0.3 & Correlations < 0.3)
+
+            list(
+                mean = dplyr::summarize(high_cor, mean = format_rounding(mean(Correlations))),
+                rng = dplyr::summarize(high_cor, rng = paste0(min(Correlations),
+                                                              ' to ', max(Correlations))),
+                fa = table(low_cor$Vars1, low_cor$Vars2) %>%
+                    tibble::as_data_frame() %>%
+                    dplyr::group_by(Var2) %>%
+                    dplyr::summarize(n = sum(n)) %>%
+                    dplyr::arrange(dplyr::desc(n))
+            )
+        }
+}

### 2016-09-01 R/variables.R
Covar change to binary ethnicity

@@ -2,4 +2,4 @@ tg_conc <- grep('^tg\\d+', vars, value = TRUE)
 tg_pct <- grep('^pct_tg\\d+', vars, value = TRUE)
 tg_totals <- c('TotalTG', 'TAG')
 outcomes <- c('linvHOMA', 'lISI', 'lIGIIR', 'lISSI2')
-covariates <- c('VN', 'Waist', 'BaseAge', 'Ethnicity', 'Sex', 'ALT', 'MET', 'TotalNE')
+covariates <- c('VN', 'Waist', 'BaseAge', 'BiEthnicity', 'Sex', 'ALT', 'MET', 'TotalNE')

### 2016-09-06 R/tables.R
Added function to make QIC table in manuscript

@@ -70,3 +70,103 @@ table_qic <- function(data, caption = NULL) {
                       Delta = format_rounding(Delta, 1)) %>%
         pander::pander(missing = '', caption = caption)
 }
+
+#' QIC table for manuscript.
+#'
+#' @param data Project data.
+#' @param caption Caption for table.
+#'
+#' @export
+table_qic_main <- function(data = project_data, caption = NULL) {
+    prep_qic <- data %>%
+        prep_gee_data() %>%
+        dplyr::select(
+            SID,
+            VN,
+            lISSI2,
+            lISI,
+            TotalTG,
+            TotalNE,
+            Sex,
+            Ethnicity,
+            MET,
+            ALT,
+            BaseAge,
+            Waist,
+            BMI,
+            AlcoholPerWk,
+            TobaccoUse,
+            FamHistDiab
+        ) %>%
+        stats::na.omit()
+
+    qic_is <- prep_qic %>%
+        dplyr::select(-lISSI2)
+
+    qic_bcf <- prep_qic %>%
+        dplyr::select(-lISI)
+    M0_ar_is <- geepack::geeglm(
+        lISI ~ TotalTG + VN,
+        data = qic_is,
+        id = SID,
+        family = stats::gaussian,
+        corstr = 'ar1'
+    )
+    M0 <- M0_ar_is
+    Full <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET +
+                AlcoholPerWk + FamHistDiab + TobaccoUse + TotalNE)
+    M1 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
+    M2 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT)
+    M3 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET)
+    M4 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk)
+    M5 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk +
+                          FamHistDiab)
[…]

### 2016-09-06 doc/manuscript.Rmd
Edited methods section in manuscript

@@
         - Look over the entire model (see which variables are significant, etc)
+        - Sensitivity analysis with Carb intake? And with adiponectin? (which is
+        involved in TAG metabolism)
     - PLS with ISI and HOMA-IR
@@
 
-Participants from London and Toronto, Canada, were recruited into the 
-Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
-for recruitment into PROMISE required having one or more risk factors for type 2
-diabetes mellitus, including obesity, hypertension, family history of diabetes, 
-and/or a history of gestational diabetes or birth of a macrosomic infant. 
-Participants aged 30 years and older (n=736) attended the baseline visit
-between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
-examination visits completed to date (2004-2006, 2007-2009, and 2010-2013). 
-Participants are contacted annually by telephone. The
-current study sample used data on participants who did not have diabetes at baseline,
-who returned for one or more of the follow-up examinations, and who had samples
-available for fatty acid measurements (n=477). A diagram of the sample size at
-each visit is shown in `r cite_sf('consort')`. At each examination,
-participants undergo metabolic characterization, anthropometric measurements,
-and questionnaires on lifestyle and sociodemographics. Research ethics approval
-was obtained from Mount Sinai Hospital and the University of Western Ontario,
-and all participants provided written informed consent.
-Data collection methods were standardized across the 2 centres and research
-nurses were centrally trained.
+{{ Use an external source via DOI for reference to these methods? }}
 
-## Blood measure assessments
+Recruitment for the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
+cohort baseline visit took place between 2004-2006 from London and Toronto, 
+Canada. Individuals were selected to participant if they met the eligibility 
+criteria of having one or more risk factors for type 2 diabetes mellitus, 
+including obesity, hypertension, family history of diabetes, and/or a history of
+gestational diabetes or birth of a macrosomic infant. A total of 736 individuals
+attended the baseline visit. Subsequent examinations occurred every three years,
+with three examination visits completed to date (2004-2006, 2007-2009, and
+2010-2013). Participants are contacted annually by telephone. The current study
+sample used data on participants who did not have diabetes at baseline, who
+returned for one or more of the follow-up examinations, and who had samples
+available for fatty acid measurements (n=477; see the CONSORT diagram in `r
+cite_sf('consort')`. Metabolic characterization, anthropometric measurements,
+and questionnaires on lifestyle and sociodemographics are collected at each
+examination visit. Research ethics approval was obtained from Mount Sinai
+Hospital and the University of Western Ontario, and all participants provided
+written informed consent. Data collection methods were standardized across the 2
+centres and research nurses were centrally trained.
 
-At each examination, an 8-12 hour fasting blood sample was drawn from each
-participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
-minute and 2 hour blood draw. All blood samples were processed and frozen at
--70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
+## Metabolic characterization
+
+Participants completed an 8-12 hour 75g oral glucose tolerance test (OGTT) at 
+each examination visit, with blood samples during at fasting, 30 min, and 2 hr 
+post-glucose load. Samples were subsequently processed and frozen at -70°C.
+Alanine aminotransferase (ALT) was measured using standard laboratory 
 procedures. Cholesterol, HDL, and clinically-measured triacylglycerides (TAG) 
@@
 ON).  Both insulin and glucose were derived from the OGTT at fasting, 30 minute,
-and 2 hour time points. Specific insulin was measured using the Elecsys 1010
+and 2 hour time points. Specific insulin was measured with the Elecsys 1010 
 (Roche Diagnostics, Basel, Switzerland) immunoassay analyzer and 
-electrochemiluminescence immunoassay. This assay shows 0.05% cross-reactivity to
+electrochemiluminescence immunoassay, which shows 0.05% cross-reactivity to 
 intact human pro-insulin and the Des 31,32 circulating split form (Linco Res. 
-Inc), and has a coefficient of variation (CV) of 9.3%. Glucose was determined
+Inc) and has a coefficient of variation (CV) of 9.3%. Glucose was determined 
 using an enzymatic hexokinase (Roche Modular, Roche Diagnostics) with a 
-detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The inter-assay %CV is <1.1%
-and intra-assay %CV is < 1.9%.  All assays were performed at the Banting and
-Best Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose
-(IFG), impaired glucose tolerance (IGT), and diabetes were categorized using the
-2006 WHO criteria [@WHO2006].
+detection range of 0.11 to 41.6 mmol/L and an inter-assay %CV of <1.1% and an
+intra-assay %CV of < 1.9%.  All assays were performed at the Banting and Best
+Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose (IFG),
+impaired glucose tolerance (IGT), and diabetes were categorized using the 2006
+WHO criteria [@WHO2006].
 
@@
 stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
-of heptadecanoic acid (17:0) was added to the serum as an internal standard
-prior to extracting total lipids according to the method of Folch [@Folch1957a].
-Each serum lipid fraction---non-esterified fatty acids (NEFA), cholesteryl
-ester, phospholipid, and TAGFA---was isolated using thin layer chromatography;
-each fraction was visualized under UV light after lightly spraying with
-8-anilino-1-naphthalene sulfonic acid (0.1% wt/vol) and then converted to fatty
-acid methyl esters with 14% boron trifluoride in methanol at 100°C for 1 h.
-Fatty acid methyl esters were separated and quantified using a Varian-430 gas
-chromatograph (Varian, Lake Forest, CA, USA) equipped with a Varian Factor Four
-capillary column and a flame ionization detector, which were injected in
-splitless mode. Fatty acid concentrations (nmol/ml) were calculated by
-proportional comparison of gas chromatography peak areas to that of the internal
-standards [@Nishi2014a]. There were 22 fatty acids measured in the TAGFA
-fraction. Given their diverse biology, as well as the complexity of the
-analyses, findings for other lipid fractions in this cohort are reported
-separately (see ref @Johnston2016a and for the analysis of the phospholipid and
-cholesteryl ester fractions and {{next citation}} for the analysis of the NEFA
-fraction).
+of {{ confirm with Richard }} heptadecanoic acid (17:0) was added as an internal
+standard prior to extracting total lipids according to the method of Folch 
+[@Folch1957a]. Each serum lipid fraction---non-esterified fatty acids (NEFA), 
+cholesteryl ester, phospholipid, and TAGFA---was isolated using thin layer 
+chromatography. Fatty acid methyl esters were separated and quantified using a 
+Varian-430 gas chromatograph (Varian, Lake Forest, CA, USA) equipped with a 
+Varian Factor Four capillary column and a flame ionization detector. Fatty acid 
+concentrations (nmol/ml) were calculated by proportional comparison of gas
+chromatography peak areas to that of the internal standards [@Nishi2014a]. There
+were 22 fatty acids measured in the TAGFA fraction. Given that each lipid
+fraction differs in biology and given the complexity of the analyses, findings
+for other lipid fractions in this cohort are reported separately (see ref
+@Johnston2016a and for the analysis of the phospholipid and cholesteryl ester
+fractions and {{next citation}} for the analysis of the NEFA fraction).
 
-## Outcome variables
+## Anthropometrics and sociodemographics
+
+Height, weight, and waist circumference (WC) were measured at all clinic 
+examinations using standard procedures.  WC was measured at the natural waist, 
+defined as the narrowest part of the torso between the umbilicus and the xiphoid
+process. BMI was calculated by dividing weight (kg) by height (m) squared.
+Questionnaires administered at each examination determined sociodemographics.
+A version of the Modifiable Activity Questionnaire (MAQ) [@Kriska1990]
+determined estimated physical activity. The MAQ collects information on leisure
+and occupational activity, including intensity, frequency, and duration, over
+the past year. Each reported activity from the MAQ was weighted by its
+metabolic intensity allowing for the estimation of MET-hours per week.
+
+## Variable calculation and statistical analysis
 
@@
 
-## Anthropometrics and sociodemographics
-
-Height, weight, and waist circumference (WC) were measured at all clinic 
-examinations using standard procedures.  WC was measured at the natural waist, 
-defined as the narrowest part of the torso between the umbilicus and the xiphoid
-process. BMI was calculated by dividing weight (kg) by height (m) squared.
-Sociodemographic information, including age, sex, and ethnicity, were determined
-using questionnaires administered at each examination. In the lifestyle
-questionnaire, physical activity was determined using a version of the
-Modifiable Activity Questionnaire (MAQ) [@Kriska1990].  The MAQ collects 
-information on leisure and occupational activity, including intensity, 
-frequency, and duration, over the past year.  Each reported activity from the 
-MAQ was weighted by its metabolic intensity allowing for the estimation of 
-MET-hours per week.
-
-## Statistical analysis
-
 The primary outcome variables for this analysis were 
@@
 
-For the primary analysis, generalized estimating equation (GEE) models 
-[@Zeger1986a] were used to determine the longitudinal associations between the 
-outcome variables and the predictor variables. Given the longitudinal design 
-an auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE models, 
-though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
-is well suited to longitudinal cohort studies given it's capacity to handle
-missed visits. The predictor variables and continuous covariates were scaled
-(mean centered and standardized). The TAGFA variables were classified as
-*time-independent* (held constant) as they were measured only at the baseline
-visit, while the outcome variables and covariates were set as *time-dependent*.
-No imputation was conducted on missing values.
+Generalized estimating equation (GEE) models [@Zeger1986a] were used in the
+primary analysis to determine the longitudinal associations between the outcome
+variables and the predictor variables. An auto-regressive of order 1 (AR1) 
+correlation matrix was chosen for the GEE models as it had the best model fit
+and given the longitudinal design, though other matrices (eg. exchangeable) had
+similar fit (data not shown). GEE is well suited to longitudinal cohort studies
+given it's capacity to handle missed visits. The predictor variables and
+continuous covariates were scaled (mean centered and standardized). The TAGFA
+variables were classified as *time-independent* (held constant) as they were
+measured only at the baseline visit, while the outcome variables and covariates
+were set as *time-dependent*. No imputation was conducted on missing values.
 
-Covariate selection was based on previous literature, directed acyclic graph 
-[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
+Covariates were selected based on previous studies, from directed acyclic graph 
+[@Greenland1999a] recommendations, and from quasi-likelihood information criterion 
 (QIC) {{citation?}}. `r cite_st('qic')` shows the covariates compared using QIC. While the
@@
 outcome_chg <- calc_outcome_changes()
-calc_discr_npct(ds0yr$Sex, 'Female')
-calc_discr_npct(ds0yr$Ethnicity, 'European')
+female <- calc_discr_npct(ds0yr$Sex, 'Female')
+euro <- calc_discr_npct(ds0yr$Ethnicity, 'European')
 estci <- calc_gee_estci(gee_df)
@@
 Basic characteristics/demographics
+
+`r tab_basic` shows some basic character
 
@@
 
+[@Lee2015a] 16:1n-7 may be good marker of liver DNL.
+
 # Acknowledgements
@@
 
+```{r table_qic}
+table_qic_main()
+```
+
 ```{r fig_gee}

### 2016-09-14 R/calculate.R
Made calc_cor output easier to extract

@@ -146,7 +146,8 @@ calc_cor <- function(data = cor_df) {
             list(
                 mean = dplyr::summarize(high_cor, mean = format_rounding(mean(Correlations))),
                 rng = dplyr::summarize(high_cor, rng = paste0(min(Correlations),
-                                                              ' to ', max(Correlations))),
+                                                              ' to ', max(Correlations))) %>%
+                    tidyr::spread(Vars1, rng),
                 fa = table(low_cor$Vars1, low_cor$Vars2) %>%
                     tibble::as_data_frame() %>%
                     dplyr::group_by(Var2) %>%

### 2016-09-14 R/prep_data.R
Typo with extracting data for dys

@@ -31,7 +31,7 @@ prep_dys_data <- function(data) {
                                                     DetailedConvert))
                 )
         ) %>%
-        dplyr::filter(!is.na(TotalNE))
+        dplyr::filter(!is.na(TotalTG))
 
     return(dysgly.data)
 }

### 2016-09-14 R/calculate.R
Added mean estimates/outcome for GEE calc

@@ -114,21 +114,27 @@ calc_gee_ave <- function(results) {
 #'
 #' @export
 calc_gee_estci <- function(results) {
-    results %>%
+    b_ci <- results %>%
         dplyr::mutate(
-            e = paste0(
-                '(beta: ',
-                format_rounding(estimate),
-                ', CI: ',
+            est = format_rounding(estimate),
+            ci = paste0(
                 format_rounding(conf.low),
                 ' to ',
-                format_rounding(conf.high),
-                ')'
-            )
+                format_rounding(conf.high)
+            ),
+            eci = paste0('(beta: ', est, ', CI: ', ci, ')')
         ) %>%
         dplyr::filter(p.value <= 0.05) %>%
         dplyr::arrange(Yterms, dplyr::desc(estimate)) %>%
-        dplyr::select(Yterms, Xterms, e)
+        dplyr::select(Yterms, Xterms, est, eci)
+
+    outcome_b <- results %>%
+        dplyr::filter(p.value <= 0.05) %>%
+        dplyr::mutate(dir = ifelse(estimate < 0, 'Neg', 'Pos')) %>%
+        dplyr::group_by(Yterms, unit, dir) %>%
+        dplyr::summarise(rng = aide::min_max(estimate))
+
+    list(b_ci = b_ci, outcome_b = outcome_b)
 }
 
 #' Extracts and computes various information from the correlation results.

### 2016-09-14 R/pls.R
For calc cor pls, pasted together r and p

@@ -66,7 +66,8 @@ calc_pred_corr <- function(model, test, ncomps = 1) {
         stats::model.frame(formula(model), data = test))
         )
     corr <- broom::tidy(stats::cor.test(predicted, measured))[c(1, 3)]
-    list(r = format_rounding(corr[1], 2),
-         p = format_p(corr[2]))
+    r <- format_rounding(corr[1], 2)
+    p <- format_p(corr[2])
+    list(r = r, p = p, r_p = paste0('r=', r, ', p', p))
 
 }

### 2016-09-14 doc/manuscript.Rmd
Added CONSORT fig in manuscript + image file
Added results to the TAGFA manuscript

@@
 
+- Run PLS with just those FA that loaded strongly to see the predictive ability.
+
 Outline:
@@
 - Primary analysis:
+    - Look over medications, how they influence the TAGFA
     - Heatmap
     - GEE
+        - SA for the large outlier for the TAGFA (see dist plot)
         - Include # of both uncorrected and corrected (FDR) significant associations
@@
         - dysgly with clinical TAG vs the 'best' fatty acid to compare to
+
+- Make sure to include comments from NEFA paper
+- Add stats to basic char table... (don't really want to. But I should ..?)
 
@@
 ```{r analyze_gee, cache=TRUE}
-gee_df <- analyze_gee()
+gee_adj_df <- analyze_gee()
+gee_unadj_df <- analyze_gee(covars = "VN")
 ```
@@
 euro <- calc_discr_npct(ds0yr$Ethnicity, 'European')
-estci <- calc_gee_estci(gee_df)
+fhd <- calc_discr_npct(ds0yr$FamHistDiab, 'Yes')
+est0 <- calc_gee_estci(gee_unadj_df)
+est1 <- calc_gee_estci(gee_adj_df)
 fa_cor <- calc_cor()
@@
 
-Basic characteristics/demographics
+`r tab_basic` shows some basic characteristics of the PROMISE cohort. There were
+`r female$npct` females and `r euro$npct` who had European-ancestry, with a mean
+age in years of `r aide::ave_sd(ds0yr$Age)` and a mean BMI
+of `r aide::ave_sd(ds0yr$BMI)`. Given the eligibility criteria, the majority of
+participants (`r fhd$npct`) had a family history of diabetes.
+Between the baseline visit and the 6-year visit in this sample, insulin
+sensitivity and beta-cell function measures had a significant median decline
+between `r outcome_chg$chg` (p`r outcome_chg$p`; n=`r outcome_chg$n`). There 
+were `r dys$dm` and `r dys$pre` incident cases of diabetes and
+pre-diabetes (IFG and IGT), respectively. {{ confirm this is incidence }}
 
-`r tab_basic` shows some basic character
-
-The outcome variables significantly decreased over the 6 years by a median 
-`r outcome_chg$chg` (p`r outcome_chg$p`; n=`r outcome_chg$n`).
-
-`r fig_tagfa` shows the distribution of each TAGFA (see `r stab_tagfa` for mean
-and SD {{ confirm }}). Three TAGFA contributed 
-`r format_rounding(sum(tagfa$pct))`% to the total TAG concentration:
-`r tagfa$c[1]`; `r tagfa$c[2]`; and, `r tagfa$c[3]`.
-
+`r fig_tagfa` shows the composition of each FA in the TAG fraction (see 
+`r stab_tagfa` for raw numerical values of the mean and SD {{ confirm }}). 
+Three TAGFA contributed `r format_rounding(sum(tagfa$pct))`% to the total 
+TAG concentration: `r tagfa$c[1]`; `r tagfa$c[2]`; and, `r tagfa$c[3]`.
 `r fig_corr` shows a heatmap of the correlation of individual TAGFA with the
@@
 As expected, nearly all TAGFA had very strong positive correlations (r=
-`r fa_cor$rng$TAG`) with clinically-measured TAG. There was also moderate
+`r fa_cor$rng$TAG`) with clinically-measured TAG and moderate positive 
+correlations with WC (r=`r fa_cor$rng$Waist`). There was also moderate
 negative correlations with HDL (r=`r fa_cor$rng$HDL`). For the outcome variables,
@@
 
-Main effect
+There were `r nrow(dplyr::filter(gee_unadj_df, p.value <= 0.05))` FDR-corrected 
+significant associations in the unadjusted GEE model (see {{ `r fig_gee_unadj` 
+}}) and `r nrow(dplyr::filter(gee_adj_df, p.value <= 0.05))` FDR-corrected
+significant associations in the adjusted GEE model as shown in `r fig_gee_adj`. 
+The majority of associations with beta-cell function measures were attenuated
+after full model adjustment, while nearly all associations with insulin
+sensitivity remained significant. {{ confirm }} Subsequent analysis revealed
+that the attenuation with beta-cell function was due primarily to adjustment for
+{{ waist? ALT? }}.
 
-There were {{ ## }} significant associations in the unadjusted GEE model shown
-in {{ `r sfig_gee` }}. Only a few associations were attenuated after full model
-adjustment primarily with the beta-cell function outcomes, as shown in 
-`r fig_gee`. As a concentration, nearly all TAGFA had a strong negative
-association on 1/HOMA-IR and ISI, and a few had strong negative associations
-with IGI/IR and ISSI-2. As a mol%, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7)
-had negative associations with 1/HOMA-IR and ISI (between {{ ## }}% lower
-insulin sensitivity for every SD increase in the TAGFA), while several more TAGFA had
-positive associations with 1/HOMA-IR and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9,
-18:2n-6, 20:2n-6, 20:4n-6, and 22:5n-3) predicting a {{ ## }}% higher insulin
-sensitivity for every SD increase in the TAGFA. 
-One TAGFA, 20:2n-6, had a very strong positive association with the
-insulin sensitivity measures, with a {{ ## }}% higher insulin sensitivity for
-every SD increase in 20:2n-6.
+As a concentration, nearly all TAGFA had a strong negative association on
+1/HOMA-IR {{ change to HOMA2-IS }} and ISI (estimates ranging from 
+`r est1$outcome_b$rng[3]` and `r est1$outcome_b$rng[7]`, respectively), and a
+few had strong negative associations with IGI/IR and ISSI-2 (estimates ranging 
+from `r est1$outcome_b$rng[10]` and `r est1$outcome_b$rng[13]`, respectively).
+As a mol%, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7) had negative
+associations with 1/HOMA-IR and ISI (between `r est1$outcome_b$rng[1]` and 
+`r est1$outcome_b$rng[5]`%, respectively, lower insulin sensitivity for every SD
+increase in the TAGFA), while several more TAGFA had positive associations with
+1/HOMA-IR and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9, 18:2n-6, 20:2n-6, 20:4n-6,
+and 22:5n-3) predicting between `r est1$outcome_b$rng[2]` and 
+`r est1$outcome_b$rng[6]`%, respectively, higher insulin sensitivity for every
+SD increase in the TAGFA. One TAGFA, 20:2n-6, had a very strong positive
+association with the insulin sensitivity measures, with a `r est1$b_ci$est[1]`
+to `r est1$b_ci$est[36]`% higher insulin sensitivity for every SD increase in 
+20:2n-6.
 
 Both clinically-measured TAG and total TAGFA concentration had very strong
-negative associations with all outcome variables.
-
-There were no significant interactions by time, sex, or ethnicity for any of the
-TAGFA (data not shown).
+negative associations with all outcome variables. There were no significant
+interactions by time, sex, or ethnicity for any of the TAGFA (data not shown).
 
@@
 
-Accuracy of predictive ability of PLS components with real measure {{use test maybe?}}.
-Correlation of pred vs real in test set.
-
-The predictive ability of the PLS analysis using the TAGFA composition
-
-`r fig_pls`
-
-The correlation between the predicted values of the outcome variables from the
-PLS analysis compared to the actual value was `r cor_pls_homa$r` 
-(p`r cor_pls_homa$p`) for 1/HOMA-IR and `r cor_pls_isi$r` (p`r cor_pls_isi$p`) 
-for ISI, suggested TAGFA composition has moderate predictive ability for 
-insulin sensitivity measures.
-
-`r calc_pred_corr()`
+Clustering analysis using PLS, shown in `r fig_pls`, revealed that 14:0, 14:1n-7,
+16:0, and 16:1n-7 loaded strongly (meaning the FA fell between the dashed and
+solid circles) and negatively on 1/HOMA-IR and ISI, suggesting this cluster of
+TAGFA tracks together with lower insulin sensitivity. The TAGFA 20:2n-6,
+20:5n-3, 22:5n-3, and 22:6n-3 loaded positively on both insulin sensitivity 
+measures. No other TAGFA loaded strongly, and the results of which remained
+after running the PLS model on the testing set. The predictive ability of the 
+PLS model, determined by correlating the predicted outcome values with the actual 
+values, were good (1/HOMA-IR: `r cor_pls_homa$r_p`; ISI: `r cor_pls_isi$r_p`).
 
@@
 
+![`r fig_consort`](../img/flowDiagramSample.png)
+
 ```{r table_basic}
@@
 ```{r fig_gee}
-plot_gee_main(gee_df)
+plot_gee_main(gee_unadj_df)
+plot_gee_main(gee_adj_df)
 ```
@@
 ```{r table_gee_main}
-table_gee_main(gee_df)
+table_gee_main(gee_unadj_df)
+table_gee_main(gee_adj_df)
 ```

### 2016-09-16 doc/manuscript.Rmd
Rearranged tabs and figs, put placeholoder captions in

@@
 cite_sf <- pryr::partial(sfig, display = 'cite')
+
 tab <- captioner::captioner(prefix = 'Table')
@@
 
-fig_gee <- fig('gee', 'caption here')
+tab_qic <- tab('qic', 'caption here')
+tab_basic <- tab('basic', 'caption here')
+tab_tagfa <- tab('tagfa', 'caption here')
+tab_gee <- tab('gee', 'caption here')
+
 fig_consort <- fig('consort', 'caption here')
+fig_tagfa <- fig('tagfa', 'caption here')
+fig_heatmap <- fig('heatmap', 'caption here')
+fig_gee_unadj <- fig('gee_unadj', 'caption here')
+fig_gee_adj <- fig('gee_adj', 'caption here')
+fig_pls <- fig('pls', 'caption here')
 ```
@@
 ```{r analyze_gee, cache=TRUE}
+gee_unadj_df <- analyze_gee(covars = "VN")
 gee_adj_df <- analyze_gee()
-gee_unadj_df <- analyze_gee(covars = "VN")
 ```
@@
 
-# Tables and figures
+# Tables
 
-![`r fig_consort`](../img/flowDiagramSample.png)
+```{r table_qic}
+table_qic_main()
+```
 
@@
 
-```{r fig_dist}
-plot_tagfa()
+```{r table_tagfa}
+table_tagfa()
 ```
 
-```{r table_dist}
-table_tagfa()
+```{r table_gee}
+table_gee_main(gee_unadj_df)
+table_gee_main(gee_adj_df)
+```
+
+# Figures
+
+![`r fig_consort`](../img/flowDiagramSample.png)
+
+```{r fig_tagfa}
+plot_tagfa()
 ```
@@
 
-```{r table_corr}
-#table_corr(cor_results)
+```{r fig_gee_unadj}
+plot_gee_main(gee_unadj_df)
 ```
 
-```{r table_qic}
-table_qic_main()
-```
-
-```{r fig_gee}
-plot_gee_main(gee_unadj_df)
+```{r fig_gee_adj}
 plot_gee_main(gee_adj_df)
-```
-
-```{r table_gee_main}
-table_gee_main(gee_unadj_df)
-table_gee_main(gee_adj_df)
 ```

### 2016-09-20 doc/manuscript.Rmd
Detailed PLS results more (incl BCF) + edit methods.

@@
 - Run PLS with just those FA that loaded strongly to see the predictive ability.
+- SA with/without NEFA
+- SA with/without WC
 
@@
     - Look over medications, how they influence the TAGFA
-    - Heatmap
     - GEE
@@
     - PLS with ISI and HOMA-IR
-        - CV for determining number of comps (best is 2)
-        - Loadings (of TAGFA) (tho all are the same, use HOMA/ISI. BCF have
-        lower corr between measured and predicted) (Use COrr of both components
-        (1 and 2))
-        - Prediction (of predicted vs real) (in the pls.rmd file, not in
-        manuscript (or maybe in ESM?))
         - Use extracted PLS components in GEE (?)
-        - Train and test (CV) to see how it predicts in a new dataset (ie. a
-        future visit) (done. In pls.rmd file. As ESM?)
 - Secondary analysis:
-    - PLS-DA (dysgly)
-        - Train and test (CV)
-        - Loadings (of TAGFA) 
-        - Groupings
-        - Prediction (misclassification)
     - AROC
@@
 pls_isi_t <- analyze_pls(y = 'lISI')
+pls_igi_t <- analyze_pls(y = 'lIGIIR')
+pls_issi2_t <- analyze_pls(y = 'lISSI2')
 
@@
 against gold standard measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].
+Specific formulas for each OGTT-derived measure can be found in the {{
+Supplementary Material }}.
 
-The primary outcome variables for this analysis were 
-1/HOMA-IR, ISI, IGI/IR, and ISSI-2, which were log-transformed for
-the statistical modeling. The primary predictor variables for this analysis were
-22 individual TAGFA as mole percent (mol%) of the total fraction and as
-concentration (nmol/mL). Pearson correlation coefficients were computed to
-assess the relationships of individual TAGFA with other continuous variables.
-
-Of note is clinically measured TAG was only quantified on a portion of the
-6-year visit (n = ~120 {{ find true number }}).
+The primary outcome variables for this analysis were 1/HOMA-IR, ISI, IGI/IR, and
+ISSI-2, which were log-transformed for the statistical modeling. The primary
+predictor variables for this analysis were 22 individual TAGFA as mole percent
+(mol%) of the total fraction and as concentration (nmol/mL). Clinically-measured
+TAG is also included as a primary predictor to serve as a comparison against the
+TAGFA composition. A limitation with the clinically-measured TAG is that it was
+only quantified on a portion of the 6-year visit (n = ~120 {{ find true number
+}}). Pearson correlation coefficients were computed to assess the relationships
+of individual TAGFA with other continuous variables.
 
@@
 cor_pls_isi <- calc_pred_corr(pls_isi_t, pls_isi_t$test_data)
+cor_pls_igiir <- calc_pred_corr(pls_igiir_t, pls_igiir_t$test_data)
+cor_pls_issi2 <- calc_pred_corr(pls_issi2_t, pls_issi2_t$test_data)
 ```
@@
 
-Clustering analysis using PLS, shown in `r fig_pls`, revealed that 14:0, 14:1n-7,
-16:0, and 16:1n-7 loaded strongly (meaning the FA fell between the dashed and
-solid circles) and negatively on 1/HOMA-IR and ISI, suggesting this cluster of
-TAGFA tracks together with lower insulin sensitivity. The TAGFA 20:2n-6,
-20:5n-3, 22:5n-3, and 22:6n-3 loaded positively on both insulin sensitivity 
-measures. No other TAGFA loaded strongly, and the results of which remained
-after running the PLS model on the testing set. The predictive ability of the 
-PLS model, determined by correlating the predicted outcome values with the actual 
-values, were good (1/HOMA-IR: `r cor_pls_homa$r_p`; ISI: `r cor_pls_isi$r_p`).
+To determine which components to use, internal CV showed that the first two
+components gave the highest amount of explained variance (data not shown).
+Confirming the findings from the GEE models, PLS analysis on the beta-cell
+function measures revealed that there was a low correlation (though significant
+at p`r cor_pls_igi$p`) between the predicted values and the actual values (r=
+`r cor_pls_igi$r` to `r cor_pls_issi2$r`), suggesting poor predictability. While 
+the clustering of TAGFA from the PLS results of the beta-cell function measures 
+were similar to the insulin sensitivity measures, only the insulin sensitivity 
+measures are presented given the low predictability.
+
+The PLS results for the insulin sensitivity (`r fig_pls`) revealed a clustering 
+of the FA 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly (meaning
+the FA fell between the dashed and solid circles) and negatively on 1/HOMA-IR
+and ISI, suggesting this cluster of TAGFA tracks together with lower insulin
+sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3, and 22:6n-3 loaded positively
+on both insulin sensitivity measures. No other TAGFA loaded strongly, the
+results of which remained after running the PLS model on the cross-validated
+testing set. The predictive ability of the PLS model, determined by correlating
+the predicted outcome values with the actual values, were good (1/HOMA-IR: 
+`r cor_pls_homa$r_p`; ISI: `r cor_pls_isi$r_p`).
 
@@
 [@Lee2015a] 16:1n-7 may be good marker of liver DNL.
+
+[@Barrows2006a] NEFA contributes 60-80% of newly synth VLDL-TAG during fasting
+and ~40% during fed. Dietary FA contribute ~25% to newly synth VLDL-TAG.
+VLDL-TAG is not cleared (lipolyzed) as much as chylomicron and so stays in
+circulation longer (hence why there is bigger TAG pool vs NEFA)
 
@@
 
+Include DAGitty image
+
 ```{r fig_tagfa}

### 2016-09-21 doc/manuscript.Rmd
Moved inline chunk up, edited/dev methods section

@@
 
+```{r inline}
+n_tag <- project_data %>% 
+    dplyr::filter(VN == 2) %>% 
+    dplyr::select(TAG) %>% 
+    stats::na.omit() %>% 
+    nrow()
+
+dys <- project_data %>% 
+    prep_dys_data() %>% 
+    calc_convert_dys()
+
+tagfa <- calc_tagfa_percent()
+outcome_chg <- calc_outcome_changes()
+female <- calc_discr_npct(ds0yr$Sex, 'Female')
+euro <- calc_discr_npct(ds0yr$Ethnicity, 'European')
+fhd <- calc_discr_npct(ds0yr$FamHistDiab, 'Yes')
+est0 <- calc_gee_estci(gee_unadj_df)
+est1 <- calc_gee_estci(gee_adj_df)
+fa_cor <- calc_cor()
+cor_pls_homa <- calc_pred_corr(pls_homa_t, pls_homa_t$test_data)
+cor_pls_isi <- calc_pred_corr(pls_isi_t, pls_isi_t$test_data)
+cor_pls_igiir <- calc_pred_corr(pls_igiir_t, pls_igiir_t$test_data)
+cor_pls_issi2 <- calc_pred_corr(pls_issi2_t, pls_issi2_t$test_data)
+```
+
 {{ Use an external source via DOI for reference to these methods? }}
@@
 returned for one or more of the follow-up examinations, and who had samples
-available for fatty acid measurements (n=477; see the CONSORT diagram in `r
-cite_sf('consort')`. Metabolic characterization, anthropometric measurements,
+available for fatty acid measurements (n=477; see the CONSORT diagram in 
+`r cite_sf('consort')`. Metabolic characterization, anthropometric measurements,
 and questionnaires on lifestyle and sociodemographics are collected at each
@@
 TAGFA composition. A limitation with the clinically-measured TAG is that it was
-only quantified on a portion of the 6-year visit (n = ~120 {{ find true number
-}}). Pearson correlation coefficients were computed to assess the relationships
-of individual TAGFA with other continuous variables.
+only quantified on a portion of the participants at the 6-year visit (n = `r n_tag`).
+Pearson correlation coefficients were computed to assess the relationships of
+individual TAGFA with other continuous variables.
 
@@
 primary analysis to determine the longitudinal associations between the outcome
-variables and the predictor variables. An auto-regressive of order 1 (AR1) 
-correlation matrix was chosen for the GEE models as it had the best model fit
-and given the longitudinal design, though other matrices (eg. exchangeable) had
-similar fit (data not shown). GEE is well suited to longitudinal cohort studies
-given it's capacity to handle missed visits. The predictor variables and
-continuous covariates were scaled (mean centered and standardized). The TAGFA
-variables were classified as *time-independent* (held constant) as they were
-measured only at the baseline visit, while the outcome variables and covariates
-were set as *time-dependent*. No imputation was conducted on missing values.
+variables and the predictor variables. Given the longitudinal design, an
+auto-regressive of order 1 (AR1) working correlation matrix was chosen for the GEE
+models as it had the best model fit when assessed using quasi-likelihood
+information criteria (QIC), though other matrices (eg. exchangeable)
+had similar fit (data not shown). GEE is well suited to longitudinal cohort 
+studies given it's capacity to handle missed visits. The predictor variables and
+continuous covariates were scaled (mean-centered and standardized). The TAGFA,
+total NEFA, sex, ethnicity, and baseline age were classified as 
+*time-independent* (held constant) as they were measured only at the baseline 
+visit or do not change throughout the study, while the outcome variables and 
+remaining covariates were set as *time-dependent*. No imputation was conducted 
+on missing values.
 
-Covariates were selected based on previous studies, from directed acyclic graph 
-[@Greenland1999a] recommendations, and from quasi-likelihood information criterion 
-(QIC) {{citation?}}. `r cite_st('qic')` shows the covariates compared using QIC. While the
-final GEE model selected as best fitting differed between insulin sensitivity
-and beta-cell function measures, the differences in QIC values were less than
-10 between the models, suggesting similar fit. As such, we selected the model
-that had the fewest covariates and that had similar fit between the outcome
-measures ({{##}}). The final GEE model was adjusted for {{covariates}}. After scaling,
-log-transforming, and exponentiating, the GEE estimates are
+Covariates were selected based on the previous literature, from directed acyclic
+graph [@Greenland1999a] recommendations (see `r cite_st('dag')`), and from 
+QIC {{citation?}}. `r cite_st('qic')` shows the covariates compared using QIC.
+While the final GEE model selected as best fitting differed between insulin
+sensitivity and beta-cell function measures, the differences in QIC values were
+less than 10 between the models, suggesting similar fit. As such, we selected
+the model that had the fewest covariates and that had similar fit between the
+outcome measures ({{##}}). The final GEE model was adjusted for {{covariates}}.
+After scaling, log-transforming, and exponentiating, the GEE estimates are 
 interpreted as an expected percent difference in the outcome variable for every 
@@
 composition of the TAGFA fraction. To confirm the GEE results in a multivariate 
-environment (all TAGFA), partial least squares regression (PLS) {{citation}} was used to
-identify the patterns of TAGFA composition against insulin sensitivity and
-beta-cell function. For more detailed explanation of GEE and PLS, please see the
-{{supplemental}} Methods.
+environment (all TAGFA), partial least squares regression (PLS) {{citation}} was
+used to identify the patterns of TAGFA composition against insulin sensitivity
+and beta-cell function. 
+For more detailed explanation of GEE and PLS, please see
+the {{supplemental}} Methods.
 
@@
 
-The number of components to extract from PLS was determined by 10-fold CV.
-
-Dataset was split into a training set and a testing set to determine the
-predictability of the given PLS model. Final results presented use the full dataset.a
-
 # Results
-
-```{r inline}
-dys <- project_data %>% 
-    prep_dys_data() %>% 
-    calc_convert_dys()
-
-tagfa <- calc_tagfa_percent()
-outcome_chg <- calc_outcome_changes()
-female <- calc_discr_npct(ds0yr$Sex, 'Female')
-euro <- calc_discr_npct(ds0yr$Ethnicity, 'European')
-fhd <- calc_discr_npct(ds0yr$FamHistDiab, 'Yes')
-est0 <- calc_gee_estci(gee_unadj_df)
-est1 <- calc_gee_estci(gee_adj_df)
-fa_cor <- calc_cor()
-cor_pls_homa <- calc_pred_corr(pls_homa_t, pls_homa_t$test_data)
-cor_pls_isi <- calc_pred_corr(pls_isi_t, pls_isi_t$test_data)
-cor_pls_igiir <- calc_pred_corr(pls_igiir_t, pls_igiir_t$test_data)
-cor_pls_issi2 <- calc_pred_corr(pls_issi2_t, pls_issi2_t$test_data)
-```
 
@@
 
+The number of components to extract from PLS was determined by 10-fold CV.
+Dataset was split into a training set and a testing set to determine the
+predictability of the given PLS model. Final results presented use the full dataset.a
+
 # References

### 2016-09-21 doc/manuscript.Rmd
(very) rough draft of discussion

@@
 - SA with/without WC
+- Interactions of VN unadjusted
 
@@
 
+We found that in a Canadian population of mainly female and European ancestry, 
+the specific TAGFA composition was strongly associated with insulin sensitivity
+and moderately associated with beta-cell function. In particular, TAGFA myristic 
+acid (14:0), 7-tetradecenoic acid {{ confirm }} (14:1n-7), palmitic acid (16:0),
+and palmitoleic acid (16:1n-7) all strongly and negatively predicted lower
+insulin sensitivity. While
+
+Limitations:
+- While GEE is well suited to longitudinal datasets.. threre is a possibility
+given teh complex biology that the temporal associations can not be adequately
+analyzed using current methodologies. IN particular, an important issue si taht
+we did not quantify TAGFA at each of the three visits and so can not directed
+infer changes in TAGFA with changes in IR. However, given potential feedback
+between IR and TAGFA, this may in fact be a strength for this particular
+association given we are trying to predict how a FA at one time point can
+predict future values.
+- While PLS is a well-established technique for constructing predictive models
+of high dimensionality data strcutures, it can not adjust for potential
+confounders and other effect modifiers. Likewise, PLS is not able to handle
+longitudinal data so only the baseline visit was used in the PLS analysis.
+However, given the limitation, there results are concordant between GEE and PLS
+models.
+- Our population was predocminiately female and of euro ancestry, who were at 
+riskj for diabetes at study enrollment, and as such our results may not be
+generalizable to other populations.
+- Given the complex biology underlying the variables of interest, there many
+some residual confounding we are unaware of or have not been able to adjust for.
+However, we have taken extensive precautions for trying to ID confounders
+through the DAG models and QIC computations.
+
+Interpretation:
+
+- previous research has shown that CHO intake increases DNL {{ cite }}. In
+particular, DNL from CHO sources increases especially the 14-16 chain TAGFA as
+well as the 18 chain TAGFA. Prev lit has also shown that higher DNL can increase
+risk for DM {{ cite }} and that higher refined CHO can increase risk of DM {{
+cite }}. Our study found that the 14-16 TAGFA all strongly predicted lower
+insulin sensitivity, reinforcing this finding. While these FA also had a
+significant association with beta-cell function, the results were not as strong
+as with the IR. {{ why is it that DNL increases risk? }}
+- TAGFA that tended to positively associated with IR and BCF also tended to be
+the very long chain FA such as eicosapentaenoic acid (20:5n-3) and
+docosahexaenoic acid (22:6n-3), which have potent anti-inflammatory properties
+{{ cite }}. Biologically {{ why these FA and how help with IR }}
+- {{ compare models with and without NEFA }} NEFA have been shown contribute
+substantially to TAGFA composition and concentration. Adjusting for ...
+- Dietary FA have been shown to contribute approx 25% of the newly synth
+VLDL-TAG {{ cite }}. This suggests that our results may not entirely reflect
+diet, but also adipose stores. {{ discuss results of NEFA paper }} However, in
+an analysis of the NEFA pool {{ cite }}, we found that total NEFA, but not the
+specific composition associated with beta-cell function. There may be something
+unique about the specific composition of the TAGFA fraction that may allow it to
+enter IR tissues more easily than NEFA. It may be that VLDL receptors allow FA
+to transition to IR tissue compared to NEFA bound to albumin {{ confirm and cite
+}}.
+
+Previous lit and limits:
+
+-
+
+New things:
+
+- No longitudinal study has examined all these TAGFA using detailed metabolic,
+OGTT-derived measures. No study has used these statistical techniques, which
+were designed with these types of data in mind, and as such allows appropriate
+conclusions to be reached from this data.
+- That 14-16 FA in the TAGFA fraction predict lower insulin sensitivity and that
+this particular cluster is also generally observed with higher CHO intake.
+
 [@Rhee2011a] fasting TAG status better as discriminating between DM case vs

### 2017-01-05 doc/manuscript.Rmd
Added basic skeleton of discussion (with refs)

@@
 
-Title:
+Title: Fatty acids involved in de novo lipogenesis within triacylglycerol fraction strongly predict lower insulin sensitivity.
 
@@
 
-Limitations:
-- While GEE is well suited to longitudinal datasets.. threre is a possibility
-given teh complex biology that the temporal associations can not be adequately
-analyzed using current methodologies. IN particular, an important issue si taht
-we did not quantify TAGFA at each of the three visits and so can not directed
-infer changes in TAGFA with changes in IR. However, given potential feedback
-between IR and TAGFA, this may in fact be a strength for this particular
-association given we are trying to predict how a FA at one time point can
-predict future values.
-- While PLS is a well-established technique for constructing predictive models
-of high dimensionality data strcutures, it can not adjust for potential
-confounders and other effect modifiers. Likewise, PLS is not able to handle
-longitudinal data so only the baseline visit was used in the PLS analysis.
-However, given the limitation, there results are concordant between GEE and PLS
-models.
-- Our population was predocminiately female and of euro ancestry, who were at 
-riskj for diabetes at study enrollment, and as such our results may not be
-generalizable to other populations.
-- Given the complex biology underlying the variables of interest, there many
-some residual confounding we are unaware of or have not been able to adjust for.
-However, we have taken extensive precautions for trying to ID confounders
-through the DAG models and QIC computations.
-
 Interpretation:
@@
 }}.
+- Challenge longitudinally since muscle IR can promote DNL + TAG Flannery2012a
+- Or that post-prandial TAG is not suppressed in IR Sondergaard2012a
+- Differences with 
+- {{ confirm }} greater flux of saturated fatty acids into tissues may encourage
+lipotoxic environment {{ cite }}
+
+Limitations:
+- While GEE is well suited to longitudinal datasets.. threre is a possibility
+given teh complex biology that the temporal associations can not be adequately
+analyzed using current methodologies. IN particular, an important issue si taht
+we did not quantify TAGFA at each of the three visits and so can not directed
+infer changes in TAGFA with changes in IR. However, given potential feedback
+between IR and TAGFA, this may in fact be a strength for this particular
+association given we are trying to predict how a FA at one time point can
+predict future values.
+- While PLS is a well-established technique for constructing predictive models
+of high dimensionality data strcutures, it can not adjust for potential
+confounders and other effect modifiers. Likewise, PLS is not able to handle
+longitudinal data so only the baseline visit was used in the PLS analysis.
+However, given the limitation, there results are concordant between GEE and PLS
+models.
+- Our population was predocminiately female and of euro ancestry, who were at 
+riskj for diabetes at study enrollment, and as such our results may not be
+generalizable to other populations.
+- Given the complex biology underlying the variables of interest, there many
+some residual confounding we are unaware of or have not been able to adjust for.
+However, we have taken extensive precautions for trying to ID confounders
+through the DAG models and QIC computations.
+- TAGFA may fluctuate daily, but we had participants come in fasted in the morning
+
+[@Barrows2006a] NEFA contributes 60-80% of newly synth VLDL-TAG during fasting
+and ~40% during fed. Dietary FA contribute ~25% to newly synth VLDL-TAG.
+VLDL-TAG is not cleared (lipolyzed) as much as chylomicron and so stays in
+circulation longer (hence why there is bigger TAG pool vs NEFA)
 
 Previous lit and limits:
+- A few epidemiological studies have examined the role of the TAGFA composition
+on insulin resistance and diabetes risk. 
+- A weight loss intervention trial showed that changes in the TAGFA composition
+of 19 participants accompanied changes in the weight and insulin sensitivity over
+the 33 week period [@Schwab2008a]. Specifically, the TAGFA 14:0, 14:1, 16:0,
+16:1, 18:0, 18:3n-3, and 18:3n-6 were all lower after weight loss and correlated
+with improved insulin sensitivity.
+- In a short (24hr) feeding trial of 24 subjects [@Lee2015a], palmitoleic acid
+(16:1n-7) was shown to be strongly correlated with isotopically measured DNL and
+was 45% higher in individuals with a greater amount of hepatic fat. However, in
+that study only the TAGFA 18:0 was correlated with HOMA-IR, likely due to the
+small sample size.
+- To our knowledge, there have been two prospective studies that had a similar
+research objective as our cohort. One was a nested case-control cohort of the 
+Framingham offspring study [@Rhee2015a], which found that TAGFA with a lower
+carbon chain and less double bonds (e.g. 14:0, 16:0) had higher OR for diabetes
+after 12-years while higher carbon chain and more double bond TAGFA had a lower
+OR for diabetes. This pattern of TAGFA with diabetes risk remained the same when
+they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
+was a prospective cohort of males in Finland [@Lankinen2015a], of which they had
+TAGFA data on 831 participants after 6-years from the baseline visit. This
+cohort has similar outcomes as PROMISE, however, OGTT data was only available at
+the 6-year visit. A beta-cell function measure was computed, but only included 
+calculations up to 30 minutes of the OGTT. As with our results, they found a
+larger effect size for insulin sensitivity compared to the beta-cell function.
+The specific TAGFA that associated with the metabolic measures were similar 
+between studies, except they did not find 
+- The primary limit to these studies is the mainly small sample size and short
+study duration. Two studies had larger sample sizes of 831 [@Lankinen2015a] and 
+`r 189+189` [@Rhee2011a] and both were prospective cohorts. In addition, the
+Framingham offspring study [@Rhee2011a] did not have data on beta-cell function 
+while the Finnish cohort [@Lankinen2015a] used {{ how do I say this politely Tony? 
+They're methods and article was ... questionable... }} simple statistical 
+techniques to analysis the longitudinal data, which do not adequately take into
+consideration auto-correlation of longitudinal values and the high dimensionality
+of the TAGFA composition data structure.
 
--
+- Refs:
+    - Lankinen2015a (poor study, but is unfortunately relevant...) (god... I
+    really don't want to cite/talk about this...)
+    - Lee2015a (16:1n7 as marker of DNL) (n=24, feeding trial, 16:1n7 was 45%
+    higher in high liver fat group, direct dnl (isotope) was 2.5 fold higher in
+    high liver fat, 16:1n7 was corr (0.56) with direct DNL, TAG 18:0 corr (.50)
+    with HOMA)
+    - Zulyniak2012a (only vaccenic acid in TAG) (n=20, TLC-GC, HOMA-IR/HOMA-B,
+    two groups: normogly and hypergly. results in general showed no diff between
+    groups, only cis-vaccenic was higher in hyper, 'trend' with palmitic, did
+    not present results on whole fraction for IR)
+    - Rhee2011a (good paper on TAGFA composition) (Offspring of Framingham...,
+    nested case-control, n=189 case, n=189 control, MS/MS, OGTT, differences in
+    fasting TAGFA were more apparent than at 2-hr ogtt, more lower chain/lower
+    desat in cases + higher OR, while higher chain/desat had lower OR, lower
+    chain/desat pos corr with HOMA-IR, MS/MS (tandem) showed that 12-18 SFA and
+    MUFA inc risk of diabetes)
+    - Kotronen2009a (TAGFA composition) (n=16, 18-60, mostly female, lipidomics,
+    GC, HOMA-IR, total esterified (of which TAGFA makes up the majority) 16,
+    161n7, 181n9 pos corr with homa, 182n6 neg with homa)
+    - Schwab2008a (TAGFA composition and insulin sensitivity) (n=75 (but only
+    n=19 for lipidomics), overweight or obese, weght loss intervention (33
+    weeks), weight, FSIGT, GC, UPLC-MS, after weight loss 14, 141, 16, 161, 18,
+    183n3, 183n6 were lower, lower TAGFA that contained 14, 14:1, 16, 161, 18,
+    181 were correlated with improved Si)
+
+[@Rhee2011a] fasting TAG status better as discriminating between DM case vs
+control than 2hr-OGTT TAG
+[@Kotronen2009a] (this wasn't studied, but discussed) higher hepatic DNL may
+increase amount of 'shorter' chain FA (16:0, 14:0, etc.) into VLDL and hence
+into circulation. So higher of these FA may indicate higher DNL -> problems with
+liver function -> maybe hepatic IR -> cyclical?
+[@Lee2015a] 16:1n-7 may be good marker of liver DNL.
+
+Wilke2009a - contributions to DNL (e.g. 70% 16:0 to newly synth)
+The four FA that cluster together on the PLS also appear to be the same FA that
+make up most of DNL, especially in context of higher carb/refined carb in diet
+[@Wilke2009a].
 
@@
 this particular cluster is also generally observed with higher CHO intake.
-
-[@Rhee2011a] fasting TAG status better as discriminating between DM case vs
-control than 2hr-OGTT TAG
-
-[@Kotronen2009a] (this wasn't studied, but discussed) higher hepatic DNL may
-increase amount of 'shorter' chain FA (16:0, 14:0, etc.) into VLDL and hence
-into circulation. So higher of these FA may indicate higher DNL -> problems with
-liver function -> maybe hepatic IR -> cyclical?
-
-Discuss my PL and CE paper as well. Maybe also the NEFA one if it is published
-by then...
-
-The four FA that cluster together on the PLS also appear to be the same FA that
-make up most of DNL, especially in context of higher carb/refined carb in diet
-[@Wilke2009a].
-
-[@Lee2015a] 16:1n-7 may be good marker of liver DNL.
-
-[@Barrows2006a] NEFA contributes 60-80% of newly synth VLDL-TAG during fasting
-and ~40% during fed. Dietary FA contribute ~25% to newly synth VLDL-TAG.
-VLDL-TAG is not cleared (lipolyzed) as much as chylomicron and so stays in
-circulation longer (hence why there is bigger TAG pool vs NEFA)
 

### 2017-01-05 R/fetch_data.R
Used HOMA2 instead of HOMA-IR

@@ -12,7 +12,7 @@
 #'
 fetch_data <- function() {
     # Load the master dataset,
-    ds.prep <- PROMISE::PROMISE_data %>%
+    ds.prep <- PROMISE::PROMISE %>%
         dplyr::filter(VN %in% c(1, 3, 6)) %>%
         ## Kick out Canoers
         dplyr::filter(is.na(Canoe)) %>%
@@ -23,7 +23,8 @@ fetch_data <- function() {
     ##' Munge and wrangle the data into the final version.
     ds <- ds.prep %>%
         dplyr::select(
-            SID, VN, BMI, Waist, HOMA, ISI, IGIIR, ISSI2, TAG, LDL, HDL, Chol,
+            SID, VN, BMI, Waist, HOMA, HOMA2IR, HOMA2_S, ISI, IGIIR, ISSI2,
+            TAG, LDL, HDL, Chol,
             ALT, CRP, FamHistDiab, dplyr::matches('meds'), Age, Sex, Ethnicity,
             IFG, IGT, DM, MET, AlcoholPerWk, TobaccoUse, SelfEdu, Occupation,
             TotalTG, dplyr::matches('^tg\\d+'), Glucose0, Glucose120, TotalNE
@@ -46,6 +47,8 @@ fetch_data <- function() {
                 as.factor(),
             invHOMA = (1 / HOMA),
             linvHOMA = log(invHOMA),
+            lHOMA2IR = log(HOMA2IR),
+            lHOMA2_S = log(HOMA2_S),
             lISI = log(ISI),
             lIGIIR = log(IGIIR),
             lISSI2 = log(ISSI2),

### 2017-01-05 R/gee.R

@@ -102,9 +102,9 @@ analyze_gee <- function(data = project_data,
         dplyr::arrange(desc(order1)) %>%
         dplyr::mutate(Yterms = factor(
             Yterms,
-            levels = c('log(1/HOMA-IR)', 'log(ISI)',
+            levels = c('log(HOMA2-IR)', 'log(HOMA2-%S)', 'log(ISI)',
                        'log(IGI/IR)', 'log(ISSI-2)'),
-            labels = c('log(1/HOMA-IR)', 'log(ISI)',
+            labels = c('log(HOMA2-IR)', 'log(HOMA2-%S)', 'log(ISI)',
                        'log(IGI/IR)', 'log(ISSI-2)'),
             ordered = TRUE),
             Xterms = factor(Xterms, unique(Xterms))) %>%

### 2017-01-05 R/renaming.R

@@ -5,6 +5,8 @@ renaming_table_rows <- function(x) {
         gsub('IGIIR', 'IGI/IR', .) %>%
         gsub('ISSI2', 'ISSI-2', .) %>%
         gsub('HOMA', 'HOMA-IR', .) %>%
+        gsub('HOMA2IR', 'HOMA2-IR', .) %>%
+        gsub('HOMA2_S', 'HOMA2-%S', .) %>%
         gsub('TAG', 'TAG (mmol/L)', .) %>%
         gsub('Chol', 'Chol (mmol/L)', .) %>%
         gsub('LDL', 'LDL (mmol/L)', .) %>%
@@ -18,6 +20,8 @@ renaming_table_rows <- function(x) {
 renaming_outcomes <- function(x) {
     x %>%
         gsub('linvHOMA', 'log(1/HOMA-IR)', .) %>%
+        gsub('lHOMA2IR', 'log(HOMA2-IR)', .) %>%
+        gsub('lHOMA2_S', 'log(HOMA2-%S)', .) %>%
         gsub('lISI', 'log(ISI)', .) %>%
         gsub('lIGIIR', 'log(IGI/IR)', .) %>%
         gsub('lISSI2', 'log(ISSI-2)', .) %>%
@@ -38,7 +42,7 @@ renaming_fats <- function(x) {
 
 renaming_fraction <- function(x) {
     x %>%
-        gsub('ne', 'Non-esterified', .)
+        gsub('tg', 'Triacylglycerol', .)
 }
 
 renaming_list <- function(x) {

### 2017-01-05 R/variables.R

@@ -1,5 +1,5 @@
 tg_conc <- grep('^tg\\d+', vars, value = TRUE)
 tg_pct <- grep('^pct_tg\\d+', vars, value = TRUE)
 tg_totals <- c('TotalTG', 'TAG')
-outcomes <- c('linvHOMA', 'lISI', 'lIGIIR', 'lISSI2')
+outcomes <- c('lHOMA2_S', 'lISI', 'lIGIIR', 'lISSI2')
 covariates <- c('VN', 'Waist', 'BaseAge', 'BiEthnicity', 'Sex', 'ALT', 'MET', 'TotalNE')

### 2017-01-05 R/fetch_data.R
Used MonthsFromBaseline rather than VN for GEE

@@ -24,7 +24,7 @@ fetch_data <- function() {
     ds <- ds.prep %>%
         dplyr::select(
             SID, VN, BMI, Waist, HOMA, HOMA2IR, HOMA2_S, ISI, IGIIR, ISSI2,
-            TAG, LDL, HDL, Chol,
+            TAG, LDL, HDL, Chol, MonthsFromBaseline,
             ALT, CRP, FamHistDiab, dplyr::matches('meds'), Age, Sex, Ethnicity,
             IFG, IGT, DM, MET, AlcoholPerWk, TobaccoUse, SelfEdu, Occupation,
             TotalTG, dplyr::matches('^tg\\d+'), Glucose0, Glucose120, TotalNE

### 2017-01-05 R/variables.R

@@ -2,4 +2,4 @@ tg_conc <- grep('^tg\\d+', vars, value = TRUE)
 tg_pct <- grep('^pct_tg\\d+', vars, value = TRUE)
 tg_totals <- c('TotalTG', 'TAG')
 outcomes <- c('lHOMA2_S', 'lISI', 'lIGIIR', 'lISSI2')
-covariates <- c('VN', 'Waist', 'BaseAge', 'BiEthnicity', 'Sex', 'ALT', 'MET', 'TotalNE')
+covariates <- c('MonthsFromBaseline', 'Waist', 'BaseAge', 'BiEthnicity', 'Sex', 'ALT', 'MET', 'TotalNE')

### 2017-01-05 R/gee.R
Switched GEE fig labels to top

@@ -163,5 +163,8 @@ plot_gee_main <- function(results) {
             xlab = 'Percent difference with 95% CI in the outcomes\nfor each SD increase in fatty acid',
             ylab = 'Triacylglycerol fatty acids'
             ) +
-        graph_theme(ticks = FALSE)
+        graph_theme(ticks = FALSE) +
+        ggplot2::facet_grid(unit~Yterms, switch = 'y',
+                            scales = 'free_y',
+                            space = 'free_y')
 }

### 2017-01-06 R/gee.R
Confirmed/fixed that incident diabetes cases were not used in GEE

@@ -7,6 +7,9 @@
 #' @param data project data
 #' @export
 prep_gee_data <- function(data) {
+    data <- data %>%
+        dplyr::filter(DM != 1)
+
     no_fattyacids <- data %>%
         dplyr::select(-dplyr::matches('pct_tg\\d+|^tg\\d+'),
                       -TotalNE,-TotalTG, -BaseTAG, -lBaseTAG)

### 2017-01-06 doc/manuscript.Rmd
Forgot to add monthsfrombaseline to manuscript
Created and added PLS fig (using HOMA2)

@@
 ```{r analyze_gee, cache=TRUE}
-gee_unadj_df <- analyze_gee(covars = "VN")
+gee_unadj_df <- analyze_gee(covars = "MonthsFromBaseline")
 gee_adj_df <- analyze_gee()
@@
 # CV results are shown in pls.Rmd vignette
-pls_homa <- analyze_pls(y = 'linvHOMA', cv = FALSE)
+pls_homa2 <- analyze_pls(y = 'lHOMA2_S', cv = FALSE)
 pls_isi <- analyze_pls(y = 'lISI', cv = FALSE)
@@
 # Correlation between predicted and actual
-pls_homa_t <- analyze_pls(y = 'linvHOMA')
+pls_homa2_t <- analyze_pls(y = 'lHOMA2_S')
 pls_isi_t <- analyze_pls(y = 'lISI')
@@
 fa_cor <- calc_cor()
-cor_pls_homa <- calc_pred_corr(pls_homa_t, pls_homa_t$test_data)
+cor_pls_homa2 <- calc_pred_corr(pls_homa2_t, pls_homa2_t$test_data)
 cor_pls_isi <- calc_pred_corr(pls_isi_t, pls_isi_t$test_data)
@@
 the predicted outcome values with the actual values, were good (1/HOMA-IR: 
-`r cor_pls_homa$r_p`; ISI: `r cor_pls_isi$r_p`).
+`r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).
 
@@
 gridExtra::grid.arrange(
-    plot_pls(pls_homa) + ggplot2::ggtitle('A: 1/HOMA-IR'), 
+    plot_pls(pls_homa2) + ggplot2::ggtitle('A: HOMA2-%S'), 
     plot_pls(pls_isi) + ggplot2::ggtitle('B: ISI'), 

### 2017-01-06 R/corr.R
Fix outcomes in heatmap and add axis tilt

@@ -24,7 +24,7 @@ analyze_corr <-
         mason::scrub() %>%
         mason::polish_renaming(renaming_fats, 'Vars2') %>%
         mason::polish_renaming(function(x)
-            gsub('l(ALT|TAG|IGIIR|invHOMA|ISI|ISSI2)', '\\1', x) %>%
+            gsub('l(ALT|TAG|IGIIR|HOMA2_S|ISI|ISSI2)', '\\1', x) %>%
                 renaming_outcomes(), 'Vars1') %>%
         dplyr::mutate(
             order1 = substr(Vars2, nchar(Vars2), nchar(Vars2)),
@@ -54,5 +54,6 @@ plot_heatmap <- function(results) {
             ylab = 'Triacylglycerol fatty acids (nmol/mL)',
             number.colours = 5,
             values.size = 4) +
-        graph_theme(ticks = FALSE, legend.pos = 'right')
+        graph_theme(ticks = FALSE, legend.pos = 'right') +
+        ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, vjust = 1, hjust = 1))
 }

### 2017-01-06 R/renaming.R

@@ -25,7 +25,7 @@ renaming_outcomes <- function(x) {
         gsub('lISI', 'log(ISI)', .) %>%
         gsub('lIGIIR', 'log(IGI/IR)', .) %>%
         gsub('lISSI2', 'log(ISSI-2)', .) %>%
-        gsub('^invHOMA$', '1/HOMA-IR', .) %>%
+        gsub('^HOMA2_S$', 'HOMA2-%S', .) %>%
         gsub('^ISSI2$', 'ISSI-2', .) %>%
         gsub('^IGIIR$', 'IGI/IR', .)
 }

### 2017-01-06 R/corr.R
Created and added heatmap fig (with just TAGFA)

@@ -39,6 +39,30 @@ analyze_corr <-
                       Correlations = round(Correlations, 2))
     }
 
+#' Correlation analysis of just TAGFA.
+#'
+#' @param data Project data
+#' @param x The TAGFA
+#'
+#' @export
+analyze_corr_tagfa <-
+    function(data = project_data,
+             x = tg_pct) {
+
+    data %>%
+        dplyr::filter(VN == 0) %>%
+        mason::design('cor') %>%
+        mason::add_settings(method = 'pearson', use = 'complete.obs', hclust.order = TRUE) %>%
+        mason::add_variables('xvars', x) %>%
+        mason::construct() %>%
+        mason::scrub() %>%
+        mason::polish_renaming(renaming_fats, 'Vars2') %>%
+        mason::polish_renaming(renaming_fats, 'Vars1') %>%
+        dplyr::mutate(Vars2 = factor(Vars2, unique(Vars2)),
+                      Vars1 = factor(Vars1, unique(Vars1)))
+    }
+
+
 # Plotting ----------------------------------------------------------------
 
 #' Correlation heatmap plot.
@@ -46,13 +70,14 @@ analyze_corr <-
 #' @param results Correlation results
 #'
 #' @export
-plot_heatmap <- function(results) {
+plot_heatmap <- function(results, text = TRUE) {
      results %>%
         seer::view_heatmap(
             y = 'Vars2',
             x = 'Vars1',
             ylab = 'Triacylglycerol fatty acids (nmol/mL)',
             number.colours = 5,
+            values.text = text,
             values.size = 4) +
         graph_theme(ticks = FALSE, legend.pos = 'right') +
         ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, vjust = 1, hjust = 1))

### 2017-01-06 doc/manuscript.Rmd
Created and added heatmap fig (with just TAGFA)
Included number of participants with 1, 2, 3 visits

@@
 cor_df <- analyze_corr()
+cor_df_tagfa <- analyze_corr_tagfa()
 ```
@@
 available for fatty acid measurements (n=477; see the CONSORT diagram in 
-`r cite_sf('consort')`. Metabolic characterization, anthropometric measurements,
-and questionnaires on lifestyle and sociodemographics are collected at each
-examination visit. Research ethics approval was obtained from Mount Sinai
-Hospital and the University of Western Ontario, and all participants provided
-written informed consent. Data collection methods were standardized across the 2
-centres and research nurses were centrally trained.
+`r cite_sf('consort')`; a total of `r calc_n_for_visits()[1]` attended all three
+visits and had complete data). Metabolic characterization, anthropometric
+measurements, and questionnaires on lifestyle and sociodemographics are
+collected at each examination visit. Research ethics approval was obtained from
+Mount Sinai Hospital and the University of Western Ontario, and all participants
+provided written informed consent. Data collection methods were standardized
+across the 2 centres and research nurses were centrally trained.
 
@@
 ```
+```{r fig_heatmap_tagfa}
+plot_heatmap(cor_df_tagfa, text = FALSE)
+```
 

### 2017-01-06 R/calculate.R

@@ -162,3 +162,33 @@ calc_cor <- function(data = cor_df) {
             )
         }
 }
+
+#' Calculate the sample size at each visit collection.
+#'
+#' May not be needed.
+#'
+#' @param data Project data
+#'
+#' @export
+calc_n_at_each_visit <- function(data = project_data) {
+    data %>%
+        dplyr::select(SID, VN) %>%
+        stats::na.omit() %>%
+        dplyr::group_by(VN) %>%
+        dplyr::summarise(n = n())
+}
+
+#' Calculate the number of participants who attended one, two, or three visits.
+#'
+#' @param data Project data.
+#'
+#' @export
+calc_n_for_visits <- function(data = project_data) {
+    data %>%
+        dplyr::select(SID, VN, f.VN) %>%
+        stats::na.omit() %>%
+        tidyr::spread(f.VN, VN) %>%
+        dplyr::mutate(Visits = paste(yr0, yr3, yr6, sep = '-')) %>%
+        .$Visits %>%
+        table()
+}

### 2017-01-06 R/calculate.R
Included MonthsFromBaseline median, IQR in results

@@ -192,3 +192,16 @@ calc_n_for_visits <- function(data = project_data) {
         .$Visits %>%
         table()
 }
+
+#' Mean and SD followup time for those who attended the 6 year visit.
+#'
+#' @param data Project data
+#'
+#' @export
+calc_followup_time <- function(data = project_data) {
+    data %>%
+        dplyr::filter(VN == 2) %>%
+        dplyr::select(MonthsFromBaseline) %>%
+        dplyr::summarise(MeanFollowup = aide::ave_sd(MonthsFromBaseline / 12))
+}
+

### 2017-01-06 R/calculate.R
Clarified that most participants had a 6yr visit

@@ -205,3 +205,10 @@ calc_followup_time <- function(data = project_data) {
         dplyr::summarise(MeanFollowup = aide::ave_sd(MonthsFromBaseline / 12))
 }
 
+#' Percent of participants who attended each of the 3 visits.
+#'
+#' @export
+calc_pct_full_visits <- function() {
+    pct_val <- calc_n_for_visits()[1] / sum(calc_n_for_visits())
+    format_rounding(pct_val * 100)
+}

### 2017-01-06 doc/manuscript.Rmd
Included MonthsFromBaseline median, IQR in results
Clarified that most participants had a 6yr visit
Added handling of diabetes cases in methods

@@
 remaining covariates were set as *time-dependent*. No imputation was conducted 
-on missing values.
+on missing values. Given how the samples for TAGFA measurement were determined
+and how the outcome variables are calculated, cases of diabetes at baseline and
+followup were not included in the GEE modeling.
 
@@
 used to identify the patterns of TAGFA composition against insulin sensitivity
-and beta-cell function. 
-For more detailed explanation of GEE and PLS, please see
-the {{supplemental}} Methods.
+and beta-cell function. No prevalent diabetes cases were included in the PLS
+analysis. For more detailed explanation of GEE and PLS, please see the
+{{supplemental}} Methods.
 
@@
 
-`r tab_basic` shows some basic characteristics of the PROMISE cohort. There were
+`r tab_basic` shows some basic characteristics of the PROMISE cohort. The mean 
+follow-up time was `r calc_followup_time()`, where `r calc_pct_full_visits()`%
+of participants attended all three visits. There were
 `r female$npct` females and `r euro$npct` who had European-ancestry, with a mean

### 2017-01-09 R/calculate.R
Keep last observation for calculating median followup

@@ -200,7 +200,10 @@ calc_n_for_visits <- function(data = project_data) {
 #' @export
 calc_followup_time <- function(data = project_data) {
     data %>%
-        dplyr::filter(VN == 2) %>%
+        dplyr::arrange(SID, VN) %>%
+        dplyr::group_by(SID) %>%
+        dplyr::slice(n()) %>%
+        dplyr::ungroup() %>%
         dplyr::select(MonthsFromBaseline) %>%
         dplyr::summarise(MeanFollowup = aide::ave_sd(MonthsFromBaseline / 12))
 }

### 2017-01-09 doc/manuscript.Rmd
Wrote up figure captions

@@
 
-fig_consort <- fig('consort', 'caption here')
-fig_tagfa <- fig('tagfa', 'caption here')
-fig_heatmap <- fig('heatmap', 'caption here')
-fig_gee_unadj <- fig('gee_unadj', 'caption here')
-fig_gee_adj <- fig('gee_adj', 'caption here')
-fig_pls <- fig('pls', 'caption here')
+fig_consort <- fig('consort', 'CONSORT diagram of PROMISE participants over the 3 visits.')
+fig_tagfa <- fig('tagfa', 'Distribution of the composition of triacylglycerol fatty acids in the baseline visit of PROMISE participants (2004-2006).
+Boxplots represent the median and interquartile range of the fatty acid values.')
+fig_heatmap <- fig('heatmap', 'Pearson correlation heatmap of triacylglycerol fatty acids (nmol/mL) with continuous basic and metabolic characteristics of PROMISE participants from the baseline visit (2004-2006).
+Darker orange represents a positive correlation; darker blue represents a negative correlation.')
+fig_heatmap_tagfa <- fig('heatmap_tagfa', 'Pearson correlation heatmap of the triacylglycerol fatty acids in the PROMISE participants from the baseline visit (2004-2006).
+The correlations of fatty acids grouped using heirarchical cluster analysis; fatty acids along the x and y axis are ordered according to this analysis.
+Darker orange represents a positive correlation; darker blue represents a negative correlation.')
+fig_gee_unadj <- fig('gee_unadj', 'Time-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
+P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
+fig_gee_adj <- fig('gee_adj', 'Fully-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+Variables controlled for were follow-up time, waist circumference, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
+X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
+P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
+fig_pls <- fig('pls', 'Partial least squares (PLS) models showing the clustering of triacylglycerol fatty acids on insulin sensitivity and beta-cell function measures.
+The R value shown above the graph is the correlation between predicted and actual values, using cross-validation; a higher value suggests the fatty acids strongly predict the metabolic outcome.
+The percent explained variance of each component is shown in brackets on each axis.
+The solid line represents an explained variance of 100% while the dashed line represents an explained variance of 50%.
+Fatty acids between these lines represent variables that strongly explain the underlying structure of the data.
+See the Supplemental Methods for a description of PLS analysis and an explanation of interpreting this plot.')
 ```

### 2017-01-09 doc/manuscript.Rmd
Wrote up table captions

@@
 
-tab_qic <- tab('qic', 'caption here')
-tab_basic <- tab('basic', 'caption here')
-tab_tagfa <- tab('tagfa', 'caption here')
-tab_gee <- tab('gee', 'caption here')
+tab_qic <- tab('qic', 'Comparison of GEE model fitness for variable selection using quasi-likelihood information criteria.')
+tab_basic <- tab('basic', 'Basic characteristics of PROMISE participants at each of the 3 clinic visits.')
+tab_tagfa <- tab('tagfa', 'Concentration (nmol/mL) and relative percent (mol%) values of triacylglycerol fatty acids in PROMISE participants at the baseline visit (2004-2006).')
+tab_gee_unadj <- tab('gee_unadj', 'Raw estimates and confidence interval values for *time*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+Estimates represent a percent difference in the outcome per SD increase in the fatty acid.
+P-values were adjusted for the BH false discovery rate, with an asterisk (*) denoting a significant (p<0.05) association.')
+tab_gee_adj <- tab('gee_adj', 'Raw estimates and confidence interval values for *fully*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+Variables controlled for were follow-up time, waist circumference, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
+Estimates represent a percent difference in the outcome per SD increase in the fatty acid.
+P-values were adjusted for the BH false discovery rate, with an asterisk (*) denoting a significant (p<0.05) association.')
 

### 2017-01-09 doc/manuscript.Rmd
Assigned the captions to figures and tables

@@
 ```{r table_qic}
-table_qic_main()
+table_qic_main(caption = tab_qic)
 ```
@@
 ```{r table_basic}
-table_basic()
+table_basic(caption = tab_basic)
 ```
@@
 ```{r table_tagfa}
-table_tagfa()
+table_tagfa(caption = tab_tagfa)
 ```
@@
 ```{r table_gee}
-table_gee_main(gee_unadj_df)
-table_gee_main(gee_adj_df)
+table_gee_main(gee_unadj_df, caption = tab_gee_unadj)
+table_gee_main(gee_adj_df, caption = tab_gee_adj)
 ```
@@
 
-```{r fig_tagfa}
+```{r fig_tagfa, fig.cap=fig_tagfa}
 plot_tagfa()
@@
 
-```{r fig_heatmap}
+```{r fig_heatmap, fig.cap=fig_heatmap}
 plot_heatmap(cor_df)
 ```
-```{r fig_heatmap_tagfa}
+
+```{r fig_heatmap_tagfa, fig.cap=fig_heatmap_tagfa}
 plot_heatmap(cor_df_tagfa, text = FALSE)
@@
 
-```{r fig_gee_unadj}
+```{r fig_gee_unadj, fig.cap=fig_gee_unadj}
 plot_gee_main(gee_unadj_df)
@@
 
-```{r fig_gee_adj}
+```{r fig_gee_adj, fig.cap=fig_gee_adj}
 plot_gee_main(gee_adj_df)
@@
 
-```{r fig_pls}
+```{r fig_pls, fig.cap=fig_pls}
 gridExtra::grid.arrange(

### 2017-01-10 R/calculate.R
Changed monthsfrombaseline to yearsfrombaseline in GEE models

@@ -204,8 +204,7 @@ calc_followup_time <- function(data = project_data) {
         dplyr::group_by(SID) %>%
         dplyr::slice(n()) %>%
         dplyr::ungroup() %>%
-        dplyr::select(MonthsFromBaseline) %>%
-        dplyr::summarise(MeanFollowup = aide::ave_sd(MonthsFromBaseline / 12))
+        dplyr::summarise(MeanFollowup = aide::ave_sd(YearsFromBaseline))
 }
 
 #' Percent of participants who attended each of the 3 visits.

### 2017-01-10 R/fetch_data.R

@@ -30,6 +30,7 @@ fetch_data <- function() {
             TotalTG, dplyr::matches('^tg\\d+'), Glucose0, Glucose120, TotalNE
         ) %>%
         dplyr::mutate(
+            YearsFromBaseline = MonthsFromBaseline / 12,
             BaseTotalNE = TotalNE,
             BaseTotalTG = TotalTG,
             BaseTAG = ifelse(VN == 1, TAG, NA),

### 2017-01-10 R/variables.R

@@ -2,4 +2,4 @@ tg_conc <- grep('^tg\\d+', vars, value = TRUE)
 tg_pct <- grep('^pct_tg\\d+', vars, value = TRUE)
 tg_totals <- c('TotalTG', 'TAG')
 outcomes <- c('lHOMA2_S', 'lISI', 'lIGIIR', 'lISSI2')
-covariates <- c('MonthsFromBaseline', 'Waist', 'BaseAge', 'BiEthnicity', 'Sex', 'ALT', 'MET', 'TotalNE')
+covariates <- c('YearsFromBaseline', 'Waist', 'BaseAge', 'BiEthnicity', 'Sex', 'ALT', 'MET', 'TotalNE')

### 2017-01-10 R/renaming.R
Fixed renaming of basic char table rows

@@ -4,14 +4,15 @@ renaming_table_rows <- function(x) {
     x %>%
         gsub('IGIIR', 'IGI/IR', .) %>%
         gsub('ISSI2', 'ISSI-2', .) %>%
-        gsub('HOMA', 'HOMA-IR', .) %>%
         gsub('HOMA2IR', 'HOMA2-IR', .) %>%
         gsub('HOMA2_S', 'HOMA2-%S', .) %>%
         gsub('TAG', 'TAG (mmol/L)', .) %>%
         gsub('Chol', 'Chol (mmol/L)', .) %>%
         gsub('LDL', 'LDL (mmol/L)', .) %>%
+        gsub('ALT', 'ALT (U/L)', .) %>%
         gsub('HDL', 'HDL (mmol/L)', .) %>%
         gsub('BaseTotalTG', 'TAGFA (nmol/mL)', .) %>%
+        gsub('BaseTotalNE', 'NEFA (nmol/mL)', .) %>%
         gsub('Age', 'Age (yrs)', .) %>%
         gsub('BMI', 'BMI (kg/m^2^)', .) %>%
         gsub('Waist', 'WC (cm)', .)

### 2017-01-10 R/tables.R

@@ -13,8 +13,8 @@ table_basic <- function(data = project_data, caption = NULL) {
                       Sex = ifelse(VN == 0, as.character(Sex), NA),
                       Sex = as.factor(Sex)) %>%
         carpenter::outline_table('f.VN') %>%
-        carpenter::add_rows(c('HOMA', 'ISI', 'IGIIR', 'ISSI2'), carpenter::stat_medianIQR, digits = 1) %>%
-        carpenter::add_rows(c('ALT', 'TAG', 'Chol', 'HDL', 'BaseTotalNE', 'MET',
+        carpenter::add_rows(c('HOMA2_S', 'ISI', 'IGIIR', 'ISSI2'), carpenter::stat_medianIQR, digits = 1) %>%
+        carpenter::add_rows(c('ALT', 'TAG', 'Chol', 'HDL', 'BaseTotalTG', 'BaseTotalNE', 'MET',
                               'Age', 'BMI', 'Waist'),
                             carpenter::stat_meanSD, digits = 1) %>%
         carpenter::add_rows(c('Ethnicity', 'Sex'), carpenter::stat_nPct, digits = 0) %>%

### 2017-01-10 doc/manuscript.Rmd
Included stats method for hclust of tagfa heatmap
Included/fixed QIC table

@@
 Pearson correlation coefficients were computed to assess the relationships of
-individual TAGFA with other continuous variables.
+individual TAGFA with other continuous variables. Correlations were also
+computed for TAGFA against each other, followed by using hierarchical clustering
+analysis to identify clusters within the correlation matrix of the TAGFA
+composition.
 
@@
 ```{r table_qic}
-table_qic_main(caption = tab_qic)
+analyze_qic() %>% 
+    table_qic(caption = tab_qic)
 ```

### 2017-01-10 R/prep_data.R

@@ -35,15 +35,3 @@ prep_dys_data <- function(data) {
 
     return(dysgly.data)
 }
-
-#' Clean data output from the QIC model selection functions.
-#'
-#' @param data QIC output data.
-#'
-#' @export
-clean_qic <- function(data) {
-    data %>%
-        tibble::rownames_to_column() %>%
-        dplyr::select(Model = rowname, QIC = IC, Delta = delta) %>%
-        dplyr::mutate(id = seq_len(nrow(.)))
-}

### 2017-01-10 R/qic.R

@@ -1,0 +1,159 @@
+
+# Prepare data for QIC ----------------------------------------------------
+
+#' Prepare the data for using in the QIC model comparisons.
+#'
+#' @param data Project data
+#'
+prep_qic_data <- function(data) {
+    data %>%
+        prep_gee_data() %>%
+        dplyr::select(
+            SID,
+            VN,
+            YearsFromBaseline,
+            lISSI2,
+            lISI,
+            TotalTG,
+            TotalNE,
+            Sex,
+            Ethnicity,
+            MET,
+            ALT,
+            BaseAge,
+            Waist,
+            BMI,
+            AlcoholPerWk,
+            TobaccoUse,
+            FamHistDiab
+        ) %>%
+        stats::na.omit()
+}
+
+# Clean QIC output --------------------------------------------------------
+
+#' Clean data output from the QIC model selection functions.
+#'
+#' @param data QIC output data.
+#'
+#' @export
+clean_qic <- function(data) {
+    data %>%
+        tibble::rownames_to_column() %>%
+        dplyr::select(Model = rowname, QIC = IC, Delta = delta) %>%
+        dplyr::mutate(id = seq_len(nrow(.)))
+}
+
+# Analyze using QIC -------------------------------------------------------
+
+#' QIC output for insulin sensitivity.
+#'
+#' @param data Project data.
+#'
+#' @export
+qic_is <- function(data = project_data) {
+    # need to force into global environment because model.sel doesn't work
+    qic_data <<- data %>%
+        prep_qic_data() %>%
+        dplyr::select(-lISSI2) %>%
+        as.data.frame()
+
+    M0 <- geepack::geeglm(
+        lISI ~ TotalTG + YearsFromBaseline,
+        data = qic_data,
+        id = SID,
+        family = stats::gaussian,
+        corstr = 'ar1'
+    )
+
+    Int <- stats::update(M0, . ~ . + TotalTG:YearsFromBaseline)
+    Full <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET +
[…]

### 2017-01-10 R/tables.R

@@ -55,118 +55,3 @@ table_corr <- function(results, caption = NULL) {
         dplyr::rename(TAGFA = Vars1) %>%
         pander::pander()
 }
-
-#' Create a table of the QIC from the QIC cleaned output.
-#'
-#' @param data Cleaned QIC output.
-#' @param caption Table caption.
-#'
-#' @export
-table_qic <- function(data, caption = NULL) {
-    data %>%
-        dplyr::arrange(id) %>%
-        dplyr::select(-id) %>%
-        dplyr::mutate(QIC = format_rounding(QIC, 1),
-                      Delta = format_rounding(Delta, 1)) %>%
-        pander::pander(missing = '', caption = caption)
-}
-
-#' QIC table for manuscript.
-#'
-#' @param data Project data.
-#' @param caption Caption for table.
-#'
-#' @export
-table_qic_main <- function(data = project_data, caption = NULL) {
-    prep_qic <- data %>%
-        prep_gee_data() %>%
-        dplyr::select(
-            SID,
-            VN,
-            lISSI2,
-            lISI,
-            TotalTG,
-            TotalNE,
-            Sex,
-            Ethnicity,
-            MET,
-            ALT,
-            BaseAge,
-            Waist,
-            BMI,
-            AlcoholPerWk,
-            TobaccoUse,
-            FamHistDiab
-        ) %>%
-        stats::na.omit()
-
-    qic_is <- prep_qic %>%
-        dplyr::select(-lISSI2)
-
-    qic_bcf <- prep_qic %>%
-        dplyr::select(-lISI)
-    M0_ar_is <- geepack::geeglm(
-        lISI ~ TotalTG + VN,
-        data = qic_is,
-        id = SID,
-        family = stats::gaussian,
-        corstr = 'ar1'
-    )
-    M0 <- M0_ar_is
-    Full <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET +
-                AlcoholPerWk + FamHistDiab + TobaccoUse + TotalNE)
-    M1 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
[…]

### 2017-01-10 doc/manuscript.Rmd
Changed HOMA-IR methods to HOMA2-%S

@@
 Insulin sensitivity and beta-cell function indices were computed using the OGTT 
-glucose and insulin data. Insulin sensitivity was assessed using 1/HOMA-IR (1
-divided by HOMA-IR) [@Matthews1985] and the Insulin Sensitivity Index (ISI)
-[@Matsuda1999]. HOMA-IR largely reflects hepatic insulin resistance, while ISI
+glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S
+[@Levy1998a] using the HOMA2 Calculator and the Insulin Sensitivity Index (ISI)
+[@Matsuda1999]. HOMA largely reflects hepatic insulin resistance, while ISI
 reflects whole-body insulin sensitivity [@AbdulGhani2007]. Beta-cell function
-was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR)
+was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR [@Matthews1985] (IGI/IR)
 and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
@@
 the disposition index (but using OGTT values). Each index has been validated
-against gold standard measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].
+against gold standard measures [@Matthews1985; @Hermans1999a; @Matsuda1999; @Retnakaran2009].
 Specific formulas for each OGTT-derived measure can be found in the {{
@@
 
-The primary outcome variables for this analysis were 1/HOMA-IR, ISI, IGI/IR, and
+The primary outcome variables for this analysis were HOMA2-%S, ISI, IGI/IR, and
 ISSI-2, which were log-transformed for the statistical modeling. The primary

### 2017-01-10 doc/manuscript.Rmd
Fixed results after HOMA-IR changed to HOMA2

@@
 ```{r analyze_gee, cache=TRUE}
-gee_unadj_df <- analyze_gee(covars = "MonthsFromBaseline")
+gee_unadj_df <- analyze_gee(covars = "YearsFromBaseline")
 gee_adj_df <- analyze_gee()
@@
 sensitivity and beta-cell function measures had a significant median decline
-between `r outcome_chg$chg` (p`r outcome_chg$p`; n=`r outcome_chg$n`). There 
+between `r outcome_chg$chg` (p`r outcome_chg$p` from GEE; n=`r outcome_chg$n`). There 
 were `r dys$dm` and `r dys$pre` incident cases of diabetes and
-pre-diabetes (IFG and IGT), respectively. {{ confirm this is incidence }}
+pre-diabetes (IFG and IGT), respectively.
 
@@
 the correlations for the insulin sensitivity measures were generally higher 
-(1/HOMA-IR: r=`r fa_cor$rng['1/HOMA-IR']`, ISI: r=`r fa_cor$rng$ISI`) than 
+(1/HOMA-IR: r=`r fa_cor$rng['HOMA2-%S']`, ISI: r=`r fa_cor$rng$ISI`) than 
 for the beta-cell function measures (all r<0.30).
@@
 As a concentration, nearly all TAGFA had a strong negative association on
-1/HOMA-IR {{ change to HOMA2-IS }} and ISI (estimates ranging from 
+HOMA2-%S and ISI (estimates ranging from 
 `r est1$outcome_b$rng[3]` and `r est1$outcome_b$rng[7]`, respectively), and a
 few had strong negative associations with IGI/IR and ISSI-2 (estimates ranging 
-from `r est1$outcome_b$rng[10]` and `r est1$outcome_b$rng[13]`, respectively).
+from `r est1$outcome_b$rng[11]` and `r est1$outcome_b$rng[15]`, respectively).
 As a mol%, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7) had negative
-associations with 1/HOMA-IR and ISI (between `r est1$outcome_b$rng[1]` and 
+associations with HOMA2-%S and ISI (between `r est1$outcome_b$rng[1]` and 
 `r est1$outcome_b$rng[5]`%, respectively, lower insulin sensitivity for every SD
 increase in the TAGFA), while several more TAGFA had positive associations with
-1/HOMA-IR and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9, 18:2n-6, 20:2n-6, 20:4n-6,
+HOMA2-%S and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9, 18:2n-6, 20:2n-6, 20:4n-6,
 and 22:5n-3) predicting between `r est1$outcome_b$rng[2]` and 
@@
 of the FA 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly (meaning
-the FA fell between the dashed and solid circles) and negatively on 1/HOMA-IR
+the FA fell between the dashed and solid circles) and negatively on HOMA2-%S
 and ISI, suggesting this cluster of TAGFA tracks together with lower insulin
@@
 testing set. The predictive ability of the PLS model, determined by correlating
-the predicted outcome values with the actual values, were good (1/HOMA-IR: 
+the predicted outcome values with the actual values, were good (HOMA2-%S: 
 `r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).

### 2017-01-10 doc/manuscript.Rmd
Polish up methods

@@
 
-{{ Use an external source via DOI for reference to these methods? }}
+<!-- Use an external source via DOI for reference to these methods? -->
 
-Recruitment for the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
-cohort baseline visit took place between 2004-2006 from London and Toronto, 
+Recruitment for the baseline visit of the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
+cohort took place between 2004-2006 from London and Toronto, 
 Canada. Individuals were selected to participant if they met the eligibility 
@@
 `r cite_sf('consort')`; a total of `r calc_n_for_visits()[1]` attended all three
-visits and had complete data). Metabolic characterization, anthropometric
+visits). Metabolic characterization, anthropometric
 measurements, and questionnaires on lifestyle and sociodemographics are
@@
 Participants completed an 8-12 hour 75g oral glucose tolerance test (OGTT) at 
-each examination visit, with blood samples during at fasting, 30 min, and 2 hr 
+each examination visit, with blood samples taken during fasting, 30 min, and 2 hr 
 post-glucose load. Samples were subsequently processed and frozen at -70°C.
@@
 stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
-of {{ confirm with Richard }} heptadecanoic acid (17:0) was added as an internal
+of {{ confirm with Richard the correct name}} heptadecanoic acid (17:0) was added as an internal
 standard prior to extracting total lipids according to the method of Folch 
@@
 for other lipid fractions in this cohort are reported separately (see ref
-@Johnston2016a and for the analysis of the phospholipid and cholesteryl ester
-fractions and {{next citation}} for the analysis of the NEFA fraction).
+@Johnston2016a for the analysis of the phospholipid and cholesteryl ester
+fractions). 
+<!--and {{next citation}} for the analysis of the NEFA fraction).-->
 
@@
 graph [@Greenland1999a] recommendations (see `r cite_st('dag')`), and from 
-QIC {{citation?}}. `r cite_st('qic')` shows the covariates compared using QIC.
+QIC. `r cite_st('qic')` shows the covariates compared using QIC.
 While the final GEE model selected as best fitting differed between insulin
 sensitivity and beta-cell function measures, the differences in QIC values were
-less than 10 between the models, suggesting similar fit. As such, we selected
+less than 10 between many of the models, suggesting similar fit. As such, we selected
 the model that had the fewest covariates and that had similar fit between the
@@
 standard deviation (SD) increase in the predictor variable given the covariates 
-are held constant. Lastly for the GEE models, we tested for an interaction with 
+are held constant (including time). Lastly for the GEE models, we tested for an interaction with 
 sex, ethnicity, or time by the predictor term for each outcome variable.
@@
 composition of the TAGFA fraction. To confirm the GEE results in a multivariate 
-environment (all TAGFA), partial least squares regression (PLS) {{citation}} was
+environment (all TAGFA), partial least squares regression (PLS) was
 used to identify the patterns of TAGFA composition against insulin sensitivity
@@
 packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a] and pls
-`r packageVersion('pls')` for PLS {{cite?}}. The R code and extra analyses for
+`r packageVersion('pls')` for PLS. The R code and extra analyses for
 this manuscript is available at {{code doi}}. Results were considered

### 2017-01-10 doc/manuscript.Rmd
First draft of results

@@
 fig_consort <- fig('consort', 'CONSORT diagram of PROMISE participants over the 3 visits.')
+fig_dag <- fig('dag', 'Directed acyclic graphic output from the DAGitty online software.')
 fig_tagfa <- fig('tagfa', 'Distribution of the composition of triacylglycerol fatty acids in the baseline visit of PROMISE participants (2004-2006).
@@
 cor_pls_isi <- calc_pred_corr(pls_isi_t, pls_isi_t$test_data)
-cor_pls_igiir <- calc_pred_corr(pls_igiir_t, pls_igiir_t$test_data)
+cor_pls_igi <- calc_pred_corr(pls_igi_t, pls_igi_t$test_data)
 cor_pls_issi2 <- calc_pred_corr(pls_issi2_t, pls_issi2_t$test_data)
@@
 Covariates were selected based on the previous literature, from directed acyclic
-graph [@Greenland1999a] recommendations (see `r cite_st('dag')`), and from 
-QIC. `r cite_st('qic')` shows the covariates compared using QIC.
+graph [@Greenland1999a] recommendations (see `r cite_t('tab_dag')`), and from 
+QIC. `r cite_t('tab_qic')` shows the covariates compared using QIC.
 While the final GEE model selected as best fitting differed between insulin
@@
 
-`r tab_basic` shows some basic characteristics of the PROMISE cohort. The mean 
+`r cite_t('tab_basic')` shows some basic characteristics of the PROMISE cohort. The mean 
 follow-up time was `r calc_followup_time()`, where `r calc_pct_full_visits()`%
@@
 
-`r fig_tagfa` shows the composition of each FA in the TAG fraction (see 
-`r stab_tagfa` for raw numerical values of the mean and SD {{ confirm }}). 
+`r cite_f('fig_tagfa')` shows the composition of each FA in the TAG fraction (see 
+`r cite_t('tab_tagfa')` for a tabular presentation of the values). 
 Three TAGFA contributed `r format_rounding(sum(tagfa$pct))`% to the total 
@@
 the correlations for the insulin sensitivity measures were generally higher 
-(1/HOMA-IR: r=`r fa_cor$rng['HOMA2-%S']`, ISI: r=`r fa_cor$rng$ISI`) than 
+(HOMA2-%S: r=`r fa_cor$rng['HOMA2-%S']`, ISI: r=`r fa_cor$rng$ISI`) than 
 for the beta-cell function measures (all r<0.30).
@@
 There were `r nrow(dplyr::filter(gee_unadj_df, p.value <= 0.05))` FDR-corrected 
-significant associations in the unadjusted GEE model (see {{ `r fig_gee_unadj` 
-}}) and `r nrow(dplyr::filter(gee_adj_df, p.value <= 0.05))` FDR-corrected
-significant associations in the adjusted GEE model as shown in `r fig_gee_adj`. 
+significant associations in the unadjusted GEE model (see `r cite_f('fig_gee_unadj')`) and 
+`r nrow(dplyr::filter(gee_adj_df, p.value <= 0.05))` FDR-corrected
+significant associations in the adjusted GEE model as shown in `r cite_f('fig_gee_adj')`. 
 The majority of associations with beta-cell function measures were attenuated
@@
 association with the insulin sensitivity measures, with a `r est1$b_ci$est[1]`
-to `r est1$b_ci$est[36]`% higher insulin sensitivity for every SD increase in 
+to `r est1$b_ci$est[37]`% higher insulin sensitivity for every SD increase in 
 20:2n-6.
@@
 interactions by time, sex, or ethnicity for any of the TAGFA (data not shown).
+A tabular presentation of the GEE results is shown in `r cite_t("tab_gee_unadj")` 
+for unadjusted models and `r cite_t("tab_gee_adj")` for adjusted models.
 
@@
 were similar to the insulin sensitivity measures, only the insulin sensitivity 
-measures are presented given the low predictability.
+measures are presented given the higher predictive ability.
 
-The PLS results for the insulin sensitivity (`r fig_pls`) revealed a clustering 
+The PLS results for the insulin sensitivity (`r cite_f('fig_pls')`) revealed a clustering 
 of the FA 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly (meaning
@@
 results of which remained after running the PLS model on the cross-validated
-testing set. The predictive ability of the PLS model, determined by correlating
-the predicted outcome values with the actual values, were good (HOMA2-%S: 
-`r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).
+testing set. The PLS model for insulin sensitivity had good predictive ability, 
+determined by correlating the predicted outcome values with the actual values
+(HOMA2-%S: `r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).
 

### 2017-01-11 doc/manuscript.Rmd
First (rough) draft of intro

@@
 
-Outline:
+Hypertriglyceridemia is an established and well known factor involved in the
+dysregulation of metabolic function and subsequent negative health impacts. It
+is a component of the metabolic syndrome {{ cite }}, a strong risk factor for
+cardiovascular disease {{cite}}, and can contribute to worsening pathogenesis of
+diabetes through it's harmful health effects {{ clarify }}. Triacylglcerol (TAG)
+is a common blood measure taken during routine clinical visits and has {{
+historically?}} been a target of pharmaceutical intervention {{cite}}, though evidence on
+the {{efficacy? effectiveness?}} of such interventions are often underwhelming
+on their effects on disease {{risk?}} reduction {{cite}}.
 
-- TAG strong predictor of diabetes and CVD risk
-- Not well studied how composition plays a role
-- Limited longitudinal data
+TAG as a pharmaceutical target and as a measure of metabolic health is a single
+measure of all fatty acids within the TAG fraction. However, given the 
+increasing appreciation for the role of the specific fatty acid (FA) composition
+of various blood lipid fractions on health outcomes, reducing the complex FA 
+metabolism within the TAG fraction to a single measure may contribute to the 
+less than expected {{?}} changes in risk reduction.
+
+In additiona, the {{interplay?}} between TAG and IR is complex and has the
+potential to act as a {{feed forward?}} loop. Greater resistance to insulin,
+particularly in the liver, can result in greater accumulation of hepatic fat and
+a potential associated increased production in TAG which are secreted into
+circulation with VLDL [@Yu2012a] {{review on mechanisms of TAG, IR, etc.}} {{not
+always the case, dependent on ethnicity}}. However, higher circulating TAG can
+exacerbate or contribute to insulin resistance in various tissues, creating a
+vicious cycle {{cite}}. Few studies have been conducted that have examined the
+longitudinal effect of TAGFA composition on diabetes pathophysiology.
+
+Several prospective studies have identified to date the association of 
+hypertriglyceridemia on incident diabetes {{[@Chien2008a; @DAgostino2004a;
+@Schulze2009a]}}, but no longitudinal studies have been conducted{{another 
+word?}} on the pathophysiological factors underlying diabetes.
 
 Our objective was to examine the longitudinal role of the specific composition
-of the TAG serum fraction on OGTT-derived measures of insulin sensitivity and
-beta-cell function.
-
-NEFA may be a mediating factor in association, but that adjustment will not
-attenuate this assoc since TAGFA composition is also dictated by other sources
-for DNL (e.g. recent meal, DNL from glucose stores, etc.).
+of the serum TAG fraction on OGTT-derived measures of insulin sensitivity and
+beta-cell function in an Canadian population at risk for diabetes over 6 years.
 

### 2017-01-11 doc/manuscript.Rmd
Included DAG (from DAGitty) as Supplement

@@
 fig_consort <- fig('consort', 'CONSORT diagram of PROMISE participants over the 3 visits.')
-fig_dag <- fig('dag', 'Directed acyclic graphic output from the DAGitty online software.')
+fig_dag_is <- fig('dag_is', 'Directed acyclic graphic output from the DAGitty online software for insulin sensitivity.')
+fig_dag_bcf <- fig('dag_bcf', 'Directed acyclic graphic output from the DAGitty online software for beta-cell function.')
 fig_tagfa <- fig('tagfa', 'Distribution of the composition of triacylglycerol fatty acids in the baseline visit of PROMISE participants (2004-2006).
@@
 Covariates were selected based on the previous literature, from directed acyclic
-graph [@Greenland1999a] recommendations (see `r cite_t('tab_dag')`), and from 
+graph [@Greenland1999a] recommendations (see `r cite_f('dag_is')` and `r cite_f('dag_bcf')`), and from 
 QIC. `r cite_t('tab_qic')` shows the covariates compared using QIC.
@@
 
-Include DAGitty image
+![`r fig_dag_is`](../img/dagitty-model-is.png)
+
+![`r fig_dag_bcf`](../img/dagitty-model-bcf.png)
 

### 2017-01-11 doc/manuscript.Rmd
Included citation for DAGitty (or link)

@@
 Covariates were selected based on the previous literature, from directed acyclic
-graph [@Greenland1999a] recommendations (see `r cite_f('dag_is')` and `r cite_f('dag_bcf')`), and from 
+graph [@Greenland1999a] recommendations (using the DAGitty software
+[@Textor2011a], http://dagitty.net/; see `r cite_f('dag_is')` and `r cite_f('dag_bcf')`), and from
 QIC. `r cite_t('tab_qic')` shows the covariates compared using QIC.

### 2017-01-12 doc/manuscript.Rmd
First (rough) draft of discussion - limitations

@@
 
-Limitations:
-- While GEE is well suited to longitudinal datasets.. threre is a possibility
-given teh complex biology that the temporal associations can not be adequately
-analyzed using current methodologies. IN particular, an important issue si taht
-we did not quantify TAGFA at each of the three visits and so can not directed
-infer changes in TAGFA with changes in IR. However, given potential feedback
-between IR and TAGFA, this may in fact be a strength for this particular
-association given we are trying to predict how a FA at one time point can
-predict future values.
-- While PLS is a well-established technique for constructing predictive models
-of high dimensionality data strcutures, it can not adjust for potential
-confounders and other effect modifiers. Likewise, PLS is not able to handle
-longitudinal data so only the baseline visit was used in the PLS analysis.
-However, given the limitation, there results are concordant between GEE and PLS
-models.
-- Our population was predocminiately female and of euro ancestry, who were at 
-riskj for diabetes at study enrollment, and as such our results may not be
-generalizable to other populations.
-- Given the complex biology underlying the variables of interest, there many
-some residual confounding we are unaware of or have not been able to adjust for.
-However, we have taken extensive precautions for trying to ID confounders
-through the DAG models and QIC computations.
-- TAGFA may fluctuate daily, but we had participants come in fasted in the morning
+## Limitations
+
+There are a few important limitations to our study. In addition to the
+observational nature of the cohort, the biology underlying the interaction
+between TAGFA, insulin secretion, and insulin sensitivity is complex and
+multifacetted {{?}} and as such there may be some residual confounding we were
+unaware of or had not been able to adjust for. However, we have taken extensive,
+empirically based precautions in identifying potential confounders and mediators
+through the use of the DAG modeling, relying on previous literature, and through
+QIC modeling fitting methods.
+
+TAGFA can fluctuate substantially throughout the day and to control for this all
+PROMISE participants came for their clinic visits in the morning in a fasted
+state. Some of the inherent variability in TAGFA fluctuation can be managed by
+standardizing the data collection time, though there may still be some
+{{residual error? random error?}}.
+
+Since TAGFA were only quantified at the baseline visit and as such we cannot
+investigate whether there are concomitant changes in TAGFA and the metabolic
+measures over time. In relation to this limitation, while GEE is well suited to
+longitudinal datasets, the complex biology and temporal nature of the
+associations may restrict the ability of GEE to appropriately analyze and
+interpret the data. However, to maximize the potential of GEE, we used the model
+to infer that a given value of TAGFA could predict values of insulin sensitivity
+or beta-cell function over a 6 year period. This in our view is a strength of
+our analysis, as it reduces the chance of reverse causality given the tight
+integration of the glucose and fatty acid metabolism pathways, as well as uses
+GEE modeling appropriately {{? another word}}.
+
+PLS is a well-established technique for constructing predictive models
+of high dimensionality data structures (i.e. fatty acid composition), however a
+limitation is that the initial models analyzed through PLS and the final
+computed scores are not able to control for potential confounders and other
+effect modifiers. Likewise, PLS is not able to handle longitudinal data so only
+the baseline visit was used in the PLS analysis. In light of these limitations,
+we still observe concordant results between the PLS and GEE analyses {{ add more }}.
+
+Even with these limitations, our study also has several major strengths,
+including the longitudinal design and the rigorous statistical techniques and
+methods applied in analyzing the data. These statistical techniques are
+specifically suited to investigating temporal relationships and to handling the
+multivariate nature of the data. Lastly, our cohort contains highly detailed and
+comprehensive variable measurements at each collection visit, as well as having
+both concentration and mol% data for the fatty acids.
 

### 2017-01-12 doc/manuscript.Rmd
Added some notes to maybe include in discussion

@@
 
+NEFA may be a mediating factor in association, but that adjustment will not
+attenuate this assoc since TAGFA composition is also dictated by other sources
+for DNL (e.g. recent meal, DNL from glucose stores, etc.).
+
+Since no interaction, no loop? Not enough sample, already IR? maybe?
+
+Concentration data is not
+
+Discuss well done paper (very relevant) Kotronen2009a
+
+Prospective cohorts:
+    - Chien K, et al. A prediction model for type 2 diabetes risk among Chinese people. Diabetologia. 2009;52(3):443–450.
+    - D’Agostino RB Jr, Hamman RF, Karter AJ, Mykkanen L, Wagenknecht LE, Haffner SM.
+Cardiovascular disease risk factors predict the development of type 2 diabetes: the insulin resistance atherosclerosis study.
+Diabetes Care.
+2004;27(9):2234–2240.
+    - Schulze MB, et al. Use of multiple metabolic and genetic markers to improve the prediction of type 2 diabetes: the EPIC-Potsdam Study. Diabetes Care. 2009;32(11):2116–2119.
+
 # Acknowledgements

### 2017-01-12 doc/manuscript.Rmd
First draft acknowledgements, conflicts, funding

@@
 Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,6), PhD
+
+{{Bernie?}}
 
@@
 
-## Funding/support
+The authors thank Jan Neuman, Paula Van Nostrand, Stella Kink, and Annette
+Barnie of the Leadership Sinai Centre for Diabetes, Mount Sinai Hospital,
+Toronto, Canada and Sheila Porter and Mauricio Marin of the Centre for Studies
+in Family Medicine, University of Western Ontario, London, Canada for their 
+expert technical assistance and dedication in their work for PROMISE.
 
-## Conflict of interest
+## Funding
+
+This study was supported by grants from the Canadian Diabetes Association (CDA),
+the Canadian Institutes for Health Research, and the University of
+Toronto Banting and Best Diabetes Centre; LWJ is supported by a CDA
+Doctoral Student Research Award; RR is supported by a Heart and Stroke
+Foundation of Ontario Mid-Career Investigator Award {{confirm}}; SBH holds the
+CDA Chair in National Diabetes Management and the Ian McWhinney Chair of Family
+Medicine Studies at the University of Western Ontario {{confirm}}; RBP holds a
+Tier II Canada Research Chair in Brain Lipid Metabolism; AJH holds a Tier II
+Canada Research Chair in Diabetes Epidemiology.
+
+## Conflicts of interest
+
+The authors report no potential conflicts of interest relevant to this study.
+
+## Contribution statement
+
+The authors had the following responsibility: LWJ conducted research, analyzed
+data, and wrote the paper; RR, ZL, and SBH designed research, conducted
+research, and provided essential materials (infrastructure and clinical
+resources); RR, SBH, RPB, and AG provided intellectual feedback on the
+paper; RPB conducted research, provided essential reagents and materials; AJH
+designed research, assisted with interpretation, and provided intellectual
+feedback on all versions of the paper; LWJ and AJH had primary responsibility
+for final content. All authors read and approved the final manuscript.
+
+{{Bernie?}}
 

### 2017-01-13 doc/manuscript.Rmd
First (rough) draft of discussion - general findings

@@
 
-We found that in a Canadian population of mainly female and European ancestry, 
-the specific TAGFA composition was strongly associated with insulin sensitivity
-and moderately associated with beta-cell function. In particular, TAGFA myristic 
-acid (14:0), 7-tetradecenoic acid {{ confirm }} (14:1n-7), palmitic acid (16:0),
-and palmitoleic acid (16:1n-7) all strongly and negatively predicted lower
-insulin sensitivity. While
+## Overall conclusion
 
-Interpretation:
-
-- previous research has shown that CHO intake increases DNL {{ cite }}. In
-particular, DNL from CHO sources increases especially the 14-16 chain TAGFA as
-well as the 18 chain TAGFA. Prev lit has also shown that higher DNL can increase
-risk for DM {{ cite }} and that higher refined CHO can increase risk of DM {{
-cite }}. Our study found that the 14-16 TAGFA all strongly predicted lower
-insulin sensitivity, reinforcing this finding. While these FA also had a
-significant association with beta-cell function, the results were not as strong
-as with the IR. {{ why is it that DNL increases risk? }}
-- TAGFA that tended to positively associated with IR and BCF also tended to be
-the very long chain FA such as eicosapentaenoic acid (20:5n-3) and
-docosahexaenoic acid (22:6n-3), which have potent anti-inflammatory properties
-{{ cite }}. Biologically {{ why these FA and how help with IR }}
-- {{ compare models with and without NEFA }} NEFA have been shown contribute
-substantially to TAGFA composition and concentration. Adjusting for ...
-- Dietary FA have been shown to contribute approx 25% of the newly synth
-VLDL-TAG {{ cite }}. This suggests that our results may not entirely reflect
-diet, but also adipose stores. {{ discuss results of NEFA paper }} However, in
-an analysis of the NEFA pool {{ cite }}, we found that total NEFA, but not the
-specific composition associated with beta-cell function. There may be something
-unique about the specific composition of the TAGFA fraction that may allow it to
-enter IR tissues more easily than NEFA. It may be that VLDL receptors allow FA
-to transition to IR tissue compared to NEFA bound to albumin {{ confirm and cite
-}}.
-- Challenge longitudinally since muscle IR can promote DNL + TAG Flannery2012a
-- Or that post-prandial TAG is not suppressed in IR Sondergaard2012a
-- Differences with 
-- {{ confirm }} greater flux of saturated fatty acids into tissues may encourage
-lipotoxic environment {{ cite }}
+We found that in an at-risk for diabetes Canadian population of mainly female
+and European ancestry, the specific TAGFA composition was strongly associated
+with insulin sensitivity and moderately associated with beta-cell function. In
+particular, TAGFA myristic acid (14:0), 7-tetradecenoic acid {{ confirm }}
+(14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and
+negatively predicted lower insulin sensitivity. {{confirm}} While most TAGFA
+were not associated with beta-cell function, {{confirm after checking models}}
+two fatty acids, 16:0 and 18:1n-7, were associated negatively and positively,
+respectively, with both measures of beta-cell function. We also found that four
+TAGFA (14:0, 14:1n-7, 16:0, 16:1n-7) clustered together, and that also strongly 
+predicted lower insulin sensitivity. These four fatty acids are also involved in
+the DNL of refined and simple carbohydrates {{cite}}. Our results, which are in
+agreement with the current literature, suggest that higher intakes of simple
+carbohydrates or higher activity of DNL may increase the risk for diabetes,
+likely through worsening insulin sensitivity.
 
@@
 both concentration and mol% data for the fatty acids.
+
+## Interpretation and previous literature
+
+- previous research has shown that CHO intake increases DNL {{ cite }}. In
+particular, DNL from CHO sources increases especially the 14-16 chain TAGFA as
+well as the 18 chain TAGFA. Prev lit has also shown that higher DNL can increase
+risk for DM {{ cite }} and that higher refined CHO can increase risk of DM {{
+cite }}. Our study found that the 14-16 TAGFA all strongly predicted lower
+insulin sensitivity, reinforcing this finding. While these FA also had a
+significant association with beta-cell function, the results were not as strong
+as with the IR. {{ why is it that DNL increases risk? }}
+- TAGFA that tended to positively associated with IR and BCF also tended to be
+the very long chain FA such as eicosapentaenoic acid (20:5n-3) and
+docosahexaenoic acid (22:6n-3), which have potent anti-inflammatory properties
+{{ cite }}. Biologically {{ why these FA and how help with IR }}
+- {{ compare models with and without NEFA }} NEFA have been shown contribute
+substantially to TAGFA composition and concentration. Adjusting for ...
+- Dietary FA have been shown to contribute approx 25% of the newly synth
+VLDL-TAG {{ cite }}. This suggests that our results may not entirely reflect
+diet, but also adipose stores. {{ discuss results of NEFA paper }} However, in
+an analysis of the NEFA pool {{ cite }}, we found that total NEFA, but not the
+specific composition associated with beta-cell function. There may be something
+unique about the specific composition of the TAGFA fraction that may allow it to
+enter IR tissues more easily than NEFA. It may be that VLDL receptors allow FA
+to transition to IR tissue compared to NEFA bound to albumin {{ confirm and cite
+}}.
+- Challenge longitudinally since muscle IR can promote DNL + TAG Flannery2012a
+- Or that post-prandial TAG is not suppressed in IR Sondergaard2012a
+- Differences with 
+- {{ confirm }} greater flux of saturated fatty acids into tissues may encourage
+lipotoxic environment {{ cite }}
 

### 2017-01-17 doc/manuscript.Rmd
Re-arranged and added to discussion - prev lit and interpret

@@
 {{residual error? random error?}}.
+[@Rhee2011a] fasting TAG status better as discriminating between DM case vs
+control than 2hr-OGTT TAG
 
@@
 
-- previous research has shown that CHO intake increases DNL {{ cite }}. In
-particular, DNL from CHO sources increases especially the 14-16 chain TAGFA as
-well as the 18 chain TAGFA. Prev lit has also shown that higher DNL can increase
-risk for DM {{ cite }} and that higher refined CHO can increase risk of DM {{
-cite }}. Our study found that the 14-16 TAGFA all strongly predicted lower
-insulin sensitivity, reinforcing this finding. While these FA also had a
-significant association with beta-cell function, the results were not as strong
-as with the IR. {{ why is it that DNL increases risk? }}
-- TAGFA that tended to positively associated with IR and BCF also tended to be
-the very long chain FA such as eicosapentaenoic acid (20:5n-3) and
-docosahexaenoic acid (22:6n-3), which have potent anti-inflammatory properties
-{{ cite }}. Biologically {{ why these FA and how help with IR }}
-- {{ compare models with and without NEFA }} NEFA have been shown contribute
-substantially to TAGFA composition and concentration. Adjusting for ...
-- Dietary FA have been shown to contribute approx 25% of the newly synth
-VLDL-TAG {{ cite }}. This suggests that our results may not entirely reflect
-diet, but also adipose stores. {{ discuss results of NEFA paper }} However, in
-an analysis of the NEFA pool {{ cite }}, we found that total NEFA, but not the
-specific composition associated with beta-cell function. There may be something
-unique about the specific composition of the TAGFA fraction that may allow it to
-enter IR tissues more easily than NEFA. It may be that VLDL receptors allow FA
-to transition to IR tissue compared to NEFA bound to albumin {{ confirm and cite
-}}.
-- Challenge longitudinally since muscle IR can promote DNL + TAG Flannery2012a
-- Or that post-prandial TAG is not suppressed in IR Sondergaard2012a
-- Differences with 
-- {{ confirm }} greater flux of saturated fatty acids into tissues may encourage
-lipotoxic environment {{ cite }}
+Previous research has shown that carbohydrate intake increases DNL {{ cite }}. In
+particular, DNL from refined or simple carbohydrate sources increases the 14 to
+16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
+link between higher estimated DNL and an increased risk for diabetes {{ cite }}
+as well as studies showing higher refined carbohydrates (e.g. added sugars) and
+a higher risk for diabetes {{cite}}. Our study extends {{adds to? confirms?}}
+these findings on DNL by showing that TAGFA with 14 to 16 carbons all strongly
+predicted lower insulin sensitivity. While these FA also had a significant
+association with beta-cell function, the results were not as strong as with the
+IR. {{ why is it that DNL increases risk? }}
 
-[@Barrows2006a] NEFA contributes 60-80% of newly synth VLDL-TAG during fasting
-and ~40% during fed. Dietary FA contribute ~25% to newly synth VLDL-TAG.
-VLDL-TAG is not cleared (lipolyzed) as much as chylomicron and so stays in
-circulation longer (hence why there is bigger TAG pool vs NEFA)
+To our knowledge, no longitudinal study has examined the role of the composition
+of the TAGFA fraction on detailed OGTT-derived metabolic measures using 
+statistical techniques designed to adequately analyze ... {{shots fired? 
+necessary?}}.
 
-Previous lit and limits:
-- A few epidemiological studies have examined the role of the TAGFA composition
-on insulin resistance and diabetes risk. 
+To our knowledge, two prospective studies have been published that had similar
+research objectives as the present analysis. One was a nested case-control
+cohort (n=189 cases and n=189 controls) of the Framingham offspring study
+[@Rhee2015a], which found that TAGFA composition with a lower carbon chain and less double
+bonds (e.g. 14:0, 16:0) had a higher OR {{confirm}} for diabetes after 12-years
+while higher carbon chain and more double bond TAGFA had a lower OR for
+diabetes. This pattern of TAGFA with diabetes risk remained
+the same when they examined HOMA-IR cross-sectionally at the baseline visit. The
+other study was a prospective cohort of males in Finland [@Lankinen2015a], of
+which they had TAGFA data on 831 participants after 6-years from the baseline
+visit. This cohort has similar outcomes as PROMISE, however, OGTT data was only
+available at the 6-year visit. A beta-cell function measure was computed, but
+only included calculations up to 30 minutes of the OGTT. As with our results,
+they found a larger effect size for insulin sensitivity compared to the
+beta-cell function. The specific TAGFA that associated with the metabolic
+measures were similar between studies, except they did not find {{confirm}}
+
+In addition, the Framingham offspring study [@Rhee2011a] did not have data on
+beta-cell function while the Finnish cohort [@Lankinen2015a] used {{ how do I
+say this politely Tony? They're methods and article was ... questionable... }}
+simple statistical techniques to analysis the longitudinal data, which do not
+adequately take into consideration auto-correlation of longitudinal values and
+the high dimensionality of the TAGFA composition data structure.
+
+A smaller study (n=16) of mostly females using highly detailed lipodomic
+techniques found a positive correlation between total esterified (of which TAGFA
+make up the majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR [@Kotronen2009a].
+
+Limitations with estimated DNL or measured DNL in prev studies.
+
+Several studies have been conducted that link greater hepatic DNL with the 14
+and 16 carbon FA. A short, small feeding trial (n=24) was conducted to identify
+which biomarkers most accurately reflected DNL [@Lee2015a]. The study found that
+16:1n-7, directly measured DNL using isotopes, and liver fat were all highly
+correlated with each other, suggesting that 16:1n-7 is a good biomarker for
+hepatic DNL.
+In another small (n=14) feeding trial, meal type (high fat vs low fat) was
+tested to determine the effect on DNL and TAGFA composition [@Wilke2009a]. They
+found that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
+carbohydrate) group. These fatty acids are similar to the fatty acids we found
+that clustered together, suggesting {{another word}} that these fatty acids are
+indicative of a higher carbohydrate diet.
+Suggesting {{expand/synth}} that greater hepatic fat may also contribute to greater DNL,
+increasing the 14 and 16 carbon FA in circulation, potentially exposing tissues
+to greater lipotoxicity, resulting in further hepatic dysfunction and continuing
+the cycle.
+
+- contributions to DNL (e.g. 70% 16:0 to newly synth) {{ confirm what this is }}
+
 - A weight loss intervention trial showed that changes in the TAGFA composition
@@
 with improved insulin sensitivity.
-- In a short (24hr) feeding trial of 24 subjects [@Lee2015a], palmitoleic acid
-(16:1n-7) was shown to be strongly correlated with isotopically measured DNL and
-was 45% higher in individuals with a greater amount of hepatic fat. However, in
-that study only the TAGFA 18:0 was correlated with HOMA-IR, likely due to the
-small sample size.
-- To our knowledge, there have been two prospective studies that had a similar
-research objective as our cohort. One was a nested case-control cohort of the 
-Framingham offspring study [@Rhee2015a], which found that TAGFA with a lower
-carbon chain and less double bonds (e.g. 14:0, 16:0) had higher OR for diabetes
-after 12-years while higher carbon chain and more double bond TAGFA had a lower
-OR for diabetes. This pattern of TAGFA with diabetes risk remained the same when
-they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
-was a prospective cohort of males in Finland [@Lankinen2015a], of which they had
-TAGFA data on 831 participants after 6-years from the baseline visit. This
-cohort has similar outcomes as PROMISE, however, OGTT data was only available at
-the 6-year visit. A beta-cell function measure was computed, but only included 
-calculations up to 30 minutes of the OGTT. As with our results, they found a
-larger effect size for insulin sensitivity compared to the beta-cell function.
-The specific TAGFA that associated with the metabolic measures were similar 
-between studies, except they did not find 
-- The primary limit to these studies is the mainly small sample size and short
-study duration. Two studies had larger sample sizes of 831 [@Lankinen2015a] and 
-`r 189+189` [@Rhee2011a] and both were prospective cohorts. In addition, the
-Framingham offspring study [@Rhee2011a] did not have data on beta-cell function 
-while the Finnish cohort [@Lankinen2015a] used {{ how do I say this politely Tony? 
-They're methods and article was ... questionable... }} simple statistical 
-techniques to analysis the longitudinal data, which do not adequately take into
-consideration auto-correlation of longitudinal values and the high dimensionality
-of the TAGFA composition data structure.
-
-- Refs:
-    - Lankinen2015a (poor study, but is unfortunately relevant...) (god... I
-    really don't want to cite/talk about this...)
-    - Lee2015a (16:1n7 as marker of DNL) (n=24, feeding trial, 16:1n7 was 45%
-    higher in high liver fat group, direct dnl (isotope) was 2.5 fold higher in
-    high liver fat, 16:1n7 was corr (0.56) with direct DNL, TAG 18:0 corr (.50)
-    with HOMA)
-    - Zulyniak2012a (only vaccenic acid in TAG) (n=20, TLC-GC, HOMA-IR/HOMA-B,
-    two groups: normogly and hypergly. results in general showed no diff between
-    groups, only cis-vaccenic was higher in hyper, 'trend' with palmitic, did
-    not present results on whole fraction for IR)
-    - Rhee2011a (good paper on TAGFA composition) (Offspring of Framingham...,
-    nested case-control, n=189 case, n=189 control, MS/MS, OGTT, differences in
-    fasting TAGFA were more apparent than at 2-hr ogtt, more lower chain/lower
-    desat in cases + higher OR, while higher chain/desat had lower OR, lower
-    chain/desat pos corr with HOMA-IR, MS/MS (tandem) showed that 12-18 SFA and
-    MUFA inc risk of diabetes)
-    - Kotronen2009a (TAGFA composition) (n=16, 18-60, mostly female, lipidomics,
-    GC, HOMA-IR, total esterified (of which TAGFA makes up the majority) 16,
-    161n7, 181n9 pos corr with homa, 182n6 neg with homa)
+{{ hard to disentangle reverse causality here }}
     - Schwab2008a (TAGFA composition and insulin sensitivity) (n=75 (but only
@@
 
-[@Rhee2011a] fasting TAG status better as discriminating between DM case vs
-control than 2hr-OGTT TAG
-[@Kotronen2009a] (this wasn't studied, but discussed) higher hepatic DNL may
-increase amount of 'shorter' chain FA (16:0, 14:0, etc.) into VLDL and hence
-into circulation. So higher of these FA may indicate higher DNL -> problems with
-liver function -> maybe hepatic IR -> cyclical?
-[@Lee2015a] 16:1n-7 may be good marker of liver DNL.
-
-Wilke2009a - contributions to DNL (e.g. 70% 16:0 to newly synth)
-The four FA that cluster together on the PLS also appear to be the same FA that
-make up most of DNL, especially in context of higher carb/refined carb in diet
-[@Wilke2009a].
-
-New things:
-
-- No longitudinal study has examined all these TAGFA using detailed metabolic,
-OGTT-derived measures. No study has used these statistical techniques, which
-were designed with these types of data in mind, and as such allows appropriate
-conclusions to be reached from this data.
-- That 14-16 FA in the TAGFA fraction predict lower insulin sensitivity and that
-this particular cluster is also generally observed with higher CHO intake.
+Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
+sources of fatty acids for TAG production during fasting are stored fatty acids
+within the liver, circulating NEFA, and through DNL {{cite}}. Therefore, the
+TAGFA composition reflects not just dietary intake but also stored fatty acids
+in adipose tissue and the DNL capacity of the individual. During fasting, a
+large portion of fatty acids bound together as TAG are obtained from circulating
+NEFA. A study examining this found that NEFA contributed 60-80% of newly synth
+VLDL-TAG during fasting and ~40% during fed while dietary fatty acids
+contributed ~25% to newly synth VLDL-TAG {{fasting or fed?}} [@Barrows2006a].
+While the NEFA fraction is much smaller than the TAG fraction even though it
+contributes substantially to the TAG fraction, VLDL-TAG is not cleared
+(lipolyzed) as much as chylomicron (which also contains a large amount of TAG)
+and so has a higher half-life than NEFA, remaining in circulation longer {{hence
+why there is bigger TAG pool vs NEFA}}. In an analysis of the NEFA fraction in
+the PROMISE cohort with the OGTT-derived outcomes (currently unpublished)
+{{correct word for this??}}, we found that total NEFA, but not the specific
+composition associated with beta-cell function and not with insulin resistance.
+While NEFA may contribute substantially to the TAG fraction, there may be
+something unique about the specific composition of the TAGFA fraction that may
+allow it to enter insulin sensitive tissues more easily than NEFA. It may be
+that VLDL receptors allow FA to transition to these tissue compared to NEFA
+bound to albumin {{ confirm and cite }}, and thus able to exert potential
+lipotoxic effects. {{ comment on comparing models with and without NEFA }}
 
@@
 
+As with our previous analysis of the PL fatty acid fraction {{cite}}, we found
+that cis-vaccenic acid strongly predicted higher insulin sensitivity and
+beta-cell function. Likewise, in the TAGFA fraction we found a similar
+association with cis-vaccenic acid. {{ continue? }}
+    - Zulyniak2012a (only vaccenic acid in TAG) (n=20, TLC-GC, HOMA-IR/HOMA-B,
+    two groups: normogly and hypergly. results in general showed no diff between
+    groups, only cis-vaccenic was higher in hyper, 'trend' with palmitic, did
+    not present results on whole fraction for IR)
+
+{{ comment on unadjusted vs adjusted }}
+
+{{ commment on null interactions with time }}
+- Challenge longitudinally since muscle IR can promote DNL + TAG Flannery2012a
 Since no interaction, no loop? Not enough sample, already IR? maybe?
 
-Concentration data is not
+- Or that post-prandial TAG is not suppressed in IR Sondergaard2012a
+- {{ confirm }} greater flux of saturated fatty acids into tissues may encourage
+lipotoxic environment {{ cite }}
 
-Discuss well done paper (very relevant) Kotronen2009a
+{{ Concentration data is not }}
 

### 2017-01-22 doc/manuscript.Rmd
First (rough) draft of discussion - prev lit and interpret

@@
 measure of all fatty acids within the TAG fraction. However, given the 
-increasing appreciation for the role of the specific fatty acid (FA) composition
-of various blood lipid fractions on health outcomes, reducing the complex FA 
+increasing appreciation for the role of the specific fatty acid composition
+of various blood lipid fractions on health outcomes, reducing the complex fatty acids 
 metabolism within the TAG fraction to a single measure may contribute to the 
@@
 The PLS results for the insulin sensitivity (`r cite_f('fig_pls')`) revealed a clustering 
-of the FA 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly (meaning
-the FA fell between the dashed and solid circles) and negatively on HOMA2-%S
+of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly (meaning
+the fatty acids fell between the dashed and solid circles) and negatively on HOMA2-%S
 and ISI, suggesting this cluster of TAGFA tracks together with lower insulin
@@
 
-Previous research has shown that carbohydrate intake increases DNL {{ cite }}. In
+To our knowledge, no longitudinal study has examined the role of the composition
+of the TAGFA fraction on detailed OGTT-derived metabolic measures using 
+advanced {{another word?}} statistical techniques designed to maximally utilize
+the data. Two large prospective studies have been published that had similar
+research objectives as the present analysis. One was a nested case-control
+cohort (n=189 cases and n=189 controls) of the Framingham offspring study
+[@Rhee2015a], which found that TAGFA composition with a lower carbon chain and
+less double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
+12-years while higher carbon chain and more double bond TAGFA had a lower risk 
+for diabetes. This pattern of TAGFA with diabetes risk remained the same when 
+they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
+was a prospective cohort of males in Finland [@Lankinen2015a], of which they had
+TAGFA data on 831 participants after 6-years from the baseline visit. This 
+cohort has similar outcomes as PROMISE, however, OGTT data was only available at
+the 6-year visit. A beta-cell function measure was computed, but only included 
+calculations up to 30 minutes of the OGTT. As with our results, they found a 
+larger effect size for insulin sensitivity compared to the beta-cell function. 
+The specific TAGFA that associated with the metabolic measures were similar 
+between studies, particularly with the four DNL fatty acids. A limitation of 
+this study was the simpler statistical techniques used to analyze the data, 
+which is not able to adequately use the longitudinal and high-dimensional 
+dataset. {{ They're article and methods were ... questionable... }} In another 
+study of a much smaller (n=16) mostly female group, the authors report a 
+positive correlation between total esterified (of which TAG make up the 
+majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR [@Kotronen2009a], similar as 
+our results except for 18:1n-9.
+
+Previous research has shown that carbohydrate intake increases DNL 
+[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. In
 particular, DNL from refined or simple carbohydrate sources increases the 14 to
 16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
-link between higher estimated DNL and an increased risk for diabetes {{ cite }}
-as well as studies showing higher refined carbohydrates (e.g. added sugars) and
-a higher risk for diabetes {{cite}}. Our study extends {{adds to? confirms?}}
-these findings on DNL by showing that TAGFA with 14 to 16 carbons all strongly
-predicted lower insulin sensitivity. While these FA also had a significant
-association with beta-cell function, the results were not as strong as with the
-IR. {{ why is it that DNL increases risk? }}
+link between higher estimated DNL and an increased risk for metabolic 
+dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a] as well as
+studies showing higher refined carbohydrates (e.g. added sugars) and a higher
+risk for diabetes {{cite}}. Our study extends these findings on DNL by showing
+that TAGFA with 14 to 16 carbons all strongly predicted lower insulin
+sensitivity. While these fatty acids also had a significant association with
+beta-cell function, the results were not as strong as with the insulin
+sensitivity.
 
-To our knowledge, no longitudinal study has examined the role of the composition
-of the TAGFA fraction on detailed OGTT-derived metabolic measures using 
-statistical techniques designed to adequately analyze ... {{shots fired? 
-necessary?}}.
+The link between higher DNL and increases in specific fatty acids has been
+examined in several studies. Previous studies have examined used markers
+of DNL, such as the ratio between 18:2n-6 to 16:0 or 16:0 to 14:0 ratio
+[@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are major
+limitations to using these ratios as the fatty acids used in their calculation
+can be created from DNL, they are also obtained from the diet [@Hodson2008a]. A
+short, small feeding trial (n=24) was conducted to identify fatty acids that
+most accurately reflected DNL, as potential biomarkers [@Lee2015a]. The study
+found that palmitoleic acid (16:1n-7), directly measured DNL using isotopes, and
+liver fat were all highly correlated with each other (r>0.50), suggesting that
+16:1n-7 may be a good biomarker for hepatic DNL. In another small (n=14) feeding
+trial, meal type (high fat vs low fat) was tested to determine its effect on DNL
+and TAGFA composition [@Wilke2009a]. The authors reported that 14:0, 16:0, 16:1,
+and 18:2 were higher in the low fat (high carbohydrate) group. These fatty acids
+are similar to the fatty acids we found that clustered together, implicating 
+these fatty acids as indicative of a higher carbohydrate diet. A higher
+carbohydrate diet, particularly with simple carbohydrates, may lead to greater
+DNL in an attempt to control blood glucose, increasing hepatic fat stores, and
+increasing the amount of TAGFA in circulation. The increase in circulating 14
+and 16 carbon fatty acids may then increase the exposure of tissues to lipotoxicity,
+for instance from palmitic acid (16:0), which is well-known to have harmful
+effects on tissues [@Riserus2008a; @Iggman2010a].
 
-To our knowledge, two prospective studies have been published that had similar
-research objectives as the present analysis. One was a nested case-control
-cohort (n=189 cases and n=189 controls) of the Framingham offspring study
-[@Rhee2015a], which found that TAGFA composition with a lower carbon chain and less double
-bonds (e.g. 14:0, 16:0) had a higher OR {{confirm}} for diabetes after 12-years
-while higher carbon chain and more double bond TAGFA had a lower OR for
-diabetes. This pattern of TAGFA with diabetes risk remained
-the same when they examined HOMA-IR cross-sectionally at the baseline visit. The
-other study was a prospective cohort of males in Finland [@Lankinen2015a], of
-which they had TAGFA data on 831 participants after 6-years from the baseline
-visit. This cohort has similar outcomes as PROMISE, however, OGTT data was only
-available at the 6-year visit. A beta-cell function measure was computed, but
-only included calculations up to 30 minutes of the OGTT. As with our results,
-they found a larger effect size for insulin sensitivity compared to the
-beta-cell function. The specific TAGFA that associated with the metabolic
-measures were similar between studies, except they did not find {{confirm}}
+We found no interaction by time for any of the TAGFA on the metabolic outcomes.
+Analyzing these biological processes longitudinally is challenging due to the
+potential feedbacks involved. For instance, while greater DNL may promote muscle
+resistance, the reverse may also be true [@Flannery2012a]. Higher insulin
+resistance may encourage greater DNL to handle the higher blood glucose. To
+illustrate, in a weight loss intervention trial (n=19 with TAGFA data),
+participants who lost weight over 33 weeks showed higher insulin sensitivity and
+lower TAGFA composition indicative of lower DNL (less 14:0, 14:1, 16:0, 16:1,
+etc) [@Schwab2008a]. Given the complex biological mechanisms and feedback loops
+involved, disentangling whether insulin sensitivity influences TAGFA more
+strongly than TAGFA influencing insulin sensitivity will require complicated
+research designs and analyses. Our own null results on this question may
+tentatively point to an equal feedback mechansim between TAGFA and insulin
+sensitivity.
 
-In addition, the Framingham offspring study [@Rhee2011a] did not have data on
-beta-cell function while the Finnish cohort [@Lankinen2015a] used {{ how do I
-say this politely Tony? They're methods and article was ... questionable... }}
-simple statistical techniques to analysis the longitudinal data, which do not
-adequately take into consideration auto-correlation of longitudinal values and
-the high dimensionality of the TAGFA composition data structure.
+Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
+sources of fatty acids for TAG production during fasting are stored fatty acids
+within the liver, circulating NEFA, and through DNL [@Barrows2006a]. Therefore, the
+TAGFA composition reflects not just dietary intake but also stored fatty acids
+from the adipose tissue and the DNL capacity of the individual. During fasting, a
+large portion of fatty acids bound together as TAG are obtained from circulating
+NEFA. A study examining this found that NEFA contributed 60-80% of newly synth
+VLDL-TAG during fasting and ~40% during fed while dietary fatty acids
+contributed ~25% to newly synth VLDL-TAG [@Barrows2006a]. In an analysis of the
+serum NEFA fraction in the PROMISE cohort with the OGTT-derived outcomes
+(currently unpublished), we found that higher total NEFA but not the specific
+composition associated with lower beta-cell function and not with insulin
+resistance. While NEFA may contribute substantially to the TAG fraction, there
+may be something unique about the specific composition of the TAGFA fraction
+that may allow it to enter insulin sensitive tissues more easily than NEFA. It
+may be that VLDL receptors allow fatty acids to enter these tissue more easily than 
+NEFA bound to albumin, or it may be due to the other influences on TAGFA
+composition such as dietary intake.
 
-A smaller study (n=16) of mostly females using highly detailed lipodomic
-techniques found a positive correlation between total esterified (of which TAGFA
-make up the majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR [@Kotronen2009a].
+{{ Not sure about this... Include this? }}
+As with our previous analysis of the PL fatty acid fraction {{cite}}, we found
+that cis-vaccenic acid strongly predicted higher insulin sensitivity and
+beta-cell function. Likewise, in the TAGFA fraction we found a similar
+association with cis-vaccenic acid. {{we can discuss to include more}}
 
-Limitations with estimated DNL or measured DNL in prev studies.
+<!--
+{{ comment on comparing models with and without NEFA }} While
+NEFA may be a mediating factor, we found that adjustment for it did not
+{{confirm}} attenuate this association since TAGFA composition is also dictated
+by other sources for DNL (e.g. recent meal, DNL from glucose stores, etc.).
 
-Several studies have been conducted that link greater hepatic DNL with the 14
-and 16 carbon FA. A short, small feeding trial (n=24) was conducted to identify
-which biomarkers most accurately reflected DNL [@Lee2015a]. The study found that
-16:1n-7, directly measured DNL using isotopes, and liver fat were all highly
-correlated with each other, suggesting that 16:1n-7 is a good biomarker for
-hepatic DNL.
-In another small (n=14) feeding trial, meal type (high fat vs low fat) was
-tested to determine the effect on DNL and TAGFA composition [@Wilke2009a]. They
-found that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
-carbohydrate) group. These fatty acids are similar to the fatty acids we found
-that clustered together, suggesting {{another word}} that these fatty acids are
-indicative of a higher carbohydrate diet.
-Suggesting {{expand/synth}} that greater hepatic fat may also contribute to greater DNL,
-increasing the 14 and 16 carbon FA in circulation, potentially exposing tissues
-to greater lipotoxicity, resulting in further hepatic dysfunction and continuing
-the cycle.
+While the NEFA fraction is much smaller than the TAG fraction, even though it
+contributes substantially to the TAG fraction, the VLDL-TAG is not cleared
+(lipolyzed) as much as chylomicron (which also contains a large amount of TAG)
+and so has a higher half-life than NEFA, remaining in circulation longer {{hence
+why there is bigger TAG pool vs NEFA}} {{cite?}}.
 
@@
 
-- A weight loss intervention trial showed that changes in the TAGFA composition
-of 19 participants accompanied changes in the weight and insulin sensitivity over
-the 33 week period [@Schwab2008a]. Specifically, the TAGFA 14:0, 14:1, 16:0,
-16:1, 18:0, 18:3n-3, and 18:3n-6 were all lower after weight loss and correlated
-with improved insulin sensitivity.
-{{ hard to disentangle reverse causality here }}
-    - Schwab2008a (TAGFA composition and insulin sensitivity) (n=75 (but only
-    n=19 for lipidomics), overweight or obese, weght loss intervention (33
-    weeks), weight, FSIGT, GC, UPLC-MS, after weight loss 14, 141, 16, 161, 18,
-    183n3, 183n6 were lower, lower TAGFA that contained 14, 14:1, 16, 161, 18,
-    181 were correlated with improved Si)
-
-Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
-sources of fatty acids for TAG production during fasting are stored fatty acids
-within the liver, circulating NEFA, and through DNL {{cite}}. Therefore, the
-TAGFA composition reflects not just dietary intake but also stored fatty acids
-in adipose tissue and the DNL capacity of the individual. During fasting, a
-large portion of fatty acids bound together as TAG are obtained from circulating
-NEFA. A study examining this found that NEFA contributed 60-80% of newly synth
-VLDL-TAG during fasting and ~40% during fed while dietary fatty acids
-contributed ~25% to newly synth VLDL-TAG {{fasting or fed?}} [@Barrows2006a].
-While the NEFA fraction is much smaller than the TAG fraction even though it
-contributes substantially to the TAG fraction, VLDL-TAG is not cleared
-(lipolyzed) as much as chylomicron (which also contains a large amount of TAG)
-and so has a higher half-life than NEFA, remaining in circulation longer {{hence
-why there is bigger TAG pool vs NEFA}}. In an analysis of the NEFA fraction in
-the PROMISE cohort with the OGTT-derived outcomes (currently unpublished)
-{{correct word for this??}}, we found that total NEFA, but not the specific
-composition associated with beta-cell function and not with insulin resistance.
-While NEFA may contribute substantially to the TAG fraction, there may be
-something unique about the specific composition of the TAGFA fraction that may
-allow it to enter insulin sensitive tissues more easily than NEFA. It may be
-that VLDL receptors allow FA to transition to these tissue compared to NEFA
-bound to albumin {{ confirm and cite }}, and thus able to exert potential
-lipotoxic effects. {{ comment on comparing models with and without NEFA }}
-
-NEFA may be a mediating factor in association, but that adjustment will not
-attenuate this assoc since TAGFA composition is also dictated by other sources
-for DNL (e.g. recent meal, DNL from glucose stores, etc.).
-
-As with our previous analysis of the PL fatty acid fraction {{cite}}, we found
-that cis-vaccenic acid strongly predicted higher insulin sensitivity and
-beta-cell function. Likewise, in the TAGFA fraction we found a similar
-association with cis-vaccenic acid. {{ continue? }}
     - Zulyniak2012a (only vaccenic acid in TAG) (n=20, TLC-GC, HOMA-IR/HOMA-B,
@@
     not present results on whole fraction for IR)
-
 {{ comment on unadjusted vs adjusted }}
@@
 {{ commment on null interactions with time }}
-- Challenge longitudinally since muscle IR can promote DNL + TAG Flannery2012a
-Since no interaction, no loop? Not enough sample, already IR? maybe?
 
@@
     - Schulze MB, et al. Use of multiple metabolic and genetic markers to improve the prediction of type 2 diabetes: the EPIC-Potsdam Study. Diabetes Care. 2009;32(11):2116–2119.
+-->
 

### 2017-01-23 doc/manuscript.Rmd
Second draft of intro

@@
 
-Note: keep the story simple and clean, without getting too dense
-
-- Run PLS with just those FA that loaded strongly to see the predictive ability.
-- SA with/without NEFA
-- SA with/without WC
-- Interactions of VN unadjusted
-
 Outline:
@@
 - Primary analysis:
-    - Look over medications, how they influence the TAGFA
-    - GEE
-        - SA for the large outlier for the TAGFA (see dist plot)
-        - Include # of both uncorrected and corrected (FDR) significant associations
-        - QIC modeling
-        - DAG development
-        - Look over the entire model (see which variables are significant, etc)
-        - Sensitivity analysis with Carb intake? And with adiponectin? (which is
-        involved in TAG metabolism)
-    - PLS with ISI and HOMA-IR
-        - Use extracted PLS components in GEE (?)
-- Secondary analysis:
-    - AROC
-        - dysgly with clinical TAG vs the 'best' fatty acid to compare to
-
-- Make sure to include comments from NEFA paper
-- Add stats to basic char table... (don't really want to. But I should ..?)
-
-Possible interpretations:
-
-- muscle IR -> higher TAG
-- But longitudinal, since no interaction with time, so maybe the higher TAG is
-because of higher IR (since this population is at risk of being IR).
+        - Sensitivity analysis with Carb intake?
 
@@
 
-Title: Fatty acids involved in de novo lipogenesis within triacylglycerol fraction strongly predict lower insulin sensitivity.
+Title: Fatty acids involved in de novo lipogenesis within the serum
+triacylglycerol fraction strongly predict lower insulin sensitivity.
 
@@
 
-Hypertriglyceridemia is an established and well known factor involved in the
-dysregulation of metabolic function and subsequent negative health impacts. It
-is a component of the metabolic syndrome {{ cite }}, a strong risk factor for
-cardiovascular disease {{cite}}, and can contribute to worsening pathogenesis of
-diabetes through it's harmful health effects {{ clarify }}. Triacylglcerol (TAG)
-is a common blood measure taken during routine clinical visits and has {{
-historically?}} been a target of pharmaceutical intervention {{cite}}, though evidence on
-the {{efficacy? effectiveness?}} of such interventions are often underwhelming
-on their effects on disease {{risk?}} reduction {{cite}}.
+Hypertriglyceridemia is an extensively studied and well known factor involved in
+the dysregulation of metabolic function and subsequent negative health impacts.
+It is a component of the metabolic syndrome [@Alberti2009], a strong risk factor for
+cardiovascular disease [@DAgostino2004a], and can contribute to the pathogenesis
+of diabetes. Measuring triacylglcerides (TAG) within the blood is commonly done
+during routine clinical visits as part of health assessment. However, clinically
+measured TAG is limited as it represents all the types of fatty acids within the
+TAG fraction as a single measure. Given the increasing appreciation for the role
+of the specific fatty acid composition on health outcomes, for instance with
+differences in associations of fatty acids within the PL lipid fraction
+[@Forouhi2014a; @Ma2015a] including an analysis from the PROMISE cohort
+[@Johnston2016a], there are relatively few studies that have explored this area
+of research in the TAG fraction.
 
-TAG as a pharmaceutical target and as a measure of metabolic health is a single
-measure of all fatty acids within the TAG fraction. However, given the 
-increasing appreciation for the role of the specific fatty acid composition
-of various blood lipid fractions on health outcomes, reducing the complex fatty acids 
-metabolism within the TAG fraction to a single measure may contribute to the 
-less than expected {{?}} changes in risk reduction.
+The interaction between TAG and insulin sensitivity is complex and has 
+components of a feedback system. Greater resistance to insulin in both the liver
+and muscle may result in greater production of TAG and secretion of lipoproteins
+that transport TAG [@Yu2012a]. Likewise, greater TAGFA may contribute to 
+metabolic dysfunction and lipotoxicity in various tissues, such as the 
+beta-cells, and thus continuing the cycle. Given the complexity and temporal
+nature of the relationship, long term studies with multiple data collections are
+paramount to better understanding the underlying biology and subsequent risk.
 
-In additiona, the {{interplay?}} between TAG and IR is complex and has the
-potential to act as a {{feed forward?}} loop. Greater resistance to insulin,
-particularly in the liver, can result in greater accumulation of hepatic fat and
-a potential associated increased production in TAG which are secreted into
-circulation with VLDL [@Yu2012a] {{review on mechanisms of TAG, IR, etc.}} {{not
-always the case, dependent on ethnicity}}. However, higher circulating TAG can
-exacerbate or contribute to insulin resistance in various tissues, creating a
-vicious cycle {{cite}}. Few studies have been conducted that have examined the
-longitudinal effect of TAGFA composition on diabetes pathophysiology.
-
-Several prospective studies have identified to date the association of 
-hypertriglyceridemia on incident diabetes {{[@Chien2008a; @DAgostino2004a;
-@Schulze2009a]}}, but no longitudinal studies have been conducted{{another 
-word?}} on the pathophysiological factors underlying diabetes.
-
-Our objective was to examine the longitudinal role of the specific composition
-of the serum TAG fraction on OGTT-derived measures of insulin sensitivity and
-beta-cell function in an Canadian population at risk for diabetes over 6 years.
+While several prospective studies have identified a harmful association of 
+hypertriglyceridemia on greater incidence for type 2 diabetes 
+[@Chien2008a; @DAgostino2004a; @Schulze2009a], there are limited longitudinal
+studies that have examined the relationship between TAG and it's composition on
+the pathophysiological factors underlying diabetes, particularly in regard to
+beta-cell function. Our objective was to examine the longitudinal role of the
+specific composition of the serum TAG fraction on OGTT-derived measures of
+insulin sensitivity and beta-cell function in an Canadian population at risk for
+diabetes.
 

### 2017-01-23 doc/manuscript.Rmd
Second draft of discussion

@@
 
-Outline:
-
-- General findings
-- Limitations of study
-- Previous lit and limits of those
-- How this study fills those gaps and new things learned
-- Potential clinical application
-
 ## Overall conclusion
 
-We found that in an at-risk for diabetes Canadian population of mainly female
-and European ancestry, the specific TAGFA composition was strongly associated
+We found that in an at-risk for diabetes Canadian population who were mainly female 
+and of European ancestry, the specific TAGFA composition was strongly associated 
 with insulin sensitivity and moderately associated with beta-cell function. In 
-particular, TAGFA myristic acid (14:0), 7-tetradecenoic acid {{ confirm }}
+particular, TAGFA myristic acid (14:0), 7-tetradecenoic acid {{ confirm name }} 
 (14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and
-negatively predicted lower insulin sensitivity. {{confirm}} While most TAGFA
-were not associated with beta-cell function, {{confirm after checking models}}
-two fatty acids, 16:0 and 18:1n-7, were associated negatively and positively,
-respectively, with both measures of beta-cell function. We also found that four
-TAGFA (14:0, 14:1n-7, 16:0, 16:1n-7) clustered together, and that also strongly 
-predicted lower insulin sensitivity. These four fatty acids are also involved in
-the DNL of refined and simple carbohydrates {{cite}}. Our results, which are in
-agreement with the current literature, suggest that higher intakes of simple
-carbohydrates or higher activity of DNL may increase the risk for diabetes,
-likely through worsening insulin sensitivity.
-
-## Limitations
-
-There are a few important limitations to our study. In addition to the
-observational nature of the cohort, the biology underlying the interaction
-between TAGFA, insulin secretion, and insulin sensitivity is complex and
-multifacetted {{?}} and as such there may be some residual confounding we were
-unaware of or had not been able to adjust for. However, we have taken extensive,
-empirically based precautions in identifying potential confounders and mediators
-through the use of the DAG modeling, relying on previous literature, and through
-QIC modeling fitting methods.
-
-TAGFA can fluctuate substantially throughout the day and to control for this all
-PROMISE participants came for their clinic visits in the morning in a fasted
-state. Some of the inherent variability in TAGFA fluctuation can be managed by
-standardizing the data collection time, though there may still be some
-{{residual error? random error?}}.
-[@Rhee2011a] fasting TAG status better as discriminating between DM case vs
-control than 2hr-OGTT TAG
-
-Since TAGFA were only quantified at the baseline visit and as such we cannot
-investigate whether there are concomitant changes in TAGFA and the metabolic
-measures over time. In relation to this limitation, while GEE is well suited to
-longitudinal datasets, the complex biology and temporal nature of the
-associations may restrict the ability of GEE to appropriately analyze and
-interpret the data. However, to maximize the potential of GEE, we used the model
-to infer that a given value of TAGFA could predict values of insulin sensitivity
-or beta-cell function over a 6 year period. This in our view is a strength of
-our analysis, as it reduces the chance of reverse causality given the tight
-integration of the glucose and fatty acid metabolism pathways, as well as uses
-GEE modeling appropriately {{? another word}}.
-
-PLS is a well-established technique for constructing predictive models
-of high dimensionality data structures (i.e. fatty acid composition), however a
-limitation is that the initial models analyzed through PLS and the final
-computed scores are not able to control for potential confounders and other
-effect modifiers. Likewise, PLS is not able to handle longitudinal data so only
-the baseline visit was used in the PLS analysis. In light of these limitations,
-we still observe concordant results between the PLS and GEE analyses {{ add more }}.
-
-Even with these limitations, our study also has several major strengths,
-including the longitudinal design and the rigorous statistical techniques and
-methods applied in analyzing the data. These statistical techniques are
-specifically suited to investigating temporal relationships and to handling the
-multivariate nature of the data. Lastly, our cohort contains highly detailed and
-comprehensive variable measurements at each collection visit, as well as having
-both concentration and mol% data for the fatty acids.
+negatively predicted lower insulin sensitivity. While most TAGFA were not
+associated with beta-cell function, two fatty acids, 16:0 and 18:1n-7, were
+associated negatively and positively, respectively, with both measures of 
+beta-cell function. We also found that four TAGFA (14:0, 14:1n-7, 16:0, 16:1n-7)
+clustered together, and that also strongly predicted lower insulin sensitivity. 
+These four fatty acids are also involved in the DNL of refined and simple 
+carbohydrates. Our results, which are in agreement with the current literature,
+suggest that higher activity of DNL (potentially through higher intakes of
+simple carbohydrates) may increase the risk for diabetes, likely through
+worsening insulin sensitivity.
 
@@
 To our knowledge, no longitudinal study has examined the role of the composition
-of the TAGFA fraction on detailed OGTT-derived metabolic measures using 
-advanced {{another word?}} statistical techniques designed to maximally utilize
-the data. Two large prospective studies have been published that had similar
-research objectives as the present analysis. One was a nested case-control
-cohort (n=189 cases and n=189 controls) of the Framingham offspring study
-[@Rhee2015a], which found that TAGFA composition with a lower carbon chain and
-less double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
+of the TAGFA fraction on detailed OGTT-derived metabolic measures using advanced
+{{another word? appropriate?}} statistical techniques designed to maximally
+utilize the data. Two large prospective studies have been published that had
+similar research objectives as the present analysis. One was a nested
+case-control cohort (n=189 cases and n=189 controls) of the Framingham offspring
+study [@Rhee2015a], which found that TAGFA composition with a lower carbon chain
+and less double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
 12-years while higher carbon chain and more double bond TAGFA had a lower risk 
@@
 between studies, particularly with the four DNL fatty acids. A limitation of 
-this study was the simpler statistical techniques used to analyze the data, 
-which is not able to adequately use the longitudinal and high-dimensional 
+the study was the simpler statistical techniques used to analyze the data, 
+which are not able to adequately use the longitudinal and high-dimensional 
 dataset. {{ They're article and methods were ... questionable... }} In another 
-study of a much smaller (n=16) mostly female group, the authors report a 
-positive correlation between total esterified (of which TAG make up the 
-majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR [@Kotronen2009a], similar as 
-our results except for 18:1n-9.
+study of a much smaller (n=16) mostly female group [@Kotronen2009a], the authors
+report a positive correlation between total esterified (of which TAG make up the
+majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, similar as our results except
+for the 18:1n-9 finding.
 
@@
 link between higher estimated DNL and an increased risk for metabolic 
-dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a] as well as
-studies showing higher refined carbohydrates (e.g. added sugars) and a higher
-risk for diabetes {{cite}}. Our study extends these findings on DNL by showing
-that TAGFA with 14 to 16 carbons all strongly predicted lower insulin
-sensitivity. While these fatty acids also had a significant association with
-beta-cell function, the results were not as strong as with the insulin
-sensitivity.
+dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. Our study
+extends these findings on DNL by showing that TAGFA with 14 to 16 carbons all
+strongly predicted lower insulin sensitivity. While these fatty acids also had a
+significant association with beta-cell function, the results were not as strong
+as with the insulin sensitivity.
 
 The link between higher DNL and increases in specific fatty acids has been
-examined in several studies. Previous studies have examined used markers
-of DNL, such as the ratio between 18:2n-6 to 16:0 or 16:0 to 14:0 ratio
-[@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are major
+examined in several studies. Previous studies that have examined DNL have used
+markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
+to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are major
 limitations to using these ratios as the fatty acids used in their calculation
-can be created from DNL, they are also obtained from the diet [@Hodson2008a]. A
-short, small feeding trial (n=24) was conducted to identify fatty acids that
-most accurately reflected DNL, as potential biomarkers [@Lee2015a]. The study
-found that palmitoleic acid (16:1n-7), directly measured DNL using isotopes, and
-liver fat were all highly correlated with each other (r>0.50), suggesting that
-16:1n-7 may be a good biomarker for hepatic DNL. In another small (n=14) feeding
-trial, meal type (high fat vs low fat) was tested to determine its effect on DNL
-and TAGFA composition [@Wilke2009a]. The authors reported that 14:0, 16:0, 16:1,
-and 18:2 were higher in the low fat (high carbohydrate) group. These fatty acids
-are similar to the fatty acids we found that clustered together, implicating 
-these fatty acids as indicative of a higher carbohydrate diet. A higher
-carbohydrate diet, particularly with simple carbohydrates, may lead to greater
-DNL in an attempt to control blood glucose, increasing hepatic fat stores, and
-increasing the amount of TAGFA in circulation. The increase in circulating 14
-and 16 carbon fatty acids may then increase the exposure of tissues to lipotoxicity,
-for instance from palmitic acid (16:0), which is well-known to have harmful
-effects on tissues [@Riserus2008a; @Iggman2010a].
+can also be obtained from the diet in addition to being created through DNL
+[@Hodson2008a]. A short, small feeding trial (n=24) was conducted to identify
+fatty acids that most accurately reflected DNL, as potential biomarkers
+[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
+DNL using isotopes, and liver fat were all highly correlated with each other
+(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
+another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
+to determine its effect on DNL and TAGFA composition [@Wilke2009a]. The authors
+reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
+carbohydrate) group. These fatty acids are similar to the fatty acids we found
+that clustered together using the PLS analysis, implicating these fatty acids as
+indicative of a higher carbohydrate diet. A higher carbohydrate diet,
+particularly with simple carbohydrates, may lead to greater DNL in an attempt to
+control blood glucose, increasing hepatic fat stores, and increasing the amount
+of TAGFA in circulation. The higher concentration of circulating 14 and 16
+carbon fatty acids may then expose tissues to greater lipotoxicity, for instance
+from palmitic acid (16:0), which is well-known to have harmful effects on
+tissues [@Riserus2008a; @Iggman2010a].
 
@@
 Analyzing these biological processes longitudinally is challenging due to the
-potential feedbacks involved. For instance, while greater DNL may promote muscle
-resistance, the reverse may also be true [@Flannery2012a]. Higher insulin
-resistance may encourage greater DNL to handle the higher blood glucose. To
-illustrate, in a weight loss intervention trial (n=19 with TAGFA data),
-participants who lost weight over 33 weeks showed higher insulin sensitivity and
-lower TAGFA composition indicative of lower DNL (less 14:0, 14:1, 16:0, 16:1,
-etc) [@Schwab2008a]. Given the complex biological mechanisms and feedback loops
-involved, disentangling whether insulin sensitivity influences TAGFA more
-strongly than TAGFA influencing insulin sensitivity will require complicated
-research designs and analyses. Our own null results on this question may
-tentatively point to an equal feedback mechansim between TAGFA and insulin
-sensitivity.
+potential feedback mechanisms involved. For instance, while greater DNL may
+promote muscle insulin resistance, the reverse may also be true
+[@Flannery2012a]. Higher insulin resistance may encourage greater DNL to handle
+the higher blood glucose. To illustrate this, in a weight loss intervention trial
+(n=19 with TAGFA data), participants who lost weight over 33 weeks showed higher
+insulin sensitivity and lower TAGFA composition indicative of lower DNL (less
+14:0, 14:1, 16:0, 16:1, etc) [@Schwab2008a]. Given the complex biological
+mechanisms and feedback loops involved, disentangling whether insulin
+sensitivity influences TAGFA more strongly than TAGFA influencing insulin
+sensitivity will require more complicated research designs and analyses. Our own
+null results on this question may tentatively point to an equally influential
+feedback mechanism between TAGFA and insulin sensitivity.
 
@@
 NEFA. A study examining this found that NEFA contributed 60-80% of newly synth
-VLDL-TAG during fasting and ~40% during fed while dietary fatty acids
+VLDL-TAG during fasting and ~40% during the fed state while dietary fatty acids
 contributed ~25% to newly synth VLDL-TAG [@Barrows2006a]. In an analysis of the
@@
 that may allow it to enter insulin sensitive tissues more easily than NEFA. It 
-may be that VLDL receptors allow fatty acids to enter these tissue more easily than 
-NEFA bound to albumin, or it may be due to the other influences on TAGFA
+may be that VLDL receptors allow fatty acids to enter these tissue more easily
+than NEFA bound to albumin, or it may be due to the other influences on TAGFA 
 composition such as dietary intake.
@@
 not present results on whole fraction for IR)
+
 {{ comment on unadjusted vs adjusted }}
 
-{{ commment on null interactions with time }}
+- Potential clinical application
 
@@
 
-Prospective cohorts:
-    - Chien K, et al. A prediction model for type 2 diabetes risk among Chinese people. Diabetologia. 2009;52(3):443–450.
-    - D’Agostino RB Jr, Hamman RF, Karter AJ, Mykkanen L, Wagenknecht LE, Haffner SM.
-Cardiovascular disease risk factors predict the development of type 2 diabetes: the insulin resistance atherosclerosis study.
-Diabetes Care.
-2004;27(9):2234–2240.
-    - Schulze MB, et al. Use of multiple metabolic and genetic markers to improve the prediction of type 2 diabetes: the EPIC-Potsdam Study. Diabetes Care. 2009;32(11):2116–2119.
 -->
+
+## Limitations
+
+There are a few important limitations to our study. In addition to the
+observational nature of the cohort, the biology underlying the interaction
+between TAGFA, insulin secretion, and insulin sensitivity is complex and
+multifaceted and as such there may be some residual confounding we were
+unaware of or had not been able to adjust for. However, we have taken extensive,
+empirically based precautions in identifying potential confounders and mediators
+through the use of the DAG modeling, relying on previous literature, and through
+QIC model fit comparison methods.
+
+<!--
+TAGFA can fluctuate substantially throughout the day and to control for this all
+PROMISE participants came for their clinic visits in the morning in a fasted
+state. Some of the inherent variability in TAGFA fluctuation can be managed by
+standardizing the data collection time, though there may still be some
+{{residual error? random error?}}.
+[@Rhee2011a] fasting TAG status better as discriminating between DM case vs
+control than 2hr-OGTT TAG
+-->
+
+TAGFA were only quantified at the baseline visit and as such we cannot
+investigate whether there are concomitant changes in TAGFA and the metabolic
+measures over time. In relation to this limitation, while GEE is well suited to
+longitudinal datasets, the complex biology and temporal nature of the
+associations may restrict the ability of GEE to appropriately analyze and
+interpret the data. However, to maximize the potential of GEE, we used the model
+to infer that a given value of TAGFA could predict values of insulin sensitivity
+or beta-cell function over a 6 year period. This in our view is a strength of
+our analysis, as it reduces the chance of reverse causality given the tight
+integration of the glucose and fatty acid metabolism pathways, as well as
+maximizes the specific usage of the GEE modeling.
+
+PLS is a well-established technique for constructing predictive models of high
+dimensionality data structures (i.e. fatty acid composition), however a 
+limitation is that the initial models analyzed through PLS and the final 
+computed scores are not able to control for potential confounders and other 
+effect modifiers. Likewise, PLS is not able to handle longitudinal data so only 
+the baseline visit was used in the PLS analysis. In light of these limitations, 
+we still observe concordant results between the PLS and GEE analyses.
+
+Even with these limitations, our study has several notable strengths, including
+the longitudinal design and the rigorous statistical techniques and methods
+applied in analyzing the data. These statistical techniques take advantage of
+the longitudinal data to allow appropriate investigation of temporal
+relationships and are able to handle the multivariate nature of the data.
+Lastly, our cohort contains highly detailed and comprehensive variable
+measurements at each collection visit, as well as having both concentration and
+mol% data for the fatty acids.
 

### 2017-01-23 doc/manuscript.Rmd
First (rough) draft of discussion - conclusion

@@
 
+## Final conclusion
+
+In conclusion, we found that a TAGFA composition indicative of higher DNL
+(containing higher 14:0, 14:1n-7, 16:0, and 16:1n-7) associated strongly lower
+insulin sensitivity and moderately with lower beta-cell function. The fatty
+acids that clustered together represent fatty acids created from DNL, which is
+characteristic of higher simple carbohydrate (e.g. added sugar) intake. Our
+results, which are congruent with current evidence, suggest that higher DNL
+likely due to greater intake of simple or refined carbohydrates may increase the
+risk of diabetes through worsening of insulin sensitivity.
+
 # Acknowledgements

### 2017-01-23 doc/manuscript.Rmd
Confirmed manuscript created as expected (minor edits, fixes)

@@
     - /home/luke/Documents/bibliography/bib/bib-db/methods.bib
+csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
 output: fost::default_manuscript
@@
 
+Word count: ~4200
+
 # Abstract
@@
 
-```{r load_data, message=FALSE}
+```{r load_data, message=FALSE, echo=FALSE}
 devtools::load_all('.')
@@
 available for fatty acid measurements (n=477; see the CONSORT diagram in 
-`r cite_sf('consort')`; a total of `r calc_n_for_visits()[1]` attended all three
+`r cite_f('consort')`; a total of `r calc_n_for_visits()[1]` attended all three
 visits). Metabolic characterization, anthropometric
@@
 stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
-of {{ confirm with Richard the correct name}} heptadecanoic acid (17:0) was added as an internal
+of {{confirm with Richard the correct name}} heptadecanoic acid (17:0) was added as an internal
 standard prior to extracting total lipids according to the method of Folch 
-[@Folch1957a]. Each serum lipid fraction---non-esterified fatty acids (NEFA), 
-cholesteryl ester, phospholipid, and TAGFA---was isolated using thin layer 
+[@Folch1957a]. Each serum lipid fraction (non-esterified fatty acids (NEFA), 
+cholesteryl ester, phospholipid, and TAGFA) was isolated using thin layer 
 chromatography. Fatty acid methyl esters were separated and quantified using a 
@@
 [@Textor2011a], http://dagitty.net/; see `r cite_f('dag_is')` and `r cite_f('dag_bcf')`), and from
-QIC. `r cite_t('tab_qic')` shows the covariates compared using QIC.
+QIC. `r cite_t('qic')` shows the covariates compared using QIC.
 While the final GEE model selected as best fitting differed between insulin
@@
 the model that had the fewest covariates and that had similar fit between the
-outcome measures ({{##}}). The final GEE model was adjusted for {{covariates}}.
+outcome measures. The final GEE model was adjusted for time, waist
+circumference, baseline age, ethnicity, sex, ALT, MET, and total NEFA.
 After scaling, log-transforming, and exponentiating, the GEE estimates are 
@@
 standard deviation (SD) increase in the predictor variable given the covariates 
-are held constant (including time). Lastly for the GEE models, we tested for an interaction with 
-sex, ethnicity, or time by the predictor term for each outcome variable.
+are held constant (including time). Lastly for the GEE models, we tested for an
+interaction with sex, ethnicity, or time by the predictor term for each outcome
+variable.
 
@@
 
-`r cite_t('tab_basic')` shows some basic characteristics of the PROMISE cohort. The mean 
+`r cite_t('basic')` shows some basic characteristics of the PROMISE cohort. The mean 
 follow-up time was `r calc_followup_time()`, where `r calc_pct_full_visits()`%
@@
 
-`r cite_f('fig_tagfa')` shows the composition of each FA in the TAG fraction (see 
-`r cite_t('tab_tagfa')` for a tabular presentation of the values). 
+`r cite_f('tagfa')` shows the composition of each FA in the TAG fraction (see 
+`r cite_t('tagfa')` for a tabular presentation of the values). 
 Three TAGFA contributed `r format_rounding(sum(tagfa$pct))`% to the total 
 TAG concentration: `r tagfa$c[1]`; `r tagfa$c[2]`; and, `r tagfa$c[3]`.
-`r fig_corr` shows a heatmap of the correlation of individual TAGFA with the
+`r cite_f('heatmap')` shows a heatmap of the correlation of individual TAGFA with the
 outcome variables and several basic characteristics.
@@
 There were `r nrow(dplyr::filter(gee_unadj_df, p.value <= 0.05))` FDR-corrected 
-significant associations in the unadjusted GEE model (see `r cite_f('fig_gee_unadj')`) and 
+significant associations in the unadjusted GEE model (see `r cite_f('gee_unadj')`) and 
 `r nrow(dplyr::filter(gee_adj_df, p.value <= 0.05))` FDR-corrected
-significant associations in the adjusted GEE model as shown in `r cite_f('fig_gee_adj')`. 
+significant associations in the adjusted GEE model as shown in `r cite_f('gee_adj')`. 
 The majority of associations with beta-cell function measures were attenuated
 after full model adjustment, while nearly all associations with insulin
-sensitivity remained significant. {{ confirm }} Subsequent analysis revealed
+sensitivity remained significant. 
+<!--{{ confirm }} Subsequent analysis revealed
 that the attenuation with beta-cell function was due primarily to adjustment for
-{{ waist? ALT? }}.
+{{ waist? ALT? }}.-->
 
@@
 interactions by time, sex, or ethnicity for any of the TAGFA (data not shown).
-A tabular presentation of the GEE results is shown in `r cite_t("tab_gee_unadj")` 
-for unadjusted models and `r cite_t("tab_gee_adj")` for adjusted models.
+A tabular presentation of the GEE results is shown in `r cite_t("gee_unadj")` 
+for unadjusted models and `r cite_t("gee_adj")` for adjusted models.
 
@@
 
-The PLS results for the insulin sensitivity (`r cite_f('fig_pls')`) revealed a clustering 
+The PLS results for the insulin sensitivity (`r cite_f('pls')`) revealed a clustering 
 of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly (meaning
@@
 case-control cohort (n=189 cases and n=189 controls) of the Framingham offspring
-study [@Rhee2015a], which found that TAGFA composition with a lower carbon chain
+study [@Rhee2011a], which found that TAGFA composition with a lower carbon chain
 and less double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
@@
 which are not able to adequately use the longitudinal and high-dimensional 
-dataset. {{ They're article and methods were ... questionable... }} In another 
+dataset. {{ Their article and methods were ... questionable... }} In another 
 study of a much smaller (n=16) mostly female group [@Kotronen2009a], the authors
@@
 large portion of fatty acids bound together as TAG are obtained from circulating
-NEFA. A study examining this found that NEFA contributed 60-80% of newly synth
+NEFA. A study examining this found that NEFA contributed 60-80% of newly synthesized
 VLDL-TAG during fasting and ~40% during the fed state while dietary fatty acids
-contributed ~25% to newly synth VLDL-TAG [@Barrows2006a]. In an analysis of the
+contributed ~25% to newly synthesized VLDL-TAG [@Barrows2006a]. In an analysis of the
 serum NEFA fraction in the PROMISE cohort with the OGTT-derived outcomes
@@
 
-```{r table_qic}
-analyze_qic() %>% 
-    table_qic(caption = tab_qic)
-```
-
 ```{r table_basic}
@@
 table_gee_main(gee_adj_df, caption = tab_gee_adj)
+```
+
+```{r table_qic}
+analyze_qic() %>% 
+    table_qic(caption = tab_qic)
 ```
@@
 
+<!--
 The number of components to extract from PLS was determined by 10-fold CV.
@@
 predictability of the given PLS model. Final results presented use the full dataset.a
+-->
 

### 2017-01-23 R/setup.R

@@ -27,6 +27,7 @@ set_options <- function() {
     # For the document (knitr)
     knitr::opts_chunk$set(
         warning = FALSE, message = FALSE, collapse = TRUE,
+        echo = FALSE,
         fig.showtext = TRUE, comment = '#>'
     )
 }

### 2017-01-30 doc/manuscript.Rmd
Created and added heatmap fig using mol%

@@
 Darker orange represents a positive correlation; darker blue represents a negative correlation.')
+fig_heatmap_mol <- fig('heatmap_mol', 'Pearson correlation heatmap of triacylglycerol fatty acids (mol%) with continuous basic and metabolic characteristics of PROMISE participants from the baseline visit (2004-2006).
+Darker orange represents a positive correlation; darker blue represents a negative correlation.')
 fig_heatmap_tagfa <- fig('heatmap_tagfa', 'Pearson correlation heatmap of the triacylglycerol fatty acids in the PROMISE participants from the baseline visit (2004-2006).
@@
 cor_df <- analyze_corr()
+cor_df_mol <- analyze_corr(y = tg_pct)
 cor_df_tagfa <- analyze_corr_tagfa()
@@
 (HOMA2-%S: r=`r fa_cor$rng['HOMA2-%S']`, ISI: r=`r fa_cor$rng$ISI`) than 
-for the beta-cell function measures (all r<0.30).
+for the beta-cell function measures (all r<0.30). A heatmap of the correlations
+of individual TAGFA using mol% with the basic characteristics is shown in 
+`r cite_f('heatmap_mol')` and the inter-correlation matrix of the TAGFA fraction 
+is shown in `r cite_f('heatmap_tagfa')`.
 
@@
 
+```{r fig_heatmap, fig.cap=fig_heatmap_mol}
+plot_heatmap(cor_df_mol, unit = 'mol%')
+```
+
 ```{r fig_heatmap_tagfa, fig.cap=fig_heatmap_tagfa}

### 2017-01-30 R/corr.R

@@ -70,12 +70,12 @@ analyze_corr_tagfa <-
 #' @param results Correlation results
 #'
 #' @export
-plot_heatmap <- function(results, text = TRUE) {
+plot_heatmap <- function(results, text = TRUE, unit = 'nmol/mL') {
      results %>%
         seer::view_heatmap(
             y = 'Vars2',
             x = 'Vars1',
-            ylab = 'Triacylglycerol fatty acids (nmol/mL)',
+            ylab = paste0('Triacylglycerol fatty acids (', unit, ')'),
             number.colours = 5,
             values.text = text,
             values.size = 4) +

### 2017-01-30 doc/manuscript.Rmd
Changed title to be results agnostic

@@
 
-Title: Fatty acids involved in de novo lipogenesis within the serum
-triacylglycerol fraction strongly predict lower insulin sensitivity.
+Title: Composition of fatty acids within the serum triacylglycerol fraction and 
+the association with the pathogenesis of diabetes in a population at-risk for 
+diabetes
 

### 2017-01-31 doc/manuscript.Rmd
Removed FDR-corrected numbers in Results

@@
 
-There were `r nrow(dplyr::filter(gee_unadj_df, p.value <= 0.05))` FDR-corrected 
-significant associations in the unadjusted GEE model (see `r cite_f('gee_unadj')`) and 
-`r nrow(dplyr::filter(gee_adj_df, p.value <= 0.05))` FDR-corrected
-significant associations in the adjusted GEE model as shown in `r cite_f('gee_adj')`. 
+Results from the unadjusted GEE model are shown in `r cite_f('gee_unadj')`) and 
+for the adjusted GEE model in `r cite_f('gee_adj')`. 
 The majority of associations with beta-cell function measures were attenuated

### 2017-01-31 doc/manuscript.Rmd
Added Bernie onto author list
Moved degree before affil number
Moved more detailed GEE stats methods to Supplemental Material

@@
 
-Author: Luke W. Johnston (1), MSc; Stewart B. Harris (2), MD; Ravi Retnakaran
-(3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
-Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,6), PhD
-
-{{Bernie?}}
+Author: Luke W. Johnston, MSc (1); Stewart B. Harris, MD (2); Ravi Retnakaran, MD
+(3,4); Bernard Zinman, MD (3,4); Adria Giacca, PhD (5); Zhen Liu, PhD (1);
+Richard P. Bazinet, PhD (1); and Anthony J. Hanley, PhD (1,6)
 
@@
 primary analysis to determine the longitudinal associations between the outcome 
-variables and the predictor variables. Given the longitudinal design, an
-auto-regressive of order 1 (AR1) working correlation matrix was chosen for the GEE
-models as it had the best model fit when assessed using quasi-likelihood
-information criteria (QIC), though other matrices (eg. exchangeable)
-had similar fit (data not shown). GEE is well suited to longitudinal cohort 
-studies given it's capacity to handle missed visits. The predictor variables and
-continuous covariates were scaled (mean-centered and standardized). The TAGFA,
-total NEFA, sex, ethnicity, and baseline age were classified as 
-*time-independent* (held constant) as they were measured only at the baseline 
-visit or do not change throughout the study, while the outcome variables and 
-remaining covariates were set as *time-dependent*. No imputation was conducted 
-on missing values. Given how the samples for TAGFA measurement were determined
-and how the outcome variables are calculated, cases of diabetes at baseline and
-followup were not included in the GEE modeling.
-
-Covariates were selected based on the previous literature, from directed acyclic
-graph [@Greenland1999a] recommendations (using the DAGitty software
-[@Textor2011a], http://dagitty.net/; see `r cite_f('dag_is')` and `r cite_f('dag_bcf')`), and from
-QIC. `r cite_t('qic')` shows the covariates compared using QIC.
-While the final GEE model selected as best fitting differed between insulin
-sensitivity and beta-cell function measures, the differences in QIC values were
-less than 10 between many of the models, suggesting similar fit. As such, we selected
-the model that had the fewest covariates and that had similar fit between the
-outcome measures. The final GEE model was adjusted for time, waist
-circumference, baseline age, ethnicity, sex, ALT, MET, and total NEFA.
-After scaling, log-transforming, and exponentiating, the GEE estimates are 
-interpreted as an expected percent difference in the outcome variable for every 
-standard deviation (SD) increase in the predictor variable given the covariates 
-are held constant (including time). Lastly for the GEE models, we tested for an
-interaction with sex, ethnicity, or time by the predictor term for each outcome
-variable.
+variables and the predictor variables. The predictor variables and continuous 
+covariates were scaled (mean-centered and standardized). Given the longitudinal 
+design, an auto-regressive of order 1 (AR1) working correlation matrix was 
+specified in the GEE model. Covariates to adjust for were selected based on the
+previous literature, from directed acyclic graph [@Greenland1999a]
+recommendations, and from quasi-likelihood information criteria (QIC). The final
+GEE model was adjusted for time, waist circumference, baseline age, ethnicity,
+sex, ALT, MET, and total NEFA. The TAGFA, total NEFA, sex, ethnicity, and
+baseline age were classified as *time-independent* (held constant) as they were
+measured only at the baseline visit or do not change throughout the study, while
+the outcome variables and remaining covariates were set as *time-dependent*.
+After transformations, the GEE estimates are interpreted as an expected percent 
+difference in the outcome variable for every standard deviation (SD) increase in
+the predictor variable given the covariates are held constant (including time). 
+We also tested for an interaction with sex, ethnicity, or time by the predictor 
+term for each outcome variable. See the Supplemental Methods for an expanded 
+explanation of the GEE modeling analysis.
 
@@
 and beta-cell function. No prevalent diabetes cases were included in the PLS
-analysis. For more detailed explanation of GEE and PLS, please see the
-{{supplemental}} Methods.
+analysis. For more detailed explanation of PLS, please see the Supplemental
+Methods.
 
@@
 
+## Supplemental Methods: Statistical analysis
+
+Generalized estimating equations (GEE) is a technique similar to mixed effects
+modeling, except it calculates the marginal population averages compared to the
+individual averages in mixed effects models. GEE is well suited and commonly
+used in longitudinal cohort studies, especially given it's capacity to handle
+missed visits.
+
+The working correlation matrix for the GEE analysis was chosen based on 
+quasi-likelihood information criteria (QIC). The auto-regressive of order 1
+(AR1) matrix was chosen for the GEE models as it had the best model fit accessed
+using QIC, though other matrices (eg. exchangeable) had similar fit (data not
+shown).
+
+For the confounders, they were chosen based on previous literature, from
+directed acyclic graph (DAG) [@Greenland1999a] recommendations, and from
+QIC. The DAG recommendations were obtained from using the DAGitty software
+[@Textor2011a], http://dagitty.net/. See `r cite_f('dag_is')` and 
+`r cite_f('dag_bcf')`) for the DAG model. See `r cite_t('qic')` for the
+comparison of various models with different covariates using QIC. While the
+final GEE model selected as best fitting differed between insulin sensitivity
+and beta-cell function measures, the differences in QIC values were less than 10
+between many of the models, suggesting similar fit. As such, we selected the
+model that had the fewest covariates and that had similar fit between the
+outcome measures.
+
+No imputation was conducted on missing values. Given how the samples for TAGFA
+measurement were determined and how the outcome variables are calculated, cases
+of diabetes at baseline and followup were not included in the GEE modeling. The
+resulting GEE beta estimates were exponentiated to allow the interpretation as
+stated in the main methods.
+
 <!--

### 2017-02-01 doc/manuscript.Rmd
Moved detailed PLS results to Supplemental Material

@@
 
-To determine which components to use, internal CV showed that the first two
-components gave the highest amount of explained variance (data not shown).
 Confirming the findings from the GEE models, PLS analysis on the beta-cell
@@
 sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3, and 22:6n-3 loaded positively
-on both insulin sensitivity measures. No other TAGFA loaded strongly, the
-results of which remained after running the PLS model on the cross-validated
-testing set. The PLS model for insulin sensitivity had good predictive ability, 
-determined by correlating the predicted outcome values with the actual values
-(HOMA2-%S: `r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).
+on both insulin sensitivity measures. No other TAGFA loaded strongly. The PLS
+model for insulin sensitivity had good predictive ability, determined by
+correlating the predicted outcome values with the actual values (HOMA2-%S: `r
+cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).
 
@@
 Generalized estimating equations (GEE) is a technique similar to mixed effects
-modeling, except it calculates the marginal population averages compared to the
-individual averages in mixed effects models. GEE is well suited and commonly
+modeling, except it calculates the marginal population estimates compared to the
+subject-specific estimates in mixed effects models. GEE is well suited and commonly
 used in longitudinal cohort studies, especially given it's capacity to handle
@@
 
+We used internal 10-fold cross-validation to determine which components to
+extract from the PLS analysis. Based on the internal cross-validation, the first
+two components gave the highest amount of explained variance (data not shown),
+of which we decided to use these two components in the final results.
+
+The predictive capability of the PLS models were tested using cross-validation. 
+The data set was randomly split in half into a training set and a testing set. 
+After specifying the model on the training set, results were compared using the 
+testing set to determine whether how predictive the model was given a new
+dataset. Final results shown in the figures use the full dataset (rather than a
+training or testing set)
+
 # References

### 2017-02-02 doc/manuscript.Rmd
Moved some tables and figs into Supplemental Results

@@
 
-tab_qic <- tab('qic', 'Comparison of GEE model fitness for variable selection using quasi-likelihood information criteria.')
+tab_qic <- stab('qic', 'Comparison of GEE model fitness for variable selection using quasi-likelihood information criteria.')
 tab_basic <- tab('basic', 'Basic characteristics of PROMISE participants at each of the 3 clinic visits.')
-tab_tagfa <- tab('tagfa', 'Concentration (nmol/mL) and relative percent (mol%) values of triacylglycerol fatty acids in PROMISE participants at the baseline visit (2004-2006).')
-tab_gee_unadj <- tab('gee_unadj', 'Raw estimates and confidence interval values for *time*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+tab_tagfa <- stab('tagfa', 'Concentration (nmol/mL) and relative percent (mol%) values of triacylglycerol fatty acids in PROMISE participants at the baseline visit (2004-2006).')
+tab_gee_unadj <- stab('gee_unadj', 'Raw estimates and confidence interval values for *time*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
 Estimates represent a percent difference in the outcome per SD increase in the fatty acid.
 P-values were adjusted for the BH false discovery rate, with an asterisk (*) denoting a significant (p<0.05) association.')
-tab_gee_adj <- tab('gee_adj', 'Raw estimates and confidence interval values for *fully*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+tab_gee_adj <- stab('gee_adj', 'Raw estimates and confidence interval values for *fully*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
 Variables controlled for were follow-up time, waist circumference, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
@@
 
-fig_consort <- fig('consort', 'CONSORT diagram of PROMISE participants over the 3 visits.')
-fig_dag_is <- fig('dag_is', 'Directed acyclic graphic output from the DAGitty online software for insulin sensitivity.')
-fig_dag_bcf <- fig('dag_bcf', 'Directed acyclic graphic output from the DAGitty online software for beta-cell function.')
+fig_consort <- sfig('consort', 'CONSORT diagram of PROMISE participants over the 3 visits.')
+fig_dag_is <- sfig('dag_is', 'Directed acyclic graphic output from the DAGitty online software for insulin sensitivity.')
+fig_dag_bcf <- sfig('dag_bcf', 'Directed acyclic graphic output from the DAGitty online software for beta-cell function.')
 fig_tagfa <- fig('tagfa', 'Distribution of the composition of triacylglycerol fatty acids in the baseline visit of PROMISE participants (2004-2006).
@@
 available for fatty acid measurements (n=477; see the CONSORT diagram in 
-`r cite_f('consort')`; a total of `r calc_n_for_visits()[1]` attended all three
+`r cite_sf('consort')`; a total of `r calc_n_for_visits()[1]` attended all three
 visits). Metabolic characterization, anthropometric
@@
 `r cite_f('tagfa')` shows the composition of each FA in the TAG fraction (see 
-`r cite_t('tagfa')` for a tabular presentation of the values). 
+`r cite_st('tagfa')` for a tabular presentation of the values). 
 Three TAGFA contributed `r format_rounding(sum(tagfa$pct))`% to the total 
@@
 interactions by time, sex, or ethnicity for any of the TAGFA (data not shown).
-A tabular presentation of the GEE results is shown in `r cite_t("gee_unadj")` 
-for unadjusted models and `r cite_t("gee_adj")` for adjusted models.
+A tabular presentation of the GEE results is shown in `r cite_st("gee_unadj")` 
+for unadjusted models and `r cite_st("gee_adj")` for adjusted models.
 
@@
 
-```{r table_tagfa}
-table_tagfa(caption = tab_tagfa)
-```
-
-```{r table_gee}
-table_gee_main(gee_unadj_df, caption = tab_gee_unadj)
-table_gee_main(gee_adj_df, caption = tab_gee_adj)
-```
-
-```{r table_qic}
-analyze_qic() %>% 
-    table_qic(caption = tab_qic)
-```
-
 # Figures
-
-![`r fig_consort`](../img/flowDiagramSample.png)
-
-![`r fig_dag_is`](../img/dagitty-model-is.png)
-
-![`r fig_dag_bcf`](../img/dagitty-model-bcf.png)
 
@@
 
-```{r fig_heatmap, fig.cap=fig_heatmap_mol}
+```{r fig_heatmap_mol, fig.cap=fig_heatmap_mol}
 plot_heatmap(cor_df_mol, unit = 'mol%')
@@
 QIC. The DAG recommendations were obtained from using the DAGitty software
-[@Textor2011a], http://dagitty.net/. See `r cite_f('dag_is')` and 
-`r cite_f('dag_bcf')`) for the DAG model. See `r cite_t('qic')` for the
+[@Textor2011a], http://dagitty.net/. See `r cite_sf('dag_is')` and 
+`r cite_sf('dag_bcf')`) for the DAG model. See `r cite_st('qic')` for the
 comparison of various models with different covariates using QIC. While the
@@
 
+## Supplemental Tables
+
+```{r table_qic}
+analyze_qic() %>% 
+    table_qic(caption = tab_qic)
+```
+
+```{r table_tagfa}
+table_tagfa(caption = tab_tagfa)
+```
+
+```{r table_gee}
+table_gee_main(gee_unadj_df, caption = tab_gee_unadj)
+table_gee_main(gee_adj_df, caption = tab_gee_adj)
+```
+
+## Supplemental Figures
+
+![`r fig_consort`](../img/flowDiagramSample.png)
+
+![`r fig_dag_is`](../img/dagitty-model-is.png)
+
+![`r fig_dag_bcf`](../img/dagitty-model-bcf.png)
+
 # References

### 2017-02-02 R/qic.R
Updated QIC with models excluding waist

@@ -17,7 +17,7 @@ prep_qic_data <- function(data) {
             TotalTG,
             TotalNE,
             Sex,
-            Ethnicity,
+            BiEthnicity,
             MET,
             ALT,
             BaseAge,
@@ -67,22 +67,25 @@ qic_is <- function(data = project_data) {
     )
 
     Int <- stats::update(M0, . ~ . + TotalTG:YearsFromBaseline)
-    Full <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET +
+    Full <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT + MET +
                 AlcoholPerWk + FamHistDiab + TobaccoUse + TotalNE)
-    M1 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
-    M2 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT)
-    M3 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET)
-    M4 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk)
-    M5 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk +
+    M1 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex)
+    M2 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT)
+    M3 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT + MET)
+    M4 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT + MET + AlcoholPerWk)
+    M5 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT + MET + AlcoholPerWk +
                           FamHistDiab)
-    M6 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TobaccoUse)
[…]

### 2017-02-03 doc/manuscript.Rmd
Added another figure of adj with and without waist to manuscript

@@
 Variables controlled for were follow-up time, waist circumference, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
+X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
+P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
+fig_gee_adj_nowaist <- sfig('gee_adj_nowaist', 'Fully-adjusted (without waist size) GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+Variables controlled for were follow-up time, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
 X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
@@
 interactions by time, sex, or ethnicity for any of the TAGFA (data not shown).
-A tabular presentation of the GEE results is shown in `r cite_st("gee_unadj")` 
-for unadjusted models and `r cite_st("gee_adj")` for adjusted models.
+We ran sensitivity analyses to identify which covariate attenuated the beta-cell
+function associations from the unadjusted model. We found that including waist
+circumference in the model lead to the attenuation of associations, shown in 
+`r cite_st("gee_adj_nowaist")`. A tabular presentation of the GEE results is
+shown in `r cite_st("gee_unadj")` for unadjusted models and 
+`r cite_st("gee_adj")` for adjusted models.
 
@@
 
+```{r fig_gee_adj_nowaist, fig.cap=fig_gee_adj_nowaist}
+plot_gee_main(analyze_gee(covars = covariates[which(covariates != 'Waist')]))
+```
+
 # References

### 2017-02-05 doc/manuscript.Rmd
Edits as per Tony's suggestions from doc # Create branch

@@
 (2) Centre for Studies in Family Medicine, University of Western Ontario, London, Ontario, Canada.
-(3) Division of Endocrinology, University of Toronto, Toronto, ON, Canada.
+(3) Leadership Sinai Centre for Diabetes, Division of Endocrinology, University of Toronto, Toronto, ON, Canada.
 (4) Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto, Ontario, Canada.
@@
 Hypertriglyceridemia is an extensively studied and well known factor involved in
-the dysregulation of metabolic function and subsequent negative health impacts.
+the dysregulation of metabolic function and subsequent negative health outcomes {{cite}}.
 It is a component of the metabolic syndrome [@Alberti2009], a strong risk factor for
-cardiovascular disease [@DAgostino2004a], and can contribute to the pathogenesis
-of diabetes. Measuring triacylglcerides (TAG) within the blood is commonly done
+cardiovascular disease [@DAgostino2004a], and has been shown to contribute to the pathogenesis
+of diabetes {{cite}}. The measurement of circulating triacylglcerides (TAG) is commonly done
 during routine clinical visits as part of health assessment. However, clinically
@@
 The interaction between TAG and insulin sensitivity is complex and has 
-components of a feedback system. Greater resistance to insulin in both the liver
+components of a feedback system {{cite}}. Greater resistance to insulin in both the liver
 and muscle may result in greater production of TAG and secretion of lipoproteins
@@
 metabolic dysfunction and lipotoxicity in various tissues, such as the 
-beta-cells, and thus continuing the cycle. Given the complexity and temporal
+beta-cells, and thus continuing the cycle {{cite}}. Given the complexity and temporal
 nature of the relationship, long term studies with multiple data collections are
@@
 
-While several prospective studies have identified a harmful association of 
-hypertriglyceridemia on greater incidence for type 2 diabetes 
+While several prospective studies have documented a prospective association of 
+hypertriglyceridemia with incident type 2 diabetes 
 [@Chien2008a; @DAgostino2004a; @Schulze2009a], there are limited longitudinal
@@
 specific composition of the serum TAG fraction on OGTT-derived measures of
-insulin sensitivity and beta-cell function in an Canadian population at risk for
-diabetes.
+insulin sensitivity and beta-cell function compared to clinically measured TAG
+in an Canadian population at risk for diabetes.
 
@@
 Recruitment for the baseline visit of the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
-cohort took place between 2004-2006 from London and Toronto, 
+cohort took place between 2004-2006 in London and Toronto, 
 Canada. Individuals were selected to participant if they met the eligibility 
@@
 2010-2013). Participants are contacted annually by telephone. The current study
-sample used data on participants who did not have diabetes at baseline, who
+used data on participants who did not have diabetes at baseline, who
 returned for one or more of the follow-up examinations, and who had samples
@@
 measurements, and questionnaires on lifestyle and sociodemographics are
-collected at each examination visit. Research ethics approval was obtained from
+administered at each examination visit. Research ethics approval was obtained from
 Mount Sinai Hospital and the University of Western Ontario, and all participants
@@
 
-Participants completed an 8-12 hour 75g oral glucose tolerance test (OGTT) at 
-each examination visit, with blood samples taken during fasting, 30 min, and 2 hr 
+After 8-12 hours of fasting, participants completed a 75g oral glucose tolerance test (OGTT) at 
+each examination visit, with blood samples taken at fasting, 30 min, and 2 hr 
 post-glucose load. Samples were subsequently processed and frozen at -70°C.
 Alanine aminotransferase (ALT) was measured using standard laboratory 
-procedures. Cholesterol, HDL, and clinically-measured triacylglycerides (TAG) 
+procedures. Cholesterol, HDL, and clinically-measured TAG 
 were measured using Roche Modular's enzymatic colorimetric tests (Mississauga, 
-ON).  Both insulin and glucose were derived from the OGTT at fasting, 30 minute,
+ON). At the 6-year visit only, clinically-measured TAG was only quantified on a
+subset of the participants (n=`r n_tag`). Both insulin and glucose were
+measured from OGTT blood samples at fasting, 30 minute,
 and 2 hour time points. Specific insulin was measured with the Elecsys 1010 
@@
 the past year. Each reported activity from the MAQ was weighted by its
-metabolic intensity allowing for the estimation of MET-hours per week.
+metabolic intensity allowing for the estimation of MET-hours per week [@Kriska1990].
 
@@
 ISSI-2, which were log-transformed for the statistical modeling. The primary
-predictor variables for this analysis were 22 individual TAGFA as mole percent
-(mol%) of the total fraction and as concentration (nmol/mL). Clinically-measured
-TAG is also included as a primary predictor to serve as a comparison against the
-TAGFA composition. A limitation with the clinically-measured TAG is that it was
-only quantified on a portion of the participants at the 6-year visit (n = `r n_tag`).
+predictor variables for this analysis were 22 individual TAGFA included as either mole percent
+(mol%) of the total fraction or as a concentration (nmol/mL). Clinically-measured
+TAG was also included as a primary predictor to allow us to test the hypothesis
+that specific TAGFA better predicted outcomes compared to clinical TAG. 
 Pearson correlation coefficients were computed to assess the relationships of
@@
 composition of the TAGFA fraction. To confirm the GEE results in a multivariate 
-environment (all TAGFA), partial least squares regression (PLS) was
+environment (i.e. all TAGFA analyzed collectively), partial least squares regression (PLS) was
 used to identify the patterns of TAGFA composition against insulin sensitivity
-and beta-cell function. No prevalent diabetes cases were included in the PLS
-analysis. For more detailed explanation of PLS, please see the Supplemental
-Methods.
+and beta-cell function. For a detailed explanation of PLS see the Supplemental
+Methods, but briefly, PLS is a technique that extracts latent structures
+(clusters) underlying a set of predictor variables conditional on a response
+variable(s) (i.e. the outcome variables). These extracted latent structures can
+be used to calculated the predicted value of the outcome values, determining how
+well TAGFA predict metabolic function by comparing to the measured metabolic function.
 
@@
 
-`r cite_t('basic')` shows some basic characteristics of the PROMISE cohort. The mean 
-follow-up time was `r calc_followup_time()`, where `r calc_pct_full_visits()`%
+`r cite_t('basic')` shows basic characteristics of the PROMISE cohort. The mean 
+follow-up time was `r calc_followup_time()` years, where `r calc_pct_full_visits()`%
 of participants attended all three visits. There were
@@
 age in years of `r aide::ave_sd(ds0yr$Age)` and a mean BMI
-of `r aide::ave_sd(ds0yr$BMI)`. Given the eligibility criteria, the majority of
+of `r aide::ave_sd(ds0yr$BMI)` kg/m^2^. As expected from the study's eligibility criteria, the majority of
 participants (`r fhd$npct`) had a family history of diabetes.
@@
 were `r dys$dm` and `r dys$pre` incident cases of diabetes and
-pre-diabetes (IFG and IGT), respectively.
+pre-diabetes (IFG and IGT), respectively, over the 6-years, of whom were excluded from later analyses.
 
@@
 Results from the unadjusted GEE model are shown in `r cite_f('gee_unadj')`) and 
-for the adjusted GEE model in `r cite_f('gee_adj')`. 
-The majority of associations with beta-cell function measures were attenuated
-after full model adjustment, while nearly all associations with insulin
-sensitivity remained significant. 
-<!--{{ confirm }} Subsequent analysis revealed
-that the attenuation with beta-cell function was due primarily to adjustment for
-{{ waist? ALT? }}.-->
+for the adjusted GEE model in `r cite_f('gee_adj')`. The majority of
+associations with beta-cell function measures were attenuated after full model
+adjustment, while nearly all associations with insulin sensitivity remained
+significant for both mol% and nmol/mL results. Subsequent analysis revealed that
+the attenuation with beta-cell function was due primarily to adjustment for
+waist.
 
-As a concentration, nearly all TAGFA had a strong negative association on
-HOMA2-%S and ISI (estimates ranging from 
+In analyses using TAGFA concentration values, nearly all TAGFA had a strong
+negative association on HOMA2-%S and ISI (estimates ranging from
 `r est1$outcome_b$rng[3]` and `r est1$outcome_b$rng[7]`, respectively), and a
@@
 from `r est1$outcome_b$rng[11]` and `r est1$outcome_b$rng[15]`, respectively).
-As a mol%, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7) had negative
+In analyses using TAGFA mol% values, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7) had negative
 associations with HOMA2-%S and ISI (between `r est1$outcome_b$rng[1]` and 
@@
 
-Confirming the findings from the GEE models, PLS analysis on the beta-cell
-function measures revealed that there was a low correlation (though significant
-at p`r cor_pls_igi$p`) between the predicted values and the actual values (r=
-`r cor_pls_igi$r` to `r cor_pls_issi2$r`), suggesting poor predictability. While 
-the clustering of TAGFA from the PLS results of the beta-cell function measures 
-were similar to the insulin sensitivity measures, only the insulin sensitivity 
-measures are presented given the higher predictive ability.
+The PLS analysis corroborated the findings from the GEE models. 
+The insulin sensitivity (`r cite_f('pls')`) PLS results showed a clustering 
+of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly
+and negatively on HOMA2-%S and ISI, suggesting this cluster of TAGFA tracks
+together with lower insulin sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3,
+and 22:6n-3 loaded positively on both insulin sensitivity measures. No other
+TAGFA loaded strongly. The PLS model for insulin sensitivity had good predictive
+ability, with a high correlation between the predicted outcome values against
+the actual values (HOMA2-%S: `r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).
 
-The PLS results for the insulin sensitivity (`r cite_f('pls')`) revealed a clustering 
-of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly (meaning
-the fatty acids fell between the dashed and solid circles) and negatively on HOMA2-%S
-and ISI, suggesting this cluster of TAGFA tracks together with lower insulin
-sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3, and 22:6n-3 loaded positively
-on both insulin sensitivity measures. No other TAGFA loaded strongly. The PLS
-model for insulin sensitivity had good predictive ability, determined by
-correlating the predicted outcome values with the actual values (HOMA2-%S: `r
-cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).
+The beta-cell function PLS results showed clustering of similar fatty acids,
+however there there was a low correlation (though significant at 
+p`r cor_pls_igi$p`) between the predicted values and the actual values 
+(r=`r cor_pls_igi$r` to `r cor_pls_issi2$r`), suggesting poor predictability.
+Given the low predictability, only the insulin sensitivity measures are
+presented.
 
@@
 negatively predicted lower insulin sensitivity. While most TAGFA were not
-associated with beta-cell function, two fatty acids, 16:0 and 18:1n-7, were
+associated with beta-cell function, two fatty acids, palmitic acid (16:0) and *cis*-vaccenic acid (18:1n-7), were
 associated negatively and positively, respectively, with both measures of 
 beta-cell function. We also found that four TAGFA (14:0, 14:1n-7, 16:0, 16:1n-7)
-clustered together, and that also strongly predicted lower insulin sensitivity. 
-These four fatty acids are also involved in the DNL of refined and simple 
-carbohydrates. Our results, which are in agreement with the current literature,
+clustered together, and this cluster also strongly predicted lower insulin sensitivity. 
+These four fatty acids are also involved in the *de novo* lipogenesis (DNL) of refined and simple 
+carbohydrates [@Hodson2008a; @Kawano2013a]. Our results, which are in agreement with the current literature,
 suggest that higher activity of DNL (potentially through higher intakes of
@@
 
-To our knowledge, no longitudinal study has examined the role of the composition
+To our knowledge, no longitudinal study to date has examined the role of the composition
 of the TAGFA fraction on detailed OGTT-derived metabolic measures using advanced
-{{another word? appropriate?}} statistical techniques designed to maximally
+{{another word? maybe 'appropriate'?}} statistical techniques designed to maximally
 utilize the data. Two large prospective studies have been published that had
 similar research objectives as the present analysis. One was a nested
-case-control cohort (n=189 cases and n=189 controls) of the Framingham offspring
-study [@Rhee2011a], which found that TAGFA composition with a lower carbon chain
-and less double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
+case-control analysis (n=189 cases and n=189 controls) within the Framingham offspring
+cohort [@Rhee2011a], which found that TAGFA composition characterized by a lower carbon chain
+and fewer double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
 12-years while higher carbon chain and more double bond TAGFA had a lower risk 
@@
 they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
-was a prospective cohort of males in Finland [@Lankinen2015a], of which they had
-TAGFA data on 831 participants after 6-years from the baseline visit. This 
+was a prospective cohort of males in Finland [@Lankinen2015a], for which 
+TAGFA data were available for 831 participants after 6-years of follow-up. This 
 cohort has similar outcomes as PROMISE, however, OGTT data was only available at
-the 6-year visit. A beta-cell function measure was computed, but only included 
-calculations up to 30 minutes of the OGTT. As with our results, they found a 
-larger effect size for insulin sensitivity compared to the beta-cell function. 
-The specific TAGFA that associated with the metabolic measures were similar 
-between studies, particularly with the four DNL fatty acids. A limitation of 
-the study was the simpler statistical techniques used to analyze the data, 
-which are not able to adequately use the longitudinal and high-dimensional 
-dataset. {{ Their article and methods were ... questionable... }} In another 
-study of a much smaller (n=16) mostly female group [@Kotronen2009a], the authors
-report a positive correlation between total esterified (of which TAG make up the
+the 6-year visit. As with our results, they found a larger effect size for
+insulin sensitivity compared to the beta-cell function. There were some
+similarities in results between the present analysis and the Finnish study,
+specifically for the saturated fatty acids, docosapentaenoic acid (22:5n-3),
+eicosapentaenoic acid (20:5n-3), and arachidonic acid (20:4n-6), though they did
+not find any associations with 7-tetradecenoic acid (14:1n-7) and palmitoleic
+acid (16:1n-7). Our study extends these findings by using multiple measurements
+of metabolic function and using statistical techniques that adequately make use
+of the longitudinal and high-dimensional dataset. In another study of a much
+smaller (n=16) mostly female group [@Kotronen2009a], the authors report a
+positive correlation between total esterified (of which TAG make up the
 majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, similar as our results except
@@
 dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. Our study
-extends these findings on DNL by showing that TAGFA with 14 to 16 carbons all
+extends these findings by showing that TAGFA with 14 to 16 carbons clustered together and all
 strongly predicted lower insulin sensitivity. While these fatty acids also had a
@@
 can also be obtained from the diet in addition to being created through DNL
-[@Hodson2008a]. A short, small feeding trial (n=24) was conducted to identify
+[@Hodson2008a]. A small, short feeding trial (n=24) was conducted to identify
 fatty acids that most accurately reflected DNL, as potential biomarkers
@@
 indicative of a higher carbohydrate diet. A higher carbohydrate diet, 
-particularly with simple carbohydrates, may lead to greater DNL in an attempt to
-control blood glucose, increasing hepatic fat stores, and increasing the amount
-of TAGFA in circulation. The higher concentration of circulating 14 and 16
-carbon fatty acids may then expose tissues to greater lipotoxicity, for instance
-from palmitic acid (16:0), which is well-known to have harmful effects on
-tissues [@Riserus2008a; @Iggman2010a].
+particularly one characterized by a predominance of simple carbohydrates, may
+lead to greater DNL in an attempt to control blood glucose, thus increasing hepatic
+fat stores and consequently increasing the amount of TAGFA in circulation {{cite}}. The higher
+concentration of circulating 14 and 16 carbon fatty acids may then expose
+tissues to greater lipotoxicity, for instance from palmitic acid (16:0), which
+is well-known to have harmful effects on tissues [@Riserus2008a; @Iggman2010a].
 
@@
 null results on this question may tentatively point to an equally influential
-feedback mechanism between TAGFA and insulin sensitivity.
+feedback mechanism between TAGFA and insulin sensitivity {{rephrase after exploring time interaction}}.
 
@@
 serum NEFA fraction in the PROMISE cohort with the OGTT-derived outcomes
-(currently unpublished), we found that higher total NEFA but not the specific
+(under review), we found that higher total NEFA but not the specific
 composition associated with lower beta-cell function and not with insulin 
@@
 than NEFA bound to albumin, or it may be due to the other influences on TAGFA 
-composition such as dietary intake.
+composition such as dietary intake {{cite?}}.
 
+<!--
 {{ Not sure about this... Include this? }}
@@
 
-<!--
 {{ comment on comparing models with and without NEFA }} While
@@
 investigate whether there are concomitant changes in TAGFA and the metabolic
-measures over time. In relation to this limitation, while GEE is well suited to
-longitudinal datasets, the complex biology and temporal nature of the
-associations may restrict the ability of GEE to appropriately analyze and
-interpret the data. However, to maximize the potential of GEE, we used the model
-to infer that a given value of TAGFA could predict values of insulin sensitivity
-or beta-cell function over a 6 year period. This in our view is a strength of
-our analysis, as it reduces the chance of reverse causality given the tight
-integration of the glucose and fatty acid metabolism pathways, as well as
-maximizes the specific usage of the GEE modeling.
+measures over time. However, to optimally use GEE to analyze the data and for
+interpretation, we used the model to infer that a given value of TAGFA could
+predict values of insulin sensitivity or beta-cell function over a 6 year
+period. This in our view is a strength of our analysis, as it reduces the chance
+of reverse causality given the tight integration of the glucose and fatty acid
+metabolism pathways, as well as maximizes the specific usage of the GEE
+modeling.
 
@@
 effect modifiers. Likewise, PLS is not able to handle longitudinal data so only 
-the baseline visit was used in the PLS analysis. In light of these limitations, 
-we still observe concordant results between the PLS and GEE analyses.
+the baseline visit was used in the PLS analysis, though we analyze the extracted
+scores using the GEE modeling to overcome this limitation. In light of these
+limitations, we still observe concordant results between the PLS and GEE analyses.
 
@@
 In conclusion, we found that a TAGFA composition indicative of higher DNL
-(containing higher 14:0, 14:1n-7, 16:0, and 16:1n-7) associated strongly lower
+(containing higher 14:0, 14:1n-7, 16:0, and 16:1n-7) associated strongly with lower
 insulin sensitivity and moderately with lower beta-cell function. The fatty
@@
 characteristic of higher simple carbohydrate (e.g. added sugar) intake. Our
-results, which are congruent with current evidence, suggest that higher DNL
-likely due to greater intake of simple or refined carbohydrates may increase the
+results, which are congruent with current evidence, suggest that higher DNL,
+likely due to greater intake of simple or refined carbohydrates, may increase the
 risk of diabetes through worsening of insulin sensitivity.
@@
 
-The authors thank Jan Neuman, Paula Van Nostrand, Stella Kink, and Annette
+The authors thank Jan Neuman, Paula Van Nostrand, Stella Kink, Nicole Rubio, and Annette
 Barnie of the Leadership Sinai Centre for Diabetes, Mount Sinai Hospital,
@@
 subject-specific estimates in mixed effects models. GEE is well suited and commonly
-used in longitudinal cohort studies, especially given it's capacity to handle
+used in longitudinal cohort studies, especially given its capacity to handle
 missed visits.
@@
 final GEE model selected as best fitting differed between insulin sensitivity
-and beta-cell function measures, the differences in QIC values were less than 10
+and beta-cell function measures, the differences in the QIC values were less than 10
 between many of the models, suggesting similar fit. As such, we selected the
 model that had the fewest covariates and that had similar fit between the
-outcome measures.
+outcome measures. {{NEFA comment for inclusion here}}
 
-No imputation was conducted on missing values. Given how the samples for TAGFA
-measurement were determined and how the outcome variables are calculated, cases
-of diabetes at baseline and followup were not included in the GEE modeling. The
-resulting GEE beta estimates were exponentiated to allow the interpretation as
-stated in the main methods.
+No imputation was conducted on missing values. Prevalent cases of diabetes at
+baseline and incident cases at follow-up were excluded from the GEE analysis. 
+The resulting GEE beta estimates were exponentiated to allow the interpretation
+as stated in the main methods.
 
@@
 training or testing set)
+
+No prevalent diabetes cases were included in the PLS
+analysis.
 

### 2017-02-07 doc/manuscript.Rmd
Created and added GEE with PLS scores code and results

@@
 cor_df_tagfa <- analyze_corr_tagfa()
+
+pls_gee <- project_data %>%
+    analyze_gee(
+        x = c('Comp1', 'Comp2'),
+        data_prep = 'pls',
+        pls_results = pls_isi
+        ) %>% 
+    dplyr::select(Yterms, Xterms, estimate, p.value) %>% 
+    dplyr::filter(Xterms == 'Comp1') %>% 
+    dplyr::mutate(p.value = format_p(p.value),
+                  estimate = format_rounding(estimate))
 ```
@@
 Given the low predictability, only the insulin sensitivity measures are
-presented.
+presented. We used the extracted PLS scores as the predictor variable in the GEE
+models and found negative associations of the first component on all outcome
+variables, with the strongest association being with the insulin sensitivity
+variables (beta=`r pls_gee$estimate[3]`, all p`r pls_gee$p.value[1]`; using PLS
+scores constrained by ISI).
 

### 2017-02-07 R/gee.R

@@ -36,6 +36,32 @@ prep_gee_data <- function(data) {
         dplyr::arrange(SID, VN)
 }
 
+#' Add the extracted PLS scores to the data for use in GEE.
+#'
+#' @param data The project data.
+#' @param pls_results Results dataset from PLS.
+#' @export
+prep_gee_pls_data <- function(data = project_data, pls_results) {
+    sid_vn <- data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select_(.dots = c(outcomes, tg_pct, 'SID', covariates)) %>%
+        stats::na.omit() %>%
+        dplyr::full_join(dplyr::select(project_data, SID, VN)) %>%
+        dplyr::select_(.dots = c(outcomes, 'SID', 'VN'))
+
+    pls_results %>%
+    {
+        dplyr::bind_cols(tibble::as_data_frame(.$model$Y),
+                         tibble::as_data_frame(unclass(.$scores))[1:2])
+    } %>%
+        dplyr::rename(Comp1 = `Comp 1`, Comp2 = `Comp 2`) %>%
+        dplyr::full_join(sid_vn) %>%
+        dplyr::select(SID, VN, Comp1, Comp2) %>%
+        dplyr::full_join(data) %>%
+        dplyr::arrange(SID, VN) %>%
+        tidyr::fill(Comp1, Comp2)
+}
+
 # Analyze -----------------------------------------------------------------
 
 #' Run GEE models on the prepared project data.
@@ -43,10 +69,15 @@ prep_gee_data <- function(data) {
 #' @param data The project data
 #' @param y outcomes (IS, BCF)
 #' @param x predictors (TAGFA)
-#' @param covariates to adjust for
 #' @param intvar interaction variable
 #' @param rename_x Function to rename x variables
+#' @param covars Covariates to include in the model
+#' @param data_prep Whether to clean the dataset before hand (either basic prep
+#'   or extracting PLS scores).
+#' @param pls_results Optional; if doing the data_prep for PLS, need to include
+#'   the pls results dataset.
 #' @param rename_y Function to rename y variables
+#'
 #' @export
 analyze_gee <- function(data = project_data,
                         y = outcomes,
@@ -58,7 +89,9 @@ analyze_gee <- function(data = project_data,
                         covars = covariates,
                         intvar = NULL,
                         rename_x = renaming_fats,
-                        rename_y = renaming_outcomes) {
+                        rename_y = renaming_outcomes,
[…]

### 2017-02-08 R/setup.R
Minor edits to settings

@@ -27,7 +27,6 @@ set_options <- function() {
     # For the document (knitr)
     knitr::opts_chunk$set(
         warning = FALSE, message = FALSE, collapse = TRUE,
-        echo = FALSE,
         fig.showtext = TRUE, comment = '#>'
     )
 }

### 2017-02-08 R/interaction.R
Created plots to view interaction with time in a vignette

@@ -1,0 +1,92 @@
+# Plot --------------------------------------------------------------------
+
+#' Interaction by time plots using GEE.
+#'
+#' @param data Project data
+#' @param y Outcome variable.
+#' @param x Predictor variable (Fatty acid).
+#' @param fatty.acid.levels Whether to split into median, tertiles, or quartiles.
+#' @param covars Covariates to adjust for.
+#' @param title The title letter of the plot.
+#' @param interaction.term Which variable to use for the interaction (VN).
+#' @param is.fatty.acid Whether the x is a fatty acids.
+#' @param y.axis.label The y axis label.
+#'
+#' @export
+plot_interaction <- function(data, y, x,
+                         fatty.acid.levels = c(2, 3, 4),
+                         covars = covariates,
+                         title = 'A',
+                         interaction.term = 'VN',
+                         is.fatty.acid = TRUE,
+                         y.axis.label = NULL) {
+    fa.lvl <- fatty.acid.levels
+    if (!fa.lvl %in% c(2, 3, 4)) {
+        print('Please select either 2, 3, or 4 for the fatty.acid.levels')
+    } else if (fa.lvl == 2) {
+        lvls <- 0:2/2
+        lvl.labs <- c('<50%', '>50%')
+    } else if (fa.lvl == 3) {
+        lvls <- 0:3/3
+        lvl.labs <- c('<33%', '33-66%', '>66%')
+    } else if (fa.lvl == 4) {
+        lvls <- 0:4/4
+        lvl.labs <- c('Lowest', 'Low', 'High', 'Highest')
+    }
+
+    if (is.null(y.axis.label)) {
+        y.axis.label <- paste0('log(', substring(y, 2), ')')
+    }
+
+    fatty.acid.name <- renaming_fats(x) %>%
+      gsub('(\\D\\D)\\.', '\\U\\1 ', ., perl = TRUE)
+
+    if (is.fatty.acid) {
+        fatty.acid.name <- paste(fatty.acid.name,
+                                 ifelse(grepl('pct_', x), '(mol%)', '\n(nmol/mL)'))
+    }
+
+    geeFormula <- paste(y, 'f.fat', sep = ' ~ ') %>%
+      paste(., paste(covars, collapse = ' + '),
+            paste('f.fat', interaction.term, sep = ':'), sep = ' + ') %>%
+      as.formula()
+
[…]

### 2017-02-08 doc/manuscript.Rmd
Added interaction with time, sex, ethnicity unadjusted results in Results

@@
 devtools::load_all('.')
+knitr::opts_chunk$set(echo = FALSE)
 ds0yr <- project_data %>% 
@@
 Both clinically-measured TAG and total TAGFA concentration had very strong
-negative associations with all outcome variables. There were no significant
-interactions by time, sex, or ethnicity for any of the TAGFA (data not shown).
-We ran sensitivity analyses to identify which covariate attenuated the beta-cell
-function associations from the unadjusted model. We found that including waist
-circumference in the model lead to the attenuation of associations, shown in 
-`r cite_st("gee_adj_nowaist")`. A tabular presentation of the GEE results is
-shown in `r cite_st("gee_unadj")` for unadjusted models and 
-`r cite_st("gee_adj")` for adjusted models.
+negative associations with all outcome variables. While there were a few
+significant interactions by time in unadjusted models, after inclusion of
+covariates in the model, these interactions were attenuated (data not
+shown). There were no significant interactions by sex or ethnicity for any of
+the TAGFA (data not shown). We ran sensitivity analyses to identify which
+covariate attenuated the beta-cell function associations from the unadjusted
+model. We found that including waist circumference in the model lead to the
+attenuation of associations, shown in `r cite_st("gee_adj_nowaist")`. A tabular
+presentation of the GEE results is shown in `r cite_st("gee_unadj")` for
+unadjusted models and `r cite_st("gee_adj")` for adjusted models.
 

### 2017-02-08 doc/manuscript.Rmd
Edits as per Tony's suggestions from email (esp re: vs Tg)

@@
 Hypertriglyceridemia is an extensively studied and well known factor involved in
-the dysregulation of metabolic function and subsequent negative health outcomes {{cite}}.
-It is a component of the metabolic syndrome [@Alberti2009], a strong risk factor for
-cardiovascular disease [@DAgostino2004a], and has been shown to contribute to the pathogenesis
-of diabetes {{cite}}. The measurement of circulating triacylglcerides (TAG) is commonly done
-during routine clinical visits as part of health assessment. However, clinically
-measured TAG is limited as it represents all the types of fatty acids within the
-TAG fraction as a single measure. Given the increasing appreciation for the role
-of the specific fatty acid composition on health outcomes, for instance with
-differences in associations of fatty acids within the PL lipid fraction
-[@Forouhi2014a; @Ma2015a] including an analysis from the PROMISE cohort
-[@Johnston2016a], there are relatively few studies that have explored this area
-of research in the TAG fraction.
+the dysregulation of metabolic function and subsequent negative health outcomes [@Chehade2013a].
+It is a component of the metabolic syndrome [@Alberti2009] and a strong risk factor for
+cardiovascular disease [@DAgostino2004a], in addition to it's role in metabolic
+dysfunction [@Verges2015a; @Kawano2013a]. The measurement of circulating 
+triacylglcerides (TAG) is commonly done during routine clinical visits as part
+of health assessment. However, clinically measured TAG is limited as it
+represents all the types of fatty acids within the TAG fraction as a single
+measure. Given the increasing appreciation for the role of the specific fatty
+acid composition on health outcomes, for instance with differences in
+associations of fatty acids within the PL lipid fraction [@Forouhi2014a; @Ma2015a] 
+including an analysis from the PROMISE cohort [@Johnston2016a], there
+are relatively few studies that have explored this area of research in the TAG
+fraction.
 
 The interaction between TAG and insulin sensitivity is complex and has 
-components of a feedback system {{cite}}. Greater resistance to insulin in both the liver
+components of a feedback system [@Verges2015a]. Greater resistance to insulin in both the liver
 and muscle may result in greater production of TAG and secretion of lipoproteins
@@
 metabolic dysfunction and lipotoxicity in various tissues, such as the 
-beta-cells, and thus continuing the cycle {{cite}}. Given the complexity and temporal
+beta-cells, and thus continuing the cycle [@Verges2015a]. Given the complexity and temporal
 nature of the relationship, long term studies with multiple data collections are
@@
 against gold standard measures [@Matthews1985; @Hermans1999a; @Matsuda1999; @Retnakaran2009].
-Specific formulas for each OGTT-derived measure can be found in the {{
-Supplementary Material }}.
+Specific formulas for each OGTT-derived measure can be found in the 
+Supplementary Material.
 
@@
 significant association with beta-cell function, the results were not as strong
-as with the insulin sensitivity.
+as with the insulin sensitivity. However, while the DNL cluster had a strong
+harmful association with insulin sensitivity when we include it in the GEE
+models, the magnitude of the association was not as large as when using
+clinically measured TAG, reinforcing that the total TAG concentration still more
+strongly predicts metabolic dysfunction.
 
@@
 lead to greater DNL in an attempt to control blood glucose, thus increasing hepatic
-fat stores and consequently increasing the amount of TAGFA in circulation {{cite}}. The higher
-concentration of circulating 14 and 16 carbon fatty acids may then expose
-tissues to greater lipotoxicity, for instance from palmitic acid (16:0), which
-is well-known to have harmful effects on tissues [@Riserus2008a; @Iggman2010a].
+fat stores and consequently increasing the amount of TAGFA in circulation
+[@Chehade2013a; @Verges2015a]. The higher concentration of circulating 14 and 16
+carbon fatty acids may then expose tissues to greater lipotoxicity, for instance
+from palmitic acid (16:0), which is well-known to have harmful effects on
+tissues [@Riserus2008a; @Iggman2010a].
 
@@
 sensitivity will require more complicated research designs and analyses. Our own
-null results on this question may tentatively point to an equally influential
-feedback mechanism between TAGFA and insulin sensitivity {{rephrase after exploring time interaction}}.
+findings of no interaction by time with the TAGFA tentatively suggest that the
+influence of either TAGFA or insulin resistance within the feedback loop may be
+of relatively equal strength.
 
@@
 may be that VLDL receptors allow fatty acids to enter these tissue more easily
-than NEFA bound to albumin, or it may be due to the other influences on TAGFA 
-composition such as dietary intake {{cite?}}.
+than NEFA bound to albumin or it may be due to the other influences on TAGFA 
+composition such as dietary intake since circulating NEFA primarily come from
+adipose tissue [@Verges2015a].
 
@@
 The authors had the following responsibility: LWJ conducted research, analyzed
-data, and wrote the paper; RR, ZL, and SBH designed research, conducted
+data, and wrote the paper; RR, BZ, ZL, and SBH designed research, conducted
 research, and provided essential materials (infrastructure and clinical
-resources); RR, SBH, RPB, and AG provided intellectual feedback on the
-paper; RPB conducted research, provided essential reagents and materials; AJH
+resources); RR, BZ, SBH, RPB, and AG provided intellectual feedback on the
+paper; RPB and ZL conducted research, provided essential reagents and materials; AJH
 designed research, assisted with interpretation, and provided intellectual
@@
 for final content. All authors read and approved the final manuscript.
-
-{{Bernie?}}
 

### 2017-02-08 doc/manuscript.Rmd
Expanded on wording of PLS methods as per Tony's email

@@
 
-# Supplemental Methods
+# Supplemental Material
 
@@
 modeling, except it calculates the marginal population estimates compared to the
-subject-specific estimates in mixed effects models. GEE is well suited and commonly
+subject-specific estimates in mixed effects models. GEE is well suited to and commonly
 used in longitudinal cohort studies, especially given its capacity to handle
@@
 model that had the fewest covariates and that had similar fit between the
-outcome measures. {{NEFA comment for inclusion here}}
+outcome measures. In particular for NEFA, it is used substantially as a source
+of fatty acids in hepatic TAG production [@Barrows2006a] and may be a strong
+confounder.
 
@@
 
-<!--
-The number of components to extract from PLS was determined by 10-fold CV.
-Dataset was split into a training set and a testing set to determine the
-predictability of the given PLS model. Final results presented use the full dataset.a
--->
-
+PLS is a technique used for multivariate, high dimensionality datasets where
+there is a potential or likely underlying structure to the data. Because it uses
+a response variable (i.e. the outcome or y variable) when identifying the 
+underlying structure it is known as a supervised statistical method, which gives
+it greater predictive power when using the model against updated or new data.
+PLS generates a number of components based on the number of variables provided.
 We used internal 10-fold cross-validation to determine which components to
@@
 dataset. Final results shown in the figures use the full dataset (rather than a
-training or testing set)
-
-No prevalent diabetes cases were included in the PLS
+training or testing set). No prevalent diabetes cases were included in the PLS 
 analysis.

### 2017-02-08 doc/manuscript.Rmd
Minor edits to TODO and manuscript

@@
 csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
-output: fost::default_manuscript
+output: fost::diabetologia_manuscript
 ---
@@
 model. We found that including waist circumference in the model lead to the
-attenuation of associations, shown in `r cite_st("gee_adj_nowaist")`. A tabular
+attenuation of associations, shown in `r cite_sf("gee_adj_nowaist")`. A tabular
 presentation of the GEE results is shown in `r cite_st("gee_unadj")` for

### 2017-02-08 doc/manuscript.Rmd
Included wording on using CV for prediction

@@
 (clusters) underlying a set of predictor variables conditional on a response
-variable(s) (i.e. the outcome variables). These extracted latent structures can
-be used to calculated the predicted value of the outcome values, determining how
-well TAGFA predict metabolic function by comparing to the measured metabolic function.
+variable(s) (i.e. the outcome variables). We determined the predictive strength
+of modeling the TAGFA composition on metabolic function by using
+cross-validation on the PLS model.
 

### 2017-02-08 R/variables.R
Changed clinical TAG as only baseline for GEE

@@ -1,5 +1,5 @@
 tg_conc <- grep('^tg\\d+', vars, value = TRUE)
 tg_pct <- grep('^pct_tg\\d+', vars, value = TRUE)
-tg_totals <- c('TotalTG', 'TAG')
+tg_totals <- c('TotalTG', 'BaseTAG')
 outcomes <- c('lHOMA2_S', 'lISI', 'lIGIIR', 'lISSI2')
 covariates <- c('YearsFromBaseline', 'Waist', 'BaseAge', 'BiEthnicity', 'Sex', 'ALT', 'MET', 'TotalNE')

### 2017-02-08 doc/manuscript.Rmd
Changed Zhen to second author
Rearranged diabetes after prediabetes

@@
 
-Author: Luke W. Johnston, MSc (1); Stewart B. Harris, MD (2); Ravi Retnakaran, MD
-(3,4); Bernard Zinman, MD (3,4); Adria Giacca, PhD (5); Zhen Liu, PhD (1);
+Author: Luke W. Johnston, MSc (1); Zhen Liu, PhD (1); Ravi Retnakaran, MD
+(3,4); Bernard Zinman, MD (3,4); Adria Giacca, PhD (5); Stewart B. Harris, MD (2); 
 Richard P. Bazinet, PhD (1); and Anthony J. Hanley, PhD (1,6)
@@
 between `r outcome_chg$chg` (p`r outcome_chg$p` from GEE; n=`r outcome_chg$n`). There 
-were `r dys$dm` and `r dys$pre` incident cases of diabetes and
-pre-diabetes (IFG and IGT), respectively, over the 6-years, of whom were excluded from later analyses.
+were `r dys$pre` and `r dys$dm` incident cases of pre-diabetes (IFG and IGT) and
+diabetes, respectively, over the 6-years, of whom were excluded from later
+analyses.
 

### 2017-02-16 doc/manuscript.Rmd
Made edits based on Tony's comments (not discussion)

@@
 Title: Composition of fatty acids within the serum triacylglycerol fraction and 
-the association with the pathogenesis of diabetes in a population at-risk for 
-diabetes
+the association with the pathogenesis of diabetes
 
@@
 The interaction between TAG and insulin sensitivity is complex and has 
-components of a feedback system [@Verges2015a]. Greater resistance to insulin in both the liver
-and muscle may result in greater production of TAG and secretion of lipoproteins
-that transport TAG [@Yu2012a]. Likewise, greater TAGFA may contribute to 
-metabolic dysfunction and lipotoxicity in various tissues, such as the 
-beta-cells, and thus continuing the cycle [@Verges2015a]. Given the complexity and temporal
-nature of the relationship, long term studies with multiple data collections are
-paramount to better understanding the underlying biology and subsequent risk.
+components of a feedback system [@Verges2015a]. Greater resistance to insulin in
+both the liver and muscle may result in greater production of TAG and secretion
+of lipoproteins that transport TAG [@Yu2012a]. Likewise, greater TAGFA may
+contribute to metabolic dysfunction and lipotoxicity in various tissues, such as
+the beta-cells, and thus continuing the cycle [@Verges2015a]. Given the
+complexity and temporal nature of the relationship, long term studies with
+multiple data collections are paramount to better understanding the underlying
+biology and subsequent risk.
 
@@
 [@Chien2008a; @DAgostino2004a; @Schulze2009a], there are limited longitudinal
-studies that have examined the relationship between TAG and it's composition on
-the pathophysiological factors underlying diabetes, particularly in regard to
-beta-cell function. Our objective was to examine the longitudinal role of the
-specific composition of the serum TAG fraction on OGTT-derived measures of
-insulin sensitivity and beta-cell function compared to clinically measured TAG
-in an Canadian population at risk for diabetes.
+studies [@Rhee2011a; @Lankinen2015a] that have examined the relationship between
+TAG and it's composition on the pathophysiological factors underlying diabetes, 
+particularly in regard to beta-cell function. Our objective was to examine the 
+longitudinal role of the specific composition of the serum TAG fraction on 
+OGTT-derived measures of insulin sensitivity and beta-cell function compared to 
+clinically measured TAG in an Canadian population at risk for diabetes.
 
@@
 fig_pls <- fig('pls', 'Partial least squares (PLS) models showing the clustering of triacylglycerol fatty acids on insulin sensitivity and beta-cell function measures.
-The R value shown above the graph is the correlation between predicted and actual values, using cross-validation; a higher value suggests the fatty acids strongly predict the metabolic outcome.
+The R value shown above the graph is the correlation between predicted and observed values, using cross-validation; a higher value suggests the fatty acids strongly predict the metabolic outcome.
 The percent explained variance of each component is shown in brackets on each axis.
@@
 
-# Correlation between predicted and actual
+# Correlation between predicted and observed
 pls_homa2_t <- analyze_pls(y = 'lHOMA2_S')
@@
 ```{r inline}
-n_tag <- project_data %>% 
-    dplyr::filter(VN == 2) %>% 
-    dplyr::select(TAG) %>% 
-    stats::na.omit() %>% 
-    nrow()
-
 dys <- project_data %>% 
@@
 est1 <- calc_gee_estci(gee_adj_df)
-fa_cor <- calc_cor()
+fa_cor <- calc_cor(cor_df)
+fa_cor_mol <- calc_cor(cor_df_mol)
 cor_pls_homa2 <- calc_pred_corr(pls_homa2_t, pls_homa2_t$test_data)
@@
 with three examination visits completed to date (2004-2006, 2007-2009, and
-2010-2013). Participants are contacted annually by telephone. The current study
+2010-2013). The current study
 used data on participants who did not have diabetes at baseline, who
@@
 available for fatty acid measurements (n=477; see the CONSORT diagram in 
-`r cite_sf('consort')`; a total of `r calc_n_for_visits()[1]` attended all three
-visits). Metabolic characterization, anthropometric
-measurements, and questionnaires on lifestyle and sociodemographics are
+`r cite_sf('consort')`). Metabolic characterization, anthropometric
+measurements, and questionnaires on lifestyle and sociodemographics were
 administered at each examination visit. Research ethics approval was obtained from
@@
 
-After 8-12 hours of fasting, participants completed a 75g oral glucose tolerance test (OGTT) at 
+After 8-12 hours of overnight fasting, participants completed a 75g oral glucose tolerance test (OGTT) at 
 each examination visit, with blood samples taken at fasting, 30 min, and 2 hr 
@@
 were measured using Roche Modular's enzymatic colorimetric tests (Mississauga, 
-ON). At the 6-year visit only, clinically-measured TAG was only quantified on a
-subset of the participants (n=`r n_tag`). Both insulin and glucose were
+ON). Both insulin and glucose were
 measured from OGTT blood samples at fasting, 30 minute,
@@
 chromatography peak areas to that of the internal standards [@Nishi2014a]. There
-were 22 fatty acids measured in the TAGFA fraction. Given that each lipid
-fraction differs in biology and given the complexity of the analyses, findings
+were 22 fatty acids measured in the TAGFA fraction. 
+Findings
 for other lipid fractions in this cohort are reported separately (see ref
@@
 Insulin sensitivity and beta-cell function indices were computed using the OGTT 
-glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S
-[@Levy1998a] using the HOMA2 Calculator and the Insulin Sensitivity Index (ISI)
-[@Matsuda1999]. HOMA largely reflects hepatic insulin resistance, while ISI
+glucose and insulin data. Insulin sensitivity was assessed using the Insulin
+Sensitivity Index (ISI) [@Matsuda1999] and HOMA2-%S [@Levy1998a] using the HOMA2
+Calculator. HOMA largely reflects hepatic insulin resistance, while ISI
 reflects whole-body insulin sensitivity [@AbdulGhani2007]. Beta-cell function
@@
 is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
-the disposition index (but using OGTT values). Each index has been validated
+the disposition index (but is calculated using OGTT values). Each index has been validated
 against gold standard measures [@Matthews1985; @Hermans1999a; @Matsuda1999; @Retnakaran2009].
@@
 individual TAGFA with other continuous variables. Correlations were also
-computed for TAGFA against each other, followed by using hierarchical clustering
-analysis to identify clusters within the correlation matrix of the TAGFA
+computed for TAGFA against each other; hierarchical clustering analysis was
+applied to the correlation matrix to identify potential clusters within the TAGFA 
 composition.
@@
 limited in that it cannot analyze the inherent multivariate nature of the 
-composition of the TAGFA fraction. To confirm the GEE results in a multivariate 
+composition of the TAGFA fraction. Therfore, to confirm the GEE results in a multivariate 
 environment (i.e. all TAGFA analyzed collectively), partial least squares regression (PLS) was
 used to identify the patterns of TAGFA composition against insulin sensitivity
-and beta-cell function. For a detailed explanation of PLS see the Supplemental
-Methods, but briefly, PLS is a technique that extracts latent structures
-(clusters) underlying a set of predictor variables conditional on a response
-variable(s) (i.e. the outcome variables). We determined the predictive strength
-of modeling the TAGFA composition on metabolic function by using
-cross-validation on the PLS model.
+and beta-cell function as outcome variables. For a detailed explanation of PLS
+see the Supplemental Methods. Briefly, PLS is a technique that extracts latent
+structures (clusters) underlying a set of predictor variables conditional on a
+response variable(s) (i.e. the outcome variables). How accurately the clusters
+within the TAGFA composition predict metabolic function is determined by using
+cross-validation on the PLS models.
 
@@
 Between the baseline visit and the 6-year visit in this sample, insulin
-sensitivity and beta-cell function measures had a significant median decline
+sensitivity and beta-cell function measures had a significant median decline of
 between `r outcome_chg$chg` (p`r outcome_chg$p` from GEE; n=`r outcome_chg$n`). There 
 were `r dys$pre` and `r dys$dm` incident cases of pre-diabetes (IFG and IGT) and
-diabetes, respectively, over the 6-years, of whom were excluded from later
+diabetes, respectively, over the 6-years; these observations were excluded from GEE and PLS
 analyses.
@@
 TAG concentration: `r tagfa$c[1]`; `r tagfa$c[2]`; and, `r tagfa$c[3]`.
-`r cite_f('heatmap')` shows a heatmap of the correlation of individual TAGFA with the
-outcome variables and several basic characteristics.
+`r cite_f('heatmap')` shows a heatmap of the correlation of individual TAGFA as
+concentrations with the outcome variables and several basic characteristics.
 As expected, nearly all TAGFA had very strong positive correlations (r=
@@
 (HOMA2-%S: r=`r fa_cor$rng['HOMA2-%S']`, ISI: r=`r fa_cor$rng$ISI`) than 
-for the beta-cell function measures (all r<0.30). A heatmap of the correlations
-of individual TAGFA using mol% with the basic characteristics is shown in 
-`r cite_f('heatmap_mol')` and the inter-correlation matrix of the TAGFA fraction 
-is shown in `r cite_f('heatmap_tagfa')`.
+for the beta-cell function measures (all r<0.30). For correlations
+of individual TAGFA using mol% with the basic participant characteristics, as shown in 
+`r cite_f('heatmap_mol')`, differences in correlations between fatty acids were
+most evident for 14:0, 14:1n-7, 16:0, and 16:1n-7 that had a moderate positive
+correlation with clinical TAG while all other fatty acids had a negative association.
+Based on heirarchical cluster analysis of the inter-correlation matrix of the
+TAGFA fraction, shown in `r cite_f('heatmap_tagfa')`, four fatty acids (14:0,
+16:0, 14:1n-7, and 16:1n-7) displayed a marked separation and cluster from the
+other TAGFA.
 
@@
 
-In analyses using TAGFA concentration values, nearly all TAGFA had a strong
+In analyses using concentration values, nearly all TAGFA had a strong
 negative association on HOMA2-%S and ISI (estimates ranging from
@@
 association with the insulin sensitivity measures, with a `r est1$b_ci$est[1]`
-to `r est1$b_ci$est[37]`% higher insulin sensitivity for every SD increase in 
-20:2n-6.
+to `r est1$b_ci$est[37]`% higher insulin sensitivity for every SD increase. 
+Both clinically-measured TAG and total TAGFA concentration had very strong
+negative associations with all outcome variables.
 
-Both clinically-measured TAG and total TAGFA concentration had very strong
-negative associations with all outcome variables. While there were a few
+While there were a few
 significant interactions by time in unadjusted models, after inclusion of
@@
 shown). There were no significant interactions by sex or ethnicity for any of
-the TAGFA (data not shown). We ran sensitivity analyses to identify which
-covariate attenuated the beta-cell function associations from the unadjusted
-model. We found that including waist circumference in the model lead to the
-attenuation of associations, shown in `r cite_sf("gee_adj_nowaist")`. A tabular
-presentation of the GEE results is shown in `r cite_st("gee_unadj")` for
-unadjusted models and `r cite_st("gee_adj")` for adjusted models.
+the TAGFA (data not shown). Results of the sensitivity analyses identifying
+waist circumference as the covariate that attenuated the beta-cell function
+associations from the unadjusted model are shown in `r cite_sf("gee_adj_nowaist")`. 
+A tabular presentation of the GEE results is shown in `r cite_st("gee_unadj")`
+for unadjusted models and `r cite_st("gee_adj")` for adjusted models.
 
@@
 The PLS analysis corroborated the findings from the GEE models. 
-The insulin sensitivity (`r cite_f('pls')`) PLS results showed a clustering 
-of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly
-and negatively on HOMA2-%S and ISI, suggesting this cluster of TAGFA tracks
+The PLS results conditioned on insulin sensitivity as the outcome showed a
+clustering of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7 
+(`r cite_f('pls')`). These TAGFA loaded strongly and negatively on HOMA2-%S and ISI,
+suggesting this cluster of TAGFA tracks
 together with lower insulin sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3,
@@
 ability, with a high correlation between the predicted outcome values against
-the actual values (HOMA2-%S: `r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).
+the observed values (HOMA2-%S: `r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).
 
-The beta-cell function PLS results showed clustering of similar fatty acids,
-however there there was a low correlation (though significant at 
-p`r cor_pls_igi$p`) between the predicted values and the actual values 
-(r=`r cor_pls_igi$r` to `r cor_pls_issi2$r`), suggesting poor predictability.
+The beta-cell function PLS results showed a similar clustering of fatty acids,
+however there was a lower correlation (though significant at 
+p`r cor_pls_igi$p`) between the predicted values and the observed values 
+(r=`r cor_pls_igi$r` to `r cor_pls_issi2$r`), suggesting that TAGFA composition
+poorly predicts beta-cell function.
 Given the low predictability, only the insulin sensitivity measures are

### 2017-02-16 R/calculate.R
Cleaned up left over functions

@@ -178,21 +178,6 @@ calc_n_at_each_visit <- function(data = project_data) {
         dplyr::summarise(n = n())
 }
 
-#' Calculate the number of participants who attended one, two, or three visits.
-#'
-#' @param data Project data.
-#'
-#' @export
-calc_n_for_visits <- function(data = project_data) {
-    data %>%
-        dplyr::select(SID, VN, f.VN) %>%
-        stats::na.omit() %>%
-        tidyr::spread(f.VN, VN) %>%
-        dplyr::mutate(Visits = paste(yr0, yr3, yr6, sep = '-')) %>%
-        .$Visits %>%
-        table()
-}
-
 #' Mean and SD followup time for those who attended the 6 year visit.
 #'
 #' @param data Project data

### 2017-02-16 doc/manuscript.Rmd
Made edits based on Tony's comments (discussion only)

@@
 
-We found that in an at-risk for diabetes Canadian population who were mainly female 
-and of European ancestry, the specific TAGFA composition was strongly associated 
+In the present study, we found that in a Canadian cohort at risk for diabetes,
+several specific TAGFA and groups of TAGFA were strongly associated
 with insulin sensitivity and moderately associated with beta-cell function. In 
-particular, TAGFA myristic acid (14:0), 7-tetradecenoic acid {{ confirm name }} 
+particular, the TAGFA myristic acid (14:0), 7-tetradecenoic acid {{ confirm name }} 
 (14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and
 negatively predicted lower insulin sensitivity. While most TAGFA were not
-associated with beta-cell function, two fatty acids, palmitic acid (16:0) and *cis*-vaccenic acid (18:1n-7), were
-associated negatively and positively, respectively, with both measures of 
-beta-cell function. We also found that four TAGFA (14:0, 14:1n-7, 16:0, 16:1n-7)
-clustered together, and this cluster also strongly predicted lower insulin sensitivity. 
-These four fatty acids are also involved in the *de novo* lipogenesis (DNL) of refined and simple 
-carbohydrates [@Hodson2008a; @Kawano2013a]. Our results, which are in agreement with the current literature,
-suggest that higher activity of DNL (potentially through higher intakes of
-simple carbohydrates) may increase the risk for diabetes, likely through
-worsening insulin sensitivity.
+associated with beta-cell function, two fatty acids, palmitic acid (16:0) and
+*cis*-vaccenic acid (18:1n-7), were associated negatively and positively,
+respectively, with measures of beta-cell function. Using PLS, we also found that four
+TAGFA (14:0, 14:1n-7, 16:0, 16:1n-7) clustered together, and that this cluster
+strongly predicted lower insulin sensitivity. These four fatty acids are
+involved in the *de novo* lipogenesis (DNL) of refined and simple carbohydrates
+[@Hodson2008a; @Kawano2013a]. Our results suggest that higher activity of DNL
+(potentially through higher intakes of simple carbohydrates) may increase the
+risk for diabetes, primarily through worsening insulin sensitivity.
 
@@
 
-To our knowledge, no longitudinal study to date has examined the role of the composition
-of the TAGFA fraction on detailed OGTT-derived metabolic measures using advanced
-{{another word? maybe 'appropriate'?}} statistical techniques designed to maximally
-utilize the data. Two large prospective studies have been published that had
-similar research objectives as the present analysis. One was a nested
-case-control analysis (n=189 cases and n=189 controls) within the Framingham offspring
-cohort [@Rhee2011a], which found that TAGFA composition characterized by a lower carbon chain
-and fewer double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
-12-years while higher carbon chain and more double bond TAGFA had a lower risk 
-for diabetes. This pattern of TAGFA with diabetes risk remained the same when 
-they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
-was a prospective cohort of males in Finland [@Lankinen2015a], for which 
-TAGFA data were available for 831 participants after 6-years of follow-up. This 
-cohort has similar outcomes as PROMISE, however, OGTT data was only available at
-the 6-year visit. As with our results, they found a larger effect size for
-insulin sensitivity compared to the beta-cell function. There were some
-similarities in results between the present analysis and the Finnish study,
-specifically for the saturated fatty acids, docosapentaenoic acid (22:5n-3),
+To our knowledge, no longitudinal study to date has examined the role of the
+composition of the TAGFA fraction on detailed OGTT-derived metabolic measures 
+Two large prospective studies have been published that similarly examined TAGFA 
+composition and diabetes outcomes. Rhee *et al* presented a nested case-control
+analysis (n=189 cases and n=189 controls) within the Framingham offspring cohort
+[@Rhee2011a], which found that subjects with a TAGFA composition characterized
+by a lower carbon chain and fewer double bonds (e.g. 14:0, 16:0) had a higher
+risk for diabetes after 12-years while those with a profile characterized by
+higher carbon chain and more double bond TAGFA had a lower risk for diabetes.
+A similar pattern of TAGFA was also associationed with HOMA-IR cross-sectionally
+at the baseline visit. In addition, Lankinen *et al* reported on
+a prospective cohort of males in Finland [@Lankinen2015a], for which 
+TAGFA data were available for 831 participants after 6-years of follow-up. 
+They found that most saturated fatty acids had negative , docosapentaenoic acid (22:5n-3),
 eicosapentaenoic acid (20:5n-3), and arachidonic acid (20:4n-6), though they did
 not find any associations with 7-tetradecenoic acid (14:1n-7) and palmitoleic
-acid (16:1n-7). Our study extends these findings by using multiple measurements
-of metabolic function and using statistical techniques that adequately make use
-of the longitudinal and high-dimensional dataset. In another study of a much
+acid (16:1n-7). {{ fix up }}
+There were some
+similarities in results between the present analysis and the Finnish study,
+Our study extends these findings by using multiple measurements
+of metabolic function and as well as multivariate statistical approaches that
+allowed us to identify a cluster of TAGFA.
+In another study of a much
 smaller (n=16) mostly female group [@Kotronen2009a], the authors report a
@@
 dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. Our study 
-extends these findings by showing that TAGFA with 14 to 16 carbons clustered together and all
-strongly predicted lower insulin sensitivity. While these fatty acids also had a
-significant association with beta-cell function, the results were not as strong
-as with the insulin sensitivity. However, while the DNL cluster had a strong
-harmful association with insulin sensitivity when we include it in the GEE
-models, the magnitude of the association was not as large as when using
-clinically measured TAG, reinforcing that the total TAG concentration still more
-strongly predicts metabolic dysfunction.
+extends these findings by showing that TAGFA with 14 to 16 carbons clustered
+together and this pattern strongly predicted lower insulin sensitivity. While
+these fatty acids also had a significant association with beta-cell function,
+the magnitude of associations were more modest compared to those for insulin
+sensitivity.
 
@@
 markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
-to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are major
+to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are 
 limitations to using these ratios as the fatty acids used in their calculation
 can also be obtained from the diet in addition to being created through DNL
-[@Hodson2008a]. A small, short feeding trial (n=24) was conducted to identify
-fatty acids that most accurately reflected DNL, as potential biomarkers
+[@Hodson2008a]. A feeding trial (n=24) was conducted to identify 
+the fatty acids that most accurately reflected DNL as potential biomarkers
 [@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
@@
 
-We found no interaction by time for any of the TAGFA on the metabolic outcomes.
-Analyzing these biological processes longitudinally is challenging due to the
-potential feedback mechanisms involved. For instance, while greater DNL may
-promote muscle insulin resistance, the reverse may also be true
-[@Flannery2012a]. Higher insulin resistance may encourage greater DNL to handle
-the higher blood glucose. To illustrate this, in a weight loss intervention trial
-(n=19 with TAGFA data), participants who lost weight over 33 weeks showed higher
-insulin sensitivity and lower TAGFA composition indicative of lower DNL (less
-14:0, 14:1, 16:0, 16:1, etc) [@Schwab2008a]. Given the complex biological
-mechanisms and feedback loops involved, disentangling whether insulin
-sensitivity influences TAGFA more strongly than TAGFA influencing insulin
-sensitivity will require more complicated research designs and analyses. Our own
-findings of no interaction by time with the TAGFA tentatively suggest that the
-influence of either TAGFA or insulin resistance within the feedback loop may be
-of relatively equal strength.
+{{fix this}}
+However, while this DNL cluster had a strong negative association
+with insulin sensitivity, the magnitude of the association was slightly weaker
+compared to clinically measured TAG.
 
-Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
-sources of fatty acids for TAG production during fasting are stored fatty acids
-within the liver, circulating NEFA, and through DNL [@Barrows2006a]. Therefore, the
-TAGFA composition reflects not just dietary intake but also stored fatty acids
-from the adipose tissue and the DNL capacity of the individual. During fasting, a
-large portion of fatty acids bound together as TAG are obtained from circulating
-NEFA. A study examining this found that NEFA contributed 60-80% of newly synthesized
-VLDL-TAG during fasting and ~40% during the fed state while dietary fatty acids
-contributed ~25% to newly synthesized VLDL-TAG [@Barrows2006a]. In an analysis of the
-serum NEFA fraction in the PROMISE cohort with the OGTT-derived outcomes
-(under review), we found that higher total NEFA but not the specific
-composition associated with lower beta-cell function and not with insulin 
-resistance. While NEFA may contribute substantially to the TAG fraction, there 
-may be something unique about the specific composition of the TAGFA fraction 
-that may allow it to enter insulin sensitive tissues more easily than NEFA. It 
-may be that VLDL receptors allow fatty acids to enter these tissue more easily
-than NEFA bound to albumin or it may be due to the other influences on TAGFA 
-composition such as dietary intake since circulating NEFA primarily come from
-adipose tissue [@Verges2015a].
+The direction of association between TAGFA and insulin sensitivity is unclear 
+from cross-sectional studies due to the physiological feedback mechanisms 
+involved. For example, while greater DNL may promote muscle insulin resistance, 
+the reverse may also be true [@Flannery2012a]. Higher insulin resistance may
+encourage greater DNL to handle the higher blood glucose. To illustrate this, in
+a weight loss intervention trial (n=19 with TAGFA data), participants who lost
+weight over 33 weeks showed higher insulin sensitivity and lower TAGFA
+composition indicative of lower DNL (less 14:0, 14:1, 16:0, 16:1, etc)
+[@Schwab2008a]. Given the complex biological mechanisms and feedback loops
+involved, disentangling whether insulin sensitivity influences TAGFA more
+strongly than TAGFA influencing insulin sensitivity will require more
+complicated research designs and analyses. While further research will need to
+confirm this, the lack of a time interaction we observed in addition to the
+timing of our measures likely suggests that it is the TAGFA predicting insulin
+resistance.
 
@@
 
-There are a few important limitations to our study. In addition to the
-observational nature of the cohort, the biology underlying the interaction
-between TAGFA, insulin secretion, and insulin sensitivity is complex and
-multifaceted and as such there may be some residual confounding we were
-unaware of or had not been able to adjust for. However, we have taken extensive,
-empirically based precautions in identifying potential confounders and mediators
-through the use of the DAG modeling, relying on previous literature, and through
-QIC model fit comparison methods.
+Our study has potential limitations that need to be considered when interpreting
+the results. Firstly, this is an observational cohort and as such there may be
+some residual confounding we were not able to control for or were unaware of. 
+However, we have taken extensive, empirically based precautions in identifying
+potential confounders and mediators through the use of the DAG modeling, relying
+on previous literature, and through QIC model fit comparison methods.
 
@@
 effect modifiers. Likewise, PLS is not able to handle longitudinal data so only 
-the baseline visit was used in the PLS analysis, though we analyze the extracted
-scores using the GEE modeling to overcome this limitation. In light of these
-limitations, we still observe concordant results between the PLS and GEE analyses.
+the baseline visit was used in the PLS analysis, although we analyzed the extracted
+scores using the GEE modeling to overcome this limitation and observed
+concordant results between the PLS and GEE analyses.
 
-Even with these limitations, our study has several notable strengths, including
-the longitudinal design and the rigorous statistical techniques and methods
-applied in analyzing the data. These statistical techniques take advantage of
-the longitudinal data to allow appropriate investigation of temporal
-relationships and are able to handle the multivariate nature of the data.
-Lastly, our cohort contains highly detailed and comprehensive variable
-measurements at each collection visit, as well as having both concentration and
-mol% data for the fatty acids.
+Our study has several notable strengths, including the longitudinal design and
+the use of advanced statistical techniques for data analysis. These statistical
+techniques take advantage of the longitudinal data to allow appropriate
+investigation of temporal relationships and are able to handle the
+multidimensional nature of the data. Lastly, our cohort contains highly detailed
+and comprehensive variable measurements for the fatty acids and outcomes,
+of which were collected at each visit.
 
@@
 (containing higher 14:0, 14:1n-7, 16:0, and 16:1n-7) associated strongly with lower
-insulin sensitivity and moderately with lower beta-cell function. The fatty
+insulin sensitivity and (more moderately) with lower beta-cell function. The fatty
 acids that clustered together represent fatty acids created from DNL, which is

### 2017-02-16 doc/manuscript.Rmd
Added models to caption for QIC table

@@
 model that had the fewest covariates and that had similar fit between the
-outcome measures. In particular for NEFA, it is used substantially as a source
-of fatty acids in hepatic TAG production [@Barrows2006a] and may be a strong
-confounder.
+outcome measures. Total NEFA was included as a confounder because it is used
+substantially as a source of fatty acids in hepatic TAG production
+[@Barrows2006a].
 
@@
 
+Given the number of possible combinations of outcome and predictor variables,
+only ISI and ISSI-2 with total triacylglycerol fatty acids (nmol/mL) were used to
+compare various GEE models and to select a final model.  Baseline age was used
+as including both the original age and the time variable would result in
+collinearity.  Column names: QIC is the quasi-likelihood information criteria
+(smaller values, eg. larger negative values, indicate a better fit compared to
+other models), Delta is the QIC minus the lowest QIC (models with delta <10 are 
+considered equivalent).  Models were:
+
+- M0: log(ISSI-2) or log(ISI) = total triacylglycerol fatty acids (nmol/mL) + years from baseline
+- M1: M0 + fatty acid by time interaction
+- M2: M0 + sex + ethnicity + baseline age
+- M3: M2 + waist
+- M4: M3 + ALT
+- M5: M4 + physical activity (MET)
+- M6: M5 + total NEFA
+- M7: M6 + alcohol intake
+- M8: M7 + family history of diabetes
+- M9: M8 + smoking status
+
 ```{r table_tagfa}

### 2017-02-16 R/qic.R

@@ -66,26 +66,20 @@ qic_is <- function(data = project_data) {
         corstr = 'ar1'
     )
 
-    Int <- stats::update(M0, . ~ . + TotalTG:YearsFromBaseline)
-    Full <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT + MET +
-                AlcoholPerWk + FamHistDiab + TobaccoUse + TotalNE)
-    M1 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex)
-    M2 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT)
-    M3 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT + MET)
-    M4 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT + MET + AlcoholPerWk)
-    M5 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT + MET + AlcoholPerWk +
-                          FamHistDiab)
-    M6 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT + FamHistDiab + TobaccoUse)
-    M7 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT + FamHistDiab)
-    M8 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT + MET + FamHistDiab)
-    M9 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT + MET + TotalNE)
-    M10 <- stats::update(M0, . ~ . + Waist + BaseAge + BiEthnicity + Sex + ALT + MET + TotalNE +
-                          FamHistDiab)
-    M11 <- stats::update(M0, . ~ . + BaseAge + BiEthnicity + Sex + ALT + MET + TotalNE)
-    M12 <- stats::update(M0, . ~ . + BaseAge + BiEthnicity + Sex + ALT + MET +
-                AlcoholPerWk + FamHistDiab + TobaccoUse + TotalNE)
+    M1 <- stats::update(M0, . ~ . + TotalTG:YearsFromBaseline)
+    M2 <- stats::update(M0, . ~ . + BaseAge + BiEthnicity + Sex)
[…]

### 2017-02-16 R/calculate.R
Brought old function (n visit calc) back to life

@@ -178,6 +178,21 @@ calc_n_at_each_visit <- function(data = project_data) {
         dplyr::summarise(n = n())
 }
 
+#' Calculate the number of participants who attended one, two, or three visits.
+#'
+#' @param data Project data.
+#'
+#' @export
+calc_n_for_visits <- function(data = project_data) {
+    data %>%
+        dplyr::select(SID, VN, f.VN) %>%
+        stats::na.omit() %>%
+        tidyr::spread(f.VN, VN) %>%
+        dplyr::mutate(Visits = paste(yr0, yr3, yr6, sep = '-')) %>%
+        .$Visits %>%
+        table()
+}
+
 #' Mean and SD followup time for those who attended the 6 year visit.
 #'
 #' @param data Project data

### 2017-02-28 doc/manuscript.Rmd
Minor edits to manuscript as per Tony's suggestions

@@
 
-Word count: ~4200
+Word count: ~4000
 
@@
 both the liver and muscle may result in greater production of TAG and secretion
-of lipoproteins that transport TAG [@Yu2012a]. Likewise, greater TAGFA may
+of lipoproteins that transport TAG [@Yu2012a]. Likewise, greater TAG may
 contribute to metabolic dysfunction and lipotoxicity in various tissues, such as
@@
 Sensitivity Index (ISI) [@Matsuda1999] and HOMA2-%S [@Levy1998a] using the HOMA2
-Calculator. HOMA largely reflects hepatic insulin resistance, while ISI
+Calculator. HOMA largely reflects hepatic insulin sensitivity, while ISI
 reflects whole-body insulin sensitivity [@AbdulGhani2007]. Beta-cell function
@@
 limited in that it cannot analyze the inherent multivariate nature of the 
-composition of the TAGFA fraction. Therfore, to confirm the GEE results in a multivariate 
+composition of the TAGFA fraction. Therefore, to confirm the GEE results in a multivariate 
 environment (i.e. all TAGFA analyzed collectively), partial least squares regression (PLS) was
@@
 age in years of `r aide::ave_sd(ds0yr$Age)` and a mean BMI
-of `r aide::ave_sd(ds0yr$BMI)` kg/m^2^. As expected from the study's eligibility criteria, the majority of
-participants (`r fhd$npct`) had a family history of diabetes.
-Between the baseline visit and the 6-year visit in this sample, insulin
-sensitivity and beta-cell function measures had a significant median decline of
+of `r aide::ave_sd(ds0yr$BMI)` kg/m^2^. As expected from the study's eligibility
+criteria, the majority of participants, n=`r fhd$npct`, had a family history of
+diabetes. Between the baseline visit and the 6-year visit in this sample,
+insulin sensitivity and beta-cell function measures had a significant median decline of
 between `r outcome_chg$chg` (p`r outcome_chg$p` from GEE; n=`r outcome_chg$n`). There 
@@
 
-Results from the unadjusted GEE model are shown in `r cite_f('gee_unadj')`) and 
+Results from the unadjusted GEE model are shown in `r cite_f('gee_unadj')` and 
 for the adjusted GEE model in `r cite_f('gee_adj')`. The majority of
@@
 associations with HOMA2-%S and ISI (between `r est1$outcome_b$rng[1]` and 
-`r est1$outcome_b$rng[5]`%, respectively, lower insulin sensitivity for every SD
+`r est1$outcome_b$rng[5]`, respectively, lower insulin sensitivity for every SD
 increase in the TAGFA), while several more TAGFA had positive associations with
@@
 higher carbon chain and more double bond TAGFA had a lower risk for diabetes.
-A similar pattern of TAGFA was also associationed with HOMA-IR cross-sectionally
+A similar pattern of TAGFA was also associated with HOMA-IR cross-sectionally
 at the baseline visit. In addition, Lankinen *et al* reported on
 a prospective cohort of males in Finland [@Lankinen2015a], for which 
-TAGFA data were available for 831 participants after 6-years of follow-up. 
-They found that most saturated fatty acids had negative , docosapentaenoic acid (22:5n-3),
-eicosapentaenoic acid (20:5n-3), and arachidonic acid (20:4n-6), though they did
-not find any associations with 7-tetradecenoic acid (14:1n-7) and palmitoleic
-acid (16:1n-7). {{ fix up }}
-There were some
-similarities in results between the present analysis and the Finnish study,
-Our study extends these findings by using multiple measurements
+TAGFA data were available for 831 participants after 6-years of follow-up. In
+their cohort, OGTT data was only available at the 6-year visit.
+They cross-sectionally found that most saturated fatty acids had negative 
+associations with insulin sensitivity and beta-cell function while
+linoleic acid (18:2n-6), docosapentaenoic acid (22:5n-3), eicosapentaenoic acid
+(20:5n-3), and arachidonic acid (20:4n-6) had positive associations with insulin
+sensitivity. The magnitude of the associations were larger in the insulin
+sensitivity results compared to the beta-cell function results, similar to what
+we observed. Our study extends these findings by using multiple measurements
 of metabolic function and as well as multivariate statistical approaches that
-allowed us to identify a cluster of TAGFA.
-In another study of a much
+allowed us to identify clusters of TAGFA. In another study of a much
 smaller (n=16) mostly female group [@Kotronen2009a], the authors report a
@@
 tissues [@Riserus2008a; @Iggman2010a].
-
-{{fix this}}
-However, while this DNL cluster had a strong negative association
-with insulin sensitivity, the magnitude of the association was slightly weaker
-compared to clinically measured TAG.
 

### 2017-02-28 doc/manuscript.Rmd
Edits as per Richard's suggestions

@@
 stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
-of {{confirm with Richard the correct name}} heptadecanoic acid (17:0) was added as an internal
+of triheptadecanoin (17:0; Nu-Chek Prep, Inc Elysian, MN, USA) was added as an internal
 standard prior to extracting total lipids according to the method of Folch 
 [@Folch1957a]. Each serum lipid fraction (non-esterified fatty acids (NEFA), 
-cholesteryl ester, phospholipid, and TAGFA) was isolated using thin layer 
+cholesteryl ester, phospholipid, and TAG) was isolated using thin layer 
 chromatography. Fatty acid methyl esters were separated and quantified using a 
@@
 with insulin sensitivity and moderately associated with beta-cell function. In 
-particular, the TAGFA myristic acid (14:0), 7-tetradecenoic acid {{ confirm name }} 
+particular, the TAGFA myristic acid (14:0), 7-tetradecenoic acid 
 (14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and

### 2017-03-01 doc/manuscript.Rmd
Created new function for corr of DNL TAGFA and added to results

@@
 As expected, nearly all TAGFA had very strong positive correlations (r=
-`r fa_cor$rng$TAG`) with clinically-measured TAG and moderate positive 
-correlations with WC (r=`r fa_cor$rng$Waist`). There was also moderate
-negative correlations with HDL (r=`r fa_cor$rng$HDL`). For the outcome variables,
+`r fa_cor$rng$TAG[2]`) with clinically-measured TAG and moderate positive 
+correlations with WC (r=`r fa_cor$rng$Waist[2]`). There was also moderate
+negative correlations with HDL (r=`r fa_cor$rng$HDL[1]`). For the outcome variables,
 the correlations for the insulin sensitivity measures were generally higher 
-(HOMA2-%S: r=`r fa_cor$rng['HOMA2-%S']`, ISI: r=`r fa_cor$rng$ISI`) than 
+(HOMA2-%S: r=`r fa_cor$rng[['HOMA2-%S']][1]`, ISI: r=`r fa_cor$rng$ISI[1]`) than 
 for the beta-cell function measures (all r<0.30). For correlations
@@
 most evident for 14:0, 14:1n-7, 16:0, and 16:1n-7 that had a moderate positive
-correlation with clinical TAG while all other fatty acids had a negative association.
+correlation with clinical TAG (r=` fa_cor_mol$rng$TAG[1]`) while all other fatty
+acids had a negative association (r=`fa_cor_mol$rng$TAG[2]`). In particular,
+those fatty acids with the negative associations with clinical TAG were all the
+very long chain polyunsaturated fatty acids (e.g. 20:4n-6, 20:5n-3).
 Based on heirarchical cluster analysis of the inter-correlation matrix of the

### 2017-03-01 R/calculate.R

@@ -144,21 +144,20 @@ calc_gee_estci <- function(results) {
 #' @export
 calc_cor <- function(data = cor_df) {
     data %>%
-        dplyr::group_by(Vars1) %>%
         {
-            high_cor <- dplyr::filter(., Correlations < -0.3 | Correlations > 0.3)
-            low_cor <- dplyr::filter(., Correlations > -0.3 & Correlations < 0.3)
+            high_cor <- dplyr::filter(., !dplyr::between(Correlations, -0.3, 0.3)) %>%
+                dplyr::mutate(Direction = dplyr::if_else(Correlations > 0.3, 'Positive', 'Negative'))
 
             list(
-                mean = dplyr::summarize(high_cor, mean = format_rounding(mean(Correlations))),
-                rng = dplyr::summarize(high_cor, rng = paste0(min(Correlations),
-                                                              ' to ', max(Correlations))) %>%
+                rng = high_cor %>%
+                    dplyr::group_by(Vars1, Direction) %>%
+                    dplyr::summarize(rng = paste0(min(Correlations), ' to ', max(Correlations))) %>%
+                    dplyr::ungroup() %>%
                     tidyr::spread(Vars1, rng),
-                fa = table(low_cor$Vars1, low_cor$Vars2) %>%
-                    tibble::as_data_frame() %>%
-                    dplyr::group_by(Var2) %>%
-                    dplyr::summarize(n = sum(n)) %>%
-                    dplyr::arrange(dplyr::desc(n))
+                pos_fa = dplyr::filter(high_cor, Direction == 'Positive') %>%
+                    dplyr::arrange(dplyr::desc(Correlations)),
+                neg_fa = dplyr::filter(high_cor, Direction == 'Negative') %>%
+                    dplyr::arrange(Correlations)
             )
         }
 }

### 2017-03-13 doc/manuscript.Rmd
Added a brief section in discussion on DNL and total TAG
Forgot to add r for inline chunk

@@
 most evident for 14:0, 14:1n-7, 16:0, and 16:1n-7 that had a moderate positive
-correlation with clinical TAG (r=` fa_cor_mol$rng$TAG[1]`) while all other fatty
-acids had a negative association (r=`fa_cor_mol$rng$TAG[2]`). In particular,
+correlation with clinical TAG (r=`r fa_cor_mol$rng$TAG[1]`) while all other fatty
+acids had a negative association (r=`r fa_cor_mol$rng$TAG[2]`). In particular,
 those fatty acids with the negative associations with clinical TAG were all the
@@
 fat stores and consequently increasing the amount of TAGFA in circulation
-[@Chehade2013a; @Verges2015a]. The higher concentration of circulating 14 and 16
-carbon fatty acids may then expose tissues to greater lipotoxicity, for instance
-from palmitic acid (16:0), which is well-known to have harmful effects on
-tissues [@Riserus2008a; @Iggman2010a].
+[@Chehade2013a; @Verges2015a]. In fact, we found in our study that
+higher proportions of these four DNL TAGFA also highly correlated with a higher
+concentration of clinical TAG, reinforcing this pathway between carbohydrate
+intake, DNL, and circulating TAG. The higher concentration of circulating 14 and
+16 carbon fatty acids may then expose tissues to greater lipotoxicity, for
+instance from palmitic acid (16:0), which is well-known to have harmful effects
+on tissues [@Riserus2008a; @Iggman2010a].
 

### 2017-03-21 doc/manuscript.Rmd
Added Tony's edits to manuscript

@@
 
-Title: Composition of fatty acids within the serum triacylglycerol fraction and 
-the association with the pathogenesis of diabetes
+Title: Specific clusters of fatty acids within the serum triacylglycerol fraction 
+associate with the pathogenesis of type 2 diabetes
 
@@
 (3,4); Bernard Zinman, MD (3,4); Adria Giacca, PhD (5); Stewart B. Harris, MD (2); 
-Richard P. Bazinet, PhD (1); and Anthony J. Hanley, PhD (1,6)
+Richard P. Bazinet, PhD (1); and Anthony J. Hanley, PhD (1,3,6)
 
@@
 
-Hypertriglyceridemia is an extensively studied and well known factor involved in
+Hypertriglyceridemia is an extensively studied and well described disorder in
 the dysregulation of metabolic function and subsequent negative health outcomes [@Chehade2013a].
-It is a component of the metabolic syndrome [@Alberti2009] and a strong risk factor for
-cardiovascular disease [@DAgostino2004a], in addition to it's role in metabolic
-dysfunction [@Verges2015a; @Kawano2013a]. The measurement of circulating 
-triacylglcerides (TAG) is commonly done during routine clinical visits as part
-of health assessment. However, clinically measured TAG is limited as it
-represents all the types of fatty acids within the TAG fraction as a single
-measure. Given the increasing appreciation for the role of the specific fatty
-acid composition on health outcomes, for instance with differences in
-associations of fatty acids within the PL lipid fraction [@Forouhi2014a; @Ma2015a] 
-including an analysis from the PROMISE cohort [@Johnston2016a], there
-are relatively few studies that have explored this area of research in the TAG
-fraction.
+It is a strong risk factor for cardiovascular disease [@DAgostino2004a], in addition to other metabolic disorders such as non-is involved in metabolic dysfunction [@Verges2015a; @Kawano2013a; @Alberti2009a].
+Circulating triacylglceride (TAG) concentration is commonly measured during
+routine clinical assessment using enzymatic methods.
+However, clinically measured TAG is limited as it
+represents the full fatty acid spectrum within the TAG fraction as a summary
+measure. There is increasing appreciation for the importance of specific fatty 
+acid composition profiles in different plasma fractions on various health
+outcomes [@Forouhi2014a; @Ma2015a; @Johnston2016a], however there are relatively
+few studies that have explored the impact of the fatty acid composition in the TAG fraction [@Rhee2011a; @Lankinen2015a].
 
-The interaction between TAG and insulin sensitivity is complex and has 
+The interaction between TAG and insulin sensitivity is complex and involves 
 components of a feedback system [@Verges2015a]. Greater resistance to insulin in
@@
 complexity and temporal nature of the relationship, long term studies with
-multiple data collections are paramount to better understanding the underlying
+multiple data collection time points are paramount to better understanding the underlying
 biology and subsequent risk.
 
-While several prospective studies have documented a prospective association of 
+While several studies have documented prospective associations of 
 hypertriglyceridemia with incident type 2 diabetes 
-[@Chien2008a; @DAgostino2004a; @Schulze2009a], there are limited longitudinal
-studies [@Rhee2011a; @Lankinen2015a] that have examined the relationship between
-TAG and it's composition on the pathophysiological factors underlying diabetes, 
-particularly in regard to beta-cell function. Our objective was to examine the 
+[@Chien2008a; @DAgostino2004a; @Schulze2009a], only a limited number of longitudinal
+studies [@Rhee2011a; @Lankinen2015a] have examined the relationship between
+TAG and its composition on the pathophysiological factors underlying diabetes, 
+particularly beta-cell function. Our objective was to examine the 
 longitudinal role of the specific composition of the serum TAG fraction on 
 OGTT-derived measures of insulin sensitivity and beta-cell function compared to 
-clinically measured TAG in an Canadian population at risk for diabetes.
+clinically measured TAG in a Canadian population at risk for diabetes.
 
@@
 attended the baseline visit. Subsequent examinations occurred every three years,
-with three examination visits completed to date (2004-2006, 2007-2009, and
-2010-2013). The current study
+with data from three examination visits available for the present analysis
+(2004-2006, 2007-2009, and 2010-2013). The current study
 used data on participants who did not have diabetes at baseline, who
@@
 using an enzymatic hexokinase (Roche Modular, Roche Diagnostics) with a 
-detection range of 0.11 to 41.6 mmol/L and an inter-assay %CV of <1.1% and an
-intra-assay %CV of < 1.9%.  All assays were performed at the Banting and Best
+detection range of 0.11 to 41.6 mmol/L and an inter-assay CV of <1.1% and an
+intra-assay CV of < 1.9%.  All assays were performed at the Banting and Best
 Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose (IFG),
@@
 In analyses using concentration values, nearly all TAGFA had a strong
-negative association on HOMA2-%S and ISI (estimates ranging from
+negative association on HOMA2-%S and ISI (estimates of percent different ranging from
 `r est1$outcome_b$rng[3]` and `r est1$outcome_b$rng[7]`, respectively), and a
@@
 The PLS results conditioned on insulin sensitivity as the outcome showed a
-clustering of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7 
+clustering of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7 as mol%
 (`r cite_f('pls')`). These TAGFA loaded strongly and negatively on HOMA2-%S and ISI,
@@
 positive correlation between total esterified (of which TAG make up the
-majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, similar as our results except
-for the 18:1n-9 finding.
+majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, findings which were largely 
+similar to the present analysis.
 
@@
 The direction of association between TAGFA and insulin sensitivity is unclear 
-from cross-sectional studies due to the physiological feedback mechanisms 
+from previous cross-sectional studies due to the physiological feedback mechanisms 
 involved. For example, while greater DNL may promote muscle insulin resistance, 

### 2017-03-21 doc/manuscript.Rmd
Fix pls figure dimensions

@@
 
-```{r fig_pls, fig.cap=fig_pls}
+```{r fig_pls, fig.cap=fig_pls, fig.height=8, fig.width=4}
 gridExtra::grid.arrange(
@@
     plot_pls(pls_isi) + ggplot2::ggtitle('B: ISI'), 
-    ncol = 2)
+    ncol = 1)
 ```
+
+# References
 

### 2017-03-21 doc/manuscript.Rmd
Edits as per Richards comments

@@
 clustering of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7 as mol%
-(`r cite_f('pls')`). These TAGFA loaded strongly and negatively on HOMA2-%S and ISI,
-suggesting this cluster of TAGFA tracks
+(`r cite_f('pls')`). These TAGFA loaded strongly and negatively on HOMA2-%S and ISI
+in the first component, suggesting this cluster of TAGFA tracks
 together with lower insulin sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3,
 and 22:6n-3 loaded positively on both insulin sensitivity measures. No other
-TAGFA loaded strongly. The PLS model for insulin sensitivity had good predictive
+TAGFA loaded strongly. In the second component, 18:1n-9 and 18:1n-7 loaded postively 
+but not strongly while 20:5n-3 and 22:6n-3 loaded strongly and negatively with both
+HOMA2-%S and ISI; however, this component only explained <10% of the variance. 
+The PLS model for insulin sensitivity had good predictive
 ability, with a high correlation between the predicted outcome values against
@@
 
-## Overall conclusion
-
 In the present study, we found that in a Canadian cohort at risk for diabetes,
@@
 risk for diabetes, primarily through worsening insulin sensitivity.
-
-## Interpretation and previous literature
 
@@
 
-## Limitations
+## Limitations and strengths
 
@@
 ```
-
-# References

### 2017-03-22 doc/manuscript.Rmd
Added abstract

@@
 csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
-output: fost::diabetologia_manuscript
+output: fost::diabetes_care_manuscript
 ---
@@
 # Abstract
+
+Word count: ~281 / 250
+
+**Objective**: 
+There is limited data on the role of the specific triacylglyceride fatty acid
+(TAGFA) composition on diabetes pathogenesis. Our aim was to examine the 
+longitudinal association of TAGFA on insulin sensitivity (IS) and beta-cell
+function.
+
+**Research Design and Methods**:
+We used data from the Prospective Metabolism and Islet Cell Evaluation (PROMISE)
+cohort of adults (n=477) at-risk for diabetes with over 6-years of followup (3
+timepoints). Glucose and insulin from an OGTT were used to calculate the Matsuda
+index (ISI), HOMA2-%S, the Insulinogenic Index over HOMA-IR (IGI/IR), and the
+Insulin Secretion-Sensitivity Index-2 (ISSI-2). Gas chromatography quantified
+TAGFA. Generalized estimating equations (GEE) adjusted for confounders and
+partial least squares (PLS) were used for the analysis.
+
+**Results**: 
+The outcome variables declined by 14% to 27% over the 6-years. Most TAGFA
+modeled as a concentration had strong negative associations with IS in adjusted
+GEE models; when as a mol%, four TAGFA (14:0, 16:0, 14:1n-7, 16:1n-7) had strong
+negative associations while most others (e.g. 18:1n-7, 18:1n-9, 20:2n-6,
+20:5n-3) had strong positive associations. Few associations were seen for
+beta-cell function, except for 16:0 (negative) and 18:1n-7 (positive). The PLS
+analysis identified that the four 'harmful' TAGFA (14 and 16 carbon TAGFA)
+clustered together and strongly predicted lower IS.
+
+**Conclusions**:
+We found that a higher proportions of a cluster of four TAGFA (14:0, 14:1-7,
+16:0, 16:1n-7) strongly predicted lower IS. Previous research suggests these
+four FA are heavily derived from the de novo lipogenesis (DNL) of higher intakes
+of simple or refined carbohydrates, especially 16:1n-7. Our results suggest that
+greater simple or refined carbohydrate intake may impact risk for diabetes
+through greater DNL and higher FA associated with lipotoxicity (e.g. 16:0),
+leading to hypertriglyceridemia and lower IS.
 
@@
 
-# References
-
 # Supplemental Material
@@
 ```
+
+# References

### 2017-04-05 doc/manuscript.Rmd
Edits to manuscript abstract as per Tony's edits

--- 2017-03-21 doc/manuscript.Rmd
@@
 
+Word count: 234 / 250
+
+**Objective**: 
+Although elevated serum triglyceride (TAG) is a well-described risk factor for
+type 2 diabetes, few data are available regarding the role of the specific fatty
+acid composition within serum TAG (TAGFA). In this context, our aim was to
+examine longitudinal associations of TAGFA with insulin sensitivity (IS) and
+beta-cell function.
+
+**Research Design and Methods**:
+We used longitudinal data (3 visits over 6 years) from the Prospective
+Metabolism and Islet Cell Evaluation (PROMISE) cohort of adults (n=477) at-risk
+for diabetes. Glucose and insulin from an OGTT were used to calculate the
+Matsuda index (ISI), HOMA2-%S, the Insulinogenic Index over HOMA-IR (IGI/IR),
+and the Insulin Secretion-Sensitivity Index-2 (ISSI-2). Gas chromatography
+quantified TAGFA composition. Generalized estimating equations (GEE) adjusted for
+confounders and partial least squares (PLS) were used for the analysis.
+
+**Results**: 
+The outcome variables declined by 14% to 27% over the 6-years. In the adjusted
+GEE models with TAGFA as mol%, four TAGFA (14:0, 16:0, 14:1n-7, 16:1n-7) had
+strong negative associations while most others (e.g. 18:1n-7, 18:1n-9, 20:2n-6, 
+20:5n-3) had strong positive associations. Few associations were seen for 
+beta-cell function, except for 16:0 (negative) and 18:1n-7 (positive). PLS
+analysis indicated that four TGFA clustered together and strongly predicted
+lower IS. These four TAGFA also correlated highly (r>0.4) with clinically
+measured TAG.
+
+**Conclusions**:
+We found that higher proportions of a cluster of four TAGFA (14:0, 14:1-7,
+16:0, 16:1n-7) strongly predicted lower IS as well as hypertriglyceridemia.
+
 # Background
@@
 the dysregulation of metabolic function and subsequent negative health outcomes [@Chehade2013a].
-It is a strong risk factor for cardiovascular disease [@DAgostino2004a], in addition to other metabolic disorders such as non-is involved in metabolic dysfunction [@Verges2015a; @Kawano2013a; @Alberti2009a].
+It is a strong risk factor for cardiovascular disease [@DAgostino2004a], in addition to other metabolic disorders such as non-is involved in metabolic dysfunction [@Verges2015a; @Kawano2013a; @Alberti2009].
 Circulating triacylglceride (TAG) concentration is commonly measured during

### 2017-04-05 doc/manuscript.Rmd
Added Tony's edits to manuscript

@@
 ---
-
-<!--
-
-Outline:
-
-- Primary analysis:
-        - Sensitivity analysis with Carb intake?
-
--->
 
@@
 
-Word count: 234 / 250
+Word count: 249 / 250
 
@@
 Although elevated serum triglyceride (TAG) is a well-described risk factor for
-type 2 diabetes, few data are available regarding the role of the specific fatty
-acid composition within serum TAG (TAGFA). In this context, our aim was to
+type 2 diabetes (T2DM), few data are available regarding the role of the specific fatty
+acid composition within serum TAG (TAGFA) in the pathogenesis of T2DM. Our aim, therefore, was to
 examine longitudinal associations of TAGFA with insulin sensitivity (IS) and
@@
 We used longitudinal data (3 visits over 6 years) from the Prospective
-Metabolism and Islet Cell Evaluation (PROMISE) cohort of adults (n=477) at-risk
-for diabetes. Glucose and insulin from an OGTT were used to calculate the
+Metabolism and Islet Cell Evaluation (PROMISE) cohort of adults (n=477) who were at-risk
+for diabetes at baseline. Glucose and insulin from an OGTT were used to calculate the
 Matsuda index (ISI), HOMA2-%S, the Insulinogenic Index over HOMA-IR (IGI/IR),
@@
 The outcome variables declined by 14% to 27% over the 6-years. In the adjusted
-GEE models with TAGFA as mol%, four TAGFA (14:0, 16:0, 14:1n-7, 16:1n-7) had
-strong negative associations while most others (e.g. 18:1n-7, 18:1n-9, 20:2n-6, 
+GEE models, four TAGFA (14:0, 16:0, 14:1n-7, 16:1n-7 as mol%) had
+strong negative associations with IS while others (e.g. 18:1n-7, 18:1n-9, 20:2n-6, 
 20:5n-3) had strong positive associations. Few associations were seen for 
 beta-cell function, except for 16:0 (negative) and 18:1n-7 (positive). PLS
-analysis indicated that four TGFA clustered together and strongly predicted
+analysis indicated four TAGFA (14:0, 16:0, 14:1n-7, 16:1n-7) that are markers of de novo lipogenesis (DNL)
+clustered together and strongly predicted
 lower IS. These four TAGFA also correlated highly (r>0.4) with clinically
@@
 **Conclusions**:
-We found that higher proportions of a cluster of four TAGFA (14:0, 14:1-7,
-16:0, 16:1n-7) strongly predicted lower IS as well as hypertriglyceridemia.
+We found that higher proportions of a cluster of four DNL TAGFA 
+strongly predicted lower IS as well as hypertriglyceridemia.
 

### 2017-05-24 doc/manuscript.Rmd
Updated md and sent to Tony + richard (old commit)

@@
 the dysregulation of metabolic function and subsequent negative health outcomes [@Chehade2013a].
-It is a strong risk factor for cardiovascular disease [@DAgostino2004a], in addition to other metabolic disorders such as non-is involved in metabolic dysfunction [@Verges2015a; @Kawano2013a; @Alberti2009].
+It is a strong risk factor for cardiovascular disease [@DAgostino2004a; @Verges2015a]
+and is involved in other metabolic disorders and phenomenon such as non-alcoholic
+fatty liver disease [@Kawano2013a], the metabolic syndrome [@Alberti2009], and
+the hypertriglyceridemic waist [@Lemieux2000a].
 Circulating triacylglceride (TAG) concentration is commonly measured during
@@
 contribute to metabolic dysfunction and lipotoxicity in various tissues, such as
-the beta-cells, and thus continuing the cycle [@Verges2015a]. Given the
+the beta-cells, and thus continue the cycle [@Verges2015a]. Given the
 complexity and temporal nature of the relationship, long term studies with
@@
 TAG and its composition on the pathophysiological factors underlying diabetes, 
-particularly beta-cell function. Our objective was to examine the 
+particularly on beta-cell function. Our objective was to examine the 
 longitudinal role of the specific composition of the serum TAG fraction on 

### 2017-05-24 doc/manuscript.Rmd
Added Bernie's comments

@@
 
-Title: Specific clusters of fatty acids within the serum triacylglycerol fraction 
+Title: Specific clusters of fatty acids within the serum triacylglyceride fraction 
 associate with the pathogenesis of type 2 diabetes
@@
 **Objective**: 
-Although elevated serum triglyceride (TAG) is a well-described risk factor for
+Although elevated serum triacylglyceride (TAG) is a well-described risk factor for
 type 2 diabetes (T2DM), few data are available regarding the role of the specific fatty
-acid composition within serum TAG (TAGFA) in the pathogenesis of T2DM. Our aim, therefore, was to
+acid (FA) composition within serum TAG (TAGFA) in the pathogenesis of T2DM. Our aim, therefore, was to
 examine longitudinal associations of TAGFA with insulin sensitivity (IS) and
@@
 **Results**: 
-The outcome variables declined by 14% to 27% over the 6-years. In the adjusted
+The metabolic outcome variables declined by 14% to 27% over the 6-years. In the adjusted
 GEE models, four TAGFA (14:0, 16:0, 14:1n-7, 16:1n-7 as mol%) had

### 2017-06-09 doc/manuscript.Rmd
Incomplete addition of co-author edits

@@
 Metabolism and Islet Cell Evaluation (PROMISE) cohort of adults (n=477) who were at-risk
-for diabetes at baseline. Glucose and insulin from an OGTT were used to calculate the
-Matsuda index (ISI), HOMA2-%S, the Insulinogenic Index over HOMA-IR (IGI/IR),
-and the Insulin Secretion-Sensitivity Index-2 (ISSI-2). Gas chromatography
+for diabetes at baseline. Glucose and insulin from an OGTT were used to calculate the outcome variables
+Matsuda index (ISI), HOMA2-%S, Insulinogenic Index over HOMA-IR (IGI/IR),
+and Insulin Secretion-Sensitivity Index-2 (ISSI-2). Gas chromatography
 quantified TAGFA composition. Generalized estimating equations (GEE) adjusted for
@@
 
-Hypertriglyceridemia is an extensively studied and well described disorder in
-the dysregulation of metabolic function and subsequent negative health outcomes [@Chehade2013a].
-It is a strong risk factor for cardiovascular disease [@DAgostino2004a; @Verges2015a]
-and is involved in other metabolic disorders and phenomenon such as non-alcoholic
+Hypertriglyceridemia is a well described metabolic disorder resulting in negative health outcomes [@Chehade2013a].
+It is a risk factor for cardiovascular disease [@DAgostino2004a; @Verges2015a]
+and is associated with other metabolic disorders such as non-alcoholic
 fatty liver disease [@Kawano2013a], the metabolic syndrome [@Alberti2009], and
-the hypertriglyceridemic waist [@Lemieux2000a].
+abdominal obesity [@Lemieux2000a].
 Circulating triacylglceride (TAG) concentration is commonly measured during
@@
 of lipoproteins that transport TAG [@Yu2012a]. Likewise, greater TAG may
-contribute to metabolic dysfunction and lipotoxicity in various tissues, such as
-the beta-cells, and thus continue the cycle [@Verges2015a]. Given the
+contribute to metabolic dysfunction and lipotoxicity in various tissues,
+affecting insulin sensitivity, and thus continue the cycle [@Verges2015a]. Given the
 complexity and temporal nature of the relationship, long term studies with
@@
 studies [@Rhee2011a; @Lankinen2015a] have examined the relationship between
-TAG and its composition on the pathophysiological factors underlying diabetes, 
-particularly on beta-cell function. Our objective was to examine the 
+TAG and its composition with the pathophysiological factors underlying diabetes, 
+particularly beta-cell function. Our objective was to examine the 
 longitudinal role of the specific composition of the serum TAG fraction on 
@@
 cohort took place between 2004-2006 in London and Toronto, 
-Canada. Individuals were selected to participant if they met the eligibility 
+Canada. Individuals were selected to participate if they met the eligibility 
 criteria of having one or more risk factors for type 2 diabetes mellitus, 
@@
 Inc) and has a coefficient of variation (CV) of 9.3%. Glucose was determined 
-using an enzymatic hexokinase (Roche Modular, Roche Diagnostics) with a 
+using an enzymatic hexokinase method (Roche Modular, Roche Diagnostics) with a 
 detection range of 0.11 to 41.6 mmol/L and an inter-assay CV of <1.1% and an
@@
 @Johnston2016a for the analysis of the phospholipid and cholesteryl ester
-fractions). 
+fractions). {{put these analyses in discussion?}}
 <!--and {{next citation}} for the analysis of the NEFA fraction).-->
@@
 and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
-is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
+is a measure of the early phase of insulin secretion while ISSI-2 is analogous to
 the disposition index (but is calculated using OGTT values). Each index has been validated
@@
 Specific formulas for each OGTT-derived measure can be found in the 
-Supplementary Material.
+Supplemental Material {{insert formula}}.
 
@@
 computed for TAGFA against each other; hierarchical clustering analysis was
-applied to the correlation matrix to identify potential clusters within the TAGFA 
-composition.
+applied to the correlation matrix to identify clusters within the TAGFA 
+composition. {{expand/clarify, ref suplemental material}}
 
@@
 recommendations, and from quasi-likelihood information criteria (QIC). The final
-GEE model was adjusted for time, waist circumference, baseline age, ethnicity,
+GEE model ({{M#?}}) was adjusted for time, waist circumference, baseline age, ethnicity,
 sex, ALT, MET, and total NEFA. The TAGFA, total NEFA, sex, ethnicity, and
@@
 this manuscript is available at {{code doi}}. Results were considered
-statistically significance at p<0.05, after adjusting for multiple testing using
+statistically significant at p<0.05, after adjusting for multiple testing using
 the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a]. STROBE was used
@@
 diabetes, respectively, over the 6-years; these observations were excluded from GEE and PLS
-analyses.
+analyses due to how the outcomes were measured (OGTT was not done on those with diabetes).
 
@@
 `r fa_cor$rng$TAG[2]`) with clinically-measured TAG and moderate positive 
-correlations with WC (r=`r fa_cor$rng$Waist[2]`). There was also moderate
+correlations with WC (r=`r fa_cor$rng$Waist[2]`). There were also moderate
 negative correlations with HDL (r=`r fa_cor$rng$HDL[1]`). For the outcome variables,
@@
 16:0, 14:1n-7, and 16:1n-7) displayed a marked separation and cluster from the
-other TAGFA.
+other TAGFA. {{clarify, as well as the legend}}
 

### 2017-06-25 doc/manuscript.Rmd
Added more edits from Ravi and Adria (incomplete)

@@
 @Johnston2016a for the analysis of the phospholipid and cholesteryl ester
-fractions). {{put these analyses in discussion?}}
+fractions). <!--{{put these analyses in discussion?}} -->
 <!--and {{next citation}} for the analysis of the NEFA fraction).-->
@@
 individual TAGFA with other continuous variables. Correlations were also
-computed for TAGFA against each other; hierarchical clustering analysis was
-applied to the correlation matrix to identify clusters within the TAGFA 
-composition. {{expand/clarify, ref suplemental material}}
+computed and hierarchical clustering analysis was conducted for TAGFA against each other.
 
@@
 recommendations, and from quasi-likelihood information criteria (QIC). The final
-GEE model ({{M#?}}) was adjusted for time, waist circumference, baseline age, ethnicity,
+GEE model (M6) was adjusted for time, waist circumference, baseline age, ethnicity,
 sex, ALT, MET, and total NEFA. The TAGFA, total NEFA, sex, ethnicity, and
@@
 very long chain polyunsaturated fatty acids (e.g. 20:4n-6, 20:5n-3).
-Based on heirarchical cluster analysis of the inter-correlation matrix of the
-TAGFA fraction, shown in `r cite_f('heatmap_tagfa')`, four fatty acids (14:0,
-16:0, 14:1n-7, and 16:1n-7) displayed a marked separation and cluster from the
-other TAGFA. {{clarify, as well as the legend}}
+As seen in `r cite_f('heatmap_tagfa')`, four fatty acids (14:0,
+16:0, 14:1n-7, and 16:1n-7) clustered together, each highly positively correlated
+with each other and negatively correlated with all other fatty acids.
 

### 2017-06-25 R/gee.R
convert to data.frame for pander

@@ -224,6 +224,7 @@ table_gee_main <- function(results, caption = NULL, digits = 1) {
         ) %>%
         dplyr::select(-unit) %>%
         dplyr::rename('Fatty acid' = Xterms) %>%
+        as.data.frame() %>%
         pander::pander(missing = '', caption = caption)
 }
 

### 2017-06-25 R/plot.R
labs to ggplot2 vs seer

@@ -23,8 +23,9 @@ plot_tagfa <- function(data = project_data) {
         dplyr::mutate(Measure = Measure %>%
                    factor(., levels = unique(.))) %>%
         seer::view_boxplots(
-            'Measure', 'Value',
-            xlab = 'Concentration (nmol/mL)',
-            ylab = 'Triacylglyceride fatty acids') +
+            'Measure', 'Value') +
+        ggplot2::labs(
+            x = 'Concentration (nmol/mL)',
+            y = 'Triacylglyceride fatty acids') +
         graph_theme(ticks = FALSE)
 }

### 2017-06-26 R/calculate.R
First commit of code used + suppl files

@@ -1,0 +1,215 @@
+# Calculate or extract for inline -----------------------------------------
+
+#' Calculate the number of individuals who converted to pre-DM or DM.
+#'
+#' @param data Data with conversion numbers.
+#'
+#' @export
+calc_convert_dys <- function(data) {
+    conversion_dm <-
+        table(data['ConvertDM'])[2] %>%
+        {
+            paste0(., ' (', round((. / 477) * 100, 0), '%)')
+        }
+
+    conversion_pre <-
+        table(data['ConvertPreDM'])[2] %>%
+        {
+            paste0(., ' (', round((. / 477) * 100, 0), '%)')
+        }
+
+    list(pre = conversion_pre, dm = conversion_dm)
+}
+
+#' Calculate the percent contribution of each TAGFA of the whole fraction.
+#'
+#' @param data Project data.
+#'
+#' @export
+calc_tagfa_percent <- function(data = project_data) {
+    over_10pct_tagfa <- data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select(dplyr::matches('pct_tg')) %>%
+        tidyr::gather(fat, value) %>%
+        dplyr::group_by(fat) %>%
+        dplyr::summarise(pct = mean(value, na.rm = TRUE)) %>%
+        dplyr::mutate(fat = renaming_fats(fat),
+               c = paste0(fat, ' (', format_rounding(pct, 1), '%)')) %>%
+        dplyr::arrange(dplyr::desc(pct)) %>%
+        dplyr::filter(pct >= 10)
+
+    return(over_10pct_tagfa)
+}
+
+#' Calculate the percent change over time for the outcome variables.
+#'
+#' @param data Project data.
+#'
+#' @export
+calc_outcome_changes <- function(data = project_data) {
+    prep.data <- data %>%
+        dplyr::select(f.VN, HOMA, ISI, IGIIR, ISSI2) %>%
+        tidyr::gather(Measure, Value,-f.VN) %>%
+        stats::na.omit() %>%
+        dplyr::group_by(Measure, f.VN) %>%
+        dplyr::summarise(med = median(Value),
+                         n = n()) %>%
+        dplyr::ungroup()
+
+    sample_size <- prep.data$n %>%
+        {paste0(min(.), '-', max(.))}
+
+    change_over_time <- prep.data %>%
+        dplyr::select(-n) %>%
+        tidyr::spread(f.VN, med) %>%
[…]

### 2017-06-26 R/corr.R

@@ -1,0 +1,83 @@
+# Functions to run the correlation analysis
+#
+# Analyze -----------------------------------------------------------------
+
+#' Correlation analysis.
+#'
+#' @param data Project data
+#' @param x The covariates and outcomes
+#' @param y The fatty acids.
+#'
+#' @export
+analyze_corr <-
+    function(data = project_data,
+             x = c(outcomes, 'BMI', 'Waist', 'Age', 'lALT', 'lTAG', 'Chol', 'HDL', 'LDL'),
+             y = tg_conc) {
+
+    data %>%
+        dplyr::filter(VN == 0) %>%
+        mason::design('cor') %>%
+        mason::add_settings(method = 'pearson', use = 'complete.obs') %>%
+        mason::add_variables('yvars', y) %>%
+        mason::add_variables('xvars', x) %>%
+        mason::construct() %>%
+        mason::scrub() %>%
+        dplyr::mutate_at('Vars2', dplyr::funs(renaming_fats)) %>%
+        dplyr::mutate_at("Vars1", function(x)
+            gsub('l(ALT|TAG|IGIIR|HOMA2_S|ISI|ISSI2)', '\\1', x) %>%
+                renaming_outcomes()) %>%
+        dplyr::mutate(
+            order1 = substr(Vars2, nchar(Vars2), nchar(Vars2)),
+            order1 = ifelse(order1 == 0, 10, order1),
+            order1 = ifelse(order1 == 'l', 20, order1),
+            order1 = as.integer(order1)
+        ) %>%
+        dplyr::arrange(dplyr::desc(order1)) %>%
+        dplyr::select(-order1) %>%
+        dplyr::mutate(Vars2 = factor(Vars2, unique(Vars2)),
+                      Vars1 = factor(Vars1, unique(Vars1)),
+                      Correlations = round(Correlations, 2))
+    }
+
+#' Correlation analysis of just TAGFA.
+#'
+#' @param data Project data
+#' @param x The TAGFA
+#'
+#' @export
+analyze_corr_tagfa <-
+    function(data = project_data,
+             x = tg_pct) {
+
+    data %>%
+        dplyr::filter(VN == 0) %>%
+        mason::design('cor') %>%
+        mason::add_settings(method = 'pearson', use = 'complete.obs', hclust.order = TRUE) %>%
+        mason::add_variables('xvars', x) %>%
+        mason::construct() %>%
+        mason::scrub() %>%
[…]

### 2017-06-26 R/fetch_data.R

@@ -1,0 +1,113 @@
+#' Fetch data from the original source
+#'
+#' This function fetchs the main dataset, keeps variables relevant to
+#' the analysis, restrict the sample size as needed, and lastly save
+#' the new dataset as an `.RData` file.
+#'
+#' @return Saves the wrangled data into the data/ folder.
+#' @export
+#'
+#' @examples
+#' fetch_data()
+#'
+fetch_data <- function() {
+    # Load the master dataset,
+    ds.prep <- PROMISE::PROMISE %>%
+        dplyr::filter(VN %in% c(1, 3, 6)) %>%
+        ## Kick out Canoers
+        dplyr::filter(is.na(Canoe)) %>%
+        dplyr::tbl_df()
+
+    print(paste0('Original dataset rows are ', dim(ds.prep)[1], ' and columns are ', dim(ds.prep)[2]))
+
+    ##' Munge and wrangle the data into the final version.
+    ds <- ds.prep %>%
+        dplyr::select(
+            SID, VN, BMI, Waist, HOMA, HOMA2IR, HOMA2_S, ISI, IGIIR, ISSI2,
+            TAG, LDL, HDL, Chol, MonthsFromBaseline,
+            ALT, CRP, FamHistDiab, dplyr::matches('meds'), Age, Sex, Ethnicity,
+            IFG, IGT, DM, MET, AlcoholPerWk, TobaccoUse, SelfEdu, Occupation,
+            TotalTG, dplyr::matches('^tg\\d+'), Glucose0, Glucose120, TotalNE
+        ) %>%
+        dplyr::mutate(
+            YearsFromBaseline = MonthsFromBaseline / 12,
+            BaseTotalNE = TotalNE,
+            BaseTotalTG = TotalTG,
+            BaseTAG = ifelse(VN == 1, TAG, NA),
+            lBaseTAG = log(BaseTAG),
+            BaseAge = ifelse(VN == 1, Age, NA),
+            FamHistDiab =
+                plyr::mapvalues(FamHistDiab, c(0, 1:12),
+                                c('No', rep('Yes', 12))) %>%
+                as.factor(),
+            AlcoholPerWk = plyr::mapvalues(
+                as.factor(AlcoholPerWk),
+                c('1', '2', '3', '4', '5', '6', '7'),
+                c('1', '2', '3', '3', '3', '3', '3')
+            ) %>%
+                as.factor(),
+            invHOMA = (1 / HOMA),
[…]

### 2017-06-26 R/gee.R

@@ -1,0 +1,254 @@
+# Functions for the GEE analysis
+#
+# Grab or combine data ----------------------------------------------------
+
+#' Prepare the project data for analysis through GEE.
+#'
+#' @param data project data
+#' @export
+prep_gee_data <- function(data) {
+    data <- data %>%
+        dplyr::filter(DM != 1)
+
+    no_fattyacids <- data %>%
+        dplyr::select(-dplyr::matches('pct_tg\\d+|^tg\\d+'),
+                      -TotalNE,-TotalTG, -BaseTAG, -lBaseTAG)
+
+    scaled_variables <- data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select(SID, TotalNE, TotalTG, BaseTAG, lBaseTAG,
+                      dplyr::matches('pct_tg\\d+|^tg\\d+')) %>%
+        dplyr::mutate_at(dplyr::vars(-SID), dplyr::funs(as.numeric(scale(.))))
+
+    dplyr::full_join(
+            no_fattyacids,
+            scaled_variables,
+            by = 'SID'
+        ) %>%
+        dplyr::group_by(VN) %>%
+        dplyr::mutate(
+            Waist = as.numeric(scale(Waist)),
+            ALT = as.numeric(scale(ALT)),
+            BaseAge = as.numeric(scale(BaseAge)),
+            MET = as.numeric(scale(MET))
+        ) %>%
+        dplyr::ungroup() %>%
+        dplyr::arrange(SID, VN)
+}
+
+#' Add the extracted PLS scores to the data for use in GEE.
+#'
+#' @param data The project data.
+#' @param pls_results Results dataset from PLS.
+#' @export
+prep_gee_pls_data <- function(data = project_data, pls_results) {
+    sid_vn <- data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select_(.dots = c(outcomes, tg_pct, 'SID', covariates)) %>%
+        stats::na.omit() %>%
+        dplyr::full_join(dplyr::select(project_data, SID, VN)) %>%
+        dplyr::select_(.dots = c(outcomes, 'SID', 'VN'))
+
+    pls_results %>%
+    {
+        dplyr::bind_cols(tibble::as_data_frame(.$model$Y),
+                         tibble::as_data_frame(unclass(.$scores))[1:2])
+    } %>%
[…]

### 2017-06-26 R/load_data.R

@@ -1,0 +1,23 @@
+#' Load the data found in data/.
+#'
+#' @param update Whether you want to update the dataset using the fetch_data
+#'   function.
+#' @param project.name Name of your project. Set the default to keep it always
+#'   working.
+#'
+#' @return Outputs the datasets in the data/ folder.
+#' @export
+#'
+#' @examples
+#' load_data()
+#' load_data(update = TRUE)
+#'
+load_data <- function(update = FALSE, project.name = 'tagDiabetes') {
+    if (is.null(project.name))
+        stop('Please change the project.name default to the name of the project.')
+    if (!dir.exists(system.file('data', package = project.name)))
+        fetch_data()
+    if (update)
+        fetch_data()
+    devtools::load_data()
+}

### 2017-06-26 R/plot.R

@@ -1,0 +1,30 @@
+# Plotting ----------------------------------------------------------------
+
+#' Plots the distribution of the TAGFA composition.
+#'
+#' @param data Project data.
+#'
+#' @export
+plot_tagfa <- function(data = project_data) {
+    data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select(SID, dplyr::matches('^tg\\d+')) %>%
+        tidyr::gather(Measure, Value,-SID) %>%
+        stats::na.omit() %>%
+        dplyr::mutate(Measure = renaming_fats(Measure)) %>%
+        dplyr::mutate(
+            order1 = substr(Measure, nchar(Measure), nchar(Measure)),
+            order1 = ifelse(order1 == 0, 10, order1),
+            order1 = ifelse(order1 == 'l', 20, order1),
+            order1 = as.integer(order1)
+        ) %>%
+        dplyr::arrange(dplyr::desc(order1)) %>%
+        dplyr::select(-order1) %>%
+        dplyr::mutate(Measure = Measure %>%
+                   factor(., levels = unique(.))) %>%
+        seer::view_boxplots(
+            'Measure', 'Value') +
+        ggplot2::labs(x = 'Concentration (nmol/mL)',
+                      y = 'Triacylglyceride fatty acids') +
+        graph_theme(ticks = FALSE)
+}

### 2017-06-26 R/pls.R

@@ -1,0 +1,73 @@
+# Functions for the PLS analysis
+#
+# Grab or combine data ----------------------------------------------------
+
+#' Prepare the project data for PLS analysis.
+#'
+#' @param data Project data.
+#' @param y Y or outcome variable.
+#' @param x X or exposure variables of interest.
+#'
+#' @export
+prep_pls_data <- function(data, y, x) {
+    data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select_(.dots = c(y, x)) %>%
+        stats::na.omit()
+}
+
+# Analyze -----------------------------------------------------------------
+
+#' Compute PLS analysis.
+#'
+#' @param data Project data.
+#' @param y Outcome variable.
+#' @param x Exposure variables.
+#' @param ncomp Number of components.
+#' @param cv Whether to use CV.
+#'
+#' @export
+analyze_pls <- function(data = project_data,
+                        y, x = tg_pct, ncomp = NULL, cv = TRUE) {
+    data %>%
+        prep_pls_data(y = y, x = x) %>%
+        mason::design('pls') %>%
+        mason::add_settings(ncomp = ncomp, validation = 'CV', cv.data = cv, cv.seed = 5436) %>%
+        mason::add_variables('yvars', y) %>%
+        mason::add_variables('xvars', x) %>%
+        mason::construct() %>%
+        mason::scrub()
+}
+
+# Plots -------------------------------------------------------------------
+
+#' Plot the X loadings of the PLS results.
+#'
+#' @param data PLS output results.
+#'
+#' @export
+plot_pls <- function(data) {
+    seer::view_pls_xloadings(data, renaming.x = renaming_fats, dot.colour = "black") +
+        graph_theme(minor.grid.lines = FALSE)
+}
+
+# Calculations ------------------------------------------------------------
+
+#' Calculate the correlation between the predicted and actual outcome.
+#'
+#' @param model The PLS output results.
+#' @param test The test data to predict from.
+#' @param ncomps The component to predict on.
+#'
+#' @export
+calc_pred_corr <- function(model, test, ncomps = 1) {
+    predicted <- stats::predict(model, ncomp = ncomps, newdata = test)
+    measured <- as.matrix(stats::model.response(
+        stats::model.frame(formula(model), data = test))
+        )
+    corr <- broom::tidy(stats::cor.test(predicted, measured))[c(1, 3)]
[…]

### 2017-06-26 R/prep_data.R

@@ -1,0 +1,37 @@
+# Grab or combine data ----------------------------------------------------
+
+#' Wrangle data to determine dysglycemia conversion.
+#'
+#' @param data The project data.
+#'
+#' @export
+prep_dys_data <- function(data) {
+    dysgly.data <-
+        dplyr::left_join(
+            data %>%
+                dplyr::filter(VN == 0),
+            data %>%
+                dplyr::filter(!is.na(TotalTG)) %>%
+                dplyr::mutate_at(c("IFG", "IGT"), dplyr::funs(ifelse(is.na(.), 0, .))) %>%
+                dplyr::mutate(PreDM = as.numeric(rowSums(.[c('IFG', 'IGT')], na.rm = TRUE))) %>%
+                dplyr::mutate(FactorDysgly = ifelse(
+                    PreDM == 1, 'PreDM',
+                    ifelse(DM == 1, 'DM',
+                           'NGT')
+                )) %>%
+                dplyr::select(SID, VN, FactorDysgly) %>%
+                tidyr::spread(VN, FactorDysgly) %>%
+                dplyr::mutate(
+                    DetailedConvert = as.factor(paste0(`0`, '-', `1`, '-', `2`)),
+                    ConvertDysgly = as.numeric(!grepl('NGT$|NGT-NA$|NGT-NGT-NA$|NGT-NA-NA$',
+                                                      DetailedConvert)),
+                    ConvertDM = as.numeric(grepl('-DM$|-DM-DM$|-DM-NA$|-DM-NGT$',
+                                                 DetailedConvert)),
+                    ConvertPreDM = as.numeric(grepl('-PreDM$|-PreDM-PreDM$|-PreDM-NA$',
+                                                    DetailedConvert))
+                )
+        ) %>%
+        dplyr::filter(!is.na(TotalTG))
+
+    return(dysgly.data)
+}

### 2017-06-26 R/qic.R

@@ -1,0 +1,153 @@
+
+# Prepare data for QIC ----------------------------------------------------
+
+#' Prepare the data for using in the QIC model comparisons.
+#'
+#' @param data Project data
+#'
+prep_qic_data <- function(data) {
+    data %>%
+        prep_gee_data() %>%
+        dplyr::select(
+            SID,
+            VN,
+            YearsFromBaseline,
+            lISSI2,
+            lISI,
+            TotalTG,
+            TotalNE,
+            Sex,
+            BiEthnicity,
+            MET,
+            ALT,
+            BaseAge,
+            Waist,
+            BMI,
+            AlcoholPerWk,
+            TobaccoUse,
+            FamHistDiab
+        ) %>%
+        stats::na.omit()
+}
+
+# Clean QIC output --------------------------------------------------------
+
+#' Clean data output from the QIC model selection functions.
+#'
+#' @param data QIC output data.
+#'
+#' @export
+clean_qic <- function(data) {
+    data %>%
+        tibble::rownames_to_column() %>%
+        dplyr::select(Model = rowname, QIC = IC, Delta = delta) %>%
+        dplyr::mutate(id = seq_len(nrow(.)))
+}
+
+# Analyze using QIC -------------------------------------------------------
+
+#' QIC output for insulin sensitivity.
+#'
+#' @param data Project data.
+#'
+#' @export
+qic_is <- function(data = project_data) {
+    # need to force into global environment because model.sel doesn't work
+    qic_data <<- data %>%
+        prep_qic_data() %>%
+        dplyr::select(-lISSI2) %>%
+        as.data.frame()
+
+    M0 <- geepack::geeglm(
+        lISI ~ TotalTG + YearsFromBaseline,
+        data = qic_data,
+        id = SID,
+        family = stats::gaussian,
+        corstr = 'ar1'
+    )
+
+    M1 <- stats::update(M0, . ~ . + TotalTG:YearsFromBaseline)
+    M2 <- stats::update(M0, . ~ . + BaseAge + BiEthnicity + Sex)
+    M3 <- stats::update(M0, . ~ . + BaseAge + BiEthnicity + Sex + Waist)
[…]

### 2017-06-26 R/renaming.R

@@ -1,0 +1,53 @@
+# Renaming ----------------------------------------------------------------
+
+renaming_table_rows <- function(x) {
+    x %>%
+        gsub('IGIIR', 'IGI/IR', .) %>%
+        gsub('ISSI2', 'ISSI-2', .) %>%
+        gsub('HOMA2IR', 'HOMA2-IR', .) %>%
+        gsub('HOMA2_S', 'HOMA2-%S', .) %>%
+        gsub('TAG', 'TAG (mmol/L)', .) %>%
+        gsub('Chol', 'Chol (mmol/L)', .) %>%
+        gsub('LDL', 'LDL (mmol/L)', .) %>%
+        gsub('ALT', 'ALT (U/L)', .) %>%
+        gsub('HDL', 'HDL (mmol/L)', .) %>%
+        gsub('BaseTotalTG', 'TAGFA (nmol/mL)', .) %>%
+        gsub('BaseTotalNE', 'NEFA (nmol/mL)', .) %>%
+        gsub('Age', 'Age (yrs)', .) %>%
+        gsub('BMI', 'BMI (kg/m^2^)', .) %>%
+        gsub('Waist', 'WC (cm)', .)
+}
+
+renaming_outcomes <- function(x) {
+    x %>%
+        gsub('linvHOMA', 'log(1/HOMA-IR)', .) %>%
+        gsub('lHOMA2IR', 'log(HOMA2-IR)', .) %>%
+        gsub('lHOMA2_S', 'log(HOMA2-%S)', .) %>%
+        gsub('lISI', 'log(ISI)', .) %>%
+        gsub('lIGIIR', 'log(IGI/IR)', .) %>%
+        gsub('lISSI2', 'log(ISSI-2)', .) %>%
+        gsub('^HOMA2_S$', 'HOMA2-%S', .) %>%
+        gsub('^ISSI2$', 'ISSI-2', .) %>%
+        gsub('^IGIIR$', 'IGI/IR', .)
+}
+
+renaming_fats <- function(x) {
+    x %>%
+        gsub('.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
+        gsub('n(\\d)$', 'n-\\1', .) %>%
+        gsub('D(\\d\\d)$', 'D-\\1', .) %>%
+        gsub('^pct_', '', .) %>%
+        gsub('TotalTG', 'Total', .) %>%
+        gsub('^TAG$', 'Clinical TAG', .)
+}
+
[…]

### 2017-06-26 R/rfigshare.R

@@ -1,0 +1,27 @@
+# Steps to take in the Git repo:
+# 1. Make new orphan branch (git checkout --orphan code)
+# 2. Keep only R files and purl('vignette/maindocument.Rmd')
+# 3. git commit the files
+# 4. git archive --format=zip --output=code-archive.zip code
+# 5. Then run the below code
+send_to_figshare <- function() {
+    rfigshare::fs_new_article(
+        "Analysis code for poster on triacylglycerol DNL on components of the Metabolic Syndrome in PROMISE",
+        "Analysis code for a poster for the American Diabetes Association Scientific Sessions in 2017",
+        type = "fileset",
+        # there was a problem with the authors...
+        # authors = c("First Last"),
+        tags = c(
+            "insulin sensitivity",
+            "beta-cell function",
+            "fatty acids",
+            "triacylglycerols",
+            "cohort",
+            "longitudinal"
+        ),
+        categories = c("Diseases", "Pathogenesis", "Epidemiology"),
+        links = "https://github.com/lwjohnst86/tagDiabetes",
+        files = "code-archive.zip",
+        visibility = "draft"
+    )
+}

### 2017-06-26 R/setup.R

@@ -1,0 +1,36 @@
+# Setup description -------------------------------------------------------
+#
+# These functions are used to setup your files so that all options are in one
+# location, and references via the function call.
+#
+# These functions can be accessed using either `library(projectname)` or
+# `devtools::load_all()` (Ctrl-Shift-L in RStudio).
+
+#' Set options for all documents and scripts.
+#'
+#' @export
+#'
+#' @examples
+#'
+#' set_options()
+#'
+set_options <- function() {
+    # Set the options here for individual packages
+
+    # For tables (uncomment below if using pander)
+    pander::panderOptions('table.split.table', Inf)
+    pander::panderOptions('table.style', 'rmarkdown')
+    pander::panderOptions('table.alignment.default',
+                  function(df)
+                      ifelse(sapply(df, is.numeric), 'center', 'left'))
+
+    # For the document (knitr)
+    knitr::opts_chunk$set(
+        warning = FALSE, message = FALSE, collapse = TRUE,
+        fig.showtext = FALSE, comment = '#>', dev = "pdf"
+    )
+}
+
+#' @importFrom magrittr %>%
+#' @export
+magrittr::`%>%`

### 2017-06-26 R/tables.R

@@ -1,0 +1,57 @@
+# Tables ------------------------------------------------------------------
+
+#' Create a table of the basic characteristics of PROMISE.
+#'
+#' @param data Project data.
+#' @param caption Table caption.
+#'
+#' @export
+table_basic <- function(data = project_data, caption = NULL) {
+    data %>%
+        dplyr::mutate(Ethnicity = ifelse(VN == 0, as.character(Ethnicity), NA),
+                      Ethnicity = as.factor(Ethnicity),
+                      Sex = ifelse(VN == 0, as.character(Sex), NA),
+                      Sex = as.factor(Sex)) %>%
+        carpenter::outline_table('f.VN') %>%
+        carpenter::add_rows(c('HOMA2_S', 'ISI', 'IGIIR', 'ISSI2'), carpenter::stat_medianIQR, digits = 1) %>%
+        carpenter::add_rows(c('ALT', 'TAG', 'Chol', 'HDL', 'BaseTotalTG', 'BaseTotalNE', 'MET',
+                              'Age', 'BMI', 'Waist'),
+                            carpenter::stat_meanSD, digits = 1) %>%
+        carpenter::add_rows(c('Ethnicity', 'Sex'), carpenter::stat_nPct, digits = 0) %>%
+        carpenter::renaming('rows', renaming_table_rows) %>%
+        carpenter::renaming("header", c('Measure', 'Baseline', '3-yr', '6-yr')) %>%
+        carpenter::build_table(caption = caption)
+}
+
+#' Create a table of the values for the conc and mol% of TAGFA.
+#'
+#' @param data Project data.
+#' @param caption Table caption.
+#'
+#' @export
+table_tagfa <- function(data = project_data, caption = NULL) {
+    fa <- function(pattern, x = c(tg_conc, 'TotalTG'))
+        grep(pattern, x, value = TRUE)
+    tgfa <- c(fa('3$'), fa('6$'), fa('7$'), fa('9$'), fa('0$'), fa('TotalTG$'))
+
+    data %>%
+        dplyr::filter(VN == 0) %>%
+        tidyr::gather(FA, Value, dplyr::matches('tg\\d')) %>%
+        dplyr::mutate(unit = as.factor(ifelse(grepl('pct_', FA), 'mol%', 'nmol/mL')),
+                      FA = gsub('pct_', '', FA)) %>%
+        tidyr::spread(FA, Value) %>%
[…]

### 2017-06-26 R/utils.R

@@ -1,0 +1,48 @@
+# Misc --------------------------------------------------------------------
+
+# Trim white space
+trim_ws <- function (x) {
+    gsub("^\\s+|\\s+$", "", x)
+}
+
+format_rounding <- function(x, digits = 1) {
+    trim_ws(format(round(x, digits = digits), nsmall = digits))
+}
+
+format_p <- function(x) {
+   format.pval(x, digits = 2, eps = 0.001)
+}
+
+graph_theme <- function(base.plot, ticks = TRUE, minor.grid.lines = FALSE, legend.pos = 'bottom') {
+    graph.theme <-
+        ggplot2::"%+replace%"(
+            ggthemes::theme_tufte(base_size = 10, base_family = 'sans'),
+            ggplot2::theme(
+                axis.line = ggplot2::element_line('black'),
+                axis.line.x = ggplot2::element_line('black'),
+                axis.line.y = ggplot2::element_line('black'),
+                legend.key.width = grid::unit(0.7, "line"),
+                legend.key.height = grid::unit(0.7, "line"),
+                strip.background = ggplot2::element_blank(),
+                plot.margin = grid::unit(c(0.5, 0, 0, 0), "cm"),
+                legend.position = legend.pos
+            )
+        )
+
+    if (!ticks) {
+        graph.theme <- ggplot2::"%+replace%"(graph.theme,
+                                             ggplot2::theme(axis.ticks.y = ggplot2::element_blank()))
+    }
+
+    if (minor.grid.lines) {
+        graph.theme <- ggplot2::"%+replace%"(
+            graph.theme,
+            ggplot2::theme(
+                panel.grid = ggplot2::element_line(),
+                panel.grid.minor = ggplot2::element_blank()
+            )
+        )
+    }
+
+    return(graph.theme)
+}

### 2017-06-26 R/variables.R

@@ -1,0 +1,5 @@
+tg_conc <- grep('^tg\\d+', vars, value = TRUE)
+tg_pct <- grep('^pct_tg\\d+', vars, value = TRUE)
+tg_totals <- c('TotalTG', 'BaseTAG')
+outcomes <- c('lHOMA2_S', 'lISI', 'lIGIIR', 'lISSI2')
+covariates <- c('YearsFromBaseline', 'Waist', 'BaseAge', 'BiEthnicity', 'Sex', 'ALT', 'MET', 'TotalNE')

### 2017-06-26 R/zzz.R

@@ -1,0 +1,4 @@
+.onLoad <- function(libname, pkgname) {
+    set_options()
+    load_data()
+}

### 2018-01-25 R/gee.R
ordering fixed for xterms

@@ -162,15 +162,18 @@ padjust_model_output <- function(gee_results) {
 order_model_output <- function(gee_results) {
     gee_results %>%
         mutate(
-            order1 = substr(Xterms, nchar(Xterms), nchar(Xterms)),
-            order1 = ifelse(order1 == 0, 10, order1),
-            order1 = ifelse(order1 == 'l', 20, order1),
-            order1 = ifelse(order1 == 'G', 30, order1),
-            order1 = as.integer(order1)
+            ordering = as.character(substr(Xterms, nchar(Xterms), nchar(Xterms))),
+            ordering = case_when(
+                ordering == "0" ~ "10",
+                ordering == "l" ~ "20",
+                ordering == "G" ~ "30",
+                TRUE ~ ordering
+            ),
+            ordering = as.integer(ordering)
         ) %>%
-        arrange(desc(order1)) %>%
-        mutate(Xterms = factor(Xterms, unique(Xterms))) %>%
-        select(-order1)
+        arrange(desc(ordering)) %>%
+        mutate(Xterms = forcats::fct_inorder(Xterms)) %>%
+        select(-ordering)
 }
 
 set_outcome_order_model_output <- function(gee_results) {

### 2018-01-25 R/setup.R

@@ -31,6 +31,7 @@ set_options <- function() {
     )
 }
 
+#' @import dplyr tidyr ggplot2 mason
 #' @importFrom magrittr %>%
 #' @export
 magrittr::`%>%`

### 2018-01-25 R/renaming.R
renaming fats fix (with promise.misc fun)

@@ -33,21 +33,13 @@ renaming_outcomes <- function(x) {
 
 renaming_fats <- function(x) {
     x %>%
-        gsub('.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
-        gsub('n(\\d)$', 'n-\\1', .) %>%
-        gsub('D(\\d\\d)$', 'D-\\1', .) %>%
-        gsub('^pct_', '', .) %>%
-        gsub('TotalTG', 'Total', .) %>%
-        gsub('^TAG$', 'Clinical TAG', .)
+        stringr::str_replace('^pct_', '') %>%
+        stringr::str_replace('TotalTG', 'Total') %>%
+        stringr::str_replace('BaseTAG', 'Clinical TAG') %>%
+        PROMISE.misc::renaming_fa(keep.fraction = FALSE)
 }
 
 renaming_fraction <- function(x) {
     x %>%
         gsub('tg', 'Triacylglycerol', .)
-}
-
-renaming_list <- function(x) {
-    x %>%
-        renaming_fa() %>%
-        renaming_outcomes()
 }

### 2018-01-25 R/utils.R
use base trimws, removed ggplot2::

@@ -1,12 +1,7 @@
 # Misc --------------------------------------------------------------------
 
-# Trim white space
-trim_ws <- function (x) {
-    gsub("^\\s+|\\s+$", "", x)
-}
-
 format_rounding <- function(x, digits = 1) {
-    trim_ws(format(round(x, digits = digits), nsmall = digits))
+    trimws(format(round(x, digits = digits), nsmall = digits))
 }
 
 format_p <- function(x) {
@@ -15,31 +10,32 @@ format_p <- function(x) {
 
 graph_theme <- function(base.plot, ticks = TRUE, minor.grid.lines = FALSE, legend.pos = 'bottom') {
     graph.theme <-
-        ggplot2::"%+replace%"(
+        "%+replace%"(
             ggthemes::theme_tufte(base_size = 10, base_family = 'sans'),
-            ggplot2::theme(
-                axis.line = ggplot2::element_line('black'),
-                axis.line.x = ggplot2::element_line('black'),
-                axis.line.y = ggplot2::element_line('black'),
+            theme(
+                axis.line = element_line('black'),
+                axis.line.x = element_line('black'),
+                axis.line.y = element_line('black'),
                 legend.key.width = grid::unit(0.7, "line"),
                 legend.key.height = grid::unit(0.7, "line"),
-                strip.background = ggplot2::element_blank(),
+                strip.background = element_blank(),
                 plot.margin = grid::unit(c(0.5, 0, 0, 0), "cm"),
-                legend.position = legend.pos
+                legend.position = legend.pos,
+                strip.placement = "outside"
             )
         )
 
     if (!ticks) {
-        graph.theme <- ggplot2::"%+replace%"(graph.theme,
-                                             ggplot2::theme(axis.ticks.y = ggplot2::element_blank()))
+        graph.theme <- "%+replace%"(graph.theme,
+                                             theme(axis.ticks.y = element_blank()))
     }
 
     if (minor.grid.lines) {
-        graph.theme <- ggplot2::"%+replace%"(
+        graph.theme <- "%+replace%"(
             graph.theme,
-            ggplot2::theme(
-                panel.grid = ggplot2::element_line(),
-                panel.grid.minor = ggplot2::element_blank()
+            theme(
+                panel.grid = element_line(),
+                panel.grid.minor = element_blank()
             )
         )
     }

### 2018-01-25 doc/manuscript.Rmd
Added code DOI
include dplyr, rename chunk to setup

@@
 
-```{r load_data, message=FALSE, echo=FALSE}
+```{r setup, message=FALSE, echo=FALSE}
 devtools::load_all('.')
+library(dplyr)
 knitr::opts_chunk$set(echo = FALSE)
 ds0yr <- project_data %>% 
-    dplyr::filter(VN == 0)
+    filter(VN == 0)
 ```
@@
 `r packageVersion('pls')` for PLS. The R code and extra analyses for
-this manuscript is available at {{code doi}}. Results were considered
+this manuscript is available at https://doi.org/10.6084/m9.figshare.5143438. Results were considered
 statistically significant at p<0.05, after adjusting for multiple testing using

### 2018-01-25 doc/manuscript.Rmd
only one dot size for signif for gee, remove legend

@@
 ```{r fig_gee_unadj, fig.cap=fig_gee_unadj}
-plot_gee_main(gee_unadj_df)
+gee_unadj_df %>% 
+    mutate(p.value = ifelse(p.value <= 0.05, 0.04, p.value)) %>% 
+    plot_gee_main() +
+    ggplot2::theme(legend.position = "none")
 ```
@@
 ```{r fig_gee_adj, fig.cap=fig_gee_adj}
-plot_gee_main(gee_adj_df)
+gee_adj_df %>% 
+    mutate(p.value = ifelse(p.value <= 0.05, 0.04, p.value)) %>% 
+    plot_gee_main() +
+    ggplot2::theme(legend.position = "none")
 ```

### 2018-02-14 doc/manuscript.Rmd
moved esm out of manuscript

@@
 
-# Supplemental Material
-
-## Supplemental Methods: Statistical analysis
-
-Generalized estimating equations (GEE) is a technique similar to mixed effects
-modeling, except it calculates the marginal population estimates compared to the
-subject-specific estimates in mixed effects models. GEE is well suited to and commonly
-used in longitudinal cohort studies, especially given its capacity to handle
-missed visits.
-
-The working correlation matrix for the GEE analysis was chosen based on 
-quasi-likelihood information criteria (QIC). The auto-regressive of order 1
-(AR1) matrix was chosen for the GEE models as it had the best model fit accessed
-using QIC, though other matrices (eg. exchangeable) had similar fit (data not
-shown).
-
-For the confounders, they were chosen based on previous literature, from
-directed acyclic graph (DAG) [@Greenland1999a] recommendations, and from
-QIC. The DAG recommendations were obtained from using the DAGitty software
-[@Textor2011a], http://dagitty.net/. See `r cite_sf('dag_is')` and 
-`r cite_sf('dag_bcf')`) for the DAG model. See `r cite_st('qic')` for the
-comparison of various models with different covariates using QIC. While the
-final GEE model selected as best fitting differed between insulin sensitivity
-and beta-cell function measures, the differences in the QIC values were less than 10
-between many of the models, suggesting similar fit. As such, we selected the
-model that had the fewest covariates and that had similar fit between the
-outcome measures. Total NEFA was included as a confounder because it is used
-substantially as a source of fatty acids in hepatic TAG production
-[@Barrows2006a].
-
-No imputation was conducted on missing values. Prevalent cases of diabetes at
-baseline and incident cases at follow-up were excluded from the GEE analysis. 
-The resulting GEE beta estimates were exponentiated to allow the interpretation
-as stated in the main methods.
-
-PLS is a technique used for multivariate, high dimensionality datasets where
-there is a potential or likely underlying structure to the data. Because it uses
-a response variable (i.e. the outcome or y variable) when identifying the 
-underlying structure it is known as a supervised statistical method, which gives
-it greater predictive power when using the model against updated or new data.
-PLS generates a number of components based on the number of variables provided.
-We used internal 10-fold cross-validation to determine which components to
-extract from the PLS analysis. Based on the internal cross-validation, the first
-two components gave the highest amount of explained variance (data not shown),
-of which we decided to use these two components in the final results.
-
-The predictive capability of the PLS models were tested using cross-validation. 
-The data set was randomly split in half into a training set and a testing set. 
-After specifying the model on the training set, results were compared using the 
-testing set to determine whether how predictive the model was given a new
-dataset. Final results shown in the figures use the full dataset (rather than a
-training or testing set). No prevalent diabetes cases were included in the PLS 
-analysis.
-
-## Supplemental Tables
-
-```{r table_qic}
-analyze_qic() %>% 
-    table_qic(caption = tab_qic)
-```
-
-Given the number of possible combinations of outcome and predictor variables,
-only ISI and ISSI-2 with total triacylglycerol fatty acids (nmol/mL) were used to
-compare various GEE models and to select a final model.  Baseline age was used
-as including both the original age and the time variable would result in
-collinearity.  Column names: QIC is the quasi-likelihood information criteria
-(smaller values, eg. larger negative values, indicate a better fit compared to
-other models), Delta is the QIC minus the lowest QIC (models with delta <10 are 
-considered equivalent).  Models were:
-
-- M0: log(ISSI-2) or log(ISI) = total triacylglycerol fatty acids (nmol/mL) + years from baseline
-- M1: M0 + fatty acid by time interaction
-- M2: M0 + sex + ethnicity + baseline age
-- M3: M2 + waist
-- M4: M3 + ALT
-- M5: M4 + physical activity (MET)
-- M6: M5 + total NEFA
-- M7: M6 + alcohol intake
-- M8: M7 + family history of diabetes
-- M9: M8 + smoking status
-
-```{r table_tagfa}
-table_tagfa(caption = tab_tagfa)
-```
-
-```{r table_gee}
-table_gee_main(gee_unadj_df, caption = tab_gee_unadj)
-table_gee_main(gee_adj_df, caption = tab_gee_adj)
-```
-
-## Supplemental Figures
-
-![`r fig_consort`](../img/flowDiagramSample.png)
-
-![`r fig_dag_is`](../img/dagitty-model-is.png)
-
-![`r fig_dag_bcf`](../img/dagitty-model-bcf.png)
-
-```{r fig_gee_adj_nowaist, fig.cap=fig_gee_adj_nowaist}
-plot_gee_main(analyze_gee(covars = covariates[which(covariates != 'Waist')]))
-```
-
 # References

### 2018-02-14 doc/manuscript.Rmd
moved analysis of gee and pls into rda files

@@
 
-```{r analyze_gee, cache=TRUE}
-gee_unadj_df <- analyze_gee(covars = "YearsFromBaseline")
-gee_adj_df <- analyze_gee()
-```
-
 ```{r analyze_others}
-# CV results are shown in pls.Rmd vignette
-pls_homa2 <- analyze_pls(y = 'lHOMA2_S', cv = FALSE)
-pls_isi <- analyze_pls(y = 'lISI', cv = FALSE)
-
 # Correlation between predicted and observed
@@
 
-cor_df <- analyze_corr()
-cor_df_mol <- analyze_corr(y = tg_pct)
-cor_df_tagfa <- analyze_corr_tagfa()
-
 pls_gee <- project_data %>%
@@
         data_prep = 'pls',
-        pls_results = pls_isi
+        pls_results = pls_results$pls_isi_df
         ) %>% 
@@
     dplyr::filter(Xterms == 'Comp1') %>% 
-    dplyr::mutate(p.value = format_p(p.value),
-                  estimate = format_rounding(estimate))
+    dplyr::mutate(p.value = aide::format_pval(p.value),
+                  estimate = aide::format_round(estimate))
 ```

### 2018-02-16 doc/manuscript.Rmd
ggplot2 and patchwork to setup chunk

@@
 devtools::load_all('.')
+library(ggplot2)
+library(patchwork)
 library(dplyr)

### 2018-02-16 doc/manuscript.Rmd
captions to esm

@@
 
-fig_consort <- sfig('consort', 'CONSORT diagram of PROMISE participants over the 3 visits.')
-fig_dag_is <- sfig('dag_is', 'Directed acyclic graphic output from the DAGitty online software for insulin sensitivity.')
-fig_dag_bcf <- sfig('dag_bcf', 'Directed acyclic graphic output from the DAGitty online software for beta-cell function.')
 fig_tagfa <- fig('tagfa', 'Distribution of the composition of triacylglycerol fatty acids in the baseline visit of PROMISE participants (2004-2006).
@@
 Variables controlled for were follow-up time, waist circumference, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
-X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
-fig_gee_adj_nowaist <- sfig('gee_adj_nowaist', 'Fully-adjusted (without waist size) GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
-Variables controlled for were follow-up time, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
 X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.

### 2018-02-16 doc/esm.Rmd
esm material as own file

---
output: fost::diabetologia_manuscript
---

```{r setup, include=FALSE}
devtools::load_all()
set_options()
knitr::opts_chunk$set(results = "asis", echo = FALSE)

# Captions
sfig <- captioner::captioner(prefix = 'Supplemental Figure')
cite_sf <- pryr::partial(sfig, display = 'cite')

tab <- captioner::captioner(prefix = 'Table')
cite_t <- pryr::partial(tab, display = 'cite')
stab <- captioner::captioner(prefix = 'Supplemental Table')
cite_st <- pryr::partial(stab, display = 'cite')

fig_consort <- sfig('consort', 'CONSORT diagram of PROMISE participants over the 3 visits.')
fig_dag_is <- sfig('dag_is', 'Directed acyclic graphic output from the DAGitty online software for insulin sensitivity.')
fig_dag_bcf <- sfig('dag_bcf', 'Directed acyclic graphic output from the DAGitty online software for beta-cell function.')
fig_gee_adj_nowaist <- sfig('gee_adj_nowaist', 'Fully-adjusted (without waist size) GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort. Variables controlled for were follow-up time, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
```

# Supplemental Methods

## Equations for outcomes

The equations for calculating the outcome variables are listed below. HOMA2-%S was 
calculated using the Excel found at https://www.dtu.ox.ac.uk/homacalculator/.

```{r}
cat(PROMISE.methods::insert_formula("ISI"))
```

```{r}
cat(PROMISE.methods::insert_formula("HOMA"))
```

```{r}
cat(PROMISE.methods::insert_formula("IGIIR"))
```

```{r}
cat(PROMISE.methods::insert_formula("ISSI2"))
```

Area under the curve (AUC) for both glucose (G) and insulin (I) was computed
using the trapezoidal rule. ISI~OGTT~ (ISI) uses all the glucose and insulin
values from the oral glucose tolerance (the overall mean).

## Statistical methods description

These methods are also described in our article on the NEFA fraction
[@Johnston2017a].

Generalised estimating equations (GEE) is a technique similar to mixed effects 
modelling, except it calculates the marginal population estimates compared to the
subject-specific estimates in mixed effects models. GEE is well suited to and
commonly used in longitudinal cohort studies, especially given its capacity to
handle missed visits and for the correlation of variables across time. 
GEE is able to handle the correlation of longitudinal data due to the
estimating of a working correlation matrix, which was chosen based on
quasi-likelihood information criteria (QIC; more detail on QIC below). The auto-regressive of order 1
(AR1) matrix was chosen for the GEE models as it had the best model fit accessed
using QIC, though other matrices (eg. exchangeable) had similar fit (data not
shown). While the choice of the correlation matrix is important, it should be noted
that even with misspecification GEE still computes robust estimates and standard errors.
One main advantage of using methods such as GEE is the inclusion of the
working correlation matrix that takes into account the inherent covariance
between values measured at different time points when computing model estimates
and standard errors. AR1 is the most appropriate correlation matrix for
longitudinal data since the correlation matrix assumes lower intercorrelation of
values within an individual as time passes. For example, BMI measured at the
baseline visit would be more highly correlated to BMI measured at the 3rd year
visit than compared to the 6th year visit.

For the covariates, they were chosen based on previous literature, from directed
acyclic graph (DAG) recommendations, and from QIC. The DAG recommendations were
obtained from using the DAGitty software, http://dagitty.net/. DAG is a
technique that allows for identifying potential confounding causal pathways and
to limit potential bias. DAGs are composed on nodes (variables) connected by
edges (arrows; causal pathway between two variables). Using DAGs to identify
variables to adjust for is done algorithmically, as implemented in the DAGitty
software. DAGitty calculates possible sets of variables to include in a model to
adjust for potential confounding in the causal network. Use of DAGs in this
manner can identify potential sources of bias, such as from inclusion of
colliding variables. A collider variable is a variable that has two (or more)
'parent' nodes/variables (connected causally by DAG edges pointing into the
'child' node, which is the collider). The collider variable and parent variables
are directed linked to the exposure and the outcome. Including all variables
involved with the collider could introduce bias as it does not generate an
unconditional association between all involved variables in the model.

QIC is a technique that balances the goodness of fit of the model with its
relative complexity (i.e. number of covariates) against other models being
compared. QIC, more so then the more common AIC, penalises the inclusion of more
variables in the model. This technique is used to compare multiple models to
identify the model with the highest fit and with the lowest complexity (variables),
*of the models compared*. QIC is similar to AIC, except AIC is based on maximum
likelihood of models such as mixed effects models. Since GEE does not use
maximum likelihood, AIC can not be used. QIC and other information
criterion methods are used in conjunction with other techniques such as
literature and DAG based. The final GEE model selected as best fitting was
negligibly different when using insulin sensitivity or beta-cell function as an
outcome and as such the same covariates were chosen for models with each of the
outcomes. The covariates medication use, WC, ALT, and MET-h/week would be considered
classical confounders while TAG would be considered a mediating variable.

Partial least squares (PLS) regression is a technique similar to principal
component analysis (PCA) that is designed to extract meaningful information from
multivariate, high dimensionality data (e.g. as in metabolomic and other -omic type 
analyses). These types of methods try to extract as much of the variance in the 
data into a smaller number of components or factors. The difference between PCA 
and PLS is that PLS is a supervised method, while PCA is not. This means that
PLS uses a response variable(s), i.e. an outcome or **Y**, to 'supervise' (or
constrain) the variation inherent in the predictors (**X**) while PCA only
describes the variation inherent in **X**. Because of this, PLS can be better at
predicting the contributions of predictor variables against an outcome variable.
In this analysis, PLS was computed on the NEFA as a proportion due to the
inherent interdependence of the values (all must equal 100%). This 
interdependence allows PLS to identify patterns in the composition, since
the concentration data is all highly positively correlated with each other.

We used internal 10-fold cross-validation to determine which components to
extract from the PLS analysis. Based on the internal cross-validation, the first
two components gave the highest amount of explained variance (data not shown),
of which we decided to use these two components in the final results.

The predictive capability of the PLS models were tested using cross-validation. 
The data set was randomly split in half into a training set and a testing set. 
After specifying the model on the training set, results were compared using the 
testing set to determine whether how predictive the model was given a new
dataset. Final results shown in the figures use the full dataset (rather than a
training or testing set). No prevalent diabetes cases were included in the PLS 
analysis.

## Supplemental Tables

```{r table_qic}
analyze_qic() %>% 
    table_qic(caption = tab_qic)
```

Given the number of possible combinations of outcome and predictor variables,
only ISI and ISSI-2 with total triacylglycerol fatty acids (nmol/mL) were used to
compare various GEE models and to select a final model.  Baseline age was used
as including both the original age and the time variable would result in
collinearity.  Column names: QIC is the quasi-likelihood information criteria
(smaller values, eg. larger negative values, indicate a better fit compared to
other models), Delta is the QIC minus the lowest QIC (models with delta <10 are 
considered equivalent).  Models were:

- M0: log(ISSI-2) or log(ISI) = total triacylglycerol fatty acids (nmol/mL) + years from baseline
- M1: M0 + fatty acid by time interaction
- M2: M0 + sex + ethnicity + baseline age
- M3: M2 + waist
- M4: M3 + ALT
- M5: M4 + physical activity (MET)
- M6: M5 + total NEFA
- M7: M6 + alcohol intake
- M8: M7 + family history of diabetes
- M9: M8 + smoking status

```{r table_tagfa}
table_tagfa(caption = tab_tagfa)
```

```{r table_gee}
table_gee_main(gee_unadj_df, caption = tab_gee_unadj)
table_gee_main(gee_adj_df, caption = tab_gee_adj)
```

## Supplemental Figures

![`r fig_consort`](../img/flowDiagramSample.png)

![`r fig_dag_is`](../img/dagitty-model-is.png)

![`r fig_dag_bcf`](../img/dagitty-model-bcf.png)

```{r fig_gee_adj_nowaist, fig.cap=fig_gee_adj_nowaist}
project_data %>% 
    analyze_gee(covars = covariates[which(covariates != 'Waist')]) %>% 
    plot_gee_main()
```

```{r fig_pls_loadings}
p1 <- plot_pls_loadings(pls_results$homa2) +
    ggtitle("PLS model loadings for HOMA2-%S")
p2 <- plot_pls_loadings(pls_results$isi) +
    ggtitle("PLS model loadings for ISI")
p1 + p2 + plot_layout(ncol = 1)
```

### 2018-02-16 R/pls.R
added several PLS functions to plot scores and loadings

@@ -1,6 +1,6 @@
 # Functions for the PLS analysis
 #
-# Grab or combine data ----------------------------------------------------
+# Grab, combine, or prep data ---------------------------------------------
 
 #' Prepare the project data for PLS analysis.
 #'
@@ -11,9 +11,64 @@
 #' @export
 prep_pls_data <- function(data, y, x) {
     data %>%
-        dplyr::filter(VN == 0) %>%
-        dplyr::select_(.dots = c(y, x)) %>%
-        stats::na.omit()
+        filter(VN == 0) %>%
+        select_(.dots = c(y, x)) %>%
+        na.omit()
+}
+
+#' Convert PLS model object to dataframe of loadings.
+#'
+#' @param .model PLS model object.
+loadings_as_df <- function(.model) {
+    loading_nums <- unclass(pls::loadings(.model))
+
+    loading_nums %>%
+        as_data_frame() %>%
+        mutate(xvariables = dimnames(loading_nums)[[1]])
+
+}
+
+#' Add a column for "large" loadings (larger than abs(0.25)).
+#'
+#' @param .loadings PLS loadings dataset.
+#'
+append_large_loadings <- function(.loadings) {
+    .loadings %>%
+        mutate(xvariables = stringr::str_replace(xvariables, "pct_", "") %>%
+                   PROMISE.misc::renaming_fa(keep.fraction = TRUE)) %>%
+        gather(components, loadings, -xvariables) %>%
+        mutate(loadings = as.numeric(loadings)) %>%
+        group_by(components) %>%
+        mutate(
+            max_loading = max(loadings),
+            min_loading = min(loadings),
+            large_loadings = !between(loadings, -0.25, 0.25),
+            xvariables = ifelse(large_loadings, xvariables, NA)
+        ) %>%
+        ungroup()
+}
+
+#' Extract scores from PLS results and include with original variables as a dataframe.
+#'
+#' @param model pls results
+#'
+scores_as_df <- function(model) {
+    .scores <- pls::scores(model)
+    attr(.scores, "explvar") <- NULL
+    .scores <- .scores %>%
+        unclass() %>%
+        as.matrix() %>%
+        as_tibble() %>%
+        rename_all(funs(stringr::str_replace(., " ", "")))
+
+    .data <- bind_cols(
+        model$model$X %>%
+            as_tibble(),
+        model$model$Y %>%
+            as_tibble()
+        )
+
[…]

### 2018-02-16 R/generate_results.R
generate the results dataframes

@@ -1,0 +1,41 @@
+#' (re-)generate the results of GEE, PLS, and correlation analyses.
+#'
+#' @return Creates several Rda files with lists of results in dataframes.
+generate_results <- function() {
+    generate_gee()
+    generate_pls()
+    generate_corr()
+}
+
+generate_gee <- function() {
+    gee_unadj_df <- analyze_gee(covars = "YearsFromBaseline")
+    gee_adj_df <- analyze_gee()
+    gee_results <- list(
+        unadj = gee_unadj_df,
+        adj = gee_adj_df
+    )
+    devtools::use_data(gee_results, overwrite = TRUE)
+}
+
+generate_pls <- function() {
+    # CV results are shown in pls.Rmd vignette
+    pls_homa2_df <- analyze_pls(y = 'lHOMA2_S', cv = FALSE)
+    pls_isi_df <- analyze_pls(y = 'lISI', cv = FALSE)
+    pls_results <- list(
+        homa2 = pls_homa2_df,
+        isi = pls_isi_df
+    )
+    devtools::use_data(pls_results, overwrite = TRUE)
+}
+
+generate_corr <- function() {
+    cor_conc_df <- analyze_corr(y = tg_conc)
+    cor_pct_df <- analyze_corr(y = tg_pct)
+    cor_tagfa_df <- analyze_corr_tagfa()
+    cor_results <- list(
+        conc = cor_conc_df,
+        pct = cor_pct_df,
+        tagfa = cor_tagfa_df
+    )
+    devtools::use_data(cor_results, overwrite = TRUE)
+}

### 2018-02-16 doc/manuscript.Rmd
jlr manuscript + vancouver csl

@@
     - /home/luke/Documents/bibliography/bib/bib-db/methods.bib
-csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
-output: fost::diabetes_care_manuscript
+csl: /home/luke/Documents/bibliography/styles/vancouver.csl
+output: fost::jlr_manuscript
 ---

### 2018-02-17 doc/manuscript.Rmd
setup and caption chunks as one

@@
     filter(VN == 0)
-```
 
-```{r caption_setup}
 ## Include captions below using `captioner` package

### 2018-02-17 doc/manuscript.Rmd
moved captions to esm

@@
 
-tab_qic <- stab('qic', 'Comparison of GEE model fitness for variable selection using quasi-likelihood information criteria.')
 tab_basic <- tab('basic', 'Basic characteristics of PROMISE participants at each of the 3 clinic visits.')
-tab_tagfa <- stab('tagfa', 'Concentration (nmol/mL) and relative percent (mol%) values of triacylglycerol fatty acids in PROMISE participants at the baseline visit (2004-2006).')
-tab_gee_unadj <- stab('gee_unadj', 'Raw estimates and confidence interval values for *time*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
-Estimates represent a percent difference in the outcome per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, with an asterisk (*) denoting a significant (p<0.05) association.')
-tab_gee_adj <- stab('gee_adj', 'Raw estimates and confidence interval values for *fully*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
-Variables controlled for were follow-up time, waist circumference, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
-Estimates represent a percent difference in the outcome per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, with an asterisk (*) denoting a significant (p<0.05) association.')
 

### 2018-02-17 doc/manuscript.Rmd
reference appropriate results dataframes

@@
         data_prep = 'pls',
-        pls_results = pls_results$pls_isi_df
+        pls_results = pls_results$isi
         ) %>% 
@@
 fhd <- calc_discr_npct(ds0yr$FamHistDiab, 'Yes')
-est0 <- calc_gee_estci(gee_unadj_df)
-est1 <- calc_gee_estci(gee_adj_df)
-fa_cor <- calc_cor(cor_df)
-fa_cor_mol <- calc_cor(cor_df_mol)
+gmest0 <- calc_gee_estci(gee_results$unadj)
+gmest1 <- calc_gee_estci(gee_results$adj)
+fa_cor <- calc_cor(cor_results$conc)
+fa_cor_mol <- calc_cor(cor_results$pct)
 cor_pls_homa2 <- calc_pred_corr(pls_homa2_t, pls_homa2_t$test_data)

### 2018-02-17 doc/manuscript.Rmd
fix plots

@@
 ```{r table_basic}
+project_data %>% 
     table_basic(caption = tab_basic)
@@
 ```{r fig_tagfa, fig.cap=fig_tagfa}
+project_data %>% 
     plot_tagfa()
@@
 ```{r fig_heatmap, fig.cap=fig_heatmap}
-plot_heatmap(cor_df)
+plot_heatmap(cor_results$conc)
 ```
@@
 ```{r fig_heatmap_mol, fig.cap=fig_heatmap_mol}
-plot_heatmap(cor_df_mol, unit = 'mol%')
+plot_heatmap(cor_results$pct, unit = 'mol%')
 ```
@@
 ```{r fig_heatmap_tagfa, fig.cap=fig_heatmap_tagfa}
-plot_heatmap(cor_df_tagfa, text = FALSE)
+plot_heatmap(cor_results$tagfa, text = FALSE)
 ```
@@
 ```{r fig_gee_unadj, fig.cap=fig_gee_unadj}
-gee_unadj_df %>% 
+gee_results$unadj %>% 
     mutate(p.value = ifelse(p.value <= 0.05, 0.04, p.value)) %>% 
@@
 ```{r fig_gee_adj, fig.cap=fig_gee_adj}
-gee_adj_df %>% 
+gee_results$adj %>% 
     mutate(p.value = ifelse(p.value <= 0.05, 0.04, p.value)) %>% 
@@
 ```{r fig_pls, fig.cap=fig_pls, fig.height=8, fig.width=4}
-gridExtra::grid.arrange(
-    plot_pls(pls_homa2) + ggplot2::ggtitle('A: HOMA2-%S'), 
-    plot_pls(pls_isi) + ggplot2::ggtitle('B: ISI'), 
-    ncol = 1)
+p1 <- plot_pls_corrcomps(pls_results$homa2) + ggplot2::ggtitle('A: HOMA2-%S')
+p2 <- plot_pls_corrcomps(pls_results$isi) + ggplot2::ggtitle('B: ISI')
+p1 + p2 + plot_layout(ncol = 1)
 ```

### 2018-02-17 doc/esm.Rmd
jlr manuscript output
caption from manuscript

@@
 ---
-output: fost::diabetologia_manuscript
+output: fost::jlr_manuscript
 ---
@@
 cite_st <- pryr::partial(stab, display = 'cite')
+
+tab_qic <- stab('qic', 'Comparison of GEE model fitness for variable selection using quasi-likelihood information criteria.')
+tab_tagfa <- stab('tagfa', 'Concentration (nmol/mL) and relative percent (mol%) values of triacylglycerol fatty acids in PROMISE participants at the baseline visit (2004-2006).')
+tab_gee_unadj <- stab('gee_unadj', 'Raw estimates and confidence interval values for *time*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+Estimates represent a percent difference in the outcome per SD increase in the fatty acid.
+P-values were adjusted for the BH false discovery rate, with an asterisk (*) denoting a significant (p<0.05) association.')
+tab_gee_adj <- stab('gee_adj', 'Raw estimates and confidence interval values for *fully*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+Variables controlled for were follow-up time, waist circumference, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
+Estimates represent a percent difference in the outcome per SD increase in the fatty acid.
+P-values were adjusted for the BH false discovery rate, with an asterisk (*) denoting a significant (p<0.05) association.')
 

### 2018-02-17 doc/esm.Rmd
ref to results dataframe

@@
 ```{r table_tagfa}
-table_tagfa(caption = tab_tagfa)
+table_tagfa(project_data, caption = tab_tagfa)
 ```
@@
 ```{r table_gee}
-table_gee_main(gee_unadj_df, caption = tab_gee_unadj)
-table_gee_main(gee_adj_df, caption = tab_gee_adj)
+table_gee_main(gee_results$unadj, caption = tab_gee_unadj)
+table_gee_main(gee_results$adj, caption = tab_gee_adj)
 ```

### 2018-02-17 doc/esm.Rmd
fix plots in esm

@@
     analyze_gee(covars = covariates[which(covariates != 'Waist')]) %>% 
-    plot_gee_main()
+    mutate(p.value = ifelse(p.value <= 0.05, 0.04, p.value)) %>% 
+    plot_gee_main() +
+    ggplot2::theme(legend.position = "none")
 ```
@@
 ```
+
+```{r, eval=FALSE}
+pls_results$homa2 %>% 
+    plot_pls_scores() +
+    theme_classic() +
+    theme(strip.background = element_blank())
+```

### 2018-02-18 doc/manuscript.Rmd
change headings to fit JLR

@@
 
-Corresponding author:
+Corresponding author contact information:
 
@@
 
-Disclaimers:
+Running title (max 60 characters):
 
-Funding support:
-
-Word count: ~4000
+Abbreviations:
 
@@
 
-Word count: 249 / 250
+Word count: 249 / 200
 
-**Objective**: 
+**Aims/hypothesis:**
 Although elevated serum triacylglyceride (TAG) is a well-described risk factor for
@@
 
-**Research Design and Methods**:
+**Methods**:
 We used longitudinal data (3 visits over 6 years) from the Prospective
@@
 
-# Background
+# Keywords
+<!-- up to 10. Five from list, five free-form. Should not appear in title or
+running title -->
+
+# Introduction
 
@@
 
-# Subjects and Methods
+# Materials and Methods
 
@@
 
-# Acknowledgements
+# Acknowledgements/grant support
 

### 2018-02-18 doc/esm.Rmd
removed most of esm material, included ref to NEFA paper

@@
 
-# Supplemental Methods
-
-## Equations for outcomes
-
-The equations for calculating the outcome variables are listed below. HOMA2-%S was 
-calculated using the Excel found at https://www.dtu.ox.ac.uk/homacalculator/.
-
-```{r}
-cat(PROMISE.methods::insert_formula("ISI"))
-```
-
-```{r}
-cat(PROMISE.methods::insert_formula("HOMA"))
-```
-
-```{r}
-cat(PROMISE.methods::insert_formula("IGIIR"))
-```
-
-```{r}
-cat(PROMISE.methods::insert_formula("ISSI2"))
-```
-
-Area under the curve (AUC) for both glucose (G) and insulin (I) was computed
-using the trapezoidal rule. ISI~OGTT~ (ISI) uses all the glucose and insulin
-values from the oral glucose tolerance (the overall mean).
-
-## Statistical methods description
-
-These methods are also described in our article on the NEFA fraction
-[@Johnston2017a].
-
-Generalised estimating equations (GEE) is a technique similar to mixed effects 
-modelling, except it calculates the marginal population estimates compared to the
-subject-specific estimates in mixed effects models. GEE is well suited to and
-commonly used in longitudinal cohort studies, especially given its capacity to
-handle missed visits and for the correlation of variables across time. 
-GEE is able to handle the correlation of longitudinal data due to the
-estimating of a working correlation matrix, which was chosen based on
-quasi-likelihood information criteria (QIC; more detail on QIC below). The auto-regressive of order 1
-(AR1) matrix was chosen for the GEE models as it had the best model fit accessed
-using QIC, though other matrices (eg. exchangeable) had similar fit (data not
-shown). While the choice of the correlation matrix is important, it should be noted
-that even with misspecification GEE still computes robust estimates and standard errors.
-One main advantage of using methods such as GEE is the inclusion of the
-working correlation matrix that takes into account the inherent covariance
-between values measured at different time points when computing model estimates
-and standard errors. AR1 is the most appropriate correlation matrix for
-longitudinal data since the correlation matrix assumes lower intercorrelation of
-values within an individual as time passes. For example, BMI measured at the
-baseline visit would be more highly correlated to BMI measured at the 3rd year
-visit than compared to the 6th year visit.
-
-For the covariates, they were chosen based on previous literature, from directed
-acyclic graph (DAG) recommendations, and from QIC. The DAG recommendations were
-obtained from using the DAGitty software, http://dagitty.net/. DAG is a
-technique that allows for identifying potential confounding causal pathways and
-to limit potential bias. DAGs are composed on nodes (variables) connected by
-edges (arrows; causal pathway between two variables). Using DAGs to identify
-variables to adjust for is done algorithmically, as implemented in the DAGitty
-software. DAGitty calculates possible sets of variables to include in a model to
-adjust for potential confounding in the causal network. Use of DAGs in this
-manner can identify potential sources of bias, such as from inclusion of
-colliding variables. A collider variable is a variable that has two (or more)
-'parent' nodes/variables (connected causally by DAG edges pointing into the
-'child' node, which is the collider). The collider variable and parent variables
-are directed linked to the exposure and the outcome. Including all variables
-involved with the collider could introduce bias as it does not generate an
-unconditional association between all involved variables in the model.
-
-QIC is a technique that balances the goodness of fit of the model with its
-relative complexity (i.e. number of covariates) against other models being
-compared. QIC, more so then the more common AIC, penalises the inclusion of more
-variables in the model. This technique is used to compare multiple models to
-identify the model with the highest fit and with the lowest complexity (variables),
-*of the models compared*. QIC is similar to AIC, except AIC is based on maximum
-likelihood of models such as mixed effects models. Since GEE does not use
-maximum likelihood, AIC can not be used. QIC and other information
-criterion methods are used in conjunction with other techniques such as
-literature and DAG based. The final GEE model selected as best fitting was
-negligibly different when using insulin sensitivity or beta-cell function as an
-outcome and as such the same covariates were chosen for models with each of the
-outcomes. The covariates medication use, WC, ALT, and MET-h/week would be considered
-classical confounders while TAG would be considered a mediating variable.
-
-Partial least squares (PLS) regression is a technique similar to principal
-component analysis (PCA) that is designed to extract meaningful information from
-multivariate, high dimensionality data (e.g. as in metabolomic and other -omic type 
-analyses). These types of methods try to extract as much of the variance in the 
-data into a smaller number of components or factors. The difference between PCA 
-and PLS is that PLS is a supervised method, while PCA is not. This means that
-PLS uses a response variable(s), i.e. an outcome or **Y**, to 'supervise' (or
-constrain) the variation inherent in the predictors (**X**) while PCA only
-describes the variation inherent in **X**. Because of this, PLS can be better at
-predicting the contributions of predictor variables against an outcome variable.
-In this analysis, PLS was computed on the NEFA as a proportion due to the
-inherent interdependence of the values (all must equal 100%). This 
-interdependence allows PLS to identify patterns in the composition, since
-the concentration data is all highly positively correlated with each other.
-
-We used internal 10-fold cross-validation to determine which components to
-extract from the PLS analysis. Based on the internal cross-validation, the first
-two components gave the highest amount of explained variance (data not shown),
-of which we decided to use these two components in the final results.
-
-The predictive capability of the PLS models were tested using cross-validation. 
-The data set was randomly split in half into a training set and a testing set. 
-After specifying the model on the training set, results were compared using the 
-testing set to determine whether how predictive the model was given a new
-dataset. Final results shown in the figures use the full dataset (rather than a
-training or testing set). No prevalent diabetes cases were included in the PLS 
-analysis.
-
-## Supplemental Tables
+# Supplemental Tables
 
@@
 
-## Supplemental Figures
+# Supplemental Figures
 

### 2018-02-18 doc/esm.Rmd
tag -> tg

@@
 tab_tagfa <- stab('tagfa', 'Concentration (nmol/mL) and relative percent (mol%) values of triacylglycerol fatty acids in PROMISE participants at the baseline visit (2004-2006).')
-tab_gee_unadj <- stab('gee_unadj', 'Raw estimates and confidence interval values for *time*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+tab_gee_unadj <- stab('gee_unadj', 'Raw estimates and confidence interval values for *time*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
 Estimates represent a percent difference in the outcome per SD increase in the fatty acid.
 P-values were adjusted for the BH false discovery rate, with an asterisk (*) denoting a significant (p<0.05) association.')
-tab_gee_adj <- stab('gee_adj', 'Raw estimates and confidence interval values for *fully*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+tab_gee_adj <- stab('gee_adj', 'Raw estimates and confidence interval values for *fully*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
 Variables controlled for were follow-up time, waist circumference, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
@@
 fig_dag_bcf <- sfig('dag_bcf', 'Directed acyclic graphic output from the DAGitty online software for beta-cell function.')
-fig_gee_adj_nowaist <- sfig('gee_adj_nowaist', 'Fully-adjusted (without waist size) GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+fig_gee_adj_nowaist <- sfig('gee_adj_nowaist', 'Fully-adjusted (without waist size) GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
 Variables controlled for were follow-up time, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.

### 2018-02-18 doc/manuscript.Rmd
removed most of esm material, included ref to NEFA paper
tag -> tg

@@
 
-Running title (max 60 characters):
+Running title (54 / 60 characters):
+Triacylglyceride fatty acids and diabetes pathogenesis
 
 Abbreviations:
+
+- TG
+- TGFA
+- IS
+- PROMISE
+- ISI
+-
 
@@
 **Aims/hypothesis:**
-Although elevated serum triacylglyceride (TAG) is a well-described risk factor for
-type 2 diabetes (T2DM), few data are available regarding the role of the specific fatty
-acid (FA) composition within serum TAG (TAGFA) in the pathogenesis of T2DM. Our aim, therefore, was to
-examine longitudinal associations of TAGFA with insulin sensitivity (IS) and
+Although elevated serum triacylglyceride (TG) is a well-described risk factor for
+type 2 diabetes (T2D), few data are available regarding the role of the specific fatty
+acid (FA) composition within serum TG (TGFA) in the pathogenesis of T2DM. Our aim, therefore, was to
+examine longitudinal associations of TGFA with insulin sensitivity (IS) and
 beta-cell function.
@@
 and Insulin Secretion-Sensitivity Index-2 (ISSI-2). Gas chromatography
-quantified TAGFA composition. Generalized estimating equations (GEE) adjusted for
+quantified TGFA composition. Generalized estimating equations (GEE) adjusted for
 confounders and partial least squares (PLS) were used for the analysis.
@@
 The metabolic outcome variables declined by 14% to 27% over the 6-years. In the adjusted
-GEE models, four TAGFA (14:0, 16:0, 14:1n-7, 16:1n-7 as mol%) had
+GEE models, four TGFA (14:0, 16:0, 14:1n-7, 16:1n-7 as mol%) had
 strong negative associations with IS while others (e.g. 18:1n-7, 18:1n-9, 20:2n-6, 
@@
 beta-cell function, except for 16:0 (negative) and 18:1n-7 (positive). PLS
-analysis indicated four TAGFA (14:0, 16:0, 14:1n-7, 16:1n-7) that are markers of de novo lipogenesis (DNL)
+analysis indicated four TGFA (14:0, 16:0, 14:1n-7, 16:1n-7) that are markers of de novo lipogenesis (DNL)
 clustered together and strongly predicted
-lower IS. These four TAGFA also correlated highly (r>0.4) with clinically
-measured TAG.
+lower IS. These four TGFA also correlated highly (r>0.4) with clinically
+measured TG.
 
 **Conclusions**:
-We found that higher proportions of a cluster of four DNL TAGFA 
+We found that higher proportions of a cluster of four DNL TGFA 
 strongly predicted lower IS as well as hypertriglyceridemia.
@@
 abdominal obesity [@Lemieux2000a].
-Circulating triacylglceride (TAG) concentration is commonly measured during
+Circulating triacylglceride (TG) concentration is commonly measured during
 routine clinical assessment using enzymatic methods.
-However, clinically measured TAG is limited as it
-represents the full fatty acid spectrum within the TAG fraction as a summary
+However, clinically measured TG is limited as it
+represents the full fatty acid spectrum within the TG fraction as a summary
 measure. There is increasing appreciation for the importance of specific fatty 
 acid composition profiles in different plasma fractions on various health
-outcomes [@Forouhi2014a; @Ma2015a; @Johnston2016a], however there are relatively
-few studies that have explored the impact of the fatty acid composition in the TAG fraction [@Rhee2011a; @Lankinen2015a].
+outcomes [@Forouhi2014a; @Ma2015a; @Johnston2016a; @Johnston2017b], however there are relatively
+few studies that have explored the impact of the fatty acid composition in the TG fraction [@Rhee2011a; @Lankinen2015a].
 
-The interaction between TAG and insulin sensitivity is complex and involves 
+The interaction between TG and insulin sensitivity is complex and involves 
 components of a feedback system [@Verges2015a]. Greater resistance to insulin in
-both the liver and muscle may result in greater production of TAG and secretion
-of lipoproteins that transport TAG [@Yu2012a]. Likewise, greater TAG may
+both the liver and muscle may result in greater production of TG and secretion
+of lipoproteins that transport TG [@Yu2012a]. Likewise, greater TG may
 contribute to metabolic dysfunction and lipotoxicity in various tissues,
@@
 studies [@Rhee2011a; @Lankinen2015a] have examined the relationship between
-TAG and its composition with the pathophysiological factors underlying diabetes, 
+TG and its composition with the pathophysiological factors underlying diabetes, 
 particularly beta-cell function. Our objective was to examine the 
-longitudinal role of the specific composition of the serum TAG fraction on 
+longitudinal role of the specific composition of the serum TG fraction on 
 OGTT-derived measures of insulin sensitivity and beta-cell function compared to 
-clinically measured TAG in a Canadian population at risk for diabetes.
+clinically measured TG in a Canadian population at risk for diabetes.
 
@@
 Darker orange represents a positive correlation; darker blue represents a negative correlation.')
-fig_gee_unadj <- fig('gee_unadj', 'Time-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+fig_gee_unadj <- fig('gee_unadj', 'Time-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
 X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
 P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
-fig_gee_adj <- fig('gee_adj', 'Fully-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TAG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+fig_gee_adj <- fig('gee_adj', 'Fully-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
 Variables controlled for were follow-up time, waist circumference, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
@@
 Alanine aminotransferase (ALT) was measured using standard laboratory 
-procedures. Cholesterol, HDL, and clinically-measured TAG 
+procedures. Cholesterol, HDL, and clinically-measured TG 
 were measured using Roche Modular's enzymatic colorimetric tests (Mississauga, 
@@
 
-TAGFA composition was quantified using stored fasting serum samples from the 
+TGFA composition was quantified using stored fasting serum samples from the 
 baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
@@
 [@Folch1957a]. Each serum lipid fraction (non-esterified fatty acids (NEFA), 
-cholesteryl ester, phospholipid, and TAG) was isolated using thin layer 
+cholesteryl ester, phospholipid, and TG) was isolated using thin layer 
 chromatography. Fatty acid methyl esters were separated and quantified using a 
@@
 chromatography peak areas to that of the internal standards [@Nishi2014a]. There
-were 22 fatty acids measured in the TAGFA fraction. 
+were 22 fatty acids measured in the TGFA fraction. 
 Findings
@@
 ISSI-2, which were log-transformed for the statistical modeling. The primary
-predictor variables for this analysis were 22 individual TAGFA included as either mole percent
+predictor variables for this analysis were 22 individual TGFA included as either mole percent
 (mol%) of the total fraction or as a concentration (nmol/mL). Clinically-measured
-TAG was also included as a primary predictor to allow us to test the hypothesis
-that specific TAGFA better predicted outcomes compared to clinical TAG. 
+TG was also included as a primary predictor to allow us to test the hypothesis
+that specific TGFA better predicted outcomes compared to clinical TG. 
 Pearson correlation coefficients were computed to assess the relationships of
-individual TAGFA with other continuous variables. Correlations were also
-computed and hierarchical clustering analysis was conducted for TAGFA against each other.
+individual TGFA with other continuous variables. Correlations were also
+computed and hierarchical clustering analysis was conducted for TGFA against each other.
 
@@
 GEE model (M6) was adjusted for time, waist circumference, baseline age, ethnicity,
-sex, ALT, MET, and total NEFA. The TAGFA, total NEFA, sex, ethnicity, and
+sex, ALT, MET, and total NEFA. The TGFA, total NEFA, sex, ethnicity, and
 baseline age were classified as *time-independent* (held constant) as they were
@@
 limited in that it cannot analyze the inherent multivariate nature of the 
-composition of the TAGFA fraction. Therefore, to confirm the GEE results in a multivariate 
-environment (i.e. all TAGFA analyzed collectively), partial least squares regression (PLS) was
-used to identify the patterns of TAGFA composition against insulin sensitivity
+composition of the TGFA fraction. Therefore, to confirm the GEE results in a multivariate 
+environment (i.e. all TGFA analyzed collectively), partial least squares regression (PLS) was
+used to identify the patterns of TGFA composition against insulin sensitivity
 and beta-cell function as outcome variables. For a detailed explanation of PLS
@@
 response variable(s) (i.e. the outcome variables). How accurately the clusters
-within the TAGFA composition predict metabolic function is determined by using
+within the TGFA composition predict metabolic function is determined by using
 cross-validation on the PLS models.
 
+A more detailed explanation of these statistical techniques and on the analysis
+process can be found in the supplemental methods for our paper in the NEFA
+fraction [@Johnston2017b].
 All analyses were performed using R `r getRversion()` [@Rbase], along with the R
@@
 
-`r cite_f('tagfa')` shows the composition of each FA in the TAG fraction (see 
+`r cite_f('tagfa')` shows the composition of each FA in the TG fraction (see 
 `r cite_st('tagfa')` for a tabular presentation of the values). 
-Three TAGFA contributed `r format_rounding(sum(tagfa$pct))`% to the total 
-TAG concentration: `r tagfa$c[1]`; `r tagfa$c[2]`; and, `r tagfa$c[3]`.
-`r cite_f('heatmap')` shows a heatmap of the correlation of individual TAGFA as
+Three TGFA contributed `r aide::format_round(sum(tagfa$pct))`% to the total 
+TG concentration: `r tagfa$c[1]`; `r tagfa$c[2]`; and, `r tagfa$c[3]`.
+`r cite_f('heatmap')` shows a heatmap of the correlation of individual TGFA as
 concentrations with the outcome variables and several basic characteristics.
-As expected, nearly all TAGFA had very strong positive correlations (r=
-`r fa_cor$rng$TAG[2]`) with clinically-measured TAG and moderate positive 
+As expected, nearly all TGFA had very strong positive correlations (r=
+`r fa_cor$rng$TG[2]`) with clinically-measured TG and moderate positive 
 correlations with WC (r=`r fa_cor$rng$Waist[2]`). There were also moderate
@@
 for the beta-cell function measures (all r<0.30). For correlations
-of individual TAGFA using mol% with the basic participant characteristics, as shown in 
+of individual TGFA using mol% with the basic participant characteristics, as shown in 
 `r cite_f('heatmap_mol')`, differences in correlations between fatty acids were
 most evident for 14:0, 14:1n-7, 16:0, and 16:1n-7 that had a moderate positive
-correlation with clinical TAG (r=`r fa_cor_mol$rng$TAG[1]`) while all other fatty
-acids had a negative association (r=`r fa_cor_mol$rng$TAG[2]`). In particular,
-those fatty acids with the negative associations with clinical TAG were all the
+correlation with clinical TG (r=`r fa_cor_mol$rng$TG[1]`) while all other fatty
+acids had a negative association (r=`r fa_cor_mol$rng$TG[2]`). In particular,
+those fatty acids with the negative associations with clinical TG were all the
 very long chain polyunsaturated fatty acids (e.g. 20:4n-6, 20:5n-3).
@@
 
-In analyses using concentration values, nearly all TAGFA had a strong
+In analyses using concentration values, nearly all TGFA had a strong
 negative association on HOMA2-%S and ISI (estimates of percent different ranging from
-`r est1$outcome_b$rng[3]` and `r est1$outcome_b$rng[7]`, respectively), and a
+`r gmest1$outcome_b$rng[3]` and `r gmest1$outcome_b$rng[7]`, respectively), and a
 few had strong negative associations with IGI/IR and ISSI-2 (estimates ranging 
-from `r est1$outcome_b$rng[11]` and `r est1$outcome_b$rng[15]`, respectively).
-In analyses using TAGFA mol% values, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7) had negative
-associations with HOMA2-%S and ISI (between `r est1$outcome_b$rng[1]` and 
-`r est1$outcome_b$rng[5]`, respectively, lower insulin sensitivity for every SD
-increase in the TAGFA), while several more TAGFA had positive associations with
+from `r gmest1$outcome_b$rng[11]` and `r gmest1$outcome_b$rng[15]`, respectively).
+In analyses using TGFA mol% values, four TGFA (14:0, 16:0, 14:1n-7, and 16:1n-7) had negative
+associations with HOMA2-%S and ISI (between `r gmest1$outcome_b$rng[1]` and 
+`r gmest1$outcome_b$rng[5]`, respectively, lower insulin sensitivity for every SD
+increase in the TGFA), while several more TGFA had positive associations with
 HOMA2-%S and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9, 18:2n-6, 20:2n-6, 20:4n-6,
-and 22:5n-3) predicting between `r est1$outcome_b$rng[2]` and 
-`r est1$outcome_b$rng[6]`%, respectively, higher insulin sensitivity for every
-SD increase in the TAGFA. One TAGFA, 20:2n-6, had a very strong positive
-association with the insulin sensitivity measures, with a `r est1$b_ci$est[1]`
-to `r est1$b_ci$est[37]`% higher insulin sensitivity for every SD increase. 
-Both clinically-measured TAG and total TAGFA concentration had very strong
+and 22:5n-3) predicting between `r gmest1$outcome_b$rng[2]` and 
+`r gmest1$outcome_b$rng[6]`%, respectively, higher insulin sensitivity for every
+SD increase in the TGFA. One TGFA, 20:2n-6, had a very strong positive
+association with the insulin sensitivity measures, with a `r gmest1$b_ci$est[1]`
+to `r gmest1$b_ci$est[37]`% higher insulin sensitivity for every SD increase. 
+Both clinically-measured TG and total TGFA concentration had very strong
 negative associations with all outcome variables.
@@
 shown). There were no significant interactions by sex or ethnicity for any of
-the TAGFA (data not shown). Results of the sensitivity analyses identifying
+the TGFA (data not shown). Results of the sensitivity analyses identifying
 waist circumference as the covariate that attenuated the beta-cell function
@@
 
-## Clustering of TAGFA by metabolic measures
+## Clustering of TGFA by metabolic measures
 
@@
 clustering of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7 as mol%
-(`r cite_f('pls')`). These TAGFA loaded strongly and negatively on HOMA2-%S and ISI
-in the first component, suggesting this cluster of TAGFA tracks
-together with lower insulin sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3,
+(`r cite_f('pls')`). These TGFA loaded strongly and negatively on HOMA2-%S and ISI
+in the first component, suggesting this cluster of TGFA tracks
+together with lower insulin sensitivity. The TGFA 20:2n-6, 20:5n-3, 22:5n-3,
 and 22:6n-3 loaded positively on both insulin sensitivity measures. No other
-TAGFA loaded strongly. In the second component, 18:1n-9 and 18:1n-7 loaded postively 
+TGFA loaded strongly. In the second component, 18:1n-9 and 18:1n-7 loaded postively 
 but not strongly while 20:5n-3 and 22:6n-3 loaded strongly and negatively with both
@@
 p`r cor_pls_igi$p`) between the predicted values and the observed values 
-(r=`r cor_pls_igi$r` to `r cor_pls_issi2$r`), suggesting that TAGFA composition
+(r=`r cor_pls_igi$r` to `r cor_pls_issi2$r`), suggesting that TGFA composition
 poorly predicts beta-cell function.
@@
 In the present study, we found that in a Canadian cohort at risk for diabetes,
-several specific TAGFA and groups of TAGFA were strongly associated
+several specific TGFA and groups of TGFA were strongly associated
 with insulin sensitivity and moderately associated with beta-cell function. In 
-particular, the TAGFA myristic acid (14:0), 7-tetradecenoic acid 
+particular, the TGFA myristic acid (14:0), 7-tetradecenoic acid 
 (14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and
-negatively predicted lower insulin sensitivity. While most TAGFA were not
+negatively predicted lower insulin sensitivity. While most TGFA were not
 associated with beta-cell function, two fatty acids, palmitic acid (16:0) and
@@
 respectively, with measures of beta-cell function. Using PLS, we also found that four
-TAGFA (14:0, 14:1n-7, 16:0, 16:1n-7) clustered together, and that this cluster
+TGFA (14:0, 14:1n-7, 16:0, 16:1n-7) clustered together, and that this cluster
 strongly predicted lower insulin sensitivity. These four fatty acids are
@@
 To our knowledge, no longitudinal study to date has examined the role of the
-composition of the TAGFA fraction on detailed OGTT-derived metabolic measures 
-Two large prospective studies have been published that similarly examined TAGFA 
+composition of the TGFA fraction on detailed OGTT-derived metabolic measures 
+Two large prospective studies have been published that similarly examined TGFA 
 composition and diabetes outcomes. Rhee *et al* presented a nested case-control
 analysis (n=189 cases and n=189 controls) within the Framingham offspring cohort
-[@Rhee2011a], which found that subjects with a TAGFA composition characterized
+[@Rhee2011a], which found that subjects with a TGFA composition characterized
 by a lower carbon chain and fewer double bonds (e.g. 14:0, 16:0) had a higher
 risk for diabetes after 12-years while those with a profile characterized by
-higher carbon chain and more double bond TAGFA had a lower risk for diabetes.
-A similar pattern of TAGFA was also associated with HOMA-IR cross-sectionally
+higher carbon chain and more double bond TGFA had a lower risk for diabetes.
+A similar pattern of TGFA was also associated with HOMA-IR cross-sectionally
 at the baseline visit. In addition, Lankinen *et al* reported on
 a prospective cohort of males in Finland [@Lankinen2015a], for which 
-TAGFA data were available for 831 participants after 6-years of follow-up. In
+TGFA data were available for 831 participants after 6-years of follow-up. In
 their cohort, OGTT data was only available at the 6-year visit.
@@
 of metabolic function and as well as multivariate statistical approaches that
-allowed us to identify clusters of TAGFA. In another study of a much
+allowed us to identify clusters of TGFA. In another study of a much
 smaller (n=16) mostly female group [@Kotronen2009a], the authors report a
-positive correlation between total esterified (of which TAG make up the
+positive correlation between total esterified (of which TG make up the
 majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, findings which were largely 
@@
 particular, DNL from refined or simple carbohydrate sources increases the 14 to
-16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
+16 chain fatty acids as well as the 18 chain TGFA. Several studies have shown a
 link between higher estimated DNL and an increased risk for metabolic 
 dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. Our study 
-extends these findings by showing that TAGFA with 14 to 16 carbons clustered
+extends these findings by showing that TGFA with 14 to 16 carbons clustered
 together and this pattern strongly predicted lower insulin sensitivity. While
@@
 another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
-to determine its effect on DNL and TAGFA composition [@Wilke2009a]. The authors
+to determine its effect on DNL and TGFA composition [@Wilke2009a]. The authors
 reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
@@
 lead to greater DNL in an attempt to control blood glucose, thus increasing hepatic
-fat stores and consequently increasing the amount of TAGFA in circulation
+fat stores and consequently increasing the amount of TGFA in circulation
 [@Chehade2013a; @Verges2015a]. In fact, we found in our study that
-higher proportions of these four DNL TAGFA also highly correlated with a higher
-concentration of clinical TAG, reinforcing this pathway between carbohydrate
-intake, DNL, and circulating TAG. The higher concentration of circulating 14 and
+higher proportions of these four DNL TGFA also highly correlated with a higher
+concentration of clinical TG, reinforcing this pathway between carbohydrate
+intake, DNL, and circulating TG. The higher concentration of circulating 14 and
 16 carbon fatty acids may then expose tissues to greater lipotoxicity, for
@@
 
-The direction of association between TAGFA and insulin sensitivity is unclear 
+The direction of association between TGFA and insulin sensitivity is unclear 
 from previous cross-sectional studies due to the physiological feedback mechanisms 
@@
 encourage greater DNL to handle the higher blood glucose. To illustrate this, in
-a weight loss intervention trial (n=19 with TAGFA data), participants who lost
-weight over 33 weeks showed higher insulin sensitivity and lower TAGFA
+a weight loss intervention trial (n=19 with TGFA data), participants who lost
+weight over 33 weeks showed higher insulin sensitivity and lower TGFA
 composition indicative of lower DNL (less 14:0, 14:1, 16:0, 16:1, etc)
 [@Schwab2008a]. Given the complex biological mechanisms and feedback loops
-involved, disentangling whether insulin sensitivity influences TAGFA more
-strongly than TAGFA influencing insulin sensitivity will require more
+involved, disentangling whether insulin sensitivity influences TGFA more
+strongly than TGFA influencing insulin sensitivity will require more
 complicated research designs and analyses. While further research will need to
 confirm this, the lack of a time interaction we observed in addition to the
-timing of our measures likely suggests that it is the TAGFA predicting insulin
+timing of our measures likely suggests that it is the TGFA predicting insulin
 resistance.
@@
 that cis-vaccenic acid strongly predicted higher insulin sensitivity and
-beta-cell function. Likewise, in the TAGFA fraction we found a similar
+beta-cell function. Likewise, in the TGFA fraction we found a similar
 association with cis-vaccenic acid. {{we can discuss to include more}}
@@
 NEFA may be a mediating factor, we found that adjustment for it did not
-{{confirm}} attenuate this association since TAGFA composition is also dictated
+{{confirm}} attenuate this association since TGFA composition is also dictated
 by other sources for DNL (e.g. recent meal, DNL from glucose stores, etc.).
 
-While the NEFA fraction is much smaller than the TAG fraction, even though it
-contributes substantially to the TAG fraction, the VLDL-TAG is not cleared
-(lipolyzed) as much as chylomicron (which also contains a large amount of TAG)
+While the NEFA fraction is much smaller than the TG fraction, even though it
+contributes substantially to the TG fraction, the VLDL-TG is not cleared
+(lipolyzed) as much as chylomicron (which also contains a large amount of TG)
 and so has a higher half-life than NEFA, remaining in circulation longer {{hence
-why there is bigger TAG pool vs NEFA}} {{cite?}}.
+why there is bigger TG pool vs NEFA}} {{cite?}}.
 
@@
 
-- Zulyniak2012a (only vaccenic acid in TAG) (n=20, TLC-GC, HOMA-IR/HOMA-B,
+- Zulyniak2012a (only vaccenic acid in TG) (n=20, TLC-GC, HOMA-IR/HOMA-B,
 two groups: normogly and hypergly. results in general showed no diff between
@@
 
-- Or that post-prandial TAG is not suppressed in IR Sondergaard2012a
+- Or that post-prandial TG is not suppressed in IR Sondergaard2012a
 - {{ confirm }} greater flux of saturated fatty acids into tissues may encourage
@@
 <!--
-TAGFA can fluctuate substantially throughout the day and to control for this all
+TGFA can fluctuate substantially throughout the day and to control for this all
 PROMISE participants came for their clinic visits in the morning in a fasted
-state. Some of the inherent variability in TAGFA fluctuation can be managed by
+state. Some of the inherent variability in TGFA fluctuation can be managed by
 standardizing the data collection time, though there may still be some
 {{residual error? random error?}}.
-[@Rhee2011a] fasting TAG status better as discriminating between DM case vs
-control than 2hr-OGTT TAG
+[@Rhee2011a] fasting TG status better as discriminating between DM case vs
+control than 2hr-OGTT TG
 -->
 
-TAGFA were only quantified at the baseline visit and as such we cannot
-investigate whether there are concomitant changes in TAGFA and the metabolic
+TGFA were only quantified at the baseline visit and as such we cannot
+investigate whether there are concomitant changes in TGFA and the metabolic
 measures over time. However, to optimally use GEE to analyze the data and for
-interpretation, we used the model to infer that a given value of TAGFA could
+interpretation, we used the model to infer that a given value of TGFA could
 predict values of insulin sensitivity or beta-cell function over a 6 year
@@
 
-In conclusion, we found that a TAGFA composition indicative of higher DNL
+In conclusion, we found that a TGFA composition indicative of higher DNL
 (containing higher 14:0, 14:1n-7, 16:0, and 16:1n-7) associated strongly with lower

### 2018-02-18 R/renaming.R

@@ -6,12 +6,12 @@ renaming_table_rows <- function(x) {
         gsub('ISSI2', 'ISSI-2', .) %>%
         gsub('HOMA2IR', 'HOMA2-IR', .) %>%
         gsub('HOMA2_S', 'HOMA2-%S', .) %>%
-        gsub('TAG', 'TAG (mmol/L)', .) %>%
+        gsub('TAG', 'TG (mmol/L)', .) %>%
         gsub('Chol', 'Chol (mmol/L)', .) %>%
         gsub('LDL', 'LDL (mmol/L)', .) %>%
         gsub('ALT', 'ALT (U/L)', .) %>%
         gsub('HDL', 'HDL (mmol/L)', .) %>%
-        gsub('BaseTotalTG', 'TAGFA (nmol/mL)', .) %>%
+        gsub('BaseTotalTG', 'TGFA (nmol/mL)', .) %>%
         gsub('BaseTotalNE', 'NEFA (nmol/mL)', .) %>%
         gsub('Age', 'Age (yrs)', .) %>%
         gsub('BMI', 'BMI (kg/m^2^)', .) %>%
@@ -35,7 +35,7 @@ renaming_fats <- function(x) {
     x %>%
         stringr::str_replace('^pct_', '') %>%
         stringr::str_replace('TotalTG', 'Total') %>%
-        stringr::str_replace('BaseTAG', 'Clinical TAG') %>%
+        stringr::str_replace('BaseTAG', 'Clinical TG') %>%
         PROMISE.misc::renaming_fa(keep.fraction = FALSE)
 }
 

### 2018-02-18 doc/manuscript.Rmd
is -> was

@@
 the Canadian Institutes for Health Research, and the University of
-Toronto Banting and Best Diabetes Centre; LWJ is supported by a CDA
+Toronto Banting and Best Diabetes Centre; LWJ was supported by a CDA
 Doctoral Student Research Award; RR is supported by a Heart and Stroke

### 2018-02-18 doc/esm.Rmd
load packages
dplyr::

@@
 ```{r setup, include=FALSE}
+library(ggplot2)
+library(patchwork)
 devtools::load_all()
@@
     analyze_gee(covars = covariates[which(covariates != 'Waist')]) %>% 
-    mutate(p.value = ifelse(p.value <= 0.05, 0.04, p.value)) %>% 
+    dplyr::mutate(p.value = ifelse(p.value <= 0.05, 0.04, p.value)) %>% 
     plot_gee_main() +

### 2018-02-18 doc/manuscript.Rmd
fix up front matter, title, authors

@@
     - /home/luke/Documents/bibliography/bib/bib-db/methods.bib
+    - /home/luke/Documents/site/cv/_includes/work.bib
 csl: /home/luke/Documents/bibliography/styles/vancouver.csl
@@
 
-Title: Specific clusters of fatty acids within the serum triacylglyceride fraction 
-associate with the pathogenesis of type 2 diabetes
+Title (109 / 110 characters): 
+Clusters of fatty acids in serum triacylglyceride fraction associate with the
+pathogenesis of type 2 diabetes
 
-Author: Luke W. Johnston, MSc (1); Zhen Liu, PhD (1); Ravi Retnakaran, MD
+Authors: Luke W. Johnston, PhD (1); Zhen Liu, PhD (1); Ravi Retnakaran, MD
 (3,4); Bernard Zinman, MD (3,4); Adria Giacca, PhD (5); Stewart B. Harris, MD (2); 
@@
 
-<!-- During time of study. -->
-Affiliation:
+Institutions:
 
-(1) Department of Nutritional Sciences, University of Toronto, Toronto, Ontario, Canada.
-(2) Centre for Studies in Family Medicine, University of Western Ontario, London, Ontario, Canada.
-(3) Leadership Sinai Centre for Diabetes, Division of Endocrinology, University of Toronto, Toronto, ON, Canada.
-(4) Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto, Ontario, Canada.
+(1) Department of Nutritional Sciences, University of Toronto, Toronto, Ontario,
+Canada.
+(2) Centre for Studies in Family Medicine, University of Western Ontario,
+London, Ontario, Canada.
+(3) Leadership Sinai Centre for Diabetes, Division of Endocrinology, University
+of Toronto, Toronto, ON, Canada.
+(4) Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto,
+Ontario, Canada.
 (5) Department of Physiology, University of Toronto, Toronto, Ontario, Canada.
-(6) Dalla Lana School of Public Health, University of Toronto, Toronto, Ontario, Canada.
+(6) Dalla Lana School of Public Health, University of Toronto, Toronto, Ontario,
+Canada.
 

### 2018-02-18 doc/manuscript.Rmd
abbreviations and fixes to abstract

@@
 
-Abbreviations:
+# Abbreviations:
 
-- TG
-- TGFA
-- IS
-- PROMISE
-- ISI
--
+- TGFA: Triacylglycerol fatty acids
+- IS: Insulin sensitivity
+- PROMISE: Prospective Metabolism and Islet Cell Evaluation cohort
+- ISI: Insulin sensitivity index
+- GEE: Generalized estimating equations
+- PLS: Partial least squares
+- HOMA2-%S: Homeostatic model of assessment 2 - percent sensitivity
+- IGI/IR: Insulinogenic index over homeostatic model of assessment for insulin
+resistance
+- ISSI-2: Insulin Secretion-Sensitivity Index-2
+- DNL: De novo lipogenesis
+- ALT: Alanine aminotransferase
+- NEFA: Non-esterified fatty acids
+- OGTT: Oral glucose tolerance test
+- CV: Coefficient of variation
+- IFG: Impaired fasting glucose
+- IGT: Impaired glucose tolerance
+- MAQ: Modified activity questionnaire
+- WC: Waist circumference
+- MET: Metabolic equivalent of task
+- SD: Standard deviation
 
@@
 
-**Aims/hypothesis:**
-Although elevated serum triacylglyceride (TG) is a well-described risk factor for
-type 2 diabetes (T2D), few data are available regarding the role of the specific fatty
-acid (FA) composition within serum TG (TGFA) in the pathogenesis of T2DM. Our aim, therefore, was to
-examine longitudinal associations of TGFA with insulin sensitivity (IS) and
-beta-cell function.
+**Aims:**
+Our aim was to examine longitudinal associations of triacylglyceride fatty acid
+(TGFA) composition with insulin sensitivity (IS) and beta-cell function.
 
 **Methods**:
-We used longitudinal data (3 visits over 6 years) from the Prospective
-Metabolism and Islet Cell Evaluation (PROMISE) cohort of adults (n=477) who were at-risk
-for diabetes at baseline. Glucose and insulin from an OGTT were used to calculate the outcome variables
-Matsuda index (ISI), HOMA2-%S, Insulinogenic Index over HOMA-IR (IGI/IR),
-and Insulin Secretion-Sensitivity Index-2 (ISSI-2). Gas chromatography
-quantified TGFA composition. Generalized estimating equations (GEE) adjusted for
-confounders and partial least squares (PLS) were used for the analysis.
+Adults at-risk for T2D (n=477) had glucose and insulin measured from a glucose
+challenge at 3 time points over 6 years. The outcome variables Matsuda index
+(ISI), HOMA2-%S, Insulinogenic Index over HOMA-IR (IGI/IR), and Insulin
+Secretion-Sensitivity Index-2 (ISSI-2) were computed from the glucose challenge.
+Gas chromatography quantified TGFA composition from the baseline. We used
+adjusted generalized estimating equations (GEE) models and partial least squares
+(PLS) regression for the analysis.
 
 **Results**: 
-The metabolic outcome variables declined by 14% to 27% over the 6-years. In the adjusted
-GEE models, four TGFA (14:0, 16:0, 14:1n-7, 16:1n-7 as mol%) had
-strong negative associations with IS while others (e.g. 18:1n-7, 18:1n-9, 20:2n-6, 
-20:5n-3) had strong positive associations. Few associations were seen for 
-beta-cell function, except for 16:0 (negative) and 18:1n-7 (positive). PLS
-analysis indicated four TGFA (14:0, 16:0, 14:1n-7, 16:1n-7) that are markers of de novo lipogenesis (DNL)
-clustered together and strongly predicted
-lower IS. These four TGFA also correlated highly (r>0.4) with clinically
+In adjusted GEE models, four TGFA (14:0, 16:0, 14:1n-7, 16:1n-7 as mol%) had
+strong negative associations with IS while others (e.g. 18:1n-7, 18:1n-9,
+20:2n-6, 20:5n-3) had strong positive associations. Few associations were seen
+for beta-cell function, except for 16:0 and 18:1n-7. PLS analysis indicated four
+TGFA (14:0, 16:0, 14:1n-7, 16:1n-7) that clustered together and strongly related
+with lower IS. These four TGFA also correlated highly (r>0.4) with clinically
 measured TG.
@@
 **Conclusions**:
-We found that higher proportions of a cluster of four DNL TGFA 
-strongly predicted lower IS as well as hypertriglyceridemia.
+We found that higher proportions of a cluster of four TGFA strongly related with
+lower IS as well as hypertriglyceridemia. Previous research suggests these four
+FA are also strongly influenced by de novo lipogenesis, particular from
+carbohydrates.
 

### 2018-02-18 doc/manuscript.Rmd
added nefa ref

@@
 were 22 fatty acids measured in the TGFA fraction. 
-Findings
-for other lipid fractions in this cohort are reported separately (see ref
-@Johnston2016a for the analysis of the phospholipid and cholesteryl ester
-fractions). <!--{{put these analyses in discussion?}} -->
-<!--and {{next citation}} for the analysis of the NEFA fraction).-->
+Findings for other lipid fractions in this cohort are reported separately (see
+ref [@Johnston2016a] for the phospholipid and cholesteryl ester fraction and
+[@Johnston2017b] for the NEFA fraction analysis).
 

### 2018-02-18 doc/manuscript.Rmd
abbrev for MET

@@
 the past year. Each reported activity from the MAQ was weighted by its
-metabolic intensity allowing for the estimation of MET-hours per week [@Kriska1990].
+metabolic intensity allowing for the estimation of metabolic equivalents of
+tasks (MET) hours per week [@Kriska1990].
 

### 2018-02-18 doc/manuscript.Rmd
remove some unnecessary words/sentences

@@
 against gold standard measures [@Matthews1985; @Hermans1999a; @Matsuda1999; @Retnakaran2009].
-Specific formulas for each OGTT-derived measure can be found in the 
-Supplemental Material {{insert formula}}.
 
@@
 covariates were scaled (mean-centered and standardized). Given the longitudinal 
-design, an auto-regressive of order 1 (AR1) working correlation matrix was 
+design, an auto-regressive of order 1 working correlation matrix was 
 specified in the GEE model. Covariates to adjust for were selected based on the
 previous literature, from directed acyclic graph [@Greenland1999a]
-recommendations, and from quasi-likelihood information criteria (QIC). The final
+recommendations, and from quasi-likelihood information criteria. The final
 GEE model (M6) was adjusted for time, waist circumference, baseline age, ethnicity,

### 2018-02-18 R/calculate.R
homa -> homa2

@@ -48,7 +48,7 @@ calc_tagfa_percent <- function(data = project_data) {
 #' @export
 calc_outcome_changes <- function(data = project_data) {
     prep.data <- data %>%
-        dplyr::select(f.VN, HOMA, ISI, IGIIR, ISSI2) %>%
+        dplyr::select(f.VN, HOMA2_S, ISI, IGIIR, ISSI2) %>%
         tidyr::gather(Measure, Value,-f.VN) %>%
         stats::na.omit() %>%
         dplyr::group_by(Measure, f.VN) %>%
@@ -70,7 +70,7 @@ calc_outcome_changes <- function(data = project_data) {
 
     pval <- mason::design(data, 'gee') %>%
         mason::add_settings(family = stats::gaussian(), corstr = 'ar1', cluster.id = 'SID') %>%
-        mason::add_variables('yvars', c('linvHOMA', 'lISI', 'lIGIIR', 'lISSI2')) %>%
+        mason::add_variables('yvars', c('lHOMA2_S', 'lISI', 'lIGIIR', 'lISSI2')) %>%
         mason::add_variables('xvars', 'VN') %>%
         mason::construct() %>%
         mason::scrub() %>%

### 2018-02-18 doc/manuscript.Rmd
remove text related to pre-diabetes, num in promise, etc

@@
 - CV: Coefficient of variation
-- IFG: Impaired fasting glucose
-- IGT: Impaired glucose tolerance
 - MAQ: Modified activity questionnaire
@@
 ```{r inline}
-dys <- project_data %>% 
-    prep_dys_data() %>% 
-    calc_convert_dys()
-
 tagfa <- calc_tagfa_percent()
@@
 intra-assay CV of < 1.9%.  All assays were performed at the Banting and Best
-Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose (IFG),
-impaired glucose tolerance (IGT), and diabetes were categorized using the 2006
-WHO criteria [@WHO2006].
+Diabetes Centre Core Lab at Mt Sinai Hospital.
 
@@
 insulin sensitivity and beta-cell function measures had a significant median decline of
-between `r outcome_chg$chg` (p`r outcome_chg$p` from GEE; n=`r outcome_chg$n`). There 
-were `r dys$pre` and `r dys$dm` incident cases of pre-diabetes (IFG and IGT) and
-diabetes, respectively, over the 6-years; these observations were excluded from GEE and PLS
-analyses due to how the outcomes were measured (OGTT was not done on those with diabetes).
+between `r outcome_chg$chg` (p`r outcome_chg$p` from GEE; n=`r outcome_chg$n`).
 

### 2018-02-18 doc/manuscript.Rmd
reformat abstract
fix code from TG -> TAG

@@
 
-Hypertriglyceridemia is a well described metabolic disorder resulting in negative health outcomes [@Chehade2013a].
-It is a risk factor for cardiovascular disease [@DAgostino2004a; @Verges2015a]
-and is associated with other metabolic disorders such as non-alcoholic
-fatty liver disease [@Kawano2013a], the metabolic syndrome [@Alberti2009], and
-abdominal obesity [@Lemieux2000a].
+Hypertriglyceridemia is a well described metabolic disorder resulting in
+negative health outcomes [@Chehade2013a]. It is a risk factor for cardiovascular
+disease [@DAgostino2004a; @Verges2015a] and is associated with other metabolic
+disorders such as non-alcoholic fatty liver disease [@Kawano2013a], the
+metabolic syndrome [@Alberti2009], and abdominal obesity [@Lemieux2000a].
 Circulating triacylglceride (TG) concentration is commonly measured during
-routine clinical assessment using enzymatic methods.
-However, clinically measured TG is limited as it
-represents the full fatty acid spectrum within the TG fraction as a summary
-measure. There is increasing appreciation for the importance of specific fatty 
-acid composition profiles in different plasma fractions on various health
-outcomes [@Forouhi2014a; @Ma2015a; @Johnston2016a; @Johnston2017b], however there are relatively
-few studies that have explored the impact of the fatty acid composition in the TG fraction [@Rhee2011a; @Lankinen2015a].
+routine clinical assessment using enzymatic methods. However, clinically
+measured TG is limited as it represents the full fatty acid spectrum within the
+TG fraction as a summary measure. There is increasing appreciation for the
+importance of specific fatty acid composition profiles in different plasma
+fractions on various health outcomes [@Forouhi2014a; @Ma2015a; @Johnston2016a;
+@Johnston2017b], however there are relatively few studies that have explored the
+impact of the fatty acid composition in the TG fraction [@Rhee2011a;
+@Lankinen2015a].
 
@@
 contribute to metabolic dysfunction and lipotoxicity in various tissues,
-affecting insulin sensitivity, and thus continue the cycle [@Verges2015a]. Given the
-complexity and temporal nature of the relationship, long term studies with
-multiple data collection time points are paramount to better understanding the underlying
-biology and subsequent risk.
+affecting insulin sensitivity, and thus continue the cycle [@Verges2015a]. Given
+the complexity and temporal nature of the relationship, long term studies with
+multiple data collection time points are paramount to better understanding the
+underlying biology and subsequent risk.
 
 While several studies have documented prospective associations of
-hypertriglyceridemia with incident type 2 diabetes 
-[@Chien2008a; @DAgostino2004a; @Schulze2009a], only a limited number of longitudinal
-studies [@Rhee2011a; @Lankinen2015a] have examined the relationship between
-TG and its composition with the pathophysiological factors underlying diabetes, 
-particularly beta-cell function. Our objective was to examine the 
-longitudinal role of the specific composition of the serum TG fraction on 
-OGTT-derived measures of insulin sensitivity and beta-cell function compared to 
-clinically measured TG in a Canadian population at risk for diabetes.
+hypertriglyceridemia with incident type 2 diabetes [@Chien2008a;
+@DAgostino2004a; @Schulze2009a], only a limited number of longitudinal studies
+[@Rhee2011a; @Lankinen2015a] have examined the relationship between TG and its
+composition with the pathophysiological factors underlying diabetes,
+particularly beta-cell function. Our objective was to examine the longitudinal
+role of the specific composition of the serum TG fraction on OGTT-derived
+measures of insulin sensitivity and beta-cell function compared to clinically
+measured TG in a Canadian population at risk for diabetes.
 
@@
 As expected, nearly all TGFA had very strong positive correlations (r=
-`r fa_cor$rng$TG[2]`) with clinically-measured TG and moderate positive 
+`r fa_cor$rng$TAG[2]`) with clinically-measured TG and moderate positive 
 correlations with WC (r=`r fa_cor$rng$Waist[2]`). There were also moderate
@@
 most evident for 14:0, 14:1n-7, 16:0, and 16:1n-7 that had a moderate positive
-correlation with clinical TG (r=`r fa_cor_mol$rng$TG[1]`) while all other fatty
-acids had a negative association (r=`r fa_cor_mol$rng$TG[2]`). In particular,
+correlation with clinical TG (r=`r fa_cor_mol$rng$TAG[1]`) while all other fatty
+acids had a negative association (r=`r fa_cor_mol$rng$TAG[2]`). In particular,
 those fatty acids with the negative associations with clinical TG were all the

### 2018-02-25 R/corr.R
fixes to naming of vars in corr

@@ -25,7 +25,10 @@ analyze_corr <-
         dplyr::mutate_at('Vars2', dplyr::funs(renaming_fats)) %>%
         dplyr::mutate_at("Vars1", function(x)
             gsub('l(ALT|TAG|IGIIR|HOMA2_S|ISI|ISSI2)', '\\1', x) %>%
-                renaming_outcomes()) %>%
+                renaming_outcomes() %>%
+                stringr::str_replace("Waist", "WC") %>%
+                stringr::str_replace("TAG", "TG")
+                ) %>%
         dplyr::mutate(
             order1 = substr(Vars2, nchar(Vars2), nchar(Vars2)),
             order1 = ifelse(order1 == 0, 10, order1),

### 2018-02-25 R/gee.R
tried to fix gee table FA ordering

@@ -209,7 +209,8 @@ table_gee_main <- function(results, caption = NULL, digits = 1) {
         ) %>%
         select(Yterms, Xterms, unit, estimate.ci) %>%
         spread(Yterms, estimate.ci) %>%
-        mutate(Xterms = as.character(Xterms))
+        mutate(Xterms = as.character(Xterms)) %>%
+        order_model_output()
 
     bind_rows(
             tibble::data_frame(Xterms = paste0('**Totals**')),

### 2018-02-25 R/setup.R
output eps and png with dpi 300

@@ -27,7 +27,9 @@ set_options <- function() {
     # For the document (knitr)
     knitr::opts_chunk$set(
         warning = FALSE, message = FALSE, collapse = TRUE,
-        fig.showtext = FALSE, comment = '#>', dev = "pdf"
+        #fig.showtext = FALSE,
+        comment = '#>', dev = c("png", "cairo_ps", "CairoPS"),
+        dpi = 300
     )
 }
 

### 2018-02-25 doc/esm.Rmd
abbrevations

@@
 tab_qic <- stab('qic', 'Comparison of GEE model fitness for variable selection using quasi-likelihood information criteria.')
-tab_tagfa <- stab('tagfa', 'Concentration (nmol/mL) and relative percent (mol%) values of triacylglycerol fatty acids in PROMISE participants at the baseline visit (2004-2006).')
-tab_gee_unadj <- stab('gee_unadj', 'Raw estimates and confidence interval values for *time*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
-Estimates represent a percent difference in the outcome per SD increase in the fatty acid.
+tab_tagfa <- stab('tagfa', 'Concentration (nmol/mL) and relative percent (mol%) values of TGFA in PROMISE participants at the baseline visit (2004-2006).')
+tab_gee_unadj <- stab('gee_unadj', 'Raw estimates and confidence interval values for *time*-adjusted GEE models of the association of the TGFA (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+Estimates represent a percent difference in the outcome per SD increase in the FA.
 P-values were adjusted for the BH false discovery rate, with an asterisk (*) denoting a significant (p<0.05) association.')
-tab_gee_adj <- stab('gee_adj', 'Raw estimates and confidence interval values for *fully*-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
-Variables controlled for were follow-up time, waist circumference, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
-Estimates represent a percent difference in the outcome per SD increase in the fatty acid.
+tab_gee_adj <- stab('gee_adj', 'Raw estimates and confidence interval values for *fully*-adjusted GEE models of the association of the TGFA (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+Variables controlled for were follow-up time, WC, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
+Estimates represent a percent difference in the outcome per SD increase in the FA.
 P-values were adjusted for the BH false discovery rate, with an asterisk (*) denoting a significant (p<0.05) association.')
@@
 fig_dag_bcf <- sfig('dag_bcf', 'Directed acyclic graphic output from the DAGitty online software for beta-cell function.')
-fig_gee_adj_nowaist <- sfig('gee_adj_nowaist', 'Fully-adjusted (without waist size) GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+fig_gee_adj_nowaist <- sfig('gee_adj_nowaist', 'Fully-adjusted (without waist size) GEE models of the association of the TGFA (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
 Variables controlled for were follow-up time, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
-X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
+X-axis values represent a percent difference in the outcome per SD increase in the FA.
 P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
+fig_pls_loadings <- sfig("pls_loadings", "PLS loadings (or weights) for each of the TGFA. A larger loading indicates a higher contribution to the PLS component score.")
 ```
@@
 Given the number of possible combinations of outcome and predictor variables,
-only ISI and ISSI-2 with total triacylglycerol fatty acids (nmol/mL) were used to
+only ISI and ISSI-2 with total TGFA (nmol/mL) were used to
 compare various GEE models and to select a final model.  Baseline age was used
@@
 
-- M0: log(ISSI-2) or log(ISI) = total triacylglycerol fatty acids (nmol/mL) + years from baseline
+- M0: log(ISSI-2) or log(ISI) = total TGFA (nmol/mL) + years from baseline
 - M1: M0 + fatty acid by time interaction
 - M2: M0 + sex + ethnicity + baseline age
-- M3: M2 + waist
+- M3: M2 + WC
 - M4: M3 + ALT
@@
 
-```{r fig_pls_loadings}
+```{r fig_pls_loadings, fig.cap=fig_pls_loadings}
 p1 <- plot_pls_loadings(pls_results$homa2) +

### 2018-02-25 doc/manuscript.Rmd
fixes in prep for submission

@@
 
-Word count: 249 / 200
+Word count: 200 / 200
 
@@
 While several studies have documented prospective associations of
-hypertriglyceridemia with incident type 2 diabetes [@Chien2008a;
+hypertriglyceridemia with incident type 2 diabetes (T2D) [@Chien2008a;
 @DAgostino2004a; @Schulze2009a], only a limited number of longitudinal studies
 [@Rhee2011a; @Lankinen2015a] have examined the relationship between TG and its
-composition with the pathophysiological factors underlying diabetes,
+composition with the pathophysiological factors underlying T2D,
 particularly beta-cell function. Our objective was to examine the longitudinal
@@
 measures of insulin sensitivity and beta-cell function compared to clinically
-measured TG in a Canadian population at risk for diabetes.
+measured TG in a Canadian population at risk for T2D.
 
@@
 
-fig_tagfa <- fig('tagfa', 'Distribution of the composition of triacylglycerol fatty acids in the baseline visit of PROMISE participants (2004-2006).
-Boxplots represent the median and interquartile range of the fatty acid values.')
-fig_heatmap <- fig('heatmap', 'Pearson correlation heatmap of triacylglycerol fatty acids (nmol/mL) with continuous basic and metabolic characteristics of PROMISE participants from the baseline visit (2004-2006).
+fig_tagfa <- fig('tagfa', 'Distribution of the composition of TGFA in the baseline visit of PROMISE participants (2004-2006). Boxplots represent the median and interquartile range of the FA values.')
+fig_heatmap <- fig('heatmap', 'Pearson correlation heatmap of TGFA (nmol/mL) with continuous basic and metabolic characteristics of PROMISE participants from the baseline visit (2004-2006).
 Darker orange represents a positive correlation; darker blue represents a negative correlation.')
-fig_heatmap_mol <- fig('heatmap_mol', 'Pearson correlation heatmap of triacylglycerol fatty acids (mol%) with continuous basic and metabolic characteristics of PROMISE participants from the baseline visit (2004-2006).
+fig_heatmap_mol <- fig('heatmap_mol', 'Pearson correlation heatmap of TGFA (mol%) with continuous basic and metabolic characteristics of PROMISE participants from the baseline visit (2004-2006).
 Darker orange represents a positive correlation; darker blue represents a negative correlation.')
-fig_heatmap_tagfa <- fig('heatmap_tagfa', 'Pearson correlation heatmap of the triacylglycerol fatty acids in the PROMISE participants from the baseline visit (2004-2006).
-The correlations of fatty acids grouped using heirarchical cluster analysis; fatty acids along the x and y axis are ordered according to this analysis.
+fig_heatmap_tagfa <- fig('heatmap_tagfa', 'Pearson correlation heatmap of TGFA in the PROMISE participants from the baseline visit (2004-2006).
+The correlations of FA grouped using heirarchical cluster analysis; FA along the x and y axis are ordered according to this analysis.
 Darker orange represents a positive correlation; darker blue represents a negative correlation.')
-fig_gee_unadj <- fig('gee_unadj', 'Time-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
-X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
+fig_gee_unadj <- fig('gee_unadj', 'Time-adjusted GEE models of the association of the TGFA (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+X-axis values represent a percent difference in the outcome per SD increase in the FA.
 P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
-fig_gee_adj <- fig('gee_adj', 'Fully-adjusted GEE models of the association of the triacylglycerol fatty acids (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
-Variables controlled for were follow-up time, waist circumference, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
-X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
+fig_gee_adj <- fig('gee_adj', 'Fully-adjusted GEE models of the association of the TGFA (mol% and nmol/mL) and total clinically-measured TG with insulin sensitivity and beta-cell function outcomes using the 6 year longitudinal data from the PROMISE cohort.
+Variables controlled for were follow-up time, WC, baseline age, ethnicity, sex, ALT, physical activity, and total NEFA.
+X-axis values represent a percent difference in the outcome per SD increase in the FA.
 P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
-fig_pls <- fig('pls', 'Partial least squares (PLS) models showing the clustering of triacylglycerol fatty acids on insulin sensitivity and beta-cell function measures.
-The R value shown above the graph is the correlation between predicted and observed values, using cross-validation; a higher value suggests the fatty acids strongly predict the metabolic outcome.
+fig_pls <- fig('pls', 'Partial least squares (PLS) models showing the clustering of TGFA on insulin sensitivity and beta-cell function measures.
 The percent explained variance of each component is shown in brackets on each axis.
 The solid line represents an explained variance of 100% while the dashed line represents an explained variance of 50%.
-Fatty acids between these lines represent variables that strongly explain the underlying structure of the data.
-See the Supplemental Methods for a description of PLS analysis and an explanation of interpreting this plot.')
+FA between these lines represent variables that strongly explain the underlying structure of the data.')
 ```
@@
 Canada. Individuals were selected to participate if they met the eligibility 
-criteria of having one or more risk factors for type 2 diabetes mellitus, 
+criteria of having one or more risk factors for T2D, 
 including obesity, hypertension, family history of diabetes, and/or a history of
@@
 (2004-2006, 2007-2009, and 2010-2013). The current study
-used data on participants who did not have diabetes at baseline, who
+used data on participants who did not have T2D at baseline, who
 returned for one or more of the follow-up examinations, and who had samples
@@
 of triheptadecanoin (17:0; Nu-Chek Prep, Inc Elysian, MN, USA) was added as an internal
-standard prior to extracting total lipids according to the method of Folch 
+standard prior to extracting total lipids according to the method of Folch et al
 [@Folch1957a]. Each serum lipid fraction (non-esterified fatty acids (NEFA), 
@@
 Pearson correlation coefficients were computed to assess the relationships of
-individual TGFA with other continuous variables. Correlations were also
-computed and hierarchical clustering analysis was conducted for TGFA against each other.
+individual TGFA with other continuous variables. Within TGFA composition correlations were also
+computed and subsequently analyzed using hierarchical clustering.
 
@@
 primary analysis to determine the longitudinal associations between the outcome 
-variables and the predictor variables. The predictor variables and continuous 
-covariates were scaled (mean-centered and standardized). Given the longitudinal 
+variables and the predictor variables. The predictor variables
+were scaled (mean-centered and standardized). Given the longitudinal 
 design, an auto-regressive of order 1 working correlation matrix was 
@@
 previous literature, from directed acyclic graph [@Greenland1999a]
-recommendations, and from quasi-likelihood information criteria. The final
-GEE model (M6) was adjusted for time, waist circumference, baseline age, ethnicity,
-sex, ALT, MET, and total NEFA. The TGFA, total NEFA, sex, ethnicity, and
+recommendations, and from quasi-likelihood information criteria. 
+The DAG structures to understand potential confounding, shown in `r cite_sf("dag_is")`
+and `r cite_sf("dag_bcf")`, were processed by the DAGitty software
+[@Textor2011a;@Shrier2008a] to generate the recommended adjustments.
+The final
+GEE model (M6; seen in `r cite_st('qic')`) was adjusted for years since baseline, WC, baseline age, ethnicity,
+sex, ALT, MET, and total NEFA. The variables TGFA, total NEFA, sex, ethnicity, and
 baseline age were classified as *time-independent* (held constant) as they were
@@
 We also tested for an interaction with sex, ethnicity, or time by the predictor 
-term for each outcome variable. See the Supplemental Methods for an expanded 
-explanation of the GEE modeling analysis.
+term for each outcome variable.
 
@@
 used to identify the patterns of TGFA composition against insulin sensitivity
-and beta-cell function as outcome variables. For a detailed explanation of PLS
-see the Supplemental Methods. Briefly, PLS is a technique that extracts latent
+and beta-cell function as outcome variables. 
+Briefly, PLS is a technique that extracts latent
 structures (clusters) underlying a set of predictor variables conditional on a
@@
 `r cite_f('tagfa')` shows the composition of each FA in the TG fraction (see 
-`r cite_st('tagfa')` for a tabular presentation of the values). 
+`r cite_st('tagfa')` for the raw values). 
 Three TGFA contributed `r aide::format_round(sum(tagfa$pct))`% to the total 
@@
 most evident for 14:0, 14:1n-7, 16:0, and 16:1n-7 that had a moderate positive
-correlation with clinical TG (r=`r fa_cor_mol$rng$TAG[1]`) while all other fatty
-acids had a negative association (r=`r fa_cor_mol$rng$TAG[2]`). In particular,
+correlation with clinical TG (r=`r fa_cor_mol$rng$TAG[2]`) while all other fatty
+acids had a negative association (r=`r fa_cor_mol$rng$TAG[1]`). In particular,
 those fatty acids with the negative associations with clinical TG were all the
@@
 the attenuation with beta-cell function was due primarily to adjustment for
-waist.
+WC.
 
 In analyses using concentration values, nearly all TGFA had a strong
-negative association on HOMA2-%S and ISI (estimates of percent different ranging from
+negative association on HOMA2-%S and ISI (estimates of percent difference ranging from
 `r gmest1$outcome_b$rng[3]` and `r gmest1$outcome_b$rng[7]`, respectively), and a
@@
 In analyses using TGFA mol% values, four TGFA (14:0, 16:0, 14:1n-7, and 16:1n-7) had negative
-associations with HOMA2-%S and ISI (between `r gmest1$outcome_b$rng[1]` and 
+associations with HOMA2-%S and ISI (estimates between `r gmest1$outcome_b$rng[1]` and 
 `r gmest1$outcome_b$rng[5]`, respectively, lower insulin sensitivity for every SD
@@
 HOMA2-%S and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9, 18:2n-6, 20:2n-6, 20:4n-6,
-and 22:5n-3) predicting between `r gmest1$outcome_b$rng[2]` and 
+and 22:5n-3) estimating between `r gmest1$outcome_b$rng[2]` and 
 `r gmest1$outcome_b$rng[6]`%, respectively, higher insulin sensitivity for every
@@
 the TGFA (data not shown). Results of the sensitivity analyses identifying
-waist circumference as the covariate that attenuated the beta-cell function
+WC as the covariate that attenuated the beta-cell function
 associations from the unadjusted model are shown in `r cite_sf("gee_adj_nowaist")`. 
@@
 and 22:6n-3 loaded positively on both insulin sensitivity measures. No other
-TGFA loaded strongly. In the second component, 18:1n-9 and 18:1n-7 loaded postively 
+TGFA loaded strongly. In the second component, 18:1n-9 and 18:1n-7 loaded positively 
 but not strongly while 20:5n-3 and 22:6n-3 loaded strongly and negatively with both
@@
 variables, with the strongest association being with the insulin sensitivity
-variables (beta=`r pls_gee$estimate[3]`, all p`r pls_gee$p.value[1]`; using PLS
-scores constrained by ISI).
+variables (estimate of `r pls_gee$estimate[3]`, all p`r pls_gee$p.value[1]`; using PLS
+scores constrained by ISI). See `r cite_sf("pls_loadings")` for a plot of the
+loadings of each TGFA on the two components.
 
@@
 
-In the present study, we found that in a Canadian cohort at risk for diabetes,
+In the present study, we found that in a Canadian cohort at risk for T2D,
 several specific TGFA and groups of TGFA were strongly associated
@@
 (potentially through higher intakes of simple carbohydrates) may increase the
-risk for diabetes, primarily through worsening insulin sensitivity.
+risk for T2D, primarily through worsening insulin sensitivity.
 
@@
 Two large prospective studies have been published that similarly examined TGFA 
-composition and diabetes outcomes. Rhee *et al* presented a nested case-control
+composition and T2D outcomes. Rhee *et al* presented a nested case-control
 analysis (n=189 cases and n=189 controls) within the Framingham offspring cohort
@@
 by a lower carbon chain and fewer double bonds (e.g. 14:0, 16:0) had a higher
-risk for diabetes after 12-years while those with a profile characterized by
-higher carbon chain and more double bond TGFA had a lower risk for diabetes.
+risk for T2D after 12-years while those with a profile characterized by
+higher carbon chain and more double bond TGFA had a lower risk for T2D.
 A similar pattern of TGFA was also associated with HOMA-IR cross-sectionally
@@
 
-## Limitations and strengths
-
 Our study has potential limitations that need to be considered when interpreting
@@
 potential confounders and mediators through the use of the DAG modeling, relying
-on previous literature, and through QIC model fit comparison methods.
+on previous literature, and through information criteria model fit comparison methods.
 
@@
 computed scores are not able to control for potential confounders and other 
-effect modifiers. Likewise, PLS is not able to handle longitudinal data so only 
+effect modifiers. PLS is also not able to handle longitudinal data so only 
 the baseline visit was used in the PLS analysis, although we analyzed the extracted
@@
 
-## Final conclusion
+## Conclusion
 
@@
 likely due to greater intake of simple or refined carbohydrates, may increase the
-risk of diabetes through worsening of insulin sensitivity.
+risk of T2D through worsening of insulin sensitivity.
 
@@
 Doctoral Student Research Award; RR is supported by a Heart and Stroke
-Foundation of Ontario Mid-Career Investigator Award {{confirm}}; SBH holds the
+Foundation of Ontario Mid-Career Investigator Award; SBH holds the
 CDA Chair in National Diabetes Management and the Ian McWhinney Chair of Family
-Medicine Studies at the University of Western Ontario {{confirm}}; RBP holds a
+Medicine Studies at the University of Western Ontario; RBP holds a
 Tier II Canada Research Chair in Brain Lipid Metabolism; AJH holds a Tier II

### 2018-03-13 doc/manuscript.Rmd
minor edit to title
rearrange abbrev

@@
 
-Title (109 / 110 characters): 
-Clusters of fatty acids in serum triacylglyceride fraction associate with the
-pathogenesis of type 2 diabetes
+Title (110 / 110 characters): 
+Clusters of fatty acids in the serum triacylglyceride fraction associate with the
+disorders of type 2 diabetes
 
@@
 
-- TGFA: Triacylglycerol fatty acids
-- IS: Insulin sensitivity
-- PROMISE: Prospective Metabolism and Islet Cell Evaluation cohort
-- ISI: Insulin sensitivity index
+- ALT: Alanine aminotransferase
+- CV: Coefficient of variation
+- DNL: De novo lipogenesis
 - GEE: Generalized estimating equations
-- PLS: Partial least squares
 - HOMA2-%S: Homeostatic model of assessment 2 - percent sensitivity
@@
 resistance
+- IS: Insulin sensitivity
+- ISI: Insulin sensitivity index
 - ISSI-2: Insulin Secretion-Sensitivity Index-2
-- DNL: De novo lipogenesis
-- ALT: Alanine aminotransferase
+- MAQ: Modified activity questionnaire
+- MET: Metabolic equivalent of task
 - NEFA: Non-esterified fatty acids
 - OGTT: Oral glucose tolerance test
-- CV: Coefficient of variation
-- MAQ: Modified activity questionnaire
+- PROMISE: Prospective Metabolism and Islet Cell Evaluation cohort
+- PLS: Partial least squares
+- SD: Standard deviation
+- TGFA: Triacylglycerol fatty acids
 - WC: Waist circumference
-- MET: Metabolic equivalent of task
-- SD: Standard deviation
 

### 2018-03-13 doc/manuscript.Rmd
typo in abstract
edits as per Tony's suggestions

@@
 lower IS as well as hypertriglyceridemia. Previous research suggests these four
-FA are also strongly influenced by de novo lipogenesis, particular from
+FA are also strongly influenced by de novo lipogenesis, particularly from
 carbohydrates.
@@
 of participants attended all three visits. There were
-`r female$npct` females and `r euro$npct` who had European-ancestry, with a mean
+`r female$npct` females and `r euro$npct` who reported European-ancestry, with a mean
 age in years of `r aide::ave_sd(ds0yr$Age)` and a mean BMI
@@
 To our knowledge, no longitudinal study to date has examined the role of the
-composition of the TGFA fraction on detailed OGTT-derived metabolic measures 
+composition of the TGFA fraction on detailed OGTT-derived metabolic measures. 
 Two large prospective studies have been published that similarly examined TGFA 
@@
 TGFA data were available for 831 participants after 6-years of follow-up. In
-their cohort, OGTT data was only available at the 6-year visit.
-They cross-sectionally found that most saturated fatty acids had negative 
+their cohort, OGTT data were only available at the 6-year visit.
+In the cross-sectional analysis they found that most saturated fatty acids had negative 
 associations with insulin sensitivity and beta-cell function while
@@
 (20:5n-3), and arachidonic acid (20:4n-6) had positive associations with insulin
-sensitivity. The magnitude of the associations were larger in the insulin
+sensitivity. The magnitude of the associations were larger for the insulin
 sensitivity results compared to the beta-cell function results, similar to what
@@
 of metabolic function and as well as multivariate statistical approaches that
-allowed us to identify clusters of TGFA. In another study of a much
-smaller (n=16) mostly female group [@Kotronen2009a], the authors report a
+allowed us to identify clusters of TGFA. In another much smaller study (n=16) of
+mostly females [@Kotronen2009a], the authors reported a
 positive correlation between total esterified (of which TG make up the

### 2018-03-14 doc/manuscript.Rmd
minor edits to submission as requested by journal staff

@@
     - /home/luke/Documents/site/cv/_includes/work.bib
-csl: /home/luke/Documents/bibliography/styles/vancouver.csl
+csl: /home/luke/Documents/bibliography/styles/journal-of-lipid-research.csl
 output: fost::jlr_manuscript
@@
 # Abstract
-
-Word count: 200 / 200
 
@@
 cite_f <- pryr::partial(fig, display = 'cite')
-sfig <- captioner::captioner(prefix = 'Supplemental Figure')
+sfig <- captioner::captioner(prefix = 'Supplemental Figure S')
 cite_sf <- pryr::partial(sfig, display = 'cite')
@@
 cite_t <- pryr::partial(tab, display = 'cite')
-stab <- captioner::captioner(prefix = 'Supplemental Table')
+stab <- captioner::captioner(prefix = 'Supplemental Table S')
 cite_st <- pryr::partial(stab, display = 'cite')

### 2018-03-14 doc/supplements.Rmd
minor edits to submission as requested by journal staff

--- 2018-02-25 doc/esm.Rmd
@@
 # Captions
-sfig <- captioner::captioner(prefix = 'Supplemental Figure')
-cite_sf <- pryr::partial(sfig, display = 'cite')
+sfig <- captioner::captioner(prefix = 'Supplemental Figure S')
 
-tab <- captioner::captioner(prefix = 'Table')
-cite_t <- pryr::partial(tab, display = 'cite')
-stab <- captioner::captioner(prefix = 'Supplemental Table')
-cite_st <- pryr::partial(stab, display = 'cite')
+stab <- captioner::captioner(prefix = 'Supplemental Table S')
 

### 2018-04-19 doc/submission/reviewer-response.md
started todo and response

---
output: fost::default_manuscript
---

<!-- add manuscript number to cover letter -->

<!--
The editorial staff will be in touch shortly to suggest revisions to your title
and abstract for you to consider. Please wait to hear from them before
submitting your revised manuscript. 
-->

# Editor

All three reviewers agreed that the paper had the strength of longitudinal
followup and would be of interest to readers of the JLR but they also noted
that there have been earlier studies with similar outcomes and one reviewer
noted that some of those papers were by the present group but were not
mentioned here. The reviewers also thought that the discussion was too
speculative and the authors need therefore, to moderate their enthusiasm. My
additional comment is that although you quote one paper by Yki-Jarvinen and
colleagues. you did not discuss/reference her more recent papers linking
shorter chain saturated fatty acids in the liver to NAFLD/NASH. You should look
at those papers and add them to the discussion. I look forward to a modified
version.

# Reviewer 1

1. This manuscript reports interesting and potentially valuable associations
between fatty acid composition and metabolic endpoints in humans over time.

2. The associations are powerful and stand alone. The authors have no data
directly implicating de novo lipogenesis, so the last sentence in the abstract
about previous research should be removed or substantially modified.

    We thank the reviewer for these comments. The suggestion is convincing, we 
    have removed the sentence from the abstract.

3. There are statements in the Discussion that are overly speculative, dated,
and misleading. At the bottom of page 13 the phrase "may lead to greater DNL in
an attempt to control blood glucose" should be removed. On page 14, lines 8-9
the sentence "Higher insulin resistance may encourage greater DNL to handle the
higher blood glucose" should be removed. In the next sentence on the same page,
the authors discuss reference 47 by implying that weight loss improved fatty
acid profiles through effects on DNL, but that is facile and does not consider
the complex effects of improved insulin sensitivity on lipoprotein clearance
associated with induction of lipoprotein lipase, so this sentence needs to be
eliminated or modified.

    These concerns and suggestions are appropriately raised. We have rephrased
    these discussions to be more cautious and less speculative.

4. While the discussion emphasizes DNL, it does not allude to how DNL can
mechanistically lead to the metabolic endpoints used in this study. It would be
helpful to indicate that DNL is complicated, and in certain tissues is involved
in the promotion of chronic inflammation leading to decreased insulin
sensitivity and other adverse metabolic events (Wei X et al. Nature 2016;
539:294-298).

    The discussion on DNL has been clarified in light of the reviewer's comments.

5. On page 9, line 11 from the bottom, the text omits the actual r values for
the relationships between fatty acids and TG (should be in the 0.8 range for
many) and waist circumference (should be in the 0.3 range for many). More
omissions are made on lines 4 and 5 from the bottom when referring to the data
in Figure 3.

    The r values have been included in the text {{confirm this is appropriate}}.

# Reviewer 2

1.	It is a frustrating that this is seemingly the third in a series of papers
that is looking at patterns in lipid fractions and association with metabolic
outcomes. Thus, the TG, NEFA and phospholipid fractions all seem to be a 'least
publishable unit'. The authors must integrate the knowledge gained from the
previous analysis of NEFA and phospholipids and discuss some of the seemingly
contradictory findings from the previous studies. If the data were examined,
would the DNL story 'hold up'?

    We thank the reviewer for these comments. We had many discussions over
    several months deciding how we can publish these results. Given the volume
    of results (figures, tables, numbers) and the complexity of the biology 
    involved in the present analysis, current journal article submissions are
    not suited for publishing a complete analysis and presentation of all the 
    results for all the lipid fractions into a single paper. We were not
    entirely satisfied with splitting the analyses up into multiple papers, 
    however it was the best course of action that we could see. The complete 
    and more integrated discussion and presentation of the results can be found
    in the first author's (LWJ) PhD thesis {{link}}. As to a combined analysis,
    we have presented these results at the European Diabetes Epidemiology Group
    Meeting in April, 2018 {{include DOI}}, and the DNL story remains and is in
    fact a strong contributor to fatty acid and diabetes pathogenesis associations.
    A short aside, these currently unpublished findings do not invalidate or 
    undermine the previous work. The combined analysis uses a powerful
    statistical technique (PLS), but it has a number of major limitations,
    not least of which is it can only handle cross-sectional data and can not
    adjust for potential confounders. The present analysis uses PLS in addition
    to the powerful longitudinal technique GEE. We have included a short 
    description expanding on

2.	The discussion of GEE and its limitations is well done. However, it seems
that the analysis suggests that total triglycerides are predictive of HOMA2-%S
and the other parameters. Do we need to subfractionate to get insight into risk
for type 2 diabetes?

    This is excellent question. For a couple of reasons, no we don't need to 
    subfractionate TAG to understand risk. Firstly, we found that the four TGFA
    also highly positively correlated with clinical TAG. Secondly, clinical TAG
    had a much higher magnitude of association compared to any individual
    TGFA. Measuring clinical TAG is enough to assess risk. However, our results
    provide insight into the potential mechanisms that higher TAG influences risk
    for diabetes, which we show could be due to higher amounts of the four TGFA
    and that higher amounts of these four TGFA alone may strongly contribute to
    the higher clinical TAG.

3.	Supplemental figures 2 and 3 are not discussed at all. They appear to be
interesting and informative in the development of the model. While the process
is described in the earlier NEFA paper, the authors should help the reader
understand what information is being provided without having to go back to a
previous paper.

    We thank the reviewer for this comment. We agree, this information is useful
    and have added a brief discussion of it to the Methods section.

# Reviewer 3

The major drawback
of the report is that some of the mechanistic conclusions are a bit overstated.
Readers would benefit from a more balanced discussion of the study limitations
and alternative interpretations.

1. The authors noted that cluster of four TGFA that correlated negatively with
IS (14:0, 16:0, 14:1n-7, 16:1n-7) are products of DNL, and are therefore likely
reflecting a connection between DNL, CHO intake and diabetes risk. This
conclusion seems a bit overstated. Clearly, a link between this cluster and DNL
is possible, but the signature could be also be reflecting dietary intake,
hepatic steatosis and/or TGFA clearance by tissues such as muscle and adipose
tissue. The authors should include a more balanced discussion of potential
explanations/mechanisms beyond DNL.

2. Are there any additional analyses that can be offered to strengthen evidence
the TGFA composition is indeed a biomarker of DNL and/or CHO intake is this
specific cohort? For example, did the PROMISE study include food recall
records?

3. Also, the discussion should mention and consider that blood samples were
acquired after an overnight fast. Does fasting/fed state impact the correlation
between TGFA composition and DNL?

Minor comments.

4. The authors should comment on how results reported here combine with
(consistencies/contradictions) those reported in their recent Diabetalogia paper
(the parallel NEFA analysis).

5. Authors underscore the strength of the statistical modeling used for the
analysis, including the use DAG modeling to identify potential confounders.
Readers might benefit from a brief comment on the basic principles and strengths
of the method.

### 2018-04-26 doc/submission/reviewer-response.md
added first draft of responses to reviewers

@@
 
-All three reviewers agreed that the paper had the strength of longitudinal
+- All three reviewers agreed that the paper had the strength of longitudinal
 followup and would be of interest to readers of the JLR but they also noted
@@
 
+    *We have made additions to the manuscript based on the editor's and reviewers'
+    suggestions, including the discussion around short chain SFA and NAFLD by
+    Yki-Jarvinen. We believe this has strengthened the paper and hope it
+    satisfactorily addresses the editor and reviewers comments.*
+
 # Reviewer 1
@@
 
-    We thank the reviewer for these comments. The suggestion is convincing, we 
-    have removed the sentence from the abstract.
+    *We thank the reviewer for these comments. The suggestion is convincing, we 
+    have removed the sentence from the abstract.*
 
@@
 
-    These concerns and suggestions are appropriately raised. We have rephrased
-    these discussions to be more cautious and less speculative.
+    *These concerns and suggestions are appropriately raised. We have rephrased
+    these discussions to be more cautious and less speculative.*
 
@@
 
-    The discussion on DNL has been clarified in light of the reviewer's comments.
+    *The discussion on DNL has been clarified in light of the reviewer's comments.*
 
@@
 
-    The r values have been included in the text {{confirm this is appropriate}}.
+    *The r values have been included in the text {{confirm this is appropriate}}.*
 
@@
 
-    We thank the reviewer for these comments. We had many discussions over
+    *We thank the reviewer for these comments. We had many discussions over
     several months deciding how we can publish these results. Given the volume
@@
     adjust for potential confounders. The present analysis uses PLS in addition
-    to the powerful longitudinal technique GEE. We have included a short 
-    description expanding on
+    to the powerful longitudinal technique GEE. {{include a short 
+    description expanding on these results here or in the manuscript?}}*
 
@@
 
-    This is excellent question. For a couple of reasons, no we don't need to 
+    *This is excellent question. For a couple of reasons, no we don't need to 
     subfractionate TAG to understand risk. Firstly, we found that the four TGFA
@@
     and that higher amounts of these four TGFA alone may strongly contribute to
-    the higher clinical TAG.
+    the higher clinical TAG.*
 
@@
 
-    We thank the reviewer for this comment. We agree, this information is useful
-    and have added a brief discussion of it to the Methods section.
+    *We thank the reviewer for this comment. We agree, this information is useful
+    and have added a brief discussion of it to the Methods section.*
 
 # Reviewer 3
-
-The major drawback
-of the report is that some of the mechanistic conclusions are a bit overstated.
-Readers would benefit from a more balanced discussion of the study limitations
-and alternative interpretations.
 
@@
 
+    *We thank the reviewer for these comments. We have revised the discussion to 
+    include alternative possibilities.*
+
 2. Are there any additional analyses that can be offered to strengthen evidence
@@
 
+    *PROMISE did collect food frequency questionnaires, however this was done at
+    the 3rd year visit while the fatty acids where measured at the baseline visit.
+    So while we could technically use this data to help understand the data,
+    we could also obtain inaccurate or biased results because of the time delay.
+    We therefore decided to not analyze this data. {{run analysis and mention it
+    here}}*
+
 3. Also, the discussion should mention and consider that blood samples were
@@
 between TGFA composition and DNL?
+
+    *Excellent suggestion. We have included this in the discussion.*
 
@@
 
+    *We have included a comment in the discussion on how our current analysis
+    fits with our previous NEFA analysis.*
+
 5. Authors underscore the strength of the statistical modeling used for the
@@
 of the method.
+
+    *We agree with this suggestion, but initially excluded it from the methods to
+    limit word count. We have included a section in the methods that briefly 
+    explains this method.*

### 2018-05-03 doc/submission/reviewer-response.md
Added edits to reviewer response

@@
 
-    *We thank the reviewer for these comments. The suggestion is convincing, we 
+    *We thank the reviewer for these comments. We agree with this suggestion and
     have removed the sentence from the abstract.*
@@
     *These concerns and suggestions are appropriately raised. We have rephrased
-    these discussions to be more cautious and less speculative.*
+    these sentences in the Discussion to be more cautious and less speculative.*
 
@@
 
-    *We thank the reviewer for these comments. We had many discussions over
-    several months deciding how we can publish these results. Given the volume
-    of results (figures, tables, numbers) and the complexity of the biology 
-    involved in the present analysis, current journal article submissions are
-    not suited for publishing a complete analysis and presentation of all the 
-    results for all the lipid fractions into a single paper. We were not
-    entirely satisfied with splitting the analyses up into multiple papers, 
-    however it was the best course of action that we could see. The complete 
-    and more integrated discussion and presentation of the results can be found
+    *We thank the reviewer for these comments. We agree that ideally the results
+    of all the pools would have been included together, but given the number of
+    fatty acids, the longitudinal visits, the depth and volume of the analyses,
+    and the complex biology involved, it simply wasn't possible to contain
+    everything in a single journal paper. We have included a brief discussion of
+    the results from the other fractions in the Discussion section. The 
+    complete and integrated discussion and presentation of the results is found
     in the first author's (LWJ) PhD thesis {{link}}. As to a combined analysis,
     we have presented these results at the European Diabetes Epidemiology Group
-    Meeting in April, 2018 {{include DOI}}, and the DNL story remains and is in
-    fact a strong contributor to fatty acid and diabetes pathogenesis associations.
+    Meeting in April, 2018 {{include DOI}}, and the proposed DNL mechanism
+    remained clearly associated and was had a strong contribution to association
+    with diabetes pathogenesis compared to other fatty acids/compositions.
     A short aside, these currently unpublished findings do not invalidate or 
@@
     TGFA. Measuring clinical TAG is enough to assess risk. However, our results
-    provide insight into the potential mechanisms that higher TAG influences risk
+    provide insight into the potential mechanisms that regarding how TAG influences risk
     for diabetes, which we show could be due to higher amounts of the four TGFA
@@
     the 3rd year visit while the fatty acids where measured at the baseline visit.
-    So while we could technically use this data to help understand the data,
-    we could also obtain inaccurate or biased results because of the time delay.
-    We therefore decided to not analyze this data. {{run analysis and mention it
-    here}}*
+    So while we could technically use these dietary intake data to help understand the fatty acid data,
+    we could also obtain inaccurate or biased results because of the time separation.
+    We therefore decided to not analyze this data. {{maybe run analysis and mention it
+    here?}}*
 

### 2018-05-07 doc/manuscript.Rmd
remove dnl bit from abstract

@@
 We found that higher proportions of a cluster of four TGFA strongly related with
-lower IS as well as hypertriglyceridemia. Previous research suggests these four
-FA are also strongly influenced by de novo lipogenesis, particularly from
-carbohydrates.
+lower IS as well as hypertriglyceridemia, suggesting only a few fatty acids within
+the TGFA composition may primarily explain lipids' role in glucose dysregulation.
 

### 2018-05-07 doc/manuscript.Rmd
expand on DAG in methods

@@
 - DNL: De novo lipogenesis
+- DAG: Directed acyclic graph
 - GEE: Generalized estimating equations
@@
 specified in the GEE model. Covariates to adjust for were selected based on the
-previous literature, from directed acyclic graph [@Greenland1999a]
+previous literature, from directed acyclic graph (DAG) [@Greenland1999a]
 recommendations, and from quasi-likelihood information criteria. 
-The DAG structures to understand potential confounding, shown in `r cite_sf("dag_is")`
-and `r cite_sf("dag_bcf")`, were processed by the DAGitty software
-[@Textor2011a;@Shrier2008a] to generate the recommended adjustments.
-The final
-GEE model (M6; seen in `r cite_st('qic')`) was adjusted for years since baseline, WC, baseline age, ethnicity,
+DAGs are used to identify the minimum adjustment necessary for a model by
+using the causal pathways to algorithmically identify potential confounding and
+colliding variables (see the [@Greenland1999a] ref for more detail about using
+DAGs). The DAG structures to understand potential confounding, shown in 
+`r cite_sf("dag_is")` and `r cite_sf("dag_bcf")`, were processed by the DAGitty
+software [@Textor2011a;@Shrier2008a] to generate the recommended adjustments.
+These DAG structures were developed based on hypothesized causal pathways
+between each variable, which were then input into the DAGitty software. The
+output from DAGitty was used to help inform the final model.
+
+The final GEE model (M6; seen in `r cite_st('qic')`) was adjusted for years
+since baseline, WC, baseline age, ethnicity,
 sex, ALT, MET, and total NEFA. The variables TGFA, total NEFA, sex, ethnicity, and

### 2018-05-08 doc/manuscript.Rmd
fix missing corr values

@@
 As expected, nearly all TGFA had very strong positive correlations (r=
-`r fa_cor$rng$TAG[2]`) with clinically-measured TG and moderate positive 
-correlations with WC (r=`r fa_cor$rng$Waist[2]`). There were also moderate
+`r fa_cor$rng$TG[2]`) with clinically-measured TG and moderate positive 
+correlations with WC (r=`r fa_cor$rng$WC[2]`). There were also moderate
 negative correlations with HDL (r=`r fa_cor$rng$HDL[1]`). For the outcome variables,
@@
 most evident for 14:0, 14:1n-7, 16:0, and 16:1n-7 that had a moderate positive
-correlation with clinical TG (r=`r fa_cor_mol$rng$TAG[2]`) while all other fatty
-acids had a negative association (r=`r fa_cor_mol$rng$TAG[1]`). In particular,
+correlation with clinical TG (r=`r fa_cor_mol$rng$TG[2]`) while all other fatty
+acids had a negative association (r=`r fa_cor_mol$rng$TG[1]`). In particular,
 those fatty acids with the negative associations with clinical TG were all the

### 2018-05-08 doc/manuscript.Rmd
started tempering dnl connection

@@
 TGFA (14:0, 14:1n-7, 16:0, 16:1n-7) clustered together, and that this cluster
-strongly predicted lower insulin sensitivity. These four fatty acids are
-involved in the *de novo* lipogenesis (DNL) of refined and simple carbohydrates
-[@Hodson2008a; @Kawano2013a]. Our results suggest that higher activity of DNL
-(potentially through higher intakes of simple carbohydrates) may increase the
-risk for T2D, primarily through worsening insulin sensitivity.
+strongly predicted lower insulin sensitivity.
 
@@
 
-Previous research has shown that carbohydrate intake increases DNL 
+There are some possible explanations to these findings.
+Previous research has shown that carbohydrate intake increases de novo lipogenesis (DNL)
 [@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. In
-particular, DNL from refined or simple carbohydrate sources increases the 14 to
+particular, DNL from simple carbohydrate sources increases the 14 to
 16 chain fatty acids as well as the 18 chain TGFA. Several studies have shown a

### 2018-05-23 doc/submission/reviewer-response.md
updated response about r values
typo fixes

@@
 
-    *The r values have been included in the text {{confirm this is appropriate}}.*
+    *We thank the reviewer for noticing that omission. The r values have been
+    included in the text.*
 
@@
     Meeting in April, 2018 {{include DOI}}, and the proposed DNL mechanism
-    remained clearly associated and was had a strong contribution to association
+    remained clearly associated and had a strong contribution to association
     with diabetes pathogenesis compared to other fatty acids/compositions.
@@
     TGFA. Measuring clinical TAG is enough to assess risk. However, our results
-    provide insight into the potential mechanisms that regarding how TAG influences risk
+    provide insight into the potential mechanisms regarding how TAG influences risk
     for diabetes, which we show could be due to higher amounts of the four TGFA

### 2018-05-23 doc/submission/reviewer-response.md
todo to done for fed/fasting for TGFA

@@
     So while we could technically use these dietary intake data to help understand the fatty acid data,
-    we could also obtain inaccurate or biased results because of the time separation.
-    We therefore decided to not analyze this data. {{maybe run analysis and mention it
-    here?}}*
+    we would likely obtain inaccurate or biased results because of the time separation.
+    We therefore decided to not analyze this data.*
 
@@
 
-    *Excellent suggestion. We have included this in the discussion.*
+    *Excellent question. In the fed state, DNL would increase due to conversion
+    of excess carbohydrates to fatty acids. This in turn would impact the composition
+    of primarily the TGFA. In order to control for this source of variability
+    in our cohort we only measured TGFA during fasting. We have included a brief
+    description of this fed vs fasted state in the limitations section.*
 

### 2018-05-23 doc/manuscript.Rmd
todo to done for fed/fasting for TGFA

@@
 {{residual error? random error?}}.
-[@Rhee2011a] fasting TG status better as discriminating between DM case vs
-control than 2hr-OGTT TG
 -->
 
-TGFA were only quantified at the baseline visit and as such we cannot
+Only fasting TGFA were quantified, and only at the baseline visit. TGFA 
+composition can fluctuate substantially throughout the day, so in order to 
+control for this, PROMISE participants came for the clinic visit in the morning
+and fasted. There is some evidence to suggest that fasting TG is better able
+to discriminate diabetes cases compared to a post-prandial state [@Rhee2011a].
+Because TGFA were only measured at the baseline visit, we cannot
 investigate whether there are concomitant changes in TGFA and the metabolic

### 2018-05-23 doc/submission/reviewer-response.md
added links in place in response file

@@
     complete and integrated discussion and presentation of the results is found
-    in the first author's (LWJ) PhD thesis {{link}}. As to a combined analysis,
+    in the first author's (LWJ) PhD thesis (http://hdl.handle.net/1807/80893). As to a combined analysis,
     we have presented these results at the European Diabetes Epidemiology Group
-    Meeting in April, 2018 {{include DOI}}, and the proposed DNL mechanism
-    remained clearly associated and had a strong contribution to association
+    Meeting in April, 2018 (https://doi.org/10.6084/m9.figshare.6159293), and the proposed DNL mechanism
+    remained clearly associated and had a strong contribution to associations
     with diabetes pathogenesis compared to other fatty acids/compositions.
     A short aside, these currently unpublished findings do not invalidate or 
-    undermine the previous work. The combined analysis uses a powerful
+    undermine the previous work and the current manuscript. The combined analysis uses a powerful
     statistical technique (PLS), but it has a number of major limitations,
     not least of which is it can only handle cross-sectional data and can not
-    adjust for potential confounders. The present analysis uses PLS in addition
-    to the powerful longitudinal technique GEE. {{include a short 
-    description expanding on these results here or in the manuscript?}}*
+    adjust for potential confounders. The present analysis in this manuscript uses PLS in addition
+    to the powerful longitudinal technique GEE.*
 

### 2018-05-28 doc/manuscript.Rmd
todo -> done, add NEFA and other fractions text

@@
 
+Given the close biological relationship between circulating NEFA and TAG, NEFA 
+may act as confounding factor and was thus adjusted for. In a previous published
+analysis of the NEFA fraction [@Johnston2017b], we found that higher total NEFA,
+but not the specific composition, associated with lower beta-cell function. This
+is in contrast to the TGFA findings that the specific composition does
+differentially associate with insulin sensitivity and beta-cell function,
+adjusting for total NEFA. There was no difference in results in models that did
+not include NEFA as a confounder (data not shown). This difference in results
+between NEFA and TGFA suggests that TGFA may independently and strongly influence the 
+pathophysiology of T2D, when compared to other lipid fraction compositions,
+including the phospholipid and cholesteryl ester fractions [@Johnston2016a]. This
+may be because TAG are biologically destined for uptake by non-hepatic tissue as
+they are found mainly in VLDL, at least during fasting. This is in contrast to
+NEFA that are mostly uptaken by the liver and used in TAG production [@Nielsen2012a].
+
 <!--
-{{ Not sure about this... Include this? }}
-As with our previous analysis of the PL fatty acid fraction {{cite}}, we found
-that cis-vaccenic acid strongly predicted higher insulin sensitivity and
-beta-cell function. Likewise, in the TGFA fraction we found a similar
-association with cis-vaccenic acid. {{we can discuss to include more}}
-
-{{ comment on comparing models with and without NEFA }} While
-NEFA may be a mediating factor, we found that adjustment for it did not
-{{confirm}} attenuate this association since TGFA composition is also dictated
-by other sources for DNL (e.g. recent meal, DNL from glucose stores, etc.).
-
-While the NEFA fraction is much smaller than the TG fraction, even though it
-contributes substantially to the TG fraction, the VLDL-TG is not cleared
-(lipolyzed) as much as chylomicron (which also contains a large amount of TG)
-and so has a higher half-life than NEFA, remaining in circulation longer {{hence
-why there is bigger TG pool vs NEFA}} {{cite?}}.
+{{ comment on comparing models with and without NEFA }} While NEFA may be a
+mediating factor, we found that adjustment for it did not attenuate this
+association since TGFA composition is also dictated by other sources for DNL
+(e.g. recent meal, DNL from glucose stores, fatty acid recycling, etc.).
 
@@
 lipotoxic environment {{ cite }}
-
-{{ Concentration data is not }}
 

### 2018-05-28 doc/manuscript.Rmd
trimming down and clarifying dnl story (incomplete)

@@
 There are some possible explanations to these findings.
+
+{{
+There are two sources for endogenous fatty acids: diet and de novo lipogenesis 
+(DNL). Given that most fatty acids can come from either source, it is difficult
+to
+
+}}
+
 Previous research has shown that carbohydrate intake increases de novo lipogenesis (DNL)
-[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. In
-particular, DNL from simple carbohydrate sources increases the 14 to
-16 chain fatty acids as well as the 18 chain TGFA. Several studies have shown a
+[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. 
+Several studies have shown a
 link between higher estimated DNL and an increased risk for metabolic 
-dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. Our study 
-extends these findings by showing that TGFA with 14 to 16 carbons clustered
-together and this pattern strongly predicted lower insulin sensitivity. While
-these fatty acids also had a significant association with beta-cell function,
-the magnitude of associations were more modest compared to those for insulin
-sensitivity.
+dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]; however, how
+DNL may influence metabolic dysfunction is not well described. One possible explanation
+may be the relationship between greater DNL and the types of fatty acids produced.
 
@@
 fat stores and consequently increasing the amount of TGFA in circulation
-[@Chehade2013a; @Verges2015a]. In fact, we found in our study that
+[@Chehade2013a; @Verges2015a]. We found that
 higher proportions of these four DNL TGFA also highly correlated with a higher
@@
 
+Our study 
+extends these findings by showing that TGFA with 14 to 16 carbons clustered
+together and this pattern strongly predicted lower insulin sensitivity. While
+these fatty acids also had a significant association with beta-cell function,
+the magnitude of associations were more modest compared to those for insulin
+sensitivity.
+
 The direction of association between TGFA and insulin sensitivity is unclear 
@@
 the reverse may also be true [@Flannery2012a]. Higher insulin resistance may
-encourage greater DNL to handle the higher blood glucose. To illustrate this, in
-a weight loss intervention trial (n=19 with TGFA data), participants who lost
-weight over 33 weeks showed higher insulin sensitivity and lower TGFA
-composition indicative of lower DNL (less 14:0, 14:1, 16:0, 16:1, etc)
-[@Schwab2008a]. Given the complex biological mechanisms and feedback loops
+encourage greater DNL to handle the higher blood glucose. {{higher insulin
+resistance may also lead to higher circulating LDL, since they are not taken
+up... cite}}
+As we found no interaction by time of TGFA on insulin sensitivity, this study
+can not determine the exact role of the feedback mechanism. Combined with the
+lack of time interaction and the consistent negative association without the
+time interaction, these results at least suggest that the feedback mechanism may
+not be strongly influential and that TGFA may predict insulin sensitivity over a
+six year period. Given the complex biological mechanisms and feedback loops
 involved, disentangling whether insulin sensitivity influences TGFA more
 strongly than TGFA influencing insulin sensitivity will require more
-complicated research designs and analyses. While further research will need to
-confirm this, the lack of a time interaction we observed in addition to the
-timing of our measures likely suggests that it is the TGFA predicting insulin
-resistance.
+complex research designs and analyses.
 
-Given the close biological relationship between circulating NEFA and TAG, NEFA 
+Given the close biological relationship between circulating NEFA and TG, NEFA 
 may act as confounding factor and was thus adjusted for. In a previous published
@@
 including the phospholipid and cholesteryl ester fractions [@Johnston2016a]. This
-may be because TAG are biologically destined for uptake by non-hepatic tissue as
+may be because TG are biologically destined for uptake by non-hepatic tissue as
 they are found mainly in VLDL, at least during fasting. This is in contrast to
-NEFA that are mostly uptaken by the liver and used in TAG production [@Nielsen2012a].
+NEFA that are mostly uptaken by the liver and used in TG production [@Nielsen2012a].
 

### 2018-05-31 doc/manuscript.Rmd
more (incomplete) edits to manuscript

@@
 
-There are some possible explanations to these findings.
+There are a few possible explanations to these findings.
 
 {{
-There are two sources for endogenous fatty acids: diet and de novo lipogenesis 
-(DNL). Given that most fatty acids can come from either source, it is difficult
-to
+Endogenous fatty acids derive from two sources: diet and de novo lipogenesis 
+(DNL). While adipose tissue also supplies circulating fatty acids, the stored
+fatty acids ultimately originate from these two sources.
+
+Since the majority of fatty acids can be obtained from either source, the specific
+composition
+
+- composition is important
+- where it comes from is difficult
+- diet can influence, just as DNL can
+- diet can also influence DNL
+    - more carbs, more DNL
+    - more fat, less DNL
+- More specific fat in diet, more specific fat in body (generally)
+- More carbs, more DNL fat
+- Dietary fat contains wide range of fat types, translates to wide range of fats
+
+irrespective of the source, individual fatty acids can have unique physiological 
+effects. For instance, 14 to 18 chain
 
 }}
-
-Previous research has shown that carbohydrate intake increases de novo lipogenesis (DNL)
-[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. 
-Several studies have shown a
-link between higher estimated DNL and an increased risk for metabolic 
-dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]; however, how
-DNL may influence metabolic dysfunction is not well described. One possible explanation
-may be the relationship between greater DNL and the types of fatty acids produced.
 
@@
 carbohydrate) group. These fatty acids are similar to the fatty acids we found
-that clustered together using the PLS analysis, implicating these fatty acids as
-indicative of a higher carbohydrate diet. A higher carbohydrate diet, 
-particularly one characterized by a predominance of simple carbohydrates, may
-lead to greater DNL in an attempt to control blood glucose, thus increasing hepatic
+that clustered together using the PLS analysis, suggesting a link to
+a higher carbohydrate diet,
+
+which may increase de novo lipogenesis (DNL)
+[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a] as one
+means of lowering blood glucose,
+[@Chehade2013a; @Verges2015a]
+
+lead to higher DNL in an attempt to control blood glucose, thus increasing hepatic
 fat stores and consequently increasing the amount of TGFA in circulation
-[@Chehade2013a; @Verges2015a]. We found that
+Several studies have shown a
+link between higher estimated DNL and an increased risk for metabolic 
+dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]; however, how
+DNL may influence metabolic dysfunction is not well described. One possible explanation
+may be the relationship between greater DNL and the types of fatty acids produced.
+
+We found that
 higher proportions of these four DNL TGFA also highly correlated with a higher
@@
 
-Our study 
-extends these findings by showing that TGFA with 14 to 16 carbons clustered
-together and this pattern strongly predicted lower insulin sensitivity. While
-these fatty acids also had a significant association with beta-cell function,
-the magnitude of associations were more modest compared to those for insulin
-sensitivity.
-
 The direction of association between TGFA and insulin sensitivity is unclear 
@@
 the reverse may also be true [@Flannery2012a]. Higher insulin resistance may
-encourage greater DNL to handle the higher blood glucose. {{higher insulin
+encourage greater DNL to handle the higher blood glucose. 
+{{higher insulin
 resistance may also lead to higher circulating LDL, since they are not taken
@@
 complex research designs and analyses.
+
+Our study 
+extends these findings by showing that TGFA with 14 to 16 carbons clustered
+together and this pattern strongly predicted lower insulin sensitivity. While
+these fatty acids also had a significant association with beta-cell function,
+the magnitude of associations were more modest compared to those for insulin
+sensitivity.
 

### 2018-06-02 doc/manuscript.Rmd
add, rearrange discussion items about dnl and alt explain

--- 2018-05-28 doc/manuscript.Rmd
@@
 
-There are some possible explanations to these findings.
-
-{{
-There are two sources for endogenous fatty acids: diet and de novo lipogenesis 
-(DNL). Given that most fatty acids can come from either source, it is difficult
-to
-
-}}
-
-Previous research has shown that carbohydrate intake increases de novo lipogenesis (DNL)
-[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. 
-Several studies have shown a
-link between higher estimated DNL and an increased risk for metabolic 
-dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]; however, how
-DNL may influence metabolic dysfunction is not well described. One possible explanation
-may be the relationship between greater DNL and the types of fatty acids produced.
+There are a few possible explanations to these findings. The composition of the
+TGFA fraction is determined by the two sources of dietary fat and de novo
+lipogenesis (DNL). And while adipose tissue also supplies circulating fatty
+acids, the stored fatty acids ultimately originate from these two sources.
+Specific dietary fat as well as carbohydrates can influence DNL activity
+{{cite}}. Since the majority of fatty acids can be obtained from either source,
+the specific source for fatty acids is extremely difficult to ascertain outside
+of highly controlled experimental settings. 
+In our findings, the four fatty acids are also highly correlated amongst each other, in addition to
+clustering together on their negative association with insulin sensitivity. These
+findings suggest that these fatty acids may derived mainly from DNL, given the
+evidence that they share a tight DNL pathway. While it may be possible that
+these fatty acids also share a common dietary source (e.g. red meat) {{cite}},
+dietary sources of these four fatty acids are also high in fatty acids such as
+18:0 and 18:1 n-9, which we did not see cluster together.
 
@@
 reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
-carbohydrate) group. These fatty acids are similar to the fatty acids we found
-that clustered together using the PLS analysis, implicating these fatty acids as
-indicative of a higher carbohydrate diet. A higher carbohydrate diet, 
-particularly one characterized by a predominance of simple carbohydrates, may
-lead to greater DNL in an attempt to control blood glucose, thus increasing hepatic
-fat stores and consequently increasing the amount of TGFA in circulation
-[@Chehade2013a; @Verges2015a]. We found that
-higher proportions of these four DNL TGFA also highly correlated with a higher
-concentration of clinical TG, reinforcing this pathway between carbohydrate
-intake, DNL, and circulating TG. The higher concentration of circulating 14 and
-16 carbon fatty acids may then expose tissues to greater lipotoxicity, for
-instance from palmitic acid (16:0), which is well-known to have harmful effects
-on tissues [@Riserus2008a; @Iggman2010a].
+carbohydrate) group. {{talk about Yki-.. study}}. These fatty acids are similar
+to the fatty acids we found that clustered together using the PLS analysis.
+
+We found that higher proportions of these four DNL TGFA also highly correlated
+with a higher concentration of clinical TG, highlighting a potential relationship
+between these four fatty acids, circulating TG, and the
+
+A higher activity in DNL may be a means to control blood glucose with higher
+dietary carbohydrate intake [@Hodson2008a; @Kawano2013a; @Harding2015a;
+@Hudgins2000a; @Parks1999a; @Chehade2013a; @Verges2015a], which could increase
+hepatic fat stores and increase the amount of TG in circulation. Several studies
+have in fact shown a link between higher estimated DNL and an increased risk for
+metabolic dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a];
+however, how DNL may influence metabolic dysfunction is not well described. One
+possible explanation may be the relationship between greater DNL and the types
+of fatty acids produced.
+
+Regardless of the exact source of these fatty acids, our results in addition to
+the available scientific evidence emphasizes the importance of the fatty acid
+composition on metabolic health, as individual fatty acids can have unique
+physiological functions {{cite}}.
+
+The higher concentration of
+circulating 14 and 16 carbon fatty acids may then expose tissues to greater
+lipotoxicity, for instance from palmitic acid (16:0), which is well-known to
+have harmful effects on tissues [@Riserus2008a; @Iggman2010a].
 
@@
 Given the close biological relationship between circulating NEFA and TG, NEFA 
-may act as confounding factor and was thus adjusted for. In a previous published
+may act as confounding factor and was thus adjusted for. In our published
 analysis of the NEFA fraction [@Johnston2017b], we found that higher total NEFA,
@@
 <!--
-{{ comment on comparing models with and without NEFA }} While NEFA may be a
-mediating factor, we found that adjustment for it did not attenuate this
-association since TGFA composition is also dictated by other sources for DNL
-(e.g. recent meal, DNL from glucose stores, fatty acid recycling, etc.).
-
 - contributions to DNL (e.g. 70% 16:0 to newly synth) {{ confirm what this is }}

### 2018-06-03 doc/manuscript.Rmd
todo to done, first draft of revision (not including citations)

--- 2018-05-28 doc/manuscript.Rmd
@@
 
-There are some possible explanations to these findings.
+There are a few possible explanations to these findings. Fatty acids, and thus
+the TGFA composition, derive from two sources: dietary fat and de novo
+lipogenesis (DNL). Adipose tissue also supplies circulating fatty acids,
+however, these also ultimately originate from the two sources. Specific dietary
+fat and carbohydrates can influence DNL activity {{cite}}. Determining the
+specific source of fatty acids is extremely difficult to ascertain outside of
+highly controlled experimental settings. However, there are sign that suggest
+one source over another.
 
-{{
-There are two sources for endogenous fatty acids: diet and de novo lipogenesis 
-(DNL). Given that most fatty acids can come from either source, it is difficult
-to
-
-}}
-
-Previous research has shown that carbohydrate intake increases de novo lipogenesis (DNL)
-[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. 
-Several studies have shown a
-link between higher estimated DNL and an increased risk for metabolic 
-dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]; however, how
-DNL may influence metabolic dysfunction is not well described. One possible explanation
-may be the relationship between greater DNL and the types of fatty acids produced.
-
-The link between higher DNL and increases in specific fatty acids has been
-examined in several studies. Previous studies that have examined DNL have used
+The link DNL activity and levels of specific fatty acids has been examined in
+several studies. Previous studies that have examined DNL have used
 markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
@@
 reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
-carbohydrate) group. These fatty acids are similar to the fatty acids we found
-that clustered together using the PLS analysis, implicating these fatty acids as
-indicative of a higher carbohydrate diet. A higher carbohydrate diet, 
-particularly one characterized by a predominance of simple carbohydrates, may
-lead to greater DNL in an attempt to control blood glucose, thus increasing hepatic
-fat stores and consequently increasing the amount of TGFA in circulation
-[@Chehade2013a; @Verges2015a]. We found that
-higher proportions of these four DNL TGFA also highly correlated with a higher
-concentration of clinical TG, reinforcing this pathway between carbohydrate
-intake, DNL, and circulating TG. The higher concentration of circulating 14 and
-16 carbon fatty acids may then expose tissues to greater lipotoxicity, for
-instance from palmitic acid (16:0), which is well-known to have harmful effects
-on tissues [@Riserus2008a; @Iggman2010a].
+carbohydrate) group. {{talk about Yki-.. study}}.  In our findings, the four
+fatty acids were highly correlated amongst each other, in addition to clustering
+together on their negative association with insulin sensitivity. If these fatty
+acids derived from mainly the diet, we would anticipate seeing more correlation
+with the 18:0, 18:1, and 18:2 fatty acids, as these fatty acids often share a
+dietary source (e.g. meat) {{cite?}} with the four clustering fatty acids. As our
+results do not show this, this suggests the common source being DNL and that the
+fatty acids may have been produced from higher DNL activity since, again, there
+was no strong correlation with 18:0 and 18:1 fatty acids, which would be the
+next stage DNL via elongation and/or desaturation of 16:0 and 16:1 fatty acids.
 
-Our study 
-extends these findings by showing that TGFA with 14 to 16 carbons clustered
-together and this pattern strongly predicted lower insulin sensitivity. While
-these fatty acids also had a significant association with beta-cell function,
-the magnitude of associations were more modest compared to those for insulin
-sensitivity.
+A higher DNL activity may result from a higher carbohydrate intake as a means to
+control blood glucose [@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a;
+@Parks1999a; @Chehade2013a; @Verges2015a], which could increase hepatic fat
+stores and increase the amount of TG in circulation. In fact, we found that
+higher proportions of these four TGFA also highly correlated with a higher
+concentration of clinical TG. Several studies have shown a link between higher
+estimated DNL and an increased risk for metabolic dysfunction [@Ma2015a;
+@Zong2013a; @Lankinen2015a; @Kroger2011a]; how DNL may influence metabolic
+dysfunction is not well understood. Possible explanations may be that higher DNL
+produces more of certain fatty acids or that higher DNL increases circulating TG,
+which itself is well documented to contribute to metabolic dysfunction {{cite?}}.
+
+Regardless of the exact source of these fatty acids, our results in addition to
+the available scientific evidence emphasizes the importance of the fatty acid
+composition on metabolic health, as individual fatty acids can have specific
+physiological functions {{cite}}. For instance, a higher concentration of
+circulating 14 and 16 carbon fatty acids may expose tissues to greater
+lipotoxicity, for instance from palmitic acid (16:0), which is well-known to
+have harmful effects on tissues [@Riserus2008a; @Iggman2010a]. Our study extends
+these findings by showing that TGFA with 14 to 16 carbons clustered together and
+this pattern strongly predicted lower insulin sensitivity. While these fatty
+acids also had a significant association with beta-cell function, the magnitude
+of associations were more modest compared to those for insulin sensitivity.
 
 The direction of association between TGFA and insulin sensitivity is unclear
-from previous cross-sectional studies due to the physiological feedback mechanisms 
-involved. For example, while greater DNL may promote muscle insulin resistance, 
-the reverse may also be true [@Flannery2012a]. Higher insulin resistance may
-encourage greater DNL to handle the higher blood glucose. {{higher insulin
-resistance may also lead to higher circulating LDL, since they are not taken
-up... cite}}
+from previous cross-sectional studies due to the physiological feedback
+mechanisms involved. For example, while greater DNL and higher TGFA may promote
+muscle insulin resistance, the reverse may also be true [@Flannery2012a]. Higher
+insulin resistance may encourage TGFA to remain in circulation longer {{cite?}}.
 As we found no interaction by time of TGFA on insulin sensitivity, this study
-can not determine the exact role of the feedback mechanism. Combined with the
-lack of time interaction and the consistent negative association without the
-time interaction, these results at least suggest that the feedback mechanism may
-not be strongly influential and that TGFA may predict insulin sensitivity over a
-six year period. Given the complex biological mechanisms and feedback loops
-involved, disentangling whether insulin sensitivity influences TGFA more
-strongly than TGFA influencing insulin sensitivity will require more
+cannot determine the exact role of the feedback mechanism. However, combining the
+with the lack of time interaction and the consistent negative association
+without the time interaction, these results at least suggest that the feedback
+mechanism may not be strongly influential and that TGFA may predict insulin
+sensitivity over a six year period. Given the complex biological mechanisms and
+feedback loops involved, disentangling whether insulin sensitivity influences
+TGFA more strongly than TGFA influencing insulin sensitivity will require more
 complex research designs and analyses.

### 2018-06-03 doc/manuscript.Rmd
removed commented text

@@
 
-<!--
-- contributions to DNL (e.g. 70% 16:0 to newly synth) {{ confirm what this is }}
-
-- Zulyniak2012a (only vaccenic acid in TG) (n=20, TLC-GC, HOMA-IR/HOMA-B,
-two groups: normogly and hypergly. results in general showed no diff between
-groups, only cis-vaccenic was higher in hyper, 'trend' with palmitic, did
-not present results on whole fraction for IR)
-
-{{ comment on unadjusted vs adjusted }}
-
-- Potential clinical application
-
-- Or that post-prandial TG is not suppressed in IR Sondergaard2012a
-- {{ confirm }} greater flux of saturated fatty acids into tissues may encourage
-lipotoxic environment {{ cite }}
-
--->
-
 Our study has potential limitations that need to be considered when interpreting
@@
 on previous literature, and through information criteria model fit comparison methods.
-
-<!--
-TGFA can fluctuate substantially throughout the day and to control for this all
-PROMISE participants came for their clinic visits in the morning in a fasted
-state. Some of the inherent variability in TGFA fluctuation can be managed by
-standardizing the data collection time, though there may still be some
-{{residual error? random error?}}.
--->
 

### 2018-06-03 doc/manuscript.Rmd
fixed final conclusion

@@
 
-In conclusion, we found that a TGFA composition indicative of higher DNL
-(containing higher 14:0, 14:1n-7, 16:0, and 16:1n-7) associated strongly with lower
-insulin sensitivity and (more moderately) with lower beta-cell function. The fatty
-acids that clustered together represent fatty acids created from DNL, which is
-characteristic of higher simple carbohydrate (e.g. added sugar) intake. Our
-results, which are congruent with current evidence, suggest that higher DNL,
-likely due to greater intake of simple or refined carbohydrates, may increase the
-risk of T2D through worsening of insulin sensitivity.
+In conclusion, we found that a TGFA composition containing higher proportions of
+14:0, 14:1n-7, 16:0, and 16:1n-7 associated strongly with lower insulin
+sensitivity and (more moderately) with lower beta-cell function. We also found
+that most other TGFA (e.g. 20:0, most omega 6 and 9 TGFA, and 20:5n-3)
+associated positively with insulin sensitivity. Only a few TGFA associated
+positively and consistently with beta-cell function (e.g. 18:1n-7, 20:2n-6). 
+These results provide more insight into how individual TGFA contribute to the
+pathogenesis of T2D while reinforcing the importance and value of clinically
+measured total TG as a indicator of metabolic health.
 

### 2018-06-03 doc/manuscript.Rmd
added citations, minor edits, and fixes for finished draft to send to tony

@@
 response variable(s) (i.e. the outcome variables). How accurately the clusters
-within the TGFA composition predict metabolic function is determined by using
-cross-validation on the PLS models.
+within the TGFA composition predict metabolic function is determined
+by using cross-validation on the PLS models.
 
@@
 (14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and
-negatively predicted lower insulin sensitivity. While most TGFA were not
-associated with beta-cell function, two fatty acids, palmitic acid (16:0) and
-*cis*-vaccenic acid (18:1n-7), were associated negatively and positively,
+negatively associated with lower insulin sensitivity. While most TGFA were not
+associated with beta-cell function, three fatty acids, palmitic acid (16:0),
+*cis*-vaccenic acid (18:1n-7), and eicosadienoic acid (20:2n-6) were associated negatively and positively,
 respectively, with measures of beta-cell function. Using PLS, we also found that four
@@
 
-There are a few possible explanations to these findings. Fatty acids, and thus
-the TGFA composition, derive from two sources: dietary fat and de novo
-lipogenesis (DNL). Adipose tissue also supplies circulating fatty acids,
-however, these also ultimately originate from the two sources. Specific dietary
-fat and carbohydrates can influence DNL activity {{cite}}. Determining the
-specific source of fatty acids is extremely difficult to ascertain outside of
-highly controlled experimental settings. However, there are sign that suggest
-one source over another.
-
-The link DNL activity and levels of specific fatty acids has been examined in
-several studies. Previous studies that have examined DNL have used
-markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
-to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are 
+There are a few possible explanations to these findings. Circulating 
+TGFA derive from three sources: adipose lyposis, dietary fat, and de novo
+lipogenesis (DNL). Dietary carbohydrates and fat can influence DNL activity
+[@Aarsland1998a;@Harding2015a;@Luukkonen2018a]. Determining the specific source
+of TGFA is extremely difficult to ascertain outside of highly controlled
+experimental settings. Many previous studies that have examined DNL
+have used markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or
+16:1n-7 to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are
 limitations to using these ratios as the fatty acids used in their calculation
 can also be obtained from the diet in addition to being created through DNL
-[@Hodson2008a]. A feeding trial (n=24) was conducted to identify 
-the fatty acids that most accurately reflected DNL as potential biomarkers
-[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
-DNL using isotopes, and liver fat were all highly correlated with each other
-(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
-another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
-to determine its effect on DNL and TGFA composition [@Wilke2009a]. The authors
+[@Hodson2008a]. An experimental feeding trial (n=24) was conducted to identify the fatty acids
+that most accurately reflected DNL as potential biomarkers [@Lee2015a]. The
+study found that palmitoleic acid (16:1n-7), directly measured DNL using
+isotopes, and liver fat were all highly correlated with each other (r>0.50),
+suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In another
+small (n=14) feeding trial, meal type (high fat vs low fat) was tested to
+determine its effect on DNL and TGFA composition [@Wilke2009a]. The authors
 reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
-carbohydrate) group. {{talk about Yki-.. study}}.  In our findings, the four
-fatty acids were highly correlated amongst each other, in addition to clustering
-together on their negative association with insulin sensitivity. If these fatty
-acids derived from mainly the diet, we would anticipate seeing more correlation
-with the 18:0, 18:1, and 18:2 fatty acids, as these fatty acids often share a
-dietary source (e.g. meat) {{cite?}} with the four clustering fatty acids. As our
-results do not show this, this suggests the common source being DNL and that the
-fatty acids may have been produced from higher DNL activity since, again, there
-was no strong correlation with 18:0 and 18:1 fatty acids, which would be the
-next stage DNL via elongation and/or desaturation of 16:0 and 16:1 fatty acids.
+carbohydrate) group. In another recent overfeeding trial, 1000 kcal of saturated
+fat, unsaturated fat, or carbohydrates over 3 weeks was given to 38 overweight
+individuals [@Luukkonen2018a] to test changes in liver fat and hepatic DNL. At the
+end of the study, the carbohydrate group had higher DNL as well as an increase
+in liver fat, though the saturated fat group had the highest increase in liver
+fat.
 
-A higher DNL activity may result from a higher carbohydrate intake as a means to
-control blood glucose [@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a;
-@Parks1999a; @Chehade2013a; @Verges2015a], which could increase hepatic fat
-stores and increase the amount of TG in circulation. In fact, we found that
-higher proportions of these four TGFA also highly correlated with a higher
-concentration of clinical TG. Several studies have shown a link between higher
-estimated DNL and an increased risk for metabolic dysfunction [@Ma2015a;
-@Zong2013a; @Lankinen2015a; @Kroger2011a]; how DNL may influence metabolic
-dysfunction is not well understood. Possible explanations may be that higher DNL
-produces more of certain fatty acids or that higher DNL increases circulating TG,
-which itself is well documented to contribute to metabolic dysfunction {{cite?}}.
+This link between carbohydrate intake and DNL activity has been well documented
+[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a; @Chehade2013a; @Verges2015a].
+In our findings, the four fatty acids were highly positively correlated
+amongst each other and negatively or neutrally with all other TGFA, in addition to
+clustering together on their negative association with insulin sensitivity. This
+may suggest that greater DNL activity may be the source of these TGFA. Several
+studies have shown a link between higher estimated DNL and an increased risk for
+metabolic dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]; how
+DNL may influence metabolic dysfunction is not well understood. Possible
+reasons may be that higher DNL produces more of certain fatty acids or that
+higher DNL increases circulating TG, which itself is well documented to
+contribute to metabolic dysfunction and which we found in our study of a high
+positive correlation between the four TGFA and clinical TG.
 
@@
 composition on metabolic health, as individual fatty acids can have specific
-physiological functions {{cite}}. For instance, a higher concentration of
+physiological functions. For instance, a higher concentration of
 circulating 14 and 16 carbon fatty acids may expose tissues to greater
@@
 these findings by showing that TGFA with 14 to 16 carbons clustered together and
-this pattern strongly predicted lower insulin sensitivity. While these fatty
+this pattern strongly associated with lower insulin sensitivity. While these fatty
 acids also had a significant association with beta-cell function, the magnitude
@@
 from previous cross-sectional studies due to the physiological feedback
-mechanisms involved. For example, while greater DNL and higher TGFA may promote
-muscle insulin resistance, the reverse may also be true [@Flannery2012a]. Higher
-insulin resistance may encourage TGFA to remain in circulation longer {{cite?}}.
+mechanisms involved. For example, while higher TGFA may promote
+muscle insulin resistance, the reverse may also be true [@Flannery2012a].
 As we found no interaction by time of TGFA on insulin sensitivity, this study
@@
 mechanism may not be strongly influential and that TGFA may predict insulin
-sensitivity over a six year period. Given the complex biological mechanisms and
+sensitivity at least over a six year period. Given the complex biological mechanisms and
 feedback loops involved, disentangling whether insulin sensitivity influences
@@
 on previous literature, and through information criteria model fit comparison methods.
-
 Only fasting TGFA were quantified, and only at the baseline visit. TGFA 

### 2018-06-04 R/gee.R
facet_grid change

@@ -244,7 +244,7 @@ plot_gee_main <- function(results) {
             ylab = 'Triacylglycerol fatty acids'
             ) +
         graph_theme(ticks = FALSE) +
-        ggplot2::facet_grid(unit~Yterms, switch = 'y',
+        ggplot2::facet_grid(cols = vars(unit), rows = vars(Yterms), switch = 'y',
                             scales = 'free_y',
                             space = 'free_y')
 }

### 2018-06-04 doc/manuscript.Rmd
edits as per Tony's suggestions

@@
 using the causal pathways to algorithmically identify potential confounding and
-colliding variables (see the [@Greenland1999a] ref for more detail about using
+colliding variables (see [@Greenland1999a] for more detail about using
 DAGs). The DAG structures to understand potential confounding, shown in 
@@
 
-There are a few possible explanations to these findings. Circulating 
+There are a few possible explanations for these findings. Circulating 
 TGFA derive from three sources: adipose lyposis, dietary fat, and de novo
@@
 
-Regardless of the exact source of these fatty acids, our results in addition to
-the available scientific evidence emphasizes the importance of the fatty acid
+Regardless of the exact source of these fatty acids, our results, in addition to
+the available scientific evidence, emphasize the importance of the fatty acid
 composition on metabolic health, as individual fatty acids can have specific
@@
 they are found mainly in VLDL, at least during fasting. This is in contrast to
-NEFA that are mostly uptaken by the liver and used in TG production [@Nielsen2012a].
+NEFA that are mostly taken up by the liver and used in TG production [@Nielsen2012a].
 
@@
 pathogenesis of T2D while reinforcing the importance and value of clinically
-measured total TG as a indicator of metabolic health.
+measured total TG as an indicator of metabolic health.
 

### 2018-06-04 doc/submission/reviewer-response.md
typo
edits as per tony's suggestions for response

@@
 
-# Editor
+# Dear Dr.
+
+We'd like to thank the editors and the reviewers for their thorough and
+insightful review of our manuscript. We have been able to make the vast majority
+of the suggested revisions, which we feel has improved the manuscript considerably.
+Below we present a point-by-point response to each of the editors' and reviewers' 
+comments.
 
@@
 
-    *We have made additions to the manuscript based on the editor's and reviewers'
+    *Thank you for this suggestion. We have made additions to the manuscript based on the editor's and reviewers'
     suggestions, including the discussion around short chain SFA and NAFLD by
@@
 
-    *We thank the reviewer for noticing that omission. The r values have been
+    *We thank the reviewer for noticing those omissions. The r values have been
     included in the text.*
@@
     *We thank the reviewer for these comments. We agree that ideally the results
-    of all the pools would have been included together, but given the number of
+    of all the fatty acid fractions would have been included together, but given the number of
     fatty acids, the longitudinal visits, the depth and volume of the analyses,
     and the complex biology involved, it simply wasn't possible to contain
-    everything in a single journal paper. We have included a brief discussion of
+    everything in a single journal paper given word count and table/figure limitations. We have included a brief discussion of
     the results from the other fractions in the Discussion section. The 
@@
     we would likely obtain inaccurate or biased results because of the time separation.
-    We therefore decided to not analyze this data.*
+    We therefore decided to not analyze these data.*
 

### 2018-06-04 doc/manuscript.Rmd
started fixing minor edits

@@
 Authors: Luke W. Johnston, PhD (1); Zhen Liu, PhD (1); Ravi Retnakaran, MD
-(3,4); Bernard Zinman, MD (3,4); Adria Giacca, PhD (5); Stewart B. Harris, MD (2); 
-Richard P. Bazinet, PhD (1); and Anthony J. Hanley, PhD (1,3,6)
+(2,3); Bernard Zinman, MD (2,3); Adria Giacca, PhD (4); Stewart B. Harris, MD (5);
+Richard P. Bazinet, PhD (1); and Anthony J. Hanley, PhD (1,2,6)
 
@@
 Canada.
-(2) Centre for Studies in Family Medicine, University of Western Ontario,
+(2) Leadership Sinai Centre for Diabetes, Division of Endocrinology, University
+of Toronto, Toronto, ON, Canada.
+(3) Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto,
+Ontario, Canada.
+(4) Department of Physiology, University of Toronto, Toronto, Ontario, Canada.
+(5) Centre for Studies in Family Medicine, University of Western Ontario,
 London, Ontario, Canada.
-(3) Leadership Sinai Centre for Diabetes, Division of Endocrinology, University
-of Toronto, Toronto, ON, Canada.
-(4) Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto,
-Ontario, Canada.
-(5) Department of Physiology, University of Toronto, Toronto, Ontario, Canada.
 (6) Dalla Lana School of Public Health, University of Toronto, Toronto, Ontario,
@@
 - Phone number: 416.978.3616
-- Fax number:
 - Email: anthony.hanley@utoronto.ca
@@
 20:2n-6, 20:5n-3) had strong positive associations. Few associations were seen
-for beta-cell function, except for 16:0 and 18:1n-7. PLS analysis indicated four
+for beta-cell function, except for 16:0, 18:1n-7, and 20:2n-6. PLS analysis indicated four
 TGFA (14:0, 16:0, 14:1n-7, 16:1n-7) that clustered together and strongly related
@@
 the TGFA composition may primarily explain lipids' role in glucose dysregulation.
-
-# Keywords
-<!-- up to 10. Five from list, five free-form. Should not appear in title or
-running title -->
 

### 2018-06-04 doc/submission/reviewer-response.md
minor edits to response

@@
 
-# Dear Dr.
+Dear Dr. Ginsberg,
 
@@
 comments.
+
+# Editor comment
 
@@
     *Thank you for this suggestion. We have made additions to the manuscript based on the editor's and reviewers'
-    suggestions, including the discussion around short chain SFA and NAFLD by
-    Yki-Jarvinen. We believe this has strengthened the paper and hope it
+    suggestions, including the discussion around SFA and NAFLD by
+    Yki-Jarvinen and colleagues. We believe this has strengthened the paper and hope it
     satisfactorily addresses the editor and reviewers comments.*

### 2018-06-05 doc/submission/reviewer-response.md
added page numbers to response

@@
 ---
-
-<!-- add manuscript number to cover letter -->
 
@@
 -->
-
-Dear Dr. Ginsberg,
-
-We'd like to thank the editors and the reviewers for their thorough and
-insightful review of our manuscript. We have been able to make the vast majority
-of the suggested revisions, which we feel has improved the manuscript considerably.
-Below we present a point-by-point response to each of the editors' and reviewers' 
-comments.
 
@@
     suggestions, including the discussion around SFA and NAFLD by
-    Yki-Jarvinen and colleagues. We believe this has strengthened the paper and hope it
+    Yki-Jarvinen and colleagues (first paragraph, page 13). We believe this has strengthened the paper and hope it
     satisfactorily addresses the editor and reviewers comments.*
@@
     *These concerns and suggestions are appropriately raised. We have rephrased
-    these sentences in the Discussion to be more cautious and less speculative.*
+    these sentences at all points in the Discussion to be more cautious and less speculative.*
 
@@
     everything in a single journal paper given word count and table/figure limitations. We have included a brief discussion of
-    the results from the other fractions in the Discussion section. The 
+    the results from the other fractions in the Discussion section (end of page 14, beginning of page 15). The 
     complete and integrated discussion and presentation of the results is found
@@
     *We thank the reviewer for this comment. We agree, this information is useful
-    and have added a brief discussion of it to the Methods section.*
+    and have added a brief discussion of it to the Methods section (end of page 7).*
 
@@
 
-    *We thank the reviewer for these comments. We have revised the discussion to 
+    *We thank the reviewer for these comments. We have revised the discussion throughout to 
     include alternative possibilities.*
@@
     in our cohort we only measured TGFA during fasting. We have included a brief
-    description of this fed vs fasted state in the limitations section.*
+    description of this fed vs fasted state in the limitations section (middle of page 15).*
 
@@
     *We have included a comment in the discussion on how our current analysis
-    fits with our previous NEFA analysis.*
+    fits with our previous NEFA analysis (end of page 14, beginning of page 15).*
 
@@
     limit word count. We have included a section in the methods that briefly 
-    explains this method.*
+    explains this method (last paragraph of page 7).*

### 2018-06-05 doc/manuscript.Rmd
minor edits and typo fixes in manuscript after final proofreading

@@
 Findings for other lipid fractions in this cohort are reported separately (see
-ref [@Johnston2016a] for the phospholipid and cholesteryl ester fraction and
+[@Johnston2016a] for the phospholipid and cholesteryl ester fraction and
 [@Johnston2017b] for the NEFA fraction analysis).
@@
 between each variable, which were then input into the DAGitty software. The
-output from DAGitty was used to help inform the final model.
+output from DAGitty was used, in conjunction with the other methods, to help inform the final model.
 
@@
 There are a few possible explanations for these findings. Circulating 
-TGFA derive from three sources: adipose lyposis, dietary fat, and de novo
+TGFA derive from three sources: adipose lipolysis, dietary fat, and de novo
 lipogenesis (DNL). Dietary carbohydrates and fat can influence DNL activity
@@
 of TGFA is extremely difficult to ascertain outside of highly controlled
-experimental settings. Many previous studies that have examined DNL
+experimental settings. Many previous studies that have examined DNL as the source
 have used markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or
@@
 individuals [@Luukkonen2018a] to test changes in liver fat and hepatic DNL. At the
-end of the study, the carbohydrate group had higher DNL as well as an increase
+end of the study, the carbohydrate group had higher DNL activity as well as an increase
 in liver fat, though the saturated fat group had the highest increase in liver
-fat.
+fat. This link between carbohydrate intake and DNL activity has been well documented
+[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a; @Chehade2013a; @Verges2015a].
 
-This link between carbohydrate intake and DNL activity has been well documented
-[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a; @Chehade2013a; @Verges2015a].
 In our findings, the four fatty acids were highly positively correlated
@@
 clustering together on their negative association with insulin sensitivity. This
-may suggest that greater DNL activity may be the source of these TGFA. Several
-studies have shown a link between higher estimated DNL and an increased risk for
-metabolic dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]; how
+may suggest that greater DNL activity is the source of these TGFA. Several
+studies have shown a link between higher estimated DNL activity and an increased risk for
+metabolic dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. How
 DNL may influence metabolic dysfunction is not well understood. Possible
@@
 higher DNL increases circulating TG, which itself is well documented to
-contribute to metabolic dysfunction and which we found in our study of a high
+contribute to metabolic dysfunction and which we found in our study as a high
 positive correlation between the four TGFA and clinical TG.
@@
 these findings by showing that TGFA with 14 to 16 carbons clustered together and
-this pattern strongly associated with lower insulin sensitivity. While these fatty
+this pattern strongly associated with lower insulin sensitivity. While some of these fatty
 acids also had a significant association with beta-cell function, the magnitude
@@
 As we found no interaction by time of TGFA on insulin sensitivity, this study
-cannot determine the exact role of the feedback mechanism. However, combining the
-with the lack of time interaction and the consistent negative association
-without the time interaction, these results at least suggest that the feedback
+cannot determine the exact role of the feedback mechanism. However, combining
+the lack of a time interaction and the consistent negative association
+in models without the time interaction, these results at least suggest that the feedback
 mechanism may not be strongly influential and that TGFA may predict insulin
@@
 Given the close biological relationship between circulating NEFA and TG, NEFA 
-may act as confounding factor and was thus adjusted for. In our published
+may act as a confounding factor and was thus adjusted for. In our published
 analysis of the NEFA fraction [@Johnston2017b], we found that higher total NEFA,
@@
 including the phospholipid and cholesteryl ester fractions [@Johnston2016a]. This
-may be because TG are biologically destined for uptake by non-hepatic tissue as
+may be due to TG being biologically destined for uptake by non-hepatic tissue as
 they are found mainly in VLDL, at least during fasting. This is in contrast to
@@
 
+Note: Tables values as either median (interquartile range), mean (SD), or n (percent).
+
 # Figures
